### 2018-04-18 doc/large_strain.tex
add LaTeX template

%\documentclass[12pt]{article}
\documentclass[preprint,12pt,times]{elsarticle}

% \usepackage{url}

% \usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}
%\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{bm}
%\usepackage{bbm}
%\usepackage{caption}
%\usepackage{color}
%\usepackage{geometry}
%\usepackage{graphicx}
%\usepackage[colorlinks,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
%\usepackage{layouts}
%\usepackage{mathrsfs}
%\usepackage{pgfkeys,pgfmath,pgfcore}
%\usepackage{srcltx}
%\usepackage[labelformat=simple]{subcaption}
%\usepackage{xspace}

%\renewcommand{\thetable}{\arabic{table}}
%\captionsetup[table]{labelformat=simple, labelsep=colon}

%\pgfkeys{/pgf/number format/.cd,1000 sep={}}

%% see http://tex.stackexchange.com/questions/2441/how-to-add-a-forced-line-break-inside-a-table-cell
%\newcommand{\specialcell}[2][c]{%
%  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

% put your own definitions here:
% \renewcommand\thesubfigure{(\alph{subfigure})}
\def\gz  #1{           \mbox{$\boldsymbol{#1}$}}
% \DeclareMathAlphabet{\bsf}{OT1}{cmss}{bx}{n}
% \def\msf  #1{           \mbox{\!\!      $\sf #1$}}
% \def\Grad #1 {{\rm Grad} #1}
% \def\grad #1 {{\rm grad} #1}
\def\Div {\mbox{Div\,}}
\def\div {\mbox{div\,}}
\def\d {\,\mbox{d}}
\def\D {\,\mbox{D}}
% \newcommand{\norm}[1]{\left\lvert #1 \right\rvert}
%\newcommand\normDouble[1]{\left\lVert#1\right\rVert}
%\newcommand{\normXY}[1]{\lvert #1 \rvert_{2d}}
%\newcommand{\dbracket}[1]{\left[\!\!\left[ #1 \right]\!\!\right]}

\def\mcl  #1{               {\cal #1}}
%\def\bcl  #1{\mbox{\boldmath$\cal #1$}}

\begin{document}

\section{Theoretical background}

A brief summary of balance equations in large strain hyperelastic continuum.

\subsection{Kinematics}

Deformation of the body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$
is defined via the mapping $\gz x = \gz \varphi (\gz X,t)$.

Linear deformation map is assumed for the macroscopic point
\begin{align}
\d \gz x = \gz F \cdot \d \gz X,
\end{align}
where $\gz F := \gz \nabla_{X} \, \gz \varphi$ is the deformation gradient.

\subsection{Kinetics}

For conservative systems the total potential energy is introduced as
\begin{align}
\mcl E = \int_{\mcl B_0} \mcl U_0(\gz \varphi, \gz F; \gz X) \,
\end{align}
that must be stationary in order for the system to be in equilibrium:
\begin{align}
\delta \mcl E = 0 \, .
\label{eq:stationary}
\end{align}

\section{Finite Element discretization}

We now introduce a FE triangulation $\mathcal{T}^h$ of $\mcl B_0$ and
the associated FE space of continuous piecewise elements of a fixed polynomial degree. % : $V^h \subset H^1 (\mcl B_0)$.
Macro- -deformation map are given in a
a vector space spanned by standard vector--valued FE basis functions $\gz N^i(\gz x)$ (e.g. polynomials with local support):
\begin{alignat}{2}
       \gz \varphi^h &=:  \sum_{i \in \mcl I_\varphi}       \varphi_i \gz N^i_\varphi (\gz X) \quad \quad \quad
\delta \gz \varphi^h &&=: \sum_{i \in \mcl I_\varphi} \delta \varphi_i \gz N^i_\varphi (\gz X) \,,
\end{alignat}
where superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom for the
macroscopic fields. For the sake of this study $\gz N^i_\varphi (\gz X)$ are such that only one component of the vector shape function is non-zero for each $i$.

\bibliographystyle{elsarticle-num}
\bibliography{bibliography}

\end{document}

### 2018-04-18 mf_elasticity.cc
take parameter file as input

@@ -134,7 +134,7 @@ namespace Cook_Membrane
 
 // @sect4{Finite Element system}
 
-// Here we specify the polynomial order used to approximate the solution.  
+// Here we specify the polynomial order used to approximate the solution.
 // The quadrature order should be adjusted accordingly.
     struct FESystem
     {
@@ -407,7 +407,7 @@ namespace Cook_Membrane
 
 // Finally we consolidate all of the above structures into a single container
 // that holds all of our run-time selections.
-    struct AllParameters : 
+    struct AllParameters :
       public AssemblyMethod,
       public FESystem,
       public Geometry,
@@ -546,7 +546,7 @@ namespace Cook_Membrane
 
 // @sect3{Compressible neo-Hookean material within a one-field formulation}
 
-// As discussed in the literature and step-44, Neo-Hookean materials are a type 
+// As discussed in the literature and step-44, Neo-Hookean materials are a type
 // of hyperelastic materials.  The entire domain is assumed to be composed of a
 // compressible neo-Hookean material.  This class defines the behaviour of
 // this material within a one-field formulation.  Compressible neo-Hookean
@@ -571,8 +571,8 @@ namespace Cook_Membrane
 // and provides a central point that one would need to modify if one were to
 // implement a different material model. For it to work, we will store one
 // object of this type per quadrature point, and in each of these objects
-// store the current state (characterized by the values or measures  of the 
-// displacement field) so that we can compute the elastic coefficients 
+// store the current state (characterized by the values or measures  of the
+// displacement field) so that we can compute the elastic coefficients
 // linearized around the current state.
   template <int dim,typename NumberType>
   class Material_Compressible_Neo_Hook_One_Field
@@ -689,9 +689,9 @@ namespace Cook_Membrane
     }
 
     // Calculate the volumetric part of the tangent $J
-    // \mathfrak{c}_\textrm{vol}$. Again, note the difference in its 
-    // definition when compared to step-44. The extra terms result from two 
-    // quantities in $\boldsymbol{\tau}_{\textrm{vol}}$ being dependent on 
+    // \mathfrak{c}_\textrm{vol}$. Again, note the difference in its
+    // definition when compared to step-44. The extra terms result from two
+    // quantities in $\boldsymbol{\tau}_{\textrm{vol}}$ being dependent on
[…]

### 2018-04-18 mf_elasticity.cc
setup a test

@@ -1001,7 +1001,7 @@ namespace Cook_Membrane
     triangulation(Triangulation<dim>::maximum_smoothing),
     time(parameters.end_time, parameters.delta_t),
     timer(std::cout,
-          TimerOutput::summary,
+          TimerOutput::never/*TimerOutput::summary*/,
           TimerOutput::wall_times),
     degree(parameters.poly_degree),
     // The Finite Element System is composed of dim continuous displacement

### 2018-04-18 mf_elasticity.cc
remove i_group/j_group

@@ -1662,7 +1662,6 @@ Point<dim> grid_y_transform (const Point<dim> &pt_in)
       const Parameters::AllParameters &parameters = data.solid->parameters;
       const Time &time = data.solid->time;
       const FESystem<dim> &fe = data.solid->fe;
-      const unsigned int &u_dof = data.solid->u_dof;
 
       // Next we assemble the Neumann contribution. We first check to see it the
       // cell face exists on a boundary on which a traction is applied and add
@@ -1694,22 +1693,16 @@ Point<dim> grid_y_transform (const Point<dim> &pt_in)
 
                 for (unsigned int i = 0; i < dofs_per_cell; ++i)
                   {
-                    const unsigned int i_group =
-                      fe.system_to_base_index(i).first.first;
+                    const unsigned int component_i =
+                      fe.system_to_component_index(i).first;
+                    const double Ni =
+                      scratch.fe_face_values_ref.shape_value(i,
+                                                              f_q_point);
+                    const double JxW = scratch.fe_face_values_ref.JxW(
+                                          f_q_point);
 
-                    if (i_group == u_dof)
-                      {
-                        const unsigned int component_i =
-                          fe.system_to_component_index(i).first;
-                        const double Ni =
-                          scratch.fe_face_values_ref.shape_value(i,
-                                                                 f_q_point);
-                        const double JxW = scratch.fe_face_values_ref.JxW(
-                                             f_q_point);
-
-                        data.cell_rhs(i) += (Ni * traction[component_i])
-                                            * JxW;
-                      }
+                    data.cell_rhs(i) += (Ni * traction[component_i])
+                                        * JxW;
                   }
               }
           }
@@ -1796,33 +1789,22 @@ Point<dim> grid_y_transform (const Point<dim> &pt_in)
           for (unsigned int i = 0; i < dofs_per_cell; ++i)
             {
               const unsigned int component_i = fe.system_to_component_index(i).first;
-              const unsigned int i_group     = fe.system_to_base_index(i).first.first;
 
-              if (i_group == u_dof)
-                data.cell_rhs(i) -= (symm_grad_Nx[i] * tau) * JxW;
-              else
-                Assert(i_group <= u_dof, ExcInternalError());
+              data.cell_rhs(i) -= (symm_grad_Nx[i] * tau) * JxW;
 
               for (unsigned int j = 0; j <= i; ++j)
                 {
[…]

### 2018-05-09 include/mf_elasticity.h
vmults are the same, huray-huray

@@ -931,15 +931,28 @@ Point<dim> grid_y_transform (const Point<dim> &pt_in)
         // now ready to go-on and assmble linearized problem around solution_n + solution_delta for this iteration.
         assemble_system();
 
-        // DEBUG: check vmult of matrix-based and matrix-free
+#ifdef DEBUG
+        // check vmult of matrix-based and matrix-free for a random vector:
         {
-          Vector<double> src(dof_handler_ref.n_dofs()), dst_mb(dof_handler_ref.n_dofs()), dst_mf(dof_handler_ref.n_dofs());
+          Vector<double> src(dof_handler_ref.n_dofs()), dst_mb(dof_handler_ref.n_dofs()), dst_mf(dof_handler_ref.n_dofs()), diff(dof_handler_ref.n_dofs());
           for (unsigned int i=0; i<dof_handler_ref.n_dofs(); ++i)
             src(i) = ((double)std::rand())/RAND_MAX;
 
+          constraints.set_zero(src);
+
           tangent_matrix.vmult(dst_mb, src);
           mf_nh_operator.vmult(dst_mf, src);
+
+          diff = dst_mb;
+          diff.add(-1, dst_mf);
+          Assert (diff.l2_norm() < 1e-10 * dst_mb.l2_norm(),
+                  ExcMessage("MF and MB are different " +
+                             std::to_string(diff.l2_norm()) +
+                             " at Newton iteration " +
+                             std::to_string(newton_iteration)
+                            ));
         }
+#endif
 
         get_error_residual(error_residual);
 

### 2018-05-09 include/mf_nh_operator.h

@@ -239,6 +239,7 @@ using namespace dealii;
   {
     // FIXME: can't use cell_loop as we need both matrix-free data objects.
     // for now do it by hand.
+    // BUT I might try cell_loop(), and simply use another MF object inside...
     FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> phi_current  (*data_current);
     FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> phi_current_s(*data_current);
     FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> phi_reference(*data_reference);
@@ -266,7 +267,7 @@ using namespace dealii;
         phi_reference.reinit(cell);
 
         // read-in total displacement and src vector and evaluate gradients
-        phi_reference.read_dof_values(*displacement);
+        phi_reference.read_dof_values_plain(*displacement);
         phi_current.  read_dof_values(src);
         phi_current_s.read_dof_values(src);
         phi_reference.evaluate (false,true,false);

### 2018-05-09 doc/large_strain.tex
outline of the paper

@@
 
+\section{Introduction}
+
+TODO: elasticity and no matrix-free. fluids used a lot.
+
 \section{Theoretical background}
@@
 
+\subsection{Linearization}
+
+TODO: linearize and show how tangent look like.
+Also show what it boils down in the case of small strain.
+
 \section{Finite Element discretization}
@@
 
+\section{Matrix-free operator evaluation}
+
+TODO[DD]: summary of what's happening, how operator is evaluated, what we do, what operations (determinant, etc), count number of FLOPS
+
+\section{Numerical examples}
+
+\subsection{Small strain linear elasticity}
+
+TODO: summarize Mathias results (head and Cook's membrane)
+
+TODO: switch head to quasi-static load instead of dynamics?
+
+\subsection{Large strain Neo-Hook elasticity}
+
+TODO: add new results from Cook's and some matrix-inclusion (matrix-pore).
+Eliptical holes are easier as we don't need heterogeneous operator.
+
+\section{Summary and Conclusions}
+
+TODO: summarize results, good/bad, where can be applied,...
+Bio-med fluid-structure ?
+
 \bibliographystyle{elsarticle-num}

### 2018-05-09 include/mf_nh_operator.h
refactor the core of MF into do_operation_on_cell()

@@ -60,6 +60,15 @@ using namespace dealii;
                       const std::pair<unsigned int,unsigned int> &cell_range) const;
     */
 
+   /**
+    * Perform operation on a cell. @p phi_current and @phi_current_s correspond to the deformed configuration
+    * where @p phi_reference is for the current configuration.
+    */
+   void do_operation_on_cell(FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> &phi_current,
+                             FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> &phi_current_s,
+                             FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> &phi_reference,
+                             const unsigned int cell) const;
+
     std::shared_ptr<const MatrixFree<dim,number>> data_current;
     std::shared_ptr<const MatrixFree<dim,number>> data_reference;
 
@@ -244,7 +253,6 @@ using namespace dealii;
     FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> phi_current_s(*data_current);
     FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> phi_reference(*data_reference);
 
-    const unsigned int n_q_points = phi_current.n_q_points;
     const unsigned int n_cells = data_current->n_macro_cells();
 
     Assert (n_cells == data_reference->n_macro_cells(), ExcInternalError());
@@ -270,56 +278,8 @@ using namespace dealii;
         phi_reference.read_dof_values_plain(*displacement);
         phi_current.  read_dof_values(src);
         phi_current_s.read_dof_values(src);
-        phi_reference.evaluate (false,true,false);
-        phi_current.  evaluate (false,true,false);
-        phi_current_s.evaluate (false,true,false);
 
-        for (unsigned int q=0; q<n_q_points; ++q)
-          {
-            // reference configuration:
-            const Tensor<2,dim,VectorizedArray<number>>         &grad_u = phi_reference.get_gradient(q);
-            const Tensor<2,dim,VectorizedArray<number>>          F      = Physics::Elasticity::Kinematics::F(grad_u);
-            const VectorizedArray<number>                        det_F  = determinant(F);
-            const Tensor<2,dim,VectorizedArray<number>>          F_bar  = Physics::Elasticity::Kinematics::F_iso(F);
-            const SymmetricTensor<2,dim,VectorizedArray<number>> b_bar  = Physics::Elasticity::Kinematics::b(F_bar);
-
[…]

### 2018-05-09 include/mf_nh_operator.h
move phi.integrate() outside of quadrature points loop

@@ -351,13 +351,11 @@ using namespace dealii;
           // phi_reference.JxW(q) / phi_current.JxW(q)
           ,q);
 
-        // actually do the contraction
-        // FIXME: that should be outside of the loop!
-        phi_current.integrate (false,true);
-        phi_current_s.integrate (false,true);
-
       } // end of the loop over quadrature points
 
+    // actually do the contraction
+    phi_current.integrate (false,true);
+    phi_current_s.integrate (false,true);
   }
 
 

### 2018-05-09 include/mf_nh_operator.h
add local_apply_cell()

@@ -53,12 +53,14 @@ using namespace dealii;
     */
 
   private:
-    /*
-    void local_apply (const MatrixFree<dim,number>    &data,
-                      Vector<double>                      &dst,
-                      const Vector<double>                &src,
-                      const std::pair<unsigned int,unsigned int> &cell_range) const;
-    */
+
+    /**
+     * Apply operator on a range of cells.
+     */
+    void local_apply_cell (const MatrixFree<dim,number>    &data,
+                           Vector<double>                      &dst,
+                           const Vector<double>                &src,
+                           const std::pair<unsigned int,unsigned int> &cell_range) const;
 
    /**
     * Perform operation on a cell. @p phi_current and @phi_current_s correspond to the deformed configuration
@@ -249,14 +251,8 @@ using namespace dealii;
     // FIXME: can't use cell_loop as we need both matrix-free data objects.
     // for now do it by hand.
     // BUT I might try cell_loop(), and simply use another MF object inside...
-    FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> phi_current  (*data_current);
-    FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> phi_current_s(*data_current);
-    FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> phi_reference(*data_reference);
 
-    const unsigned int n_cells = data_current->n_macro_cells();
-
-    Assert (n_cells == data_reference->n_macro_cells(), ExcInternalError());
-    Assert (phi_current.n_q_points == phi_reference.n_q_points, ExcInternalError());
+    Assert (data_current->n_macro_cells() == data_reference->n_macro_cells(), ExcInternalError());
 
     // MatrixFree::cell_loop() is more complicated than a simple update_ghost_values() / compress(),
     // it loops on different cells (inner without ghosts and outer) in different order
@@ -267,7 +263,38 @@ using namespace dealii;
     // src.update_ghost_values();
 
     // 2. loop over all locally owned cell blocks
-    for (unsigned int cell=0; cell<n_cells; ++cell)
+    local_apply_cell(*data_current, dst, src,
+                     std::make_pair<unsigned int,unsigned int>(0,data_current->n_macro_cells()));
+
+    // 3. communicate results with MPI
+    // dst.compress(VectorOperation::add);
+
+    // 4. constraints
[…]

### 2018-05-15 include/mf_nh_operator.h
cach jacobian to the power -1/dim

@@ -58,6 +58,11 @@ using namespace dealii;
                              const Vector<number> &src,
                              const number omega) const;
 
+    /**
+     * Cache a few things for the current displacement.
+     */
+    void cache();
+
   private:
 
     /**
@@ -95,6 +100,8 @@ using namespace dealii;
 
     std::shared_ptr<DiagonalMatrix<Vector<number>>>  inverse_diagonal_entries;
     std::shared_ptr<DiagonalMatrix<Vector<number>>>  diagonal_entries;
+
+    Table<2,VectorizedArray<number>> cached_jacobian;
 
     bool            diagonal_is_available;
   };
@@ -165,6 +172,36 @@ using namespace dealii;
     data_current = data_current_;
     data_reference = data_reference_;
     displacement = &displacement_;
+
+
+    const unsigned int n_cells = data_reference_->n_macro_cells();
+    FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> phi (*data_reference_);
+    cached_jacobian.reinit(n_cells,phi.n_q_points);
+  }
+
+
+
+  template <int dim, int fe_degree, int n_q_points_1d, typename number>
+  void
+  NeoHookOperator<dim,fe_degree,n_q_points_1d,number>::cache()
+  {
+    const unsigned int n_cells = data_reference->n_macro_cells();
+
+    FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> phi_reference(*data_reference);
+
+    for (unsigned int cell = 0; cell < n_cells; ++cell)
+      {
+        phi_reference.reinit(cell);
+        phi_reference.read_dof_values_plain(*displacement);
+        phi_reference.evaluate (false,true,false);
+        for (unsigned int q=0; q<phi_reference.n_q_points; ++q)
+          {
+            const Tensor<2,dim,VectorizedArray<number>>         &grad_u = phi_reference.get_gradient(q);
+            const Tensor<2,dim,VectorizedArray<number>>          F      = Physics::Elasticity::Kinematics::F(grad_u);
+            const VectorizedArray<number>                        det_F  = determinant(F);
+            cached_jacobian(cell,q) = std::pow(det_F,-1.0/dim);
+          }
+      }
   }
 
 
@@ -362,7 +399,7 @@ using namespace dealii;
                              FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> &phi_current,
                              FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> &phi_current_s,
[…]

### 2018-05-15 doc/large_strain.tex
mention GMG in tex
add a note on caching for large strain

@@
 
+\section{Geometric multigrid preconditioning}
+
+TODO[DD]: explain how this works with GMG (eulerian mapping on fine level gets restricted onto the coarse levels, the rest works as-is).
+Hopefully we can still use some mappings on coarse level to better represent geometric of inclusions (pores).
+
 \section{Numerical examples}
+
+TODO: question to asnwer -- how much to cache? At least detF ^ {2/3}. Other options is stress and strain at quadrature points.
+Run benchmarks and see.
 

### 2018-05-15 include/mf_nh_operator.h
spell out the whole tangent operator

@@ -401,6 +401,11 @@ using namespace dealii;
                              FEEvaluation<dim,fe_degree,n_q_points_1d,dim,number> &phi_reference,
                              const unsigned int cell) const
   {
+    typedef VectorizedArray<number> NumberType;
+    static constexpr number dim_f = dim;
+    const number kappa = material->kappa;
+    const number c_1   = material->c_1;
+
     phi_reference.evaluate (false,true,false);
     phi_current.  evaluate (false,true,false);
     phi_current_s.evaluate (false,true,false);
@@ -408,21 +413,109 @@ using namespace dealii;
     for (unsigned int q=0; q<phi_current.n_q_points; ++q)
       {
         // reference configuration:
-        const Tensor<2,dim,VectorizedArray<number>>         &grad_u = phi_reference.get_gradient(q);
-        const Tensor<2,dim,VectorizedArray<number>>          F      = Physics::Elasticity::Kinematics::F(grad_u);
-        const VectorizedArray<number>                        det_F  = determinant(F);
-        const Tensor<2,dim,VectorizedArray<number>>          F_bar  = F * cached_jacobian(cell,q);
-        const SymmetricTensor<2,dim,VectorizedArray<number>> b_bar  = Physics::Elasticity::Kinematics::b(F_bar);
+        const Tensor<2,dim,NumberType>         &grad_u = phi_reference.get_gradient(q);
+        const Tensor<2,dim,NumberType>          F      = Physics::Elasticity::Kinematics::F(grad_u);
+        const NumberType                        det_F  = determinant(F);
+        const Tensor<2,dim,NumberType>          F_bar  = F * cached_jacobian(cell,q);
+        const SymmetricTensor<2,dim,NumberType> b_bar  = Physics::Elasticity::Kinematics::b(F_bar);
 
         // current configuration
-        const Tensor<2,dim,VectorizedArray<number>>          &grad_Nx_v      = phi_current.get_gradient(q);
-        const SymmetricTensor<2,dim,VectorizedArray<number>> &symm_grad_Nx_v = phi_current.get_symmetric_gradient(q);
+        const Tensor<2,dim,NumberType>          &grad_Nx_v      = phi_current.get_gradient(q);
[…]

### 2018-05-15 include/mf_nh_operator.h
minor cleanup

@@ -403,6 +403,7 @@ using namespace dealii;
   {
     typedef VectorizedArray<number> NumberType;
     static constexpr number dim_f = dim;
+    static constexpr number two_over_dim = 2.0/dim;
     const number kappa = material->kappa;
     const number c_1   = material->c_1;
 
@@ -460,8 +461,6 @@ using namespace dealii;
           tau += tau_bar;
         }
 
-        const Tensor<2,dim,VectorizedArray<number>> tau_ns (tau);
-
         // material part of the action of tangent:
         // The action of the fourth-order material elasticity tensor in the spatial setting
         // on symmetric tensor.
@@ -500,19 +499,18 @@ using namespace dealii;
           // The isochoric Kirchhoff stress
           // $\boldsymbol{\tau}_{\textrm{iso}} =
           // \mathcal{P}:\overline{\boldsymbol{\tau}}$:
-          SymmetricTensor<2,dim,NumberType> tau_iso(b_bar);
-          tau_iso = tau_iso * (2.0 * c_1);
+          SymmetricTensor<2,dim,NumberType> tau_iso(tau_bar);
           for (unsigned int i = 0; i < dim; ++i)
             tau_iso[i][i] -= tr_tau_bar_dim;
 
           // term with deviatoric part of the tensor
-          jc_part += ((2.0 / dim) * tr_tau_bar) * dev_src;
+          jc_part += (two_over_dim * tr_tau_bar) * dev_src;
 
           // term with tau_iso_x_I + I_x_tau_iso
-          jc_part -= ((2.0 / dim) * tr) * tau_iso;
+          jc_part -= (two_over_dim * tr) * tau_iso;
           const NumberType tau_iso_src = tau_iso * symm_grad_Nx_v;
           for (unsigned int i = 0; i < dim; ++i)
-            jc_part[i][i] -= (2.0 / dim) * tau_iso_src;
+            jc_part[i][i] -= two_over_dim * tau_iso_src;
 
           // c_bar==0 so we don't have a term with it.
         }
@@ -535,6 +533,7 @@ using namespace dealii;
         // geometrical stress contribution
         // In index notation this tensor is $ [j e^{geo}]_{ijkl} = j \delta_{ik} \sigma^{tot}_{jl} = \delta_{ik} \tau^{tot}_{jl} $.
         // the product is actually  GradN * tau^T but due to symmetry of tau we can do GradN * tau
[…]

### 2018-05-18 doc/large_strain.tex
Add some text outlining the constitutive model

@@
 
+%\DeclareMathOperator{\det}{det}
+\DeclareMathOperator{\trace}{tr}
+\newcommand{\diff}{\mathop{}\!\mathrm{d}}
+
 \begin{document}
@@
 
+\subsection{Constitutive modelling}
+
+Neo-Hookean model
+\begin{gather}
+\psi \left( \mathbf{C} \right)
+  = \frac{\mu}{2} \left[ \trace{\mathbf{C}} - \trace{\mathbf{I}} - 2 \ln\left( J \right) \right]
+  - \lambda \ln^{2}\left( J \right)
+\end{gather}
+where $\mu$ and $\lambda$ respectively denote the shear modulus and Lam\'{e} parameter,
+and the volumetric Jacobian $J = \det\left(\mathbf{F}\right) = \sqrt{\det\left(\mathbf{C}\right)}$.
+\begin{gather}
+\frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}}
+  = \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1}
+\end{gather}
+\begin{gather}
+\frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
+  = \left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
+  + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
+\end{gather}
+
+Denoting $\chi\left( \bullet \right)$ as the Piola transformation, which implies
+\begin{gather}
+\chi\left( \mathbf{A} \right)_{ij}
+  = F_{iA} A_{AB} F_{jB} \\
+\chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
+  = F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD} 
+\end{gather}
+for rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$, then the Kirchhoff stress and its associated material tangent are
+\begin{gather}
+\boldsymbol{\tau}
+  \equiv J \boldsymbol{\sigma}
+  = \chi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
+  = \mu \mathbf{b} - \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{I}
+\end{gather}
+and 
+\begin{gather}
+J \boldsymbol{\mathcal{C}}
+  = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
+  = \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
+  + \frac{\lambda}{2} \mathbf{I} \otimes \mathbf{I}
+\end{gather}
+where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor.
+The action that $J \boldsymbol{\mathcal{C}}$ performs when contracted with an arbitrary rank-2 symmetric tensor is therefore
+\begin{gather}
+J \boldsymbol{\mathcal{C}} : \left( \bullet \right)
+  = \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
+  + \frac{\lambda}{2} \trace\left( \bullet \right) \mathbf{I}
+\end{gather}
+
 \section{Finite Element discretization}
@@
 
-TODO: question to asnwer -- how much to cache? At least detF ^ {2/3}. Other options is stress and strain at quadrature points.
+TODO: question to asnwer -- how much to cache? At least $detF ^ {2/3}$. Other options is stress and strain at quadrature points.
 Run benchmarks and see.

### 2018-08-01 include/mf_elasticity.h
fix missing material_id setup in Cook's example

@@ -1053,19 +1053,23 @@ namespace Cook_Membrane
         // as in the global refinement cells will be repartitioned and faces of
         // their parents should have right IDs
         for (auto cell : triangulation.active_cell_iterators())
-          for (unsigned int face = 0; face < GeometryInfo<dim>::faces_per_cell;
-              ++face)
-            if (cell->face(face)->at_boundary() == true)
-              {
-                if (std::abs(cell->face(face)->center()[0] - 0.0) < tol_boundary)
-                  cell->face(face)->set_boundary_id(1); // -X faces
-                else if (std::abs(cell->face(face)->center()[0] - 48.0) <
-                        tol_boundary)
-                  cell->face(face)->set_boundary_id(11); // +X faces
-                else if (std::abs(std::abs(cell->face(face)->center()[0]) - 0.5) <
-                        tol_boundary)
-                  cell->face(face)->set_boundary_id(2); // +Z and -Z faces
-              }
+          {
+            for (unsigned int face = 0; face < GeometryInfo<dim>::faces_per_cell;
+                ++face)
+              if (cell->face(face)->at_boundary() == true)
+                {
+                  if (std::abs(cell->face(face)->center()[0] - 0.0) < tol_boundary)
+                    cell->face(face)->set_boundary_id(1); // -X faces
+                  else if (std::abs(cell->face(face)->center()[0] - 48.0) <
+                          tol_boundary)
+                    cell->face(face)->set_boundary_id(11); // +X faces
+                  else if (std::abs(std::abs(cell->face(face)->center()[0]) - 0.5) <
+                          tol_boundary)
+                    cell->face(face)->set_boundary_id(2); // +Z and -Z faces
+                }
+            // on the coarse mesh reset material ID
+            cell->set_material_id(0);
+          }
 
         // Transform the hyper-rectangle into the beam shape
         GridTools::transform(&grid_y_transform<dim>, triangulation);

### 2018-08-01 include/mf_nh_operator.h
extend MF operator to support heterogeneity

@@ -250,11 +250,15 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::cache()
 
   for (unsigned int cell = 0; cell < n_cells; ++cell)
     {
+      const unsigned int material_id =
+        data_current->get_cell_iterator(cell, 0)->material_id();
+      const auto &cell_mat = (material_id == 0 ? material : material_inclusion);
+
       phi_reference.reinit(cell);
       phi_reference.read_dof_values_plain(*displacement);
       phi_reference.evaluate(false, true, false);
 
-      if (material->formulation == 0)
+      if (cell_mat->formulation == 0)
         {
           for (unsigned int q = 0; q < phi_reference.n_q_points; ++q)
             {
@@ -275,7 +279,7 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::cache()
               cached_scalar(cell, q) = std::pow(det_F, number(-1.0 / dim));
             }
         }
-      else if (material->formulation == 1)
+      else if (cell_mat->formulation == 1)
         {
           for (unsigned int q = 0; q < phi_reference.n_q_points; ++q)
             {
@@ -294,7 +298,7 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::cache()
                          "] is not positive: " + std::to_string(det_F[i])));
 
               cached_scalar(cell, q) =
-                material->mu - 2.0 * material->lambda * std::log(det_F);
+                cell_mat->mu - 2.0 * cell_mat->lambda * std::log(det_F);
             }
         }
       else
@@ -538,6 +542,27 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::do_operation_on_cell(
   FEEvaluation<dim, fe_degree, n_q_points_1d, dim, number> &phi_reference,
   const unsigned int                                        cell) const
 {
+  const unsigned int material_id = data_current->get_cell_iterator(cell, 0)->material_id();
+  const auto &cell_mat =
+    (material_id == 0 ?
+       material :
+       material_inclusion);
+
+  // make sure all filled cells have the same ID
+#ifdef DEBUG
+  for (unsigned int i = 1; i < data_current->n_components_filled(cell); ++i)
+    {
+      const unsigned int other_id =
+        data_current->get_cell_iterator(cell, i)->material_id();
+      Assert(other_id == material_id,
+             ExcMessage("Cell block " + std::to_string(cell) + " element " +
+                        std::to_string(i) + " has material ID " +
+                        std::to_string(other_id) +
[…]

### 2018-08-23 doc/large_strain.tex
summarize current formation=1 algorithm

@@
 %\usepackage{xspace}
+
+\usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}
 
@@
   = \frac{\mu}{2} \left[ \trace{\mathbf{C}} - \trace{\mathbf{I}} - 2 \ln\left( J \right) \right]
-  - \lambda \ln^{2}\left( J \right)
+  + \lambda \ln^{2}\left( J \right)
 \end{gather}
@@
 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
-  = \left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
+  = \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
   + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
@@
   = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
-  = \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
-  + \frac{\lambda}{2} \mathbf{I} \otimes \mathbf{I}
+  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
+  + 2 \lambda \mathbf{I} \otimes \mathbf{I}
 \end{gather}
@@
 J \boldsymbol{\mathcal{C}} : \left( \bullet \right)
-  = \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
-  + \frac{\lambda}{2} \trace\left( \bullet \right) \mathbf{I}
+  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
+  + 2 \lambda \trace\left( \bullet \right) \mathbf{I}
 \end{gather}
@@
 \section{Matrix-free operator evaluation}
+
+Due to the specifics of matrix-free operator evaluation in \texttt{deal.II}, when performing contraction with gradient of shape functions in current (deformed) configuration, the integration is also done over the deformed configuration.
+Therefore in order to arrive at the integral in undeformed configuration, additionally we have to divide by Jacobian $J$ of the mapping at a given quadrature point.
+
+\begin{algorithm}[h]
+  \SetKwInOut{Input}{Given}
+  \SetKwInOut{Output}{Return}
+  \Input{RHS FE vector $\gz x$, current FE solution $\gz u$,
+  cacheed $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
+  \Output{action of the FE tangent on $\gz x$}
+  \ForEach{ element $K \in \Omega^h$ }{
+        \ForEach{quadrature point $q$ on $K$}{
+          evaluate $\gz \nabla_X \gz u^h$ \tcp*{2nd order}
+          evaluate $\gz F = \gz I + \gz \nabla_X \gz u^h$ \tcp*{2nd order}
+          evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
+          evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
+          evaluate $\gz x_1 := \gz \nabla \gz x$ \tcp*{2nd order}
+          evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
+          evaluate $J \boldsymbol{\mathcal{G}} = x_1 \cdot \gz \tau$ \tcp*{2nd order symmetric}
+          ``submit'' contraction $\left[\gz \nabla \gz N_i : J\boldsymbol{\mathcal{G}}\right] / J$\;
+          evaluate $\gz x_2 := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
+          evaluate $J \boldsymbol{\mathcal{C}} = 2 c_1 \gz x_2 + 2 \lambda \rm{tr}( \gz x_2) \gz I $ \tcp*{2nd order symmetric}
+          ``submit'' contraction $\left[\gz \nabla_{sym} \gz N_i : J \boldsymbol{\mathcal{C}} \right]/ J$ \;
+        }
+  }
+  \caption{Matrix-free tangent cell operator, variant 1: minimum caching.}
+  \label{alg:mf_multivector}
+\end{algorithm}
 

### 2018-08-23 include/mf_elasticity.h
implement different caching strategies

@@ -292,6 +292,7 @@ namespace Cook_Membrane
       std::string  preconditioner_type;
       double       preconditioner_relaxation;
       unsigned int cond_number_cg_iterations;
+      std::string  mf_caching;
 
       static void
       declare_parameters(ParameterHandler &prm);
@@ -336,6 +337,12 @@ namespace Cook_Membrane
           "20",
           Patterns::Integer(1),
           "Number of CG iterations to estimate condition number");
+
+        prm.declare_entry("MF caching",
+                  "scalar",
+                  Patterns::Selection("scalar|tensor2|tensor4"),
+                  "Type of preconditioner");
+
       }
       prm.leave_subsection();
     }
@@ -352,6 +359,7 @@ namespace Cook_Membrane
         preconditioner_relaxation = prm.get_double("Preconditioner relaxation");
         cond_number_cg_iterations =
           prm.get_integer("Condition number CG iterations");
+        mf_caching                = prm.get("MF caching");
       }
       prm.leave_subsection();
     }
@@ -1443,7 +1451,8 @@ namespace Cook_Membrane
 
         mf_nh_operator.initialize(mf_data_current,
                                   mf_data_reference,
-                                  solution_total);
+                                  solution_total,
+                                  parameters.mf_caching);
         mf_ad_nh_operator.initialize(mf_data_current,
                                      mf_data_reference,
                                      solution_total);
@@ -1491,7 +1500,8 @@ namespace Cook_Membrane
             mg_mf_nh_operator[level].initialize(
               mg_mf_data_current[level],
               mg_mf_data_reference[level],
-              mg_solution_total[level]); // (mg_level_data, mg_constrained_dofs,
+              mg_solution_total[level],
+              parameters.mf_caching); // (mg_level_data, mg_constrained_dofs,
                                          // level);
           }
       }

### 2018-08-23 include/mf_nh_operator.h

@@ -54,7 +54,8 @@ public:
   void
   initialize(std::shared_ptr<const MatrixFree<dim, number>>    data_current,
              std::shared_ptr<const MatrixFree<dim, number>>    data_reference,
-             const LinearAlgebra::distributed::Vector<number> &displacement);
+             const LinearAlgebra::distributed::Vector<number> &displacement,
+             const std::string caching);
 
   void
   set_material(
@@ -158,9 +159,14 @@ private:
   std::shared_ptr<DiagonalMatrix<LinearAlgebra::distributed::Vector<number>>>
     diagonal_entries;
 
-  Table<2, VectorizedArray<number>> cached_scalar;
+  Table<2, VectorizedArray<number>>                 cached_scalar;
+  Table<2, VectorizedArray<number>>                 cached_second_scalar;
+  Table<2, Tensor<2, dim, VectorizedArray<number>>> cached_tensor2;
+  Table<2, SymmetricTensor<4, dim, VectorizedArray<number>>> cached_tensor4;
 
   bool diagonal_is_available;
+
+  std::string mf_caching;
 };
 
 
@@ -169,6 +175,7 @@ template <int dim, int fe_degree, int n_q_points_1d, typename number>
 NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::NeoHookOperator()
   : Subscriptor()
   , diagonal_is_available(false)
+  , mf_caching("")
 {}
 
 
@@ -224,17 +231,36 @@ void
 NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::initialize(
   std::shared_ptr<const MatrixFree<dim, number>>    data_current_,
   std::shared_ptr<const MatrixFree<dim, number>>    data_reference_,
-  const LinearAlgebra::distributed::Vector<number> &displacement_)
+  const LinearAlgebra::distributed::Vector<number> &displacement_,
+  const std::string caching)
 {
   data_current   = data_current_;
   data_reference = data_reference_;
   displacement   = &displacement_;
-
+  mf_caching     = caching;
 
   const unsigned int n_cells = data_reference_->n_macro_cells();
   FEEvaluation<dim, fe_degree, n_q_points_1d, dim, number> phi(
     *data_reference_);
-  cached_scalar.reinit(n_cells, phi.n_q_points);
+  if (mf_caching == "scalar")
+    {
+      cached_scalar.reinit(n_cells, phi.n_q_points);
+    }
[…]

### 2018-08-24 doc/large_strain.tex
add more caching algorithms plus minor cleanup

@@
 %\usepackage{caption}
-%\usepackage{color}
+\usepackage{color}
 %\usepackage{geometry}
@@
 %\usepackage{xspace}
-
 \usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}
@@
 
+{\color{red}
+[DA] Daniel, please add a brief intro about MF (readership will likely not know anything about it), maybe borrow 1-2 pictures to explain how tensor-product structure is handled.
+}
+
 Due to the specifics of matrix-free operator evaluation in \texttt{deal.II}, when performing contraction with gradient of shape functions in current (deformed) configuration, the integration is also done over the deformed configuration.
@@
           evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
-          evaluate $\gz x_1 := \gz \nabla \gz x$ \tcp*{2nd order}
+          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
           evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
-          evaluate $J \boldsymbol{\mathcal{G}} = x_1 \cdot \gz \tau$ \tcp*{2nd order symmetric}
-          ``submit'' contraction $\left[\gz \nabla \gz N_i : J\boldsymbol{\mathcal{G}}\right] / J$\;
-          evaluate $\gz x_2 := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
-          evaluate $J \boldsymbol{\mathcal{C}} = 2 c_1 \gz x_2 + 2 \lambda \rm{tr}( \gz x_2) \gz I $ \tcp*{2nd order symmetric}
-          ``submit'' contraction $\left[\gz \nabla_{sym} \gz N_i : J \boldsymbol{\mathcal{C}} \right]/ J$ \;
+          evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
+          ``submit'' contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
+          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
+          evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
+          ``submit'' contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
         }
   }
-  \caption{Matrix-free tangent cell operator, variant 1: minimum caching.}
-  \label{alg:mf_multivector}
+  \caption{Matrix-free tangent cell operator: scalar cache}
+  \label{alg:mf_scalar}
 \end{algorithm}
+
+\begin{algorithm}[h]
+  \SetKwInOut{Input}{Given}
+  \SetKwInOut{Output}{Return}
+  \Input{RHS FE vector $\gz x$,
+  cached
+  $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
+  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point}
+  \Output{action of the FE tangent on $\gz x$}
+  \ForEach{ element $K \in \Omega^h$ }{
+        \ForEach{quadrature point $q$ on $K$}{
+          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
+          evaluate $\boldsymbol{\mathcal{G}}\gz g = \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
+          ``submit'' contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
+          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
+          evaluate $\boldsymbol{\mathcal{C}} \gz g_s = c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
+          ``submit'' contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
+        }
+  }
+  \caption{Matrix-free tangent cell operator, variant 1: cache second order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate refential quantities like $\gz F$ at runtime.}
+  \label{alg:mf_tensor2}
+\end{algorithm}
+
+\begin{algorithm}[h]
+  \SetKwInOut{Input}{Given}
+  \SetKwInOut{Output}{Return}
+  \Input{RHS FE vector $\gz x$,
+  cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point}
+  \Output{action of the FE tangent on $\gz x$}
+  \ForEach{ element $K \in \Omega^h$ }{
+        \ForEach{quadrature point $q$ on $K$}{
+          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
+          evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
+          ``submit'' contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
+          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
+          evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
+          ``submit'' contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
+        }
+  }
+  \caption{Matrix-free tangent cell operator: cache fourth order material tangent $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
+  \label{alg:mf_tensor4}
+\end{algorithm}
+
+The primary difference between Algorithm \ref{alg:mf_tensor2} and Algorithm \ref{alg:mf_tensor4}, is that the former utilizes the chosen structure of free energy function and thereby ends up with a computationally cheap way of evaluation of $\boldsymbol{\mathcal{C}} \gz g_s $, whereas the latter performs the double contraction between the fourth order and second order tensors.
 

### 2018-08-29 Calculations/emmy_benchmark_intel.sh
incrase requested time more (7 hours)

@@ -1,8 +1,8 @@
 #!/bin/bash
-#   allocate 1 nodes with 40 CPU per node for 6 hours:
-#PBS -l nodes=1:ppn=40,walltime=6:00:00
+#   allocate 1 nodes with 40 CPU per node for 7 hours:
+#PBS -l nodes=1:ppn=40,walltime=7:00:00
 #   job name
-#PBS -N dealii_large_strain
+#PBS -N dealii_mf_cm
 #   stdout and stderr files:
 #PBS -o /home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/_build/bench.txt -e /home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/_build/bench_error.txt
 #   first non-empty non-comment line ends PBS options

### 2018-08-29 include/mf_elasticity.h
optionally disable output (for benchmarks)

@@ -120,6 +120,8 @@ namespace Cook_Membrane
     struct Misc
     {
       std::string output_folder;
+      bool output_solution;
+
       static void
       declare_parameters(ParameterHandler &prm);
 
@@ -136,6 +138,10 @@ namespace Cook_Membrane
                           "",
                           Patterns::Anything(),
                           "Output folder (must exist)");
+        prm.declare_entry("Output solution",
+                          "true",
+                          Patterns::Bool(),
+                          "Output solution and mesh");
       }
       prm.leave_subsection();
     }
@@ -148,6 +154,8 @@ namespace Cook_Membrane
         output_folder = prm.get("Output folder");
         if (!output_folder.empty() && output_folder.back() != '/')
           output_folder += "/";
+
+        output_solution = prm.get_bool("Output solution");
       }
       prm.leave_subsection();
     }
@@ -2660,6 +2668,9 @@ namespace Cook_Membrane
   void
   Solid<dim, degree, n_q_points_1d, NumberType>::output_results() const
   {
+    if (!parameters.output_solution)
+      return;
+
     DataOut<dim> data_out;
     std::vector<DataComponentInterpretation::DataComponentInterpretation>
       data_component_interpretation(

### 2018-08-29 doc/large_strain.tex

@@
+\RequirePackage[l2tabu, orthodox]{nag}
 %\documentclass[12pt]{article}
@@
 %\usepackage{caption}
-\usepackage{color}
+\usepackage{xcolor}
 %\usepackage{geometry}
@@
 %\usepackage[colorlinks,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
+\usepackage{listings}
+\lstset{ %
+   backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
+   basicstyle=\small,        % the size of the fonts that are used for the code
+   breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
+   breaklines=false,                 % sets automatic line breaking
+   captionpos=b,                    % sets the caption-position to bottom
+   %commentstyle=\color{mygreen},    % comment style
+   deletekeywords={...},            % if you want to delete keywords from the given language
+   escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
+   extendedchars=false,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
+   frame=single,	                   % adds a frame around the code
+   keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
+   keywordstyle=\color{blue},       % keyword style
+   language=C++,                 % the language of the code
+   otherkeywords={*,...},           % if you want to add more keywords to the set
+   numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
+   numbersep=5pt,                   % how far the line-numbers are from the code
+   numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
+   rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
+   showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
+   showstringspaces=false,          % underline spaces within strings only
+   showtabs=false,                  % show tabs within strings adding particular underscores
+   stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
+%   stringstyle=\color{mymauve},     % string literal style
+   tabsize=2,	                   % sets default tabsize to 2 spaces
+   title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
+}
 %\usepackage{layouts}
@@
 \section{Matrix-free operator evaluation}
+Classically the approach for finding a discrete solution to a linearized partial differential equation can be described as follows:
+First, ansatz spaces and a suitable bilinear form is defined. Then, a matrix and force vector corresponding to the discretization are assembled.
+Finally, a direct or iterative solver is used to solve the linear system for the discretized solution.
+On modern computer architectures however, it turns out that loading data is significantly slower than performing the arithmetic operations when solving.
+For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations in the solver on-the-fly
+rather than loading matrix elements from memory.
+For iterative solvers, this is possible since it is sufficient to compute matrix vector products. More precisely, the linear system corresponding to a bilinear form 
+$a\colon V_h\times V_h\to\mathbb{R}$ and right-hand side $f\colon V_h\to\mathbb{R}$ defined with respect to the ansatz space $V_h$ can be represented by
+\begin{align*}
+ &&a(v,u) &= f(v) && \forall v\in V_h\\
+ \Longleftrightarrow &&a(\varphi_i, u) &= f(\varphi_i) && \forall \varphi_i \\
+ \Longleftrightarrow &&\sum_j a(\varphi_i,\varphi_j) u_j &= f(\varphi_i) && \forall \varphi_i \\
+ \Longleftrightarrow &&A x &= b
+\end{align*}
+where $A_{i,j}=a(\varphi_i, \varphi_j)$ and $b_i= f(\varphi_i)$.
+Thus, the matrix-vector product $Ax$ can be expressed by
+\begin{align*}
+ (Ax)_i &= \sum_j a(\varphi_i,\varphi_j) x_j \\
+        &= \sum_K\sum_q \sum_j \tilde{a}_K(\varphi_i,\varphi_j)(x_q) x_j w_q 
+\end{align*}
+where in the last step $\tilde{a}_K(\varphi_i,\varphi_j)$ is a representation for the evaluation of the bilinear in one quadrature point and a quadrature rule is used.
+For the sake of demonstration, let us assume $A$ represents a mass matrix. In this case, it holds 
+\begin{align*}
+ \tilde{a}_K(\varphi_i,\varphi_j)(x_q) = \varphi_i(x_q)\varphi_j(x_q).
+\end{align*}
+Further assuming that the number of the one-dimensional quadrature points $n$ matches the degree of the ansatz space plus one, the evaluation of all the shape functions ($n^d$) 
+in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in case we consider a problem in $d$ space dimensions. For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$
+and this is the same as for a regular matrix-vector where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, we can actually do better
+than that. Indeed, the evaluation of all the shape functions in all the ansatz points can be rewritten as
+\begin{lstlisting}
+template<int k, int q>
+void evaluate_dof_vector_in_all_quadrature_points
+  (const double u[k][k][k],
+   const double phi[k][q],
+   double return_value[q][q][q])
+{
+  double tmp1[k][k][q] {};
+  for (int i1=0; i1<k; ++i1)
+    for (int i2=0; i2<k; ++i2)
+      for (int q3=0; q3<q; ++q3)
+        for (int i3=0; i3<k; ++i3)
+          tmp1[i1][i2][q3] += u[i1][i2][i3]*phi[i3][q3];
 
-{\color{red}
-[DA] Daniel, please add a brief intro about MF (readership will likely not know anything about it), maybe borrow 1-2 pictures to explain how tensor-product structure is handled.
+  double temp2[k][q][q] {};
+  for (int i1=0; i1<k; ++i1)
+    for (int q3=0; q3<q; ++q3)
+      for (int q2=0; q2<q; ++q2)
+        for (int i2=0; i2<k; ++i2)
+          tmp2[i1][q2][q3] += tmp1[i1][i2][q3]*phi[i2][q2];
+
+  for (int q2=0; q2<q; ++q2)
+    for (int q3=0; q3<q; ++q3)
+      for (int q1=0; q1<q; ++q1)
+        for (int i1=0; i1<k; ++i1)
+  return_value[q1][q2][q3] += tmp2[i1][q2][q3]*phi[i1][q1];
 }
+\end{lstlisting}
+in three space dimensions. In general, this reordering reduces the arithmetic complexity to $\mathcal{O}(dn^{d+1})$ and is called sum factorization.
+Hence, we can expect that such a matrix-free approach is faster than a regular matrix-base especially for 3D and high polynomial degree.
+More information on techniques to increase the computational density can be found in \cite{kronbichler12,vos10}.
 
@@
 
-TODO: question to asnwer -- how much to cache? At least $detF ^ {2/3}$. Other options is stress and strain at quadrature points.
+TODO: question to answer -- how much to cache? At least $detF ^ {2/3}$. Other options is stress and strain at quadrature points.
 Run benchmarks and see.
@@
 TODO: add new results from Cook's and some matrix-inclusion (matrix-pore).
-Eliptical holes are easier as we don't need heterogeneous operator.
+Elliptical holes are easier as we don't need heterogeneous operator.
 

### 2018-08-29 Calculations/post_process.py
add post-processing script

@@ -1,0 +1,127 @@
+import re
+import os
+import argparse
+from matplotlib.pyplot import figure, show
+import matplotlib.pyplot as plt
+import matplotlib as mp
+import numpy as np
+
+# define command line arguments
+parser = argparse.ArgumentParser(description='Post-Process timing/memory info and plot figures.')
+parser.add_argument('prefix', metavar='prefix', default='Emmy_RRZE', nargs='?',
+                    help='A folder to look for data')
+args = parser.parse_args()
+
+prefix = args.prefix if args.prefix.startswith('/') else os.path.join(os.getcwd(), args.prefix)
+
+files = [os.path.join(prefix, k,'timings.txt') for k in os.listdir(prefix) if os.path.isfile(os.path.join(prefix, k,'timings.txt'))]
+
+print 'Gather data from {0}'.format(prefix)
+print 'found {0} files'.format(len(files))
+
+pattern = r'[+\-]?(?:[0-9]\d*)(?:\.\d*)?(?:[eE][+\-]?\d+)?'
+
+# sections in timer output:
+sections = [
+    'vmult (MF)',
+    'vmult (Trilinos)',
+]
+
+start_line = 'Total wallclock time elapsed since start'
+
+data_scalar = []
+data_tensor2 = []
+data_tensor4 = []
+
+for f in files:
+    fin = open(f, 'r')
+    ready = False
+
+    # timing = ['-' for i in range(2*len(sections))]
+    timing = [np.nan for i in range(len(sections))]
+
+    for line in fin:
+        if 'dim   =' in line:
+            dim = int(re.findall(pattern,line)[0])
+
+        elif 'p     =' in line:
+            p = int(re.findall(pattern,line)[0])
+
+        elif 'q     =' in line:
+            q = int(re.findall(pattern,line)[0])
+
+        elif 'cells =' in line:
+            cells = int(re.findall(pattern,line)[0])
+
+        elif 'dofs  =' in line:
+            dofs = int(re.findall(pattern,line)[0])
+
+        elif 'Trilinos memory =' in line:
+            tr_memory = float(re.findall(pattern,line)[0])
+
+        elif 'MF cache memory =' in line:
+            mf_memory = float(re.findall(pattern,line)[0])
+
[…]

### 2018-08-29 Calculations/post_process.py
adjust where to save

@@ -104,6 +104,9 @@ mem_sc = [tup[3]/tup[1] for tup in data_scalar]
 mem_t2 = [tup[3]/tup[1] for tup in data_tensor2]
 mem_t4 = [tup[3]/tup[1] for tup in data_tensor4]
 
+# file location
+fig_prefix = os.path.join(os.getcwd(), '../doc/' + os.path.basename(os.path.normpath(prefix)) + '_')
+
 plt.plot(deg,time_tr, 'rs--', label='Trilinos')
 plt.plot(deg,time_sc, 'bo--', label='MF scalar')
 plt.plot(deg,time_t2, 'g^--', label='MF tensor2')
@@ -111,7 +114,7 @@ plt.plot(deg,time_t4, 'cv--', label='MF tensor4')
 plt.xlabel('degree')
 plt.ylabel('wall time (s) / DoF')
 leg = plt.legend(loc='best', ncol=1)
-plt.savefig('timing.eps', format='eps')
+plt.savefig(fig_prefix + 'timing.eps', format='eps')
 
 
 # clear
@@ -124,4 +127,4 @@ plt.plot(deg,mem_t4, 'cv--', label='MF tensor4')
 plt.xlabel('degree')
 plt.ylabel('memory (Mb) / DoF')
 leg = plt.legend(loc='best', ncol=1)
-plt.savefig('memory.eps', format='eps')
+plt.savefig(fig_prefix + 'memory.eps', format='eps')

### 2018-08-29 include/mf_elasticity.h
sum memory in Mb to avoid overflow

@@ -1491,7 +1491,7 @@ namespace Cook_Membrane
       << "cells = " << triangulation.n_global_active_cells() << std::endl
       << "dofs  = " << dof_handler_ref.n_dofs() << std::endl
       << std::endl
-      << "Trilinos memory = " << dealii::Utilities::MPI::sum(tangent_matrix.memory_consumption(), mpi_communicator)/1000000 << " Mb" << std::endl;
+      << "Trilinos memory = " << dealii::Utilities::MPI::sum(tangent_matrix.memory_consumption()/1000000, mpi_communicator) << " Mb" << std::endl;
 
   }
 
@@ -1606,7 +1606,7 @@ namespace Cook_Membrane
         // print memory consumption by MF
         if (print_mf_memory)
           {
-            timer_out << "MF cache memory = " << dealii::Utilities::MPI::sum(mf_nh_operator.memory_consumption(), mpi_communicator)/1000000 << " Mb" << std::endl;
+            timer_out << "MF cache memory = " << dealii::Utilities::MPI::sum(mf_nh_operator.memory_consumption()/1000000, mpi_communicator) << " Mb" << std::endl;
             print_mf_memory = false;
           }
 

### 2018-09-03 include/mf_elasticity.h
minor fixes

@@ -1190,7 +1190,7 @@ namespace Cook_Membrane
       triangulation_2d, 5, 1.0 * scale, dst, true);
 
     // we need to set manifolds:
-    Tensor<1, 3> dir_1(center_dim_1), dir_2(center_dim_2), dir_3(center_dim_3);
+    Tensor<1, 3> dir_1, dir_2, dir_3;
     dir_1[2] = 1.;
     dir_2[2] = 1.;
     dir_3[2] = 1.;
@@ -2209,7 +2209,8 @@ namespace Cook_Membrane
         soln_pt[0] = 0. * parameters.scale;
         soln_pt[1] = 0. * parameters.scale;
         if (dim == 3)
-          soln_pt[2] = 0. * parameters.scale;
+          // we extrude by 1.0 * scale:
+          soln_pt[2] = 0.5 * parameters.scale;
 
       }
     double vertical_tip_displacement       = 0.0;

### 2018-09-03 include/mf_elasticity.h
rework print_vertical_tip_displacement()

@@ -2213,54 +2213,40 @@ namespace Cook_Membrane
           soln_pt[2] = 0.5 * parameters.scale;
 
       }
-    double vertical_tip_displacement       = 0.0;
-    double vertical_tip_displacement_check = 0.0;
 
-    for (const auto &cell : dof_handler_ref.active_cell_iterators())
-      if (cell->is_locally_owned())
-        {
-          for (unsigned int v = 0; v < GeometryInfo<dim>::vertices_per_cell;
-               ++v)
-            if (cell->vertex(v).distance(soln_pt) < 1e-6)
-              {
-                // Extract y-component of solution at the given point
-                // This point is coindicent with a vertex, so we can
-                // extract it directly as we're using FE_Q finite elements
-                // that have support at the vertices
-                vertical_tip_displacement =
-                  solution_n(cell->vertex_dof_index(v, u_dof + 1));
+    double vertical_tip_displacement = 0.0;
+    unsigned int found = 0;
 
-                // Sanity check using alternate method to extract the solution
-                // at the given point. To do this, we must create an FEValues
-                // instance to help us extract the solution value at the desired
-                // point
-                const MappingQ<dim> mapping(degree);
-                const Point<dim>    qp_unit =
-                  mapping.transform_real_to_unit_cell(cell, soln_pt);
-                const Quadrature<dim> soln_qrule(qp_unit);
-                AssertThrow(soln_qrule.size() == 1, ExcInternalError());
-                FEValues<dim> fe_values_soln(fe, soln_qrule, update_values);
-                fe_values_soln.reinit(cell);
+    try
+      {
+        const MappingQ<dim> mapping(degree);
+        const auto          cell_point =
+          GridTools::find_active_cell_around_point(mapping,
+                                                   dof_handler_ref,
+                                                   soln_pt);
+        found = 1;
 
-                // Extract y-component of solution at given point
-                std::vector<Tensor<1, dim>> soln_values(soln_qrule.size());
-                fe_values_soln[u_fe].get_function_values(solution_n,
-                                                         soln_values);
-                vertical_tip_displacement_check = soln_values[0][u_dof + 1];
+        const Quadrature<dim> soln_qrule(cell_point.second);
+        AssertThrow(soln_qrule.size() == 1, ExcInternalError());
[…]

### 2018-09-03 include/mf_elasticity.h
keep TFI around hole

@@ -1436,12 +1436,20 @@ namespace Cook_Membrane
            center_dim_3);
 
          if (dim == 2) // FIXME: some issues with 3D
-           for (unsigned int i = 4; i <= 8; ++i)
-             {
-               TransfiniteInterpolationManifold<dim> transfinite_manifold;
-               transfinite_manifold.initialize(triangulation);
-               triangulation.set_manifold(i, transfinite_manifold);
-             }
+           {
+             for (unsigned int i = 4; i <= 8; ++i)
+               {
+                 TransfiniteInterpolationManifold<dim> transfinite_manifold;
+                 transfinite_manifold.initialize(triangulation);
+                 triangulation.set_manifold(i, transfinite_manifold);
+               }
+           }
+         else
+           {
+             TransfiniteInterpolationManifold<dim> transfinite_manifold;
+             transfinite_manifold.initialize(triangulation);
+             triangulation.set_manifold(4, transfinite_manifold); // hole
+           }
 
          const double tol_boundary = 1e-6*parameters.scale;
          for (auto cell : triangulation.active_cell_iterators())
@@ -1463,6 +1471,49 @@ namespace Cook_Membrane
           std::ofstream output("grid.eps");
           grid_out.write_eps(triangulation, output);
           */
+          // debug output of face data:
+          if (true && dim==3)
+            {
+              using Tuple =
+                std::tuple<Point<dim>, types::boundary_id, types::manifold_id>;
+              std::vector<Tuple> boundary_faces;
+
+              for (const auto &cell : triangulation.active_cell_iterators())
+                for (unsigned int face_n = 0;
+                     face_n < GeometryInfo<dim>::faces_per_cell;
+                     ++face_n)
+                  if (cell->face(face_n)->manifold_id() == 2 || cell->face(face_n)->manifold_id() == 3)
+                  // if (cell->face(face_n)->at_boundary() &&
+                  //     std::abs(cell->face(face_n)->center()[2])<1e-10)
+                    boundary_faces.push_back(
+                      std::make_tuple(cell->face(face_n)->center(),
+                                      cell->face(face_n)->boundary_id(),
+                                      cell->face(face_n)->manifold_id()));
+
+              // see dof_tools.cc internal::ComparisonHelper
+              std::sort(begin(boundary_faces),
+                        end(boundary_faces),
+                        [](const Tuple &t1, const Tuple &t2) {
+                          const auto &b1 = std::get<1>(t1);
[…]

### 2018-09-03 doc/large_strain.tex
add figures and use in latex
replot Emmy figures (now correct memory) and add IWR to Latex
add abstract and title

@@
 
+\usepackage[format=plain,indention=.5cm]{caption}
+\usepackage[labelformat=simple]{subcaption}
+\renewcommand\thesubfigure{(\alph{subfigure})}
+
 %\renewcommand{\thetable}{\arabic{table}}
@@
 \begin{document}
+
+\begin{frontmatter}
+  \title{
+  Large-strain matrix-free finite element solver of hyperelastic problems with geometric multigrid preconditioner
+  }
+
+  \author[a]{Denis Davydov\corref{cor}}
+  \ead{denis.davydov@fau.de}
+
+  \author[a]{Jean-Paul Pelteret}
+  \ead{jean-paul.pelteret@fau.de}
+
+  \author[b]{Daniel Arndt}
+  \ead{daniel.arndt@iwr.uni-heidelberg.de}
+
+  \author[a]{Paul Steinmann}
+  \ead{paul.steinmann@fau.de}
+
+  \cortext[cor]{Corresponding author.}
+
+  \address[a]{Chair of Applied Mechanics,
+  Friedrich-Alexander-Universit\"{a}t Erlangen-N\"{u}rnberg,
+  Egerlandstr.\ 5, 91058 Erlangen, Germany}
+
+  \address[b]{Interdisciplinary Center for Scientific Computing (IWR),
+      Heidelberg University,
+      Im Neuenheimer Feld 205,
+      69120 Heidelberg,
+      Germany}
+
+  \begin{abstract}
+    Solution of source problems on modern computer architectures is typically memory banded.
+    Loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
+    In order to improve the performance of the iterative solver, the so-called matrix-free methods are
+    widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly.
+
+    In this work we advance the application of matrix-free approaches to problems in solid mechanics.
+    We propose and numerically investigate different implementations of the large-strain hyper-elastic tangent operator.
+    In order to improve the convergence of iterative solvers, we also propose a way to construct a level tangent operator
+    and employ it to define a geometric multigrid preconditioner.
+    Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.
+    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based counterparts on the representative numerical example of heterogeneous hyper-elastic material in two and three dimensions.
+  \end{abstract}
+
+  \begin{keyword}
+      adaptive finite element method \sep
+      geometric multigrids \sep
+      large strain \sep
+      matrix-free \sep
+      hyperelasticity
+  \end{keyword}
+
+  \end{frontmatter}
 
@@
 
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.49\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{Emmy_RRZE_timing.eps}
+      \caption{vmult}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.49\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{Emmy_RRZE_memory.eps}
+      \caption{memory consumption}
+  \end{subfigure}
+  \caption{Heterogeneous 2D large-strain example (Emmy cluster, RRZE).}%
+  \label{fig:benchmark_2d_miehe}
+\end{figure}
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.49\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{IWR_timing.eps}
+      \caption{vmult}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.49\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{IWR_memory.eps}
+      \caption{memory consumption}
+  \end{subfigure}
+  \caption{Heterogeneous 2D large-strain example (IWR cluster).}%
+  \label{fig:benchmark_2d_miehe_IWR}
+\end{figure}
+
 TODO: add new results from Cook's and some matrix-inclusion (matrix-pore).

### 2018-09-03 doc/large_strain.tex
fix a typo
more

@@
   \begin{abstract}
-    Solution of source problems on modern computer architectures is typically memory banded.
+    Solution of source problems on modern computer architectures is typically memory bound.
     Loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
@@
     We propose and numerically investigate different implementations of the large-strain hyper-elastic tangent operator.
-    In order to improve the convergence of iterative solvers, we also propose a way to construct a level tangent operator
-    and employ it to define a geometric multigrid preconditioner.
+    In order to improve the convergence of iterative solvers, we also propose a way to construct level tangent operators
+    and employ them to define a geometric multigrid preconditioner.
     Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.

### 2018-09-04 include/mf_elasticity.h
add parameters for extrusion height and slices

@@ -215,6 +215,8 @@ namespace Cook_Membrane
       double       scale;
       unsigned int dim;
       unsigned int n_global_refinement;
+      unsigned int extrusion_slices;
+      double       extrusion_height;
       std::string  type;
 
       static void
@@ -244,6 +246,16 @@ namespace Cook_Membrane
                           Patterns::Double(0.0),
                           "Global grid scaling factor");
 
+        prm.declare_entry("Extrusion height",
+                          "1",
+                          Patterns::Double(0.0),
+                          "Extrusion height");
+
+        prm.declare_entry("Extrusion slices",
+                          "5",
+                          Patterns::Integer(0),
+                          "Number of extrusion slices");
+
         prm.declare_entry("Dimension",
                           "2",
                           Patterns::Integer(2, 3),
@@ -268,6 +280,8 @@ namespace Cook_Membrane
         dim                 = prm.get_integer("Dimension");
         n_global_refinement = prm.get_integer("Global refinement");
         type                = prm.get("Type");
+        extrusion_slices    = prm.get_integer("Extrusion slices");
+        extrusion_height    = prm.get_double("Extrusion height");
       }
       prm.leave_subsection();
     }
@@ -1178,7 +1192,9 @@ namespace Cook_Membrane
     const double                                                scale,
     const Point<3> &                                            center_dim_1,
     const Point<3> &                                            center_dim_2,
-    const Point<3> &                                            center_dim_3)
+    const Point<3> &                                            center_dim_3,
+    const double                                                extrusion_height,
+    const unsigned int                                          extrusion_slices)
   {
     Triangulation<2> triangulation_2d;
     GridGenerator::merge_triangulations(triangulations,
@@ -1187,7 +1203,7 @@ namespace Cook_Membrane
                                         true);
 
     GridGenerator::extrude_triangulation(
-      triangulation_2d, 5, 1.0 * scale, dst, true);
+      triangulation_2d, extrusion_slices, extrusion_height * scale, dst, true);
 
     for (const auto& cell: dst.active_cell_iterators())
       for (unsigned int face_no=0; face_no < GeometryInfo<3>::faces_per_cell; ++face_no)
@@ -1215,7 +1231,10 @@ namespace Cook_Membrane
     const double                                                scale,
     const Point<2> &                                            center_dim_1,
     const Point<2> &                                            center_dim_2,
[…]

### 2018-09-04 doc/large_strain.tex
change title and finite-strain

@@
   \title{
-  Large-strain matrix-free finite element solver of hyperelastic problems with geometric multigrid preconditioner
+  A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid
   }
@@
     In this work we advance the application of matrix-free approaches to problems in solid mechanics.
-    We propose and numerically investigate different implementations of the large-strain hyper-elastic tangent operator.
+    We propose and numerically investigate different implementations of the finite-strain hyper-elastic tangent operator.
     In order to improve the convergence of iterative solvers, we also propose a way to construct level tangent operators

### 2018-09-04 doc/large_strain.tex
more edits

@@
   \begin{abstract}
-    Solution of source problems on modern computer architectures is typically memory bound.
-    Loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
-    In order to improve the performance of the iterative solver, the so-called matrix-free methods are
+    The performance of finite element solvers on modern computer architectures is typically memory bound.
+    The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
+    In order to improve the performance of the iterative solver, so-called matrix-free methods are
     widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly.
 
-    In this work we advance the application of matrix-free approaches to problems in solid mechanics.
+    In this work, we advance the application of matrix-free approaches to problems in solid mechanics.
     We propose and numerically investigate different implementations of the finite-strain hyper-elastic tangent operator.

### 2018-09-04 doc/large_strain.tex
on the -> for a

@@
     Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.
-    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based counterparts on the representative numerical example of heterogeneous hyper-elastic material in two and three dimensions.
+    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based counterparts for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions.
   \end{abstract}

### 2018-09-04 doc/large_strain.tex
finite-strain
mention amg

@@
     Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.
-    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based counterparts for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions.
+    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions.
   \end{abstract}
@@
       geometric multigrids \sep
-      large strain \sep
+      finite-strain \sep
       matrix-free \sep

### 2018-09-04 include/mf_elasticity.h
minor speedup of matrix-based assembly
bring back geometric contribution via components, see
288ca0905eb457c3863b9818270190f9617aeb46

@@ -9,6 +9,8 @@
  */
 static const unsigned int debug_level = 0;
 
+#define COMPONENT_LESS_GEOM_TANGENT
+
 // We start by including all the necessary deal.II header files and some C++
 // related ones. They have been discussed in detail in previous tutorial
 // programs, so you need only refer to past tutorials for details.
@@ -2424,13 +2426,16 @@ namespace Cook_Membrane
                 Physics::Elasticity::Kinematics::F(grad_u);
               const SymmetricTensor<2, dim, NumberType> b =
                 Physics::Elasticity::Kinematics::b(F);
+
               const NumberType                 det_F = determinant(F);
-              const Tensor<2, dim, NumberType> F_bar =
-                Physics::Elasticity::Kinematics::F_iso(F);
-              const SymmetricTensor<2, dim, NumberType> b_bar =
-                Physics::Elasticity::Kinematics::b(F_bar);
-              const Tensor<2, dim, NumberType> F_inv = invert(F);
               Assert(det_F > NumberType(0.0), ExcInternalError());
+              const Tensor<2, dim, NumberType> F_inv = invert(F);
+
+              // don't calculate b_bar if we don't need to:
+              const SymmetricTensor<2, dim, NumberType> b_bar =
+                cell_mat->formulation == 0 ?
+                Physics::Elasticity::Kinematics::b(Physics::Elasticity::Kinematics::F_iso(F)) :
+                SymmetricTensor<2, dim, NumberType>();
 
               for (unsigned int k = 0; k < dofs_per_cell; ++k)
                 {
@@ -2450,6 +2455,10 @@ namespace Cook_Membrane
                   if (skip_assembly_on_this_cell)
                     continue;
 
+#ifndef COMPONENT_LESS_GEOM_TANGENT
+                  const unsigned int component_i = fe.system_to_component_index(i).first;
+                  const Tensor<1,dim> grad_Nx_i_comp_i_tau = grad_Nx[i][component_i] * tau_ns;
+#endif
                   for (unsigned int j = 0; j <= i; ++j)
                     {
                       // This is the $\mathsf{\mathbf{k}}_{\mathbf{u}
@@ -2464,10 +2473,17 @@ namespace Cook_Membrane
                            b,
                            symm_grad_Nx[j])) // The material contribution:
                         * JxW;
+
+#ifdef COMPONENT_LESS_GEOM_TANGENT
                       // geometrical stress contribution
[…]

### 2018-09-04 include/mf_elasticity.h
initialize vmult benchmark vectors once per Newton solver step

@@ -2044,6 +2044,16 @@ namespace Cook_Membrane
 
     print_conv_header();
 
+    // auxiliary vectors to test vmult()
+    TrilinosWrappers::MPI::Vector src_trilinos(newton_update_trilinos),
+      dst_mb(newton_update_trilinos);
+    for (unsigned int i = 0; i < locally_owned_dofs.n_elements(); ++i)
+      src_trilinos(locally_owned_dofs.nth_index_in_set(i)) =
+        ((double)std::rand()) / RAND_MAX;
+
+    LinearAlgebra::distributed::Vector<double> src(newton_update),
+      dst_mf(newton_update);
+
     // We now perform a number of Newton iterations to iteratively solve the
     // nonlinear problem.  Since the problem is fully nonlinear and we are
     // using a full Newton method, the data stored in the tangent matrix and
@@ -2077,16 +2087,7 @@ namespace Cook_Membrane
 
         // check vmult of matrix-based and matrix-free for a random vector:
         {
-          TrilinosWrappers::MPI::Vector src_trilinos(newton_update_trilinos),
-            dst_mb(newton_update_trilinos);
-          for (unsigned int i = 0; i < locally_owned_dofs.n_elements(); ++i)
-            src_trilinos(locally_owned_dofs.nth_index_in_set(i)) =
-              ((double)std::rand()) / RAND_MAX;
-
           constraints.set_zero(src_trilinos);
-
-          LinearAlgebra::distributed::Vector<double> src(newton_update),
-            dst_mf(newton_update);
           copy_trilinos(src, src_trilinos);
 
           const unsigned int n_times = 10;

### 2018-09-07 Calculations/post_process.py
plot all results

@@ -51,6 +51,8 @@ for f in files:
     # timing = ['-' for i in range(2*len(sections))]
     timing = [np.nan for i in range(len(sections))]
 
+    # reset CG iterations in case AMG did not have enough memory
+    cg_iterations = np.nan
     for line in fin:
         if 'dim   =' in line:
             dim = int(re.findall(pattern,line)[0])
@@ -73,6 +75,9 @@ for f in files:
         elif 'MF cache memory =' in line:
             mf_memory = float(re.findall(pattern,line)[0])
 
+        elif 'Average CG iter =' in line:
+            cg_iterations = int(re.findall(pattern,line)[0])
+
         if start_line in line:
             print 'p={0} q={1} cells={2} dofs={3} tr_memory={4} mf_memory={5} file={6}'.format(p, q, cells, dofs, tr_memory, mf_memory, f)
             ready = True
@@ -84,7 +89,7 @@ for f in files:
                     timing[idx] = float(nums[1]) / int(nums[0]) # total / how many times
 
     # finish processing the file, put the data
-    tp = tuple((p, dofs, tr_memory, mf_memory, timing))
+    tp = tuple((p, dofs, tr_memory, mf_memory, timing, cg_iterations))
     if 'MF_CG' in f:
         if '_scalar' in f:
             if dim == 2:
@@ -144,6 +149,16 @@ solver3d_tr = [tup[4][2]/tup[1] for tup in mb3d_data]
 solver3d_sc = [tup[4][2]/tup[1] for tup in mf3d_data_scalar]
 solver3d_t2 = [tup[4][2]/tup[1] for tup in mf3d_data_tensor2]
 solver3d_t4 = [tup[4][2]/tup[1] for tup in mf3d_data_tensor4]
+
+assembly2d_tr = [tup[4][3]/tup[1] for tup in mb2d_data]
+assembly3d_tr = [tup[4][3]/tup[1] for tup in mb3d_data]
+
+# CG iterations
+cg2d_tr = [tup[5] for tup in mb2d_data]
+cg2d_t4 = [tup[5] for tup in mf2d_data_tensor4]
+
+cg3d_tr = [tup[5] for tup in mb3d_data]
+cg3d_t4 = [tup[5] for tup in mf3d_data_tensor4]
 
[…]

### 2018-09-17 doc/large_strain.tex
plot all results
plot IWR
preliminary version of most of text

@@
 
-TODO: elasticity and no matrix-free. fluids used a lot.
+The performance of finite element (FEM) solvers on modern computer architectures is typically memory bound.
+The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
+In order to improve the performance of the iterative solver, so-called matrix-free methods are
+widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly \cite{kronbichler12}.
+
+To the best of our knowledge, the matrix-free methods are not widely adopted within the solid mechanics community.
+A possible reason is that tangent operators (coming from linearization of the nonlinear balance equations) are
+more elaborate as compared to the fluid mechanics and it is not obvious whether the matrix-free methods can be advantageous in this case.
+Another reason could be the smoothness of the underlying solution.
+FEM with linear or quadratic elements (to avoid locking) is often adopted in solid mechanics community.
+However we note that large scale computations with $10^{12}$ unknowns even with linear FEM inevitably require matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
+Therefore we believe that for large scale computations in solid mechanics it is crucial to adopt matrix-free approaches and develop efficient multigrid solvers.
+
+In this work, we advance the application of matrix-free approaches to problems in solid mechanics.
+We propose and numerically investigate different implementations of the finite-strain hyper-elastic tangent operator.
+In order to improve the convergence of iterative solvers, we also propose a way to construct level tangent operators
+and employ them to define a geometric multigrid preconditioner.
+Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.
+
+The paper is organized as follows.
+In Section \ref{sec:theory} we briefly introduce partial differential equations of finite-strain solid mechanics.
+Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the proposed matrix-free operator evaluation approaches.
+In Section \ref{sec:gmg} we propose a geometric multigrid preconditioner suitable for finite-strain linear elasticity.
+The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions in Section \ref{sec:example}.
+Summary is presented in Section \ref{sec:summary}.
 
 \section{Theoretical background}
+\label{sec:theory}
 
-A brief summary of balance equations in large strain hyperelastic continuum.
+In this section we briefly summarize balance equations of fine-strain hyperelastic continuum.
 
@@
 
-TODO: linearize and show how tangent look like.
+{\color{red}TODO: linearize and show how tangent look like.
 Also show what it boils down in the case of small strain.
+}
 
@@
 \section{Finite Element discretization}
+\label{sec:fe}
 
@@
 the associated FE space of continuous piecewise elements of a fixed polynomial degree. % : $V^h \subset H^1 (\mcl B_0)$.
-Macro- -deformation map are given in a
+The deformation map are given in a
 a vector space spanned by standard vector--valued FE basis functions $\gz N^i(\gz x)$ (e.g. polynomials with local support):
@@
 \end{alignat}
-where superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom for the
-macroscopic fields. For the sake of this study $\gz N^i_\varphi (\gz X)$ are such that only one component of the vector shape function is non-zero for each $i$.
+where superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom (DoF).
+For the sake of this study $\gz N^i_\varphi (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.
 
 \section{Matrix-free operator evaluation}
+\label{sec:mf}
 Classically the approach for finding a discrete solution to a linearized partial differential equation can be described as follows:
@@
 More information on techniques to increase the computational density can be found in \cite{kronbichler12,vos10}.
+We briefly note that the SIMD vectorization is applied at the element level, that is the operator is applied simultaneously on several several elements (called ``block'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
 
-Due to the specifics of matrix-free operator evaluation in \texttt{deal.II}, when performing contraction with gradient of shape functions in current (deformed) configuration, the integration is also done over the deformed configuration.
+Given that the matrix-free operators are applied on-the-fly, it's crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
+Below we propose three algorithms to implement tangent operator of finite-strain linear elasticity using matrix-free approach, that employ different caching strategies.
+%
+In order to obtain gradients with respect to the deformed configuration we adopted \textit{MappingQEulerian} class from \texttt{deal.II} library.
+In addition to the standard mapping from the reference (unit / isoparametric) element to the element in real space, this class adds mapping to the deformed configuration given a displacement field.
+%
+Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, when performing contraction with gradient of shape functions in current (deformed) configuration, the integration is also done over the deformed configuration.
 Therefore in order to arrive at the integral in undeformed configuration, additionally we have to divide by Jacobian $J$ of the mapping at a given quadrature point.
 
-\begin{algorithm}[h]
+\begin{algorithm}[!ht]
   \SetKwInOut{Input}{Given}
@@
 
-\begin{algorithm}[h]
+Algorithm \ref{alg:mf_scalar} caches a single scalar which involves logarithm of Jacobian -- a computationally demanding operation compared to addition and multiplications.
+As evident from Section \ref{sec:theory}, in order to apply the operator we need to re-evaluate Kirchhoff stress at each quadrature point.
+This in turn requires evaluation of deformation gradient $\gz F$ and therefore gradients with respect to the undeformed configuration.
+
+\begin{algorithm}[!ht]
   \SetKwInOut{Input}{Given}
@@
 
-\begin{algorithm}[h]
+In order to avoid re-calculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point.
+This also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
+This algorithm utilize the chosen constitutive relationship and therefore the action of the material fourth order tangent $\boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$ can be evaluated by taking a linear combination this tensor and its trace times identity tensor.
+To that end we also cache two scalars that depend on the Jacobian $J$.
+
+\begin{algorithm}[!ht]
   \SetKwInOut{Input}{Given}
@@
 
-The primary difference between Algorithm \ref{alg:mf_tensor2} and Algorithm \ref{alg:mf_tensor4}, is that the former utilizes the chosen structure of free energy function and thereby ends up with a computationally cheap way of evaluation of $\boldsymbol{\mathcal{C}} \gz g_s $, whereas the latter performs the double contraction between the fourth order and second order tensors.
-
-TODO[DD]: summary of what's happening, how operator is evaluated, what we do, what operations (determinant, etc), count number of FLOPS
+Finally Algorithm \ref{alg:mf_tensor4} represents the most general algorithm which does not assume specific constitutive relationship.
+Clearly in this case we have to cache the fourth order symmetric tensor $\boldsymbol{\mathcal{C}}$ for each element and quadrature point and perform the double contraction with the second-order symmetric tensor on-the-fly.
 
 \section{Geometric multigrid preconditioning}
+\label{sec:gmg}
 
-TODO[DD]: explain how this works with GMG (eulerian mapping on fine level gets restricted onto the coarse levels, the rest works as-is).
-Hopefully we can still use some mappings on coarse level to better represent geometric of inclusions (pores).
+Efficient and scalable matrix-vector products is not enough to result in a method with linear scaling with respect to the number of unknown DoFs $N$.
+It has been widely accepted that the geometric multigrid (GMG) preconditioner can greatly improve the performance of the Conjugate Gradient (CG) solver.
+In particular, the number of CG iterations for convergence becomes independent of the number of mesh refinements and thus leads to the $\mathcal O (N)$ solution method.
+
+For this work we adopt the matrix-free level operators in GMG and Chebyshev smoothing.
+Given a generally heterogeneous non-linear material, definition of transfer operations (restriction and prolongation) and level operators is not trivial.
+One approach is to start from a arbitrary heterogeneous material at the fine scale and employ non-linear homogenization theory to design specific  transfer operators \cite{Miehe2007}.
+In this approach the fine-scale displacement is decomposed into long-wave and short-wave contributions by additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
+In case of a homogeneous patch, the transfer operators reduce to the standard geometric transfer.
+Based on this conclusion we develop our multigrid preconditioner.
+Namely, we assume that the coarsest level can accurately describe the heterogeneous fine-strain elastic material.
+In this case the restriction and prolongation operations for patches of elements are always performed for the same material.
+In order to define the level operators we adopt the following approach.
+Recall that the fine scale tangent is obtain by linearization around the current displacement field $\gz u^h$.
+We then restrict this displacement field to multigrid levels and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
+At the coarsest level we adopt CG solver which shall reduce residual by three orders of magnitude within at most hundred of steps.
 
 \section{Numerical examples}
+\label{sec:example}
 
-TODO: question to answer -- how much to cache? At least $detF ^ {2/3}$. Other options is stress and strain at quadrature points.
-Run benchmarks and see.
+In this section we apply the proposed matrix-free operator algorithms for finite-strain hyper-elastic material as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}.
+Coarse mesh for 2D and 3D problems is depicted in Figure \ref{fig:miehe}.
+The 2D material consists of a matrix (depicted in blue), a hole and two spherical inclusions (depicted in red).
+The size of the square domain is $10^{-3} \rm{mm}$.
+The matrix material is taken to have Poisson's ratio of $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is take to be $100$ times harder.
+The 3D material is obtained by extrusion in the third direction.
+Note that a relatively coarse mesh is adopted at the coarsest multigrid level, however the geometry of inclusions is captured accurately by employing manifold description of the boundaries and interfaces.
+Both domains are fixed at the bottom surface and the distributed load is applied at the top in $(1,0)$ and $(1,1,0)$ direction for 2D and 3D problems, respectively.
+The load of $0.5\,\rm{N}$ is applied in 5 steps.
+The displacement tolerance for Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for forces $10^{-8}$.
+The relative convergence criteria for linear solver is $10^{-6}$.
+%
+Figure \ref{fig:miehe_gmg} illustrates the multigrid mesh for the 2D problem after two global refinements.
+%
+The benchmark calculations are performed for 2D and 3D problems with various combination of polynomial degrees and number of global refinements, see Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
+%
+We are interested in the following metrics:
+(i) walltime per DoF of matrix-vector multiplication;
+(ii) memory requirement per DoF;
+(iii) average number of CG iterations throughout the whole run (i.e. each Newton-Raphson iteration and each loading step);
+(iv) walltime per DoF of CG solver;
+%
+The results of the matrix-free approach with geometric-multigrid preconditioner are compared to the matrix-based approach with algebraic multigrid (AMG) preconditioner using Trilinos \cite{Heroux2005} library.
+The aggregation threshold for AMG is taken as $10^{-4}$.
+Computations are done on a single node of two clusters.
+A node on ``Emmy'' cluster has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB Shared Cache per chip and 64 GB of RAM.
+Intel compilers version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
+A node on ``IWR'' cluster
+{\color{red}Daniel please add...}
 
-\subsection{Small strain linear elasticity}
+Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the benchmarks as performed on the two clusters.
+As expected the matrix-vector multiplication becomes very expensive for higher order discretization for matrix-based approaches.
+The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
+Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly shows the influence of different caching strategies.
+Two conclusions can be drawn: scalar caching implementation is the most time consuming at each step we additionally evaluate the gradient of displacement field in reference configuration, required to evaluate the Kirchhoff stress $\gz \tau$.
+On another hand there is little difference between the approach where the fourth order material tangent tensor is cached (``tensor4'') and the one where we only cache the second order Kirchhoff tensor and use the chosen hyper-elastic constitutive equation to efficiently implement the action of the material tangent on the second order symmetric tensor.
+This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
+Consequently we can apply matrix-free approach to any fine-strain material model by caching the fourth order material tangent in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
 
-TODO: summarize Mathias results (head and Cook's membrane)
+As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind the matrix-free methods.
+Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} show that even with caching of one fourth order symmetric tensor and one second order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
 
-TODO: switch head to quasi-static load instead of dynamics?
+Next we evaluate the performance of the proposed geometric multigrid preconditioner. First we consider the number of CG iterations.
+Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_Emmy_cg3} show a very mild increase in the average number of CG iteration for different polynomial degrees.
+On another hand the black-box AGM preconditioner require much more iterations for convergence.
+Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution time polynomial degrees up to $p=5$.
+For higher orders in 2D the walltime is comparable to the AMG preconditioner, see Figure \ref{fig:benchmark_miehe_Emmy_sol2}.
+The reason for that is the coarse level solver within the GMG preconditioner. Evidently the adopted here Chebyshev smoother with the CG solver have unsatisfactory performance for higher degrees.
+A possible solution is to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.
+This should also improve the performance of the preconditioner for lower degrees.
 
-\subsection{Large strain Neo-Hook elasticity}
+\begin{table}
+  \centering
+  \begin{tabular}{ccccc}
+  \hline
+    $p$ & $q$ & $N_{gref}$ & $N_{el}$ & $N_{DoF}$ \\
+  \hline
+    1 & 2 & 7 & 1441792 & 2887680 \\
+    2 & 3 & 6 & 360448 & 2887680 \\
+    3 & 4 & 5 & 90112 & 1625088 \\
+    4 & 5 & 5 & 90112 & 2887680 \\
+    5 & 6 & 5 & 90112 & 4510720 \\
+    6 & 7 & 4 & 22528 & 1625088 \\
+    7 & 8 & 4 & 22528 & 2211328 \\
+    8 & 9 & 4 & 22528 & 2887680 \\
+  \hline
+  \end{tabular}
+  \caption{Parameters for 2D benchmark.
+  }
+  \label{tab:input_parameters_2d}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{ccccc}
+  \hline
+    $p$ & $q$ & $N_{gref}$ & $N_{el}$ & $N_{DoF}$ \\
+  \hline
+    1 & 2 & 4 & 1441792 & 4442880 \\
+    2 & 3 & 3 & 180224 & 4442880 \\
+    3 & 4 & 2 & 22528 & 1891008 \\
+    4 & 5 & 2 & 22528 & 4442880 \\
+  \hline
+  \end{tabular}
+  \caption{Parameters for 3D benchmark.
+  }
+  \label{tab:input_parameters_3d}
+\end{table}
 
@@
   \centering
-  \begin{subfigure}[b]{0.49\textwidth}
+  \begin{subfigure}[b]{0.45\textwidth}
       \centering
-      \includegraphics[width=\textwidth]{Emmy_RRZE_timing.eps}
-      \caption{vmult}
+      \includegraphics[width=\textwidth]{coarse_mesh.png}
+      \caption{2D coarse mesh}
   \end{subfigure}
-  \begin{subfigure}[b]{0.49\textwidth}
+  \begin{subfigure}[b]{0.45\textwidth}
     \centering
-      \includegraphics[width=\textwidth]{Emmy_RRZE_memory.eps}
-      \caption{memory consumption}
+    \includegraphics[width=\textwidth]{coarse_mesh_3d.png}
+    \caption{3D coarse mesh}
   \end{subfigure}
-  \caption{Heterogeneous 2D large-strain example (Emmy cluster, RRZE).}%
-  \label{fig:benchmark_2d_miehe}
+  \caption{Heterogeneous material.}%
+  \label{fig:miehe}
 \end{figure}
@@
   \centering
-  \begin{subfigure}[b]{0.49\textwidth}
+  \begin{subfigure}[b]{0.8\textwidth}
     \centering
-      \includegraphics[width=\textwidth]{IWR_timing.eps}
-      \caption{vmult}
+    \includegraphics[width=\textwidth]{gmg_2d.png}
+    % \caption{2D multigrid mesh with 2 global refinements}
   \end{subfigure}
-  \begin{subfigure}[b]{0.49\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{IWR_memory.eps}
-      \caption{memory consumption}
-  \end{subfigure}
-  \caption{Heterogeneous 2D large-strain example (IWR cluster).}%
-  \label{fig:benchmark_2d_miehe_IWR}
+  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. Color indicate the process owning the element.}%
+  \label{fig:miehe_gmg}
 \end{figure}
 
-TODO: add new results from Cook's and some matrix-inclusion (matrix-pore).
-Elliptical holes are easier as we don't need heterogeneous operator.
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{Emmy_RRZE_timing2d.eps}
+      \caption{vmult 2D}
+      \label{fig:benchmark_miehe_Emmy_vmult2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{Emmy_RRZE_timing3d.eps}
+    \caption{vmult 3D}
+    \label{fig:benchmark_miehe_Emmy_vmult3}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{Emmy_RRZE_memory2d.eps}
+      \caption{memory consumption 2D}
+      \label{fig:benchmark_miehe_Emmy_memory2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{Emmy_RRZE_memory3d.eps}
+    \caption{memory consumption 3D}
+    \label{fig:benchmark_miehe_Emmy_memory3}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{Emmy_RRZE_cg2d.eps}
+    \caption{CG iterations 2D}
+    \label{fig:benchmark_miehe_Emmy_cg2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{Emmy_RRZE_cg3d.eps}
+    \caption{CG iterations 3D}
+    \label{fig:benchmark_miehe_Emmy_cg3}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{Emmy_RRZE_solver2d.eps}
+    \caption{CG solution time 2D}
+    \label{fig:benchmark_miehe_Emmy_sol2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{Emmy_RRZE_solver3d.eps}
+    \caption{CG solution time 3D}
+    \label{fig:benchmark_miehe_Emmy_sol3}
+  \end{subfigure}
+  \caption{Emmy cluster, RRZE.}%
+  \label{fig:benchmark_miehe_Emmy}
+\end{figure}
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{IWR_timing2d.eps}
+      \caption{vmult 2D}
+      \label{fig:benchmark_miehe_IWR_vmult2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{IWR_timing3d.eps}
+    \caption{vmult 3D}
+    \label{fig:benchmark_miehe_IWR_vmult3}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{IWR_memory2d.eps}
+      \caption{memory consumption 2D}
+      \label{fig:benchmark_miehe_IWR_memory2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{IWR_memory3d.eps}
+    \caption{memory consumption 3D}
+    \label{fig:benchmark_miehe_IWR_memory3}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{IWR_cg2d.eps}
+    \caption{CG iterations 2D}
+    \label{fig:benchmark_miehe_IWR_cg2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{IWR_cg3d.eps}
+    \caption{CG iterations 3D}
+    \label{fig:benchmark_miehe_IWR_cg3}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{IWR_solver2d.eps}
+    \caption{CG solution time 2D}
+    \label{fig:benchmark_miehe_IWR_sol2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{IWR_solver3d.eps}
+    \caption{CG solution time 3D}
+    \label{fig:benchmark_miehe_IWR_sol3}
+  \end{subfigure}
+  \caption{IWR cluster.}%
+  \label{fig:benchmark_miehe_IWR}
+\end{figure}
 
 \section{Summary and Conclusions}
+\label{sec:summary}
 
-TODO: summarize results, good/bad, where can be applied,...
-Bio-med fluid-structure ?
+In this contribution we proposed and numerically investigated several matrix-free implementations of tangent operators for fine-strain elasticity of heterogeneous materials.
+The implementation which caches the fourth order material tangent together with the second order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
+This indicates that the matrix-free implementation of tangent operator can be applied any constitutive model which operates with the material tangent on quadrature point level.
+
+We also studies the performance of the GMG preconditioner with standard geometric transfer operations between each level.
+The GMG was applied to heterogeneous material assuming that the coarsest level can provide adequate discretization of heterogeneity.
+The numerical studies indicate that the proposed preconditioner lead to the solution approach that is faster than the matrix-based algebraic multigrid preconditioner for moderate polynomial degrees.
+However the considerable amount of time is spent in the coarse level solver.
+Therefore our future work will be focused on extending this proposed here matrix-free multigrid solution approach to homogenization based transfer operators \cite{Miehe2007}.
 

### 2018-09-17 doc/large_strain.tex
Add IWR processor information

@@
 Intel compilers version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
-A node on ``IWR'' cluster
-{\color{red}Daniel please add...}
+The ``IWR'' results were obtained ona single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB Shared Cache per chip.
+For the simulation, GNU 8.1.0 with flags ``-O3 -march=native'' and OpenMPI 3.0.0 were used.
 

### 2018-09-18 doc/large_strain.tex
more references

@@
 In order to improve the performance of the iterative solver, so-called matrix-free methods are
-widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly \cite{kronbichler12}.
+widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017}.
+Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016}.
 

### 2018-09-18 doc/large_strain.tex
more citations

@@
 In order to improve the performance of the iterative solver, so-called matrix-free methods are
-widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017}.
+widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010}.
 Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016}.
@@
 To the best of our knowledge, the matrix-free methods are not widely adopted within the solid mechanics community.
+We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018}.
 A possible reason is that tangent operators (coming from linearization of the nonlinear balance equations) are

### 2018-09-18 doc/large_strain.tex
acknowledgements

@@
 
+\section*{Acknowledgements}
+
+DD acknowledge the financial support of the German Science Foundation (Deutsche Forschungs-Gemeinschaft, DFG), grant DA 1664/2-1
+and the Competence Network for Technical and Scientific High Performance Computing in Bavaria (KONWIHR).
+DD, DA and JPP are grateful to Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for a fruitful discussion on matrix-free operator evaluation approaches.
+%
+{\color{red}
+DA was supported ...
+%
+JPP was supported by ...
+}
+%
+PS acknowledges the support of the Cluster of Excellence Engineering of Advanced Materials (EAM) which made this collaboration possible.
+
 \bibliographystyle{elsarticle-num}

### 2018-09-18 doc/large_strain.tex
sort and compress bib, minor fixes

@@
 \documentclass[preprint,12pt,times]{elsarticle}
+
+\biboptions{sort&compress,comma,square}
 
@@
       adaptive finite element method \sep
-      geometric multigrids \sep
+      geometric multigrid \sep
       finite-strain \sep
@@
 In order to improve the performance of the iterative solver, so-called matrix-free methods are
-widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010}.
+widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016}.
 Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016}.
 
-To the best of our knowledge, the matrix-free methods are not widely adopted within the solid mechanics community.
+To the best of our knowledge, the matrix-free methods are not yet widely adopted within the solid mechanics community.
 We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018}.
@@
 Another reason could be the smoothness of the underlying solution.
-FEM with linear or quadratic elements (to avoid locking) is often adopted in solid mechanics community.
+FEM with linear or quadratic elements (to avoid locking) are often adopted in solid mechanics community.
 However we note that large scale computations with $10^{12}$ unknowns even with linear FEM inevitably require matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.

### 2018-09-18 doc/large_strain.tex
a bit more on GMG

@@
 
-Efficient and scalable matrix-vector products is not enough to result in a method with linear scaling with respect to the number of unknown DoFs $N$.
-It has been widely accepted that the geometric multigrid (GMG) preconditioner can greatly improve the performance of the Conjugate Gradient (CG) solver.
-In particular, the number of CG iterations for convergence becomes independent of the number of mesh refinements and thus leads to the $\mathcal O (N)$ solution method.
+Efficient and scalable matrix-vector products is not enough to obtain a method with linear scaling with respect to the number of unknown DoFs $N$.
+Efficiency of standard iterative solvers deteriorate for large systems of linear equations.
+It has been widely accepted that the geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can greatly improve the performance of iterative solvers by smoothing the defect on a hierarchy of levels.
+As a result the number of iterations for convergence becomes independent of the number of mesh refinements and thus leads to the $\mathcal O (N)$ solution method.
+For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000}.
 
-For this work we adopt the matrix-free level operators in GMG and Chebyshev smoothing.
+For this work we adopt the matrix-free level operators within the GMG and Chebyshev pre- and post-smoothing on each level.
 Given a generally heterogeneous non-linear material, definition of transfer operations (restriction and prolongation) and level operators is not trivial.
-One approach is to start from a arbitrary heterogeneous material at the fine scale and employ non-linear homogenization theory to design specific  transfer operators \cite{Miehe2007}.
+One approach is to start from an arbitrary heterogeneous material at the fine scale and employ non-linear homogenization theory to design specific  transfer operators \cite{Miehe2007}.
 In this approach the fine-scale displacement is decomposed into long-wave and short-wave contributions by additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
-In case of a homogeneous patch, the transfer operators reduce to the standard geometric transfer.
-Based on this conclusion we develop our multigrid preconditioner.
-Namely, we assume that the coarsest level can accurately describe the heterogeneous fine-strain elastic material.
-In this case the restriction and prolongation operations for patches of elements are always performed for the same material.
-In order to define the level operators we adopt the following approach.
+In the case of a homogeneous patch, the transfer operators reduce to the standard geometric transfer.
+Based on this observation we adopt the following approach.
+It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous fine-strain elastic material.
+In this case the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric restriction and prolongation operations.
+We define the level operators as follows.
 Recall that the fine scale tangent is obtain by linearization around the current displacement field $\gz u^h$.
-We then restrict this displacement field to multigrid levels and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
-At the coarsest level we adopt CG solver which shall reduce residual by three orders of magnitude within at most hundred of steps.
+We then restrict this displacement field to multigrid levels and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}, essentially corresponding to the linearization around smoothed representation of displacement field.
+Note that by doing so the coarse level operator (tangent) $K^{l+1}$ on level $l+1$ is not directly related to the fine level operator (tangent) $K^{l}$ on level $l$ via $K^{l+1}=I^{l+1}_{l}  K^l I^l_{l+1}$, where $I^l_{l+1}$ and $I^{l+1}_l$ are prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
+Finally, at the coarsest level we adopt CG solver which shall reduce residual by three orders of magnitude within at most hundred of steps.
+The classical multigrid V-cycle approach with two pre- and post-smoothing is adopted.
 
@@
 Therefore our future work will be focused on extending this proposed here matrix-free multigrid solution approach to homogenization based transfer operators \cite{Miehe2007}.
+We are also interested  extending these methods to incompressible three-field formulation.
 

### 2018-09-18 doc/large_strain.tex
Update acknowledgments

@@
 
-DD acknowledge the financial support of the German Science Foundation (Deutsche Forschungs-Gemeinschaft, DFG), grant DA 1664/2-1
+D.~Davydov acknowledge the financial support of the German Research Foundation (DFG), grant DA 1664/2-1
 and the Competence Network for Technical and Scientific High Performance Computing in Bavaria (KONWIHR).
-DD, DA and JPP are grateful to Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for a fruitful discussion on matrix-free operator evaluation approaches.
-%
+
+D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for a fruitful discussion on matrix-free operator evaluation approaches.
+
+D.~Arndt was supported by the German Research Foundation (DFG) under the project ``High-order discontinuous
+Galerkin for the exa-scale'' (\mbox{ExaDG}) within the priority program ``Software
+for Exascale Computing'' (SPPEXA).
+
 {\color{red}
-DA was supported ...
-%
-JPP was supported by ...
+J-P.~Pelteret was supported by the European Research Council (ERC) through
+the Advanced Grant 289049 MOCOPOLY.(?)
 }
-%
-PS acknowledges the support of the Cluster of Excellence Engineering of Advanced Materials (EAM) which made this collaboration possible.
+
+P.~Steinmann acknowledges the support of the Cluster of Excellence Engineering of Advanced Materials (EAM) which made this collaboration possible.
 

### 2018-09-18 doc/large_strain.tex
minor in GMG

@@
 Given a generally heterogeneous non-linear material, definition of transfer operations (restriction and prolongation) and level operators is not trivial.
-One approach is to start from an arbitrary heterogeneous material at the fine scale and employ non-linear homogenization theory to design specific  transfer operators \cite{Miehe2007}.
+One approach is to start from an arbitrary heterogeneous material at the fine scale and employ non-linear homogenization theory to design specific transfer operators \cite{Miehe2007}\footnote{
+  We note that this approach is limited to the case at each quadrature point within the element the fourth order tangent operator is the same.
+}.
 In this approach the fine-scale displacement is decomposed into long-wave and short-wave contributions by additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
@@
 It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous fine-strain elastic material.
-In this case the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric restriction and prolongation operations.
+In this case the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
 We define the level operators as follows.
 Recall that the fine scale tangent is obtain by linearization around the current displacement field $\gz u^h$.
-We then restrict this displacement field to multigrid levels and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}, essentially corresponding to the linearization around smoothed representation of displacement field.
-Note that by doing so the coarse level operator (tangent) $K^{l+1}$ on level $l+1$ is not directly related to the fine level operator (tangent) $K^{l}$ on level $l$ via $K^{l+1}=I^{l+1}_{l}  K^l I^l_{l+1}$, where $I^l_{l+1}$ and $I^{l+1}_l$ are prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
+We then restrict this displacement field to multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
+Essentially level operators correspond to the linearization around smoothed representation of displacement field.
+Note that by doing so the operator (tangent) $K^{l+1}$ on level $l+1$ is not directly related to the operator (tangent) $K^{l}$ on level $l$ via $K^{l+1}=I^{l+1}_{l}  K^l I^l_{l+1}$, where $I^l_{l+1}$ and $I^{l+1}_l$ are prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
 Finally, at the coarsest level we adopt CG solver which shall reduce residual by three orders of magnitude within at most hundred of steps.

### 2018-09-18 doc/large_strain.tex
a few references on homogenization

@@
 Given a generally heterogeneous non-linear material, definition of transfer operations (restriction and prolongation) and level operators is not trivial.
-One approach is to start from an arbitrary heterogeneous material at the fine scale and employ non-linear homogenization theory to design specific transfer operators \cite{Miehe2007}\footnote{
+One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1985, Hill1972,Hashin1983,Castaneda1997} to design specific transfer operators \cite{Miehe2007}\footnote{
   We note that this approach is limited to the case at each quadrature point within the element the fourth order tangent operator is the same.

### 2018-09-18 doc/large_strain.tex
typo

--- 2018-09-18 doc/large_strain.tex
@@
 Given a generally heterogeneous non-linear material, definition of transfer operations (restriction and prolongation) and level operators is not trivial.
-One approach is to start from an arbitrary heterogeneous material at the fine scale and employ non-linear homogenization theory to design specific transfer operators \cite{Miehe2007}\footnote{
-  We note that this approach is limited to the case at each quadrature point within the element the fourth order tangent operator is the same.
+One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1985, Hill1972,Hashin1983,Castaneda1997} to design specific transfer operators \cite{Miehe2007}\footnote{
+  We note that this approach is limited to the case where for each quadrature point within an element the fourth order tangent operator is the same.
 }.

### 2018-09-18 doc/large_strain.tex
section Bibliography
a few more GMG references

@@
 As a result the number of iterations for convergence becomes independent of the number of mesh refinements and thus leads to the $\mathcal O (N)$ solution method.
-For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000}.
+For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
 
@@
 The classical multigrid V-cycle approach with two pre- and post-smoothing is adopted.
+For other approaches that combine homogenization and multigrid methods for small strain elastic materials, see \cite{Bayreuther2003,Fish1995}.
 
@@
 
+\section*{Bibliography}
+
 \bibliographystyle{elsarticle-num}

### 2018-09-18 doc/large_strain.tex
cite Mathias master thesis

@@
 To the best of our knowledge, the matrix-free methods are not yet widely adopted within the solid mechanics community.
-We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018}.
+We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a master thesis of a student of two of the authors \cite{Mentler2017}.
 A possible reason is that tangent operators (coming from linearization of the nonlinear balance equations) are

### 2018-09-18 doc/large_strain.tex
as agreed, TODO for JP ;-)
fill a bit CM

@@
 
-\subsection{Kinematics}
+{\color{red}
+TODO: JP please fill in this part based on your other papers.
+}
 
@@
 is defined via the mapping $\gz x = \gz \varphi (\gz X,t)$.
+%
+Linear deformation map is assumed
+$\d \gz x = \gz F \cdot \d \gz X$,
+where $\gz F := \gz \nabla_{X} \, \gz \varphi$ is the deformation gradient.
+For conservative systems without the volumetric load the total potential energy functional $\mcl E$ is introduced as
+\begin{align}
+\mcl E =
+\int_{\mcl B_0} \mcl \psi(\gz \varphi, \gz F; \gz X) \,
+- \int_{\partial \mcl B_0^N} \gz T \cdot \gz u \d S \, ,
+\end{align}
+where $\gz T$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in reference configuration, and $\mcl \psi$ denotes the strain-energy per unit reference volume.
 
-Linear deformation map is assumed for the macroscopic point
+The principal of stationary potential energy requires that the directional derivative with respect to the displacement $\gz u$
 \begin{align}
-\d \gz x = \gz F \cdot \d \gz X,
-\end{align}
-where $\gz F := \gz \nabla_{X} \, \gz \varphi$ is the deformation gradient.
-
-\subsection{Kinetics}
-
-For conservative systems the total potential energy is introduced as
-\begin{align}
-\mcl E = \int_{\mcl B_0} \mcl U_0(\gz \varphi, \gz F; \gz X) \,
-\end{align}
-that must be stationary in order for the system to be in equilibrium:
-\begin{align}
-\delta \mcl E = 0 \, .
+\delta \mcl E = \frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, .
 \label{eq:stationary}
 \end{align}
+vanishes for all directions $\delta \gz u$ at the equilibrium state.
+The vector field $\delta \gz u$ should be equal to zero at the Dirichlet boundary $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N$.
 

### 2018-09-18 doc/large_strain.tex
a bit more CM

@@
 % \def\msf  #1{           \mbox{\!\!      $\sf #1$}}
-% \def\Grad #1 {{\rm Grad} #1}
-% \def\grad #1 {{\rm grad} #1}
+\def\Grad #1 {{\rm Grad} #1}
+\def\grad #1 {{\rm grad} #1}
 \def\Div {\mbox{Div\,}}
@@
 $\d \gz x = \gz F \cdot \d \gz X$,
-where $\gz F := \gz \nabla_{X} \, \gz \varphi$ is the deformation gradient.
+where $\gz F := \Grad \, \gz \varphi$ is the deformation gradient.
 For conservative systems without the volumetric load the total potential energy functional $\mcl E$ is introduced as
@@
 \mcl E =
-\int_{\mcl B_0} \mcl \psi(\gz \varphi, \gz F; \gz X) \,
+\int_{\mcl B_0} \mcl \psi(\gz \varphi, \gz F; \gz X) \d V \,
 - \int_{\partial \mcl B_0^N} \gz T \cdot \gz u \d S \, ,
@@
 \begin{align}
-\delta \mcl E = \frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, .
+\delta \mcl E =
+\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, .
 \label{eq:stationary}
@@
 The vector field $\delta \gz u$ should be equal to zero at the Dirichlet boundary $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N$.
+
+The stationary conditions yield
+\begin{align}
+F(\gz u, \delta \gz u) =
+\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \d V \,
+-
+\int_{\partial \mcl B_0^N} \gz T \cdot \delta \gz u \d S
+= 0 \, ,
+\end{align}
+where $\gz P := \partial \mcl \psi / \partial \gz F$ is the Piola stress.
 

### 2018-09-18 doc/large_strain.tex
same

@@
 {\color{red}
-TODO: JP please fill in this part based on your other papers.
+TODO: JP please adjust this part to your liking
 }
@@
 is defined via the mapping $\gz x = \gz \varphi (\gz X,t)$.
+The displacement field reads $\gz u = \gz x - \gz X$.
 %
@@
 $\d \gz x = \gz F \cdot \d \gz X$,
-where $\gz F := \Grad \, \gz \varphi$ is the deformation gradient.
+where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is the deformation gradient.
 For conservative systems without the volumetric load the total potential energy functional $\mcl E$ is introduced as
@@
 \mcl E =
-\int_{\mcl B_0} \mcl \psi(\gz \varphi, \gz F; \gz X) \d V \,
+\int_{\mcl B_0} \mcl \psi(\gz F; \gz X) \d V \,
 - \int_{\partial \mcl B_0^N} \gz T \cdot \gz u \d S \, ,
 \end{align}
-where $\gz T$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in reference configuration, and $\mcl \psi$ denotes the strain-energy per unit reference volume.
+where $\gz T$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \gz u^D$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
 
-The principal of stationary potential energy requires that the directional derivative with respect to the displacement $\gz u$
+The principal of stationary potential energy requires that at the equilibriums the directional derivative with respect to the displacement $\gz u$
 \begin{align}
@@
 \end{align}
-vanishes for all directions $\delta \gz u$ at the equilibrium state.
-The vector field $\delta \gz u$ should be equal to zero at the Dirichlet boundary $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N$.
-
-The stationary conditions yield
+vanishes for all directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
+%
+This leads to the following non-linear equation
 \begin{align}

### 2018-09-18 doc/large_strain.tex
a bit more CM

@@
 
-In this section we briefly summarize balance equations of fine-strain hyperelastic continuum.
+In this section we briefly summarize balance equations of fine-strain hyperelastic continuum in Lagrangian (also known as referential) description.
 
@@
 
-Deformation of the body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$
-is defined via the mapping $\gz x = \gz \varphi (\gz X,t)$.
+%
+% Kinematics
+%
+Deformation of the body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$ at time $t$
+is defined via the mapping $\gz \varphi_t: \mcl B \rightarrow \mathbb E^3$, which places material points of $\mcl B$ in Euclidean space $\mathbb E^3$.
+The location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
 The displacement field reads $\gz u = \gz x - \gz X$.
+
+%
+% Deformation gradient
 %
@@
 $\d \gz x = \gz F \cdot \d \gz X$,
-where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is the deformation gradient.
+where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient.
+The mapping has to be one-to-one and exclude a self penetration, consequently the Jacobian has to be positive $J = \det \gz F > 0$.
+Deformation gradient can be split in the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors as $\gz F = \gz R \gz U = \gz V \gz R$.
+%
+% strain measures
+%
+In what follows we characterize the material behavior using the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$, where $\gz C$ is right Cauchy-Green tensor $\gz C := \gz F^T \cdot \gz F = \gz U^2$.
+Clearly $J = \sqrt{\det \gz C}$ and $\partial (\bullet)/\partial \gz E = 2\, \partial (\bullet)/\partial \gz C$.
+
 For conservative systems without the volumetric load the total potential energy functional $\mcl E$ is introduced as
@@
 \mcl E =
-\int_{\mcl B_0} \mcl \psi(\gz F; \gz X) \d V \,
+\int_{\mcl B_0} \mcl \psi(\gz C) \d V \,
 - \int_{\partial \mcl B_0^N} \gz T \cdot \gz u \d S \, ,
@@
 F(\gz u, \delta \gz u) =
-\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \d V \,
+\int_{\mcl B_0} \gz S : \delta \gz E \d V \,
 -
@@
 \end{align}
-where $\gz P := \partial \mcl \psi / \partial \gz F$ is the Piola stress.
+where $\gz S := 2\, \partial \mcl \psi / \partial \gz C$ is the second Piola-Kirchhoff stress tensor.
 

### 2018-09-18 doc/large_strain.tex
finally get to the correct weak form

@@
 Clearly $J = \sqrt{\det \gz C}$ and $\partial (\bullet)/\partial \gz E = 2\, \partial (\bullet)/\partial \gz C$.
+We will also use the left Cauchy-Green tensor $\gz b := \gz F \cdot \gz F^T = \gz V^2$.
 
@@
 \mcl E =
-\int_{\mcl B_0} \mcl \psi(\gz C) \d V \,
-- \int_{\partial \mcl B_0^N} \gz T \cdot \gz u \d S \, ,
+\int_{\mcl B_0} \mcl \psi(\gz F) \d V \,
+- \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \d S \, ,
 \end{align}
-where $\gz T$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \gz u^D$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
+where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
 
@@
 %
-This leads to the following non-linear equation
+This leads to the following scalar non-linear equation
 \begin{align}
 F(\gz u, \delta \gz u) =
-\int_{\mcl B_0} \gz S : \delta \gz E \d V \,
+\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \d V \,
 -
-\int_{\partial \mcl B_0^N} \gz T \cdot \delta \gz u \d S
+\int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \d S
 = 0 \, ,
 \end{align}
-where $\gz S := 2\, \partial \mcl \psi / \partial \gz C$ is the second Piola-Kirchhoff stress tensor.
+where $\gz P := \partial \mcl \psi / \partial \gz F$ is the first Piola-Kirchhoff stress tensor.
+The double contraction in the first term can be re-written in terms of the symmetric Kirchhoff stress tensor $\gz \tau := \gz P \cdot \gz F^T$ as
+\begin{align}
+\gz P : \Grad \, \delta \gz u =
+(\gz \tau \cdot \gz F^{-T}) : \Grad \, \delta \gz u =
+\gz \tau : (\Grad \, \delta \gz u : \gz F^{-1}) =
+\gz \tau : \grad \, \delta \gz u \, ,
+\end{align}
+and therefore
+\begin{align}
+  F(\gz u, \delta \gz u) =
+  \int_{\mcl B_0} \gz \tau : \grad \, \delta \gz u \d V \,
+  -
+  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \d S
+  = 0 \, .
+\end{align}
+Note that $\gz \tau$ is the Piola transformation of the second Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
+that is $\gz \tau = \gz F \cdot \gz S \cdot \gz F^T$.
 

### 2018-09-18 doc/large_strain.tex
move todo

@@
 
-In this section we briefly summarize balance equations of fine-strain hyperelastic continuum in Lagrangian (also known as referential) description.
-
-{\color{red}
-TODO: JP please adjust this part to your liking
-}
-
 %
@@
 
-{\color{red}TODO: linearize and show how tangent look like.
-Also show what it boils down in the case of small strain.
+{\color{red}TODO: JP add the linearization and material/geometric tangents.
 }

### 2018-09-18 doc/large_strain.tex
symgrad

@@
   F(\gz u, \delta \gz u) =
-  \int_{\mcl B_0} \gz \tau : \grad \, \delta \gz u \d V \,
+  \int_{\mcl B_0} \gz \tau : \grad^{s} \, \delta \gz u \d V \,
   -
   \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \d S
-  = 0 \, .
+  = 0 \, ,
 \end{align}
+where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$.
 Note that $\gz \tau$ is the Piola transformation of the second Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,

### 2018-09-18 doc/large_strain.tex
add linearization

@@
 \def\Grad #1 {{\rm Grad} #1}
-\def\grad #1 {{\rm grad} #1}
+\def\grad {\mbox{grad}}
 \def\Div {\mbox{Div\,}}
 \def\div {\mbox{div\,}}
-\def\d {\,\mbox{d}}
-\def\D {\,\mbox{D}}
+\def\d {\mbox{d}}
+\def\D {\mbox{D}}
 % \newcommand{\norm}[1]{\left\lvert #1 \right\rvert}
@@
 
+\subsection{Weak form}
 %
@@
 Deformation gradient can be split in the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors as $\gz F = \gz R \gz U = \gz V \gz R$.
+We shall denote the Piola transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
+\begin{align}
+  \chi\left( \mathbf{A} \right)_{ij}
+  &= F_{iA} A_{AB} F_{jB} \\
+\chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
+  &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}
+\end{align}
+
 %
@@
 (\gz \tau \cdot \gz F^{-T}) : \Grad \, \delta \gz u =
-\gz \tau : (\Grad \, \delta \gz u : \gz F^{-1}) =
+\gz \tau : (\Grad \, \delta \gz u \cdot \gz F^{-1}) =
 \gz \tau : \grad \, \delta \gz u \, ,
@@
   = 0 \, ,
+\label{eq:weak_form}
 \end{align}
@@
 Note that $\gz \tau$ is the Piola transformation of the second Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
-that is $\gz \tau = \gz F \cdot \gz S \cdot \gz F^T$.
+that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.
 
@@
 
-{\color{red}TODO: JP add the linearization and material/geometric tangents.
-}
+In order to apply the Newton method, linearization of \eqref{eq:weak_form} around a given solution field $\overline{\gz u}$ is required. This is given by
+\begin{align}
+F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
+F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
+\end{align}
+For conservative traction boundary conditions, the directional derivative is given by
+\begin{align}
+\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u)
+=
+\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \d V
++
+\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V
+\end{align}
+This can be simplified \cite{Wriggers2008} to
+\begin{align}
+\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
+  =
+  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \d V
+  +
+  \int_{\mcl B_0}
+  \overline{\grad}\delta \gz u :
+  \left[
+  \overline{\grad} \Delta \gz u \cdot
+  \overline{\gz \tau}
+  \right]
+  \d V
+\label{eq:tangent}
+\end{align}
+Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$ and the fourth order tangent tensor is push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
+The first term in \eqref{eq:tangent} is called material part of the tangent operator as it is related to the linearization of the Kirchhoff stress.
+Whereas the second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from linearization of the deformation gradient $\gz F$.
 
@@
 
-Neo-Hookean model
+For the current study we use the Neo-Hookean model
 \begin{gather}
@@
 \end{gather}
-where $\mu$ and $\lambda$ respectively denote the shear modulus and Lam\'{e} parameter,
-and the volumetric Jacobian $J = \det\left(\mathbf{F}\right) = \sqrt{\det\left(\mathbf{C}\right)}$.
-\begin{gather}
+where $\mu$ and $\lambda$ respectively denote the shear modulus and Lam\'{e} parameter. It can be shown that the first and second derivatives of the strain energy function are given by
+\begin{align}
 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}}
-  = \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1}
-\end{gather}
-\begin{gather}
+  &= \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1} \\
 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
-  = \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
+  &= \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
   + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
-\end{gather}
-
-Denoting $\chi\left( \bullet \right)$ as the Piola transformation, which implies
-\begin{gather}
-\chi\left( \mathbf{A} \right)_{ij}
-  = F_{iA} A_{AB} F_{jB} \\
-\chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
-  = F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}
-\end{gather}
-for rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$, then the Kirchhoff stress and its associated material tangent are
+\end{align}
+The Kirchhoff stress and its associated material tangent are
 \begin{gather}
@@
   + 2 \lambda \trace\left( \bullet \right) \mathbf{I}
+\label{eq:simplified_action}
 \end{gather}
@@
 This also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
-This algorithm utilize the chosen constitutive relationship and therefore the action of the material fourth order tangent $\boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$ can be evaluated by taking a linear combination this tensor and its trace times identity tensor.
-To that end we also cache two scalars that depend on the Jacobian $J$.
+This algorithm utilize the chosen constitutive relationship and therefore the action of the material fourth order tangent $J \boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$ is cheap to evaluate, see \eqref{eq:simplified_action}.
+Here we also cache two scalars that depend on the Jacobian $J$.
 

### 2018-09-18 doc/large_strain.tex
add system of equations form weak form and linearization

@@
 the associated FE space of continuous piecewise elements of a fixed polynomial degree. % : $V^h \subset H^1 (\mcl B_0)$.
-The deformation map are given in a
-a vector space spanned by standard vector--valued FE basis functions $\gz N^i(\gz x)$ (e.g. polynomials with local support):
+The displacement fields are given in a
+a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support):
 \begin{alignat}{2}
-       \gz \varphi^h &=:  \sum_{i \in \mcl I_\varphi}       \varphi_i \gz N^i_\varphi (\gz X) \quad \quad \quad
-\delta \gz \varphi^h &&=: \sum_{i \in \mcl I_\varphi} \delta \varphi_i \gz N^i_\varphi (\gz X) \,,
+       \gz u^h &=:  \sum_{i \in \mcl I_\varphi} u^i \gz N_i (\gz X) \quad \quad \quad
+\delta \gz u^h &&=: \sum_{i \in \mcl I_\varphi} \delta u^i \gz N_i (\gz X) \,,
 \end{alignat}
 where superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom (DoF).
-For the sake of this study $\gz N^i_\varphi (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.
+For the sake of this study $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.
+
+The Newton-Raphson solution approach is adopted for the here considered nonlinear problem.
+Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is to be obtained as a solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent})
+\begin{align}
+  K_{ij} \Delta u_j &= - F_i \\
+  K_{ij} &=
+  \int_{\mcl B_0} \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j \d V
+  +
+  \int_{\mcl B_0}
+  \overline{\grad}\gz N_i :
+  \left[
+  \overline{\grad} \gz N_j \cdot
+  \overline{\gz \tau}
+  \right]
+  \d V
+  \\
+  F_i &=
+  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \d V \,
+  -
+  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \d S
+\end{align}
+Here $\gz K$ is the discrete tangent operator and $\gz F$ is the discrete gradient.
 

### 2018-10-10 doc/large_strain.tex
spaces in equatinos
various improvements to the text

@@
 Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016}.
+{\color{red}Daniel: please add more links if you want, maybe GPU related}.
 
@@
 Another reason could be the smoothness of the underlying solution.
-FEM with linear or quadratic elements (to avoid locking) are often adopted in solid mechanics community.
+FEM with linear or quadratic elements, that avoid locking, are often adopted in solid mechanics community.
+Higher order elements are rarely employed.
 However we note that large scale computations with $10^{12}$ unknowns even with linear FEM inevitably require matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
@@
 We propose and numerically investigate different implementations of the finite-strain hyper-elastic tangent operator.
-In order to improve the convergence of iterative solvers, we also propose a way to construct level tangent operators
+In order to improve the convergence of iterative solvers, we also propose a way to construct level matrix-free tangent operators
 and employ them to define a geometric multigrid preconditioner.
@@
 \mcl E =
-\int_{\mcl B_0} \mcl \psi(\gz F) \d V \,
-- \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \d S \, ,
+\int_{\mcl B_0} \mcl \psi(\gz F) \, \d V \,
+- \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \, \d S \, ,
 \end{align}
@@
 F(\gz u, \delta \gz u) =
-\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \d V \,
+\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \, \d V \,
 -
-\int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \d S
+\int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
 = 0 \, ,
@@
   F(\gz u, \delta \gz u) =
-  \int_{\mcl B_0} \gz \tau : \grad^{s} \, \delta \gz u \d V \,
+  \int_{\mcl B_0} \gz \tau : \grad^{s} \, \delta \gz u \, \d V \,
   -
-  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \d S
+  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
   = 0 \, ,
@@
 =
-\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \d V
+\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \, \d V
 +
@@
   =
-  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \d V
+  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V
   +
@@
 The displacement fields are given in a
-a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support):
+a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support on a patch of elements):
 \begin{alignat}{2}
@@
 \begin{align}
-  K_{ij} \Delta u_j &= - F_i \\
-  K_{ij} &=
-  \int_{\mcl B_0} \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j \d V
+  A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
+  A_{ij} &=
+  \int_{\mcl B_0}
+  \underbrace{
+  \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j
   +
-  \int_{\mcl B_0}
   \overline{\grad}\gz N_i :
@@
   \right]
+  }_{\tilde{a}(\gz N_i, \gz N_j)}
   \d V
+  \label{eq:algebraic_tangent}
   \\
   F_i &=
-  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \d V \,
+  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \, \d V \,
   -
-  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \d S
+  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \, \d S
 \end{align}
-Here $\gz K$ is the discrete tangent operator and $\gz F$ is the discrete gradient.
+Here $\gz A$ is the discrete tangent operator, $\gz F$ is the discrete gradient
+and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the bilinear form for the trial solution field $\overline{\gz u^h}$.
 
@@
 \label{sec:mf}
-Classically the approach for finding a discrete solution to a linearized partial differential equation can be described as follows:
-First, ansatz spaces and a suitable bilinear form is defined. Then, a matrix and force vector corresponding to the discretization are assembled.
-Finally, a direct or iterative solver is used to solve the linear system for the discretized solution.
+Classically in order to solve \eqref{eq:linear_system}, the matrix $\gz K$ and the force vector $\gz F$ corresponding to the discretization are assembled and
+a direct or iterative solver is used to solve the linear system.
 On modern computer architectures however, it turns out that loading data is significantly slower than performing the arithmetic operations when solving.
@@
 rather than loading matrix elements from memory.
-For iterative solvers, this is possible since it is sufficient to compute matrix vector products. More precisely, the linear system corresponding to a bilinear form
-$a\colon V_h\times V_h\to\mathbb{R}$ and right-hand side $f\colon V_h\to\mathbb{R}$ defined with respect to the ansatz space $V_h$ can be represented by
+For iterative solvers, this is possible since it is sufficient to compute matrix vector products.
+%
+The matrix-vector product $A x$ can be expressed by
 \begin{align*}
- &&a(v,u) &= f(v) && \forall v\in V_h\\
- \Longleftrightarrow &&a(\varphi_i, u) &= f(\varphi_i) && \forall \varphi_i \\
- \Longleftrightarrow &&\sum_j a(\varphi_i,\varphi_j) u_j &= f(\varphi_i) && \forall \varphi_i \\
- \Longleftrightarrow &&A x &= b
+ (Ax)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
+        &= \sum_K\sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz x_q) x_j w_q
 \end{align*}
-where $A_{i,j}=a(\varphi_i, \varphi_j)$ and $b_i= f(\varphi_i)$.
-Thus, the matrix-vector product $Ax$ can be expressed by
-\begin{align*}
- (Ax)_i &= \sum_j a(\varphi_i,\varphi_j) x_j \\
-        &= \sum_K\sum_q \sum_j \tilde{a}_K(\varphi_i,\varphi_j)(x_q) x_j w_q
-\end{align*}
-where in the last step $\tilde{a}_K(\varphi_i,\varphi_j)$ is a representation for the evaluation of the bilinear in one quadrature point and a quadrature rule is used.
+where in the last step $\tilde{a}_K(\varphi_i,\varphi_j)(\gz x_q)$ is a representation for the evaluation of the bilinear in one quadrature point and a quadrature rule is used.
 For the sake of demonstration, let us assume $A$ represents a mass matrix. In this case, it holds
 \begin{align*}
- \tilde{a}_K(\varphi_i,\varphi_j)(x_q) = \varphi_i(x_q)\varphi_j(x_q).
+ \tilde{a}_K(N_i,N_j)(\gz x_q) = N_i(\gz x_q)N_j(\gz x_q).
 \end{align*}
@@
 More information on techniques to increase the computational density can be found in \cite{kronbichler12,vos10}.
-We briefly note that the SIMD vectorization is applied at the element level, that is the operator is applied simultaneously on several several elements (called ``block'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
+We briefly note that the SIMD vectorization is applied at the element level, that is the operator is applied simultaneously on several elements (called ``block'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
 
 Given that the matrix-free operators are applied on-the-fly, it's crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
-Below we propose three algorithms to implement tangent operator of finite-strain linear elasticity using matrix-free approach, that employ different caching strategies.
+Below we propose three algorithms to implement tangent operator of finite-strain linear elasticity (see \eqref{eq:algebraic_tangent}) using matrix-free approach.
 %
 In order to obtain gradients with respect to the deformed configuration we adopted \textit{MappingQEulerian} class from \texttt{deal.II} library.
-In addition to the standard mapping from the reference (unit / isoparametric) element to the element in real space, this class adds mapping to the deformed configuration given a displacement field.
+In addition to the standard mapping from the reference (unit / isoparametric) element to the element in real space, given a displacement field this class adds mapping to the deformed configuration.
 %
@@
   cacheed $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
-  \Output{action of the FE tangent on $\gz x$}
+  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
   \ForEach{ element $K \in \Omega^h$ }{
@@
 
-Algorithm \ref{alg:mf_scalar} caches a single scalar which involves logarithm of Jacobian -- a computationally demanding operation compared to addition and multiplications.
+Algorithm \ref{alg:mf_scalar} caches a single scalar which involves logarithm of Jacobian -- a computationally demanding operation compared to additions and multiplications.
 As evident from Section \ref{sec:theory}, in order to apply the operator we need to re-evaluate Kirchhoff stress at each quadrature point.
@@
   $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point}
-  \Output{action of the FE tangent on $\gz x$}
+  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
   \ForEach{ element $K \in \Omega^h$ }{
@@
 In order to avoid re-calculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point.
-This also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
-This algorithm utilize the chosen constitutive relationship and therefore the action of the material fourth order tangent $J \boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$ is cheap to evaluate, see \eqref{eq:simplified_action}.
-Here we also cache two scalars that depend on the Jacobian $J$.
+In this case we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
+This algorithm utilizes the chosen constitutive relationship and therefore the action of the material fourth order tangent $J \boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$ is cheap to evaluate, see \eqref{eq:simplified_action}.
+To that end we cache two scalars that depend on the Jacobian $J$.
 
@@
   cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point}
-  \Output{action of the FE tangent on $\gz x$}
+  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
   \ForEach{ element $K \in \Omega^h$ }{
@@
 
-Finally Algorithm \ref{alg:mf_tensor4} represents the most general algorithm which does not assume specific constitutive relationship.
+Finally Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent tensor.
 Clearly in this case we have to cache the fourth order symmetric tensor $\boldsymbol{\mathcal{C}}$ for each element and quadrature point and perform the double contraction with the second-order symmetric tensor on-the-fly.
@@
 In this approach the fine-scale displacement is decomposed into long-wave and short-wave contributions by additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
-In the case of a homogeneous patch, the transfer operators reduce to the standard geometric transfer.
+However, for a homogeneous patch of elements, the transfer operators reduce to the standard geometric transfer.
 Based on this observation we adopt the following approach.
@@
 We define the level operators as follows.
-Recall that the fine scale tangent is obtain by linearization around the current displacement field $\gz u^h$.
+Recall that the fine scale tangent \eqref{eq:algebraic_tangent} is obtain by linearization around the current displacement field $\gz u^h$.
 We then restrict this displacement field to multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
 Essentially level operators correspond to the linearization around smoothed representation of displacement field.
-Note that by doing so the operator (tangent) $K^{l+1}$ on level $l+1$ is not directly related to the operator (tangent) $K^{l}$ on level $l$ via $K^{l+1}=I^{l+1}_{l}  K^l I^l_{l+1}$, where $I^l_{l+1}$ and $I^{l+1}_l$ are prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
+Note that by doing so the operator (tangent) $A^{l+1}$ on level $l+1$ is not directly related to the operator (tangent) $A^{l}$ on level $l$ via $A^{l+1}=I^{l+1}_{l}  A^l I^l_{l+1}$, where $I^l_{l+1}$ and $I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
 Finally, at the coarsest level we adopt CG solver which shall reduce residual by three orders of magnitude within at most hundred of steps.
@@
 Note that a relatively coarse mesh is adopted at the coarsest multigrid level, however the geometry of inclusions is captured accurately by employing manifold description of the boundaries and interfaces.
+{\color{red}Daniel: do we have any paper referencing manifolds in deal.II?}
 Both domains are fixed at the bottom surface and the distributed load is applied at the top in $(1,0)$ and $(1,1,0)$ direction for 2D and 3D problems, respectively.
 The load of $0.5\,\rm{N}$ is applied in 5 steps.
-The displacement tolerance for Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for forces $10^{-8}$.
+The displacement tolerance for Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for forces is $10^{-8}$.
 The relative convergence criteria for linear solver is $10^{-6}$.
@@
 %
-The results of the matrix-free approach with geometric-multigrid preconditioner are compared to the matrix-based approach with algebraic multigrid (AMG) preconditioner using Trilinos \cite{Heroux2005} library.
+The results of the matrix-free approach with geometric-multigrid preconditioner are compared to the matrix-based approach with algebraic multigrid (AMG) preconditioner using Trilinos \cite{Heroux2005} library, version 12.12.1.
 The aggregation threshold for AMG is taken as $10^{-4}$.
@@
 Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly shows the influence of different caching strategies.
-Two conclusions can be drawn: scalar caching implementation is the most time consuming at each step we additionally evaluate the gradient of displacement field in reference configuration, required to evaluate the Kirchhoff stress $\gz \tau$.
-On another hand there is little difference between the approach where the fourth order material tangent tensor is cached (``tensor4'') and the one where we only cache the second order Kirchhoff tensor and use the chosen hyper-elastic constitutive equation to efficiently implement the action of the material tangent on the second order symmetric tensor.
+Two conclusions can be drawn: scalar caching implementation is the most time consuming, at each step we additionally evaluate the gradient of displacement field in reference configuration, required to evaluate the Kirchhoff stress $\gz \tau$.
+On another hand there is little difference between the approach where the fourth order material tangent tensor is cached (``tensor4'') and the one where we only cache the second order Kirchhoff tensor and utilize the chosen hyper-elastic constitutive equation to efficiently implement the action of the material tangent on the second order symmetric tensor.
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
@@
 Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_Emmy_cg3} show a very mild increase in the average number of CG iteration for different polynomial degrees.
-On another hand the black-box AGM preconditioner require much more iterations for convergence.
-Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution time polynomial degrees up to $p=5$.
+On another hand the black-box AGM preconditioner requires much more iterations for convergence.
+Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution time for polynomial degrees up to $p=5$.
 For higher orders in 2D the walltime is comparable to the AMG preconditioner, see Figure \ref{fig:benchmark_miehe_Emmy_sol2}.
@@
 This should also improve the performance of the preconditioner for lower degrees.
+An valid alternative is to employ matrix-based AMG at the coarsest level.
 
@@
 The implementation which caches the fourth order material tangent together with the second order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
-This indicates that the matrix-free implementation of tangent operator can be applied any constitutive model which operates with the material tangent on quadrature point level.
+This gives hope that the matrix-free implementation of tangent operator can be applied to any constitutive model which operates with the material tangent on the quadrature point level.
 
-We also studies the performance of the GMG preconditioner with standard geometric transfer operations between each level.
-The GMG was applied to heterogeneous material assuming that the coarsest level can provide adequate discretization of heterogeneity.
+We also studied the performance of the GMG preconditioner with standard geometric transfer operations between each level.
+The GMG was applied to heterogeneous material assuming that the coarsest level can provide adequate discretization of a heterogeneity.
 The numerical studies indicate that the proposed preconditioner lead to the solution approach that is faster than the matrix-based algebraic multigrid preconditioner for moderate polynomial degrees.
 However the considerable amount of time is spent in the coarse level solver.
-Therefore our future work will be focused on extending this proposed here matrix-free multigrid solution approach to homogenization based transfer operators \cite{Miehe2007}.
+Therefore our future work will be focused on extending the proposed here matrix-free multigrid solution approach to homogenization based transfer operators \cite{Miehe2007}.
 We are also interested  extending these methods to incompressible three-field formulation.

### 2018-10-14 doc/large_strain.tex
More improvements

@@
    showtabs=false,                  % show tabs within strings adding particular underscores
-   stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
+   stepnumber=2,                    % the step between two line-numbers. If it is 1, each line will be numbered
 %   stringstyle=\color{mymauve},     % string literal style
@@
     In order to improve the performance of the iterative solver, so-called matrix-free methods are
-    widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly.
+    widely adopted in the fluid mechanics community, where the matrix-vector products are computed on-the-fly.
 
-    In this work, we advance the application of matrix-free approaches to problems in solid mechanics.
-    We propose and numerically investigate different implementations of the finite-strain hyper-elastic tangent operator.
-    In order to improve the convergence of iterative solvers, we also propose a way to construct level tangent operators
+    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyper-elastic tangent operator numerically.
+    In order to improve the convergence behavior of iterative solvers, we also propose a way to construct level tangent operators
     and employ them to define a geometric multigrid preconditioner.
-    Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.
+    Our implementation employs MPI {\color{red}and Intel Threading Building Blocks parallelization} and SIMD vectorization.
     The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions.
@@
 In order to improve the performance of the iterative solver, so-called matrix-free methods are
-widely adopted in fluid mechanics community, where the matrix vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016}.
-Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016}.
-{\color{red}Daniel: please add more links if you want, maybe GPU related}.
+widely adopted in fluid mechanics community, where the matrix-vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016}.
+Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016, ljungkvist2017multigrid}.
 
-To the best of our knowledge, the matrix-free methods are not yet widely adopted within the solid mechanics community.
+To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
 We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a master thesis of a student of two of the authors \cite{Mentler2017}.
-A possible reason is that tangent operators (coming from linearization of the nonlinear balance equations) are
-more elaborate as compared to the fluid mechanics and it is not obvious whether the matrix-free methods can be advantageous in this case.
-Another reason could be the smoothness of the underlying solution.
-FEM with linear or quadratic elements, that avoid locking, are often adopted in solid mechanics community.
+A possible reason is that the tangent operators (coming from linearization of the nonlinear balance equations) are
+more elaborate as compared to the operators used in fluid mechanics and it is not obvious whether the matrix-free methods can be advantageous in this case.
+Another reason could be the use of lower order elements due to lacking smoothness of the underlying solution.
+FEM with linear or quadratic elements, that avoid locking, are often adopted in the solid mechanics community.
 Higher order elements are rarely employed.
-However we note that large scale computations with $10^{12}$ unknowns even with linear FEM inevitably require matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
-Therefore we believe that for large scale computations in solid mechanics it is crucial to adopt matrix-free approaches and develop efficient multigrid solvers.
+However, we note that large scale computations with $10^{12}$ unknowns even with linear FEM inevitably require matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
+Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.
 
-In this work, we advance the application of matrix-free approaches to problems in solid mechanics.
-We propose and numerically investigate different implementations of the finite-strain hyper-elastic tangent operator.
+In this work, we investigate different implementations of the finite-strain hyper-elastic tangent operator with respcet to their computational cost.
 In order to improve the convergence of iterative solvers, we also propose a way to construct level matrix-free tangent operators
@@
 The paper is organized as follows.
-In Section \ref{sec:theory} we briefly introduce partial differential equations of finite-strain solid mechanics.
-Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the proposed matrix-free operator evaluation approaches.
-In Section \ref{sec:gmg} we propose a geometric multigrid preconditioner suitable for finite-strain linear elasticity.
+In Section \ref{sec:theory}, we briefly introduce the partial differential equations used in  finite-strain solid mechanics.
+Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the different matrix-free operator evaluation approaches.
+In Section \ref{sec:gmg}, we propose a geometric multigrid preconditioner suitable for finite-strain linear elasticity.
 The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions in Section \ref{sec:example}.
-Summary is presented in Section \ref{sec:summary}.
+The results are summarized in Section \ref{sec:summary}.
 
@@
 %
-Deformation of the body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$ at time $t$
-is defined via the mapping $\gz \varphi_t: \mcl B \rightarrow \mathbb E^3$, which places material points of $\mcl B$ in Euclidean space $\mathbb E^3$.
+The deformation of a body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$ at time $t$
+is defined via the mapping $\gz \varphi_t: \mcl B \rightarrow \mathbb E^3$, which places material points of $\mcl B$ in the Euclidean space $\mathbb E^3$.
 The location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
@@
 %
-Linear deformation map is assumed
+We assume that the deformation map is linear such that
 $\d \gz x = \gz F \cdot \d \gz X$,
 where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient.
-The mapping has to be one-to-one and exclude a self penetration, consequently the Jacobian has to be positive $J = \det \gz F > 0$.
-Deformation gradient can be split in the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors as $\gz F = \gz R \gz U = \gz V \gz R$.
+The mapping has to be one-to-one and must exclude self-penetration.
+Consequently, the Jacobian has to be positive $J = \det \gz F > 0$.
+The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
 We shall denote the Piola transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
@@
 \chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
-  &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}
+  &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}.
 \end{align}
@@
 %
-In what follows we characterize the material behavior using the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$, where $\gz C$ is right Cauchy-Green tensor $\gz C := \gz F^T \cdot \gz F = \gz U^2$.
+In what follows, we characterize the material behavior using the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$, where $\gz C$ is the right Cauchy-Green tensor $\gz C := \gz F^T \cdot \gz F = \gz U^2$.
 Clearly $J = \sqrt{\det \gz C}$ and $\partial (\bullet)/\partial \gz E = 2\, \partial (\bullet)/\partial \gz C$.
@@
 
-For conservative systems without the volumetric load the total potential energy functional $\mcl E$ is introduced as
+For conservative systems without volumetric load, the total potential energy functional $\mcl E$ is introduced as
 \begin{align}
@@
 \end{align}
-where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
+where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
 
-The principal of stationary potential energy requires that at the equilibriums the directional derivative with respect to the displacement $\gz u$
+The principal of stationary potential energy requires for all  equilibrium states $\gz u$ that the directional derivative with respect to the displacement
 \begin{align}
@@
 
-In order to apply the Newton method, linearization of \eqref{eq:weak_form} around a given solution field $\overline{\gz u}$ is required. This is given by
+In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
 \begin{align}
 F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
-F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
+F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u).
 \end{align}
@@
 +
-\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V
+\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V.
 \end{align}
-This can be simplified \cite{Wriggers2008} to
+Following \cite{Wriggers2008}, this can be simplified to
 \begin{align}
@@
   \right]
-  \d V
+  \d V.
 \label{eq:tangent}
 \end{align}
-Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$ and the fourth order tangent tensor is push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
-The first term in \eqref{eq:tangent} is called material part of the tangent operator as it is related to the linearization of the Kirchhoff stress.
-Whereas the second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from linearization of the deformation gradient $\gz F$.
+Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
+The first term in \eqref{eq:tangent} is related to the linearization of the Kirchhoff stress and therefore called material part of the tangent operator.
+The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the the linearization of the deformation gradient $\gz F$.
 
@@
 
-For the current study we use the Neo-Hookean model
+For the current study, we use the Neo-Hookean model
 \begin{gather}
@@
 \end{gather}
-where $\mu$ and $\lambda$ respectively denote the shear modulus and Lam\'{e} parameter. It can be shown that the first and second derivatives of the strain energy function are given by
+where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively. It can be shown that the first and second derivatives of the strain energy function are given by
 \begin{align}
@@
   = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
-  + 2 \lambda \trace\left( \bullet \right) \mathbf{I}
+  + 2 \lambda \trace\left( \bullet \right) \mathbf{I}.
 \label{eq:simplified_action}
@@
 We now introduce a FE triangulation $\mathcal{T}^h$ of $\mcl B_0$ and
-the associated FE space of continuous piecewise elements of a fixed polynomial degree. % : $V^h \subset H^1 (\mcl B_0)$.
+the associated FE space of continuous piecewise elements with polynomial space of fixed degree. % : $V^h \subset H^1 (\mcl B_0)$.
 The displacement fields are given in a
@@
 \end{alignat}
-where superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom (DoF).
-For the sake of this study $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.
+where the superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom (DoF).
+%{\color{red} For the sake of this study, $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.}
 
 The Newton-Raphson solution approach is adopted for the here considered nonlinear problem.
-Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is to be obtained as a solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent})
+Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is to obtained as solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent})
 \begin{align}
@@
   -
-  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \, \d S
+  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \, \d S.
 \end{align}
@@
 rather than loading matrix elements from memory.
-For iterative solvers, this is possible since it is sufficient to compute matrix vector products.
+For iterative solvers, this is possible since it is sufficient to compute matrix-vector products.
 %
@@
 
-Given that the matrix-free operators are applied on-the-fly, it's crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
-Below we propose three algorithms to implement tangent operator of finite-strain linear elasticity (see \eqref{eq:algebraic_tangent}) using matrix-free approach.
+Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
+Below, we propose three algorithms to implement tangent operator of finite-strain linear elasticity (see \eqref{eq:algebraic_tangent}) using matrix-free approach.
 %
-In order to obtain gradients with respect to the deformed configuration we adopted \textit{MappingQEulerian} class from \texttt{deal.II} library.
-In addition to the standard mapping from the reference (unit / isoparametric) element to the element in real space, given a displacement field this class adds mapping to the deformed configuration.
+In order to obtain gradients with respect to the deformed configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II}\cite{dealII90} library.
+In addition to the standard mapping from the reference (unit / isoparametric) element to the element in real space, this class computes the mapping to the deformed configuration given a displacement field.
 %
-Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, when performing contraction with gradient of shape functions in current (deformed) configuration, the integration is also done over the deformed configuration.
-Therefore in order to arrive at the integral in undeformed configuration, additionally we have to divide by Jacobian $J$ of the mapping at a given quadrature point.
+Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II} the integration is done over the deformed configuration.
+Therefore, we have to divide by the Jacobian $J$ of the mapping at a given quadrature point additionally to express the integral in the undeformed configuration.
 
@@
 
-Algorithm \ref{alg:mf_scalar} caches a single scalar which involves logarithm of Jacobian -- a computationally demanding operation compared to additions and multiplications.
-As evident from Section \ref{sec:theory}, in order to apply the operator we need to re-evaluate Kirchhoff stress at each quadrature point.
-This in turn requires evaluation of deformation gradient $\gz F$ and therefore gradients with respect to the undeformed configuration.
+Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to additions and multiplications.
+As evident from Section \ref{sec:theory}, we need to reevaluate the Kirchhoff stress at each quadrature point in order to apply the operator.
+This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the undeformed configuration.
 
@@
 
-In order to avoid re-calculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point.
-In this case we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
-This algorithm utilizes the chosen constitutive relationship and therefore the action of the material fourth order tangent $J \boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$ is cheap to evaluate, see \eqref{eq:simplified_action}.
-To that end we cache two scalars that depend on the Jacobian $J$.
+In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point.
+In this case, we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
+This algorithm utilizes the chosen constitutive relationship.
+Therefore, the action of the material fourth order tangent $J \boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$ is cheap to evaluate, see \eqref{eq:simplified_action}.
+To that end, we cache two scalars that depend on the Jacobian $J$.
 
@@
 
-Finally Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent tensor.
-Clearly in this case we have to cache the fourth order symmetric tensor $\boldsymbol{\mathcal{C}}$ for each element and quadrature point and perform the double contraction with the second-order symmetric tensor on-the-fly.
+Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent tensor.
+In this case, we have to cache the fourth order symmetric tensor $\boldsymbol{\mathcal{C}}$ for each element and quadrature point and perform the double contraction with the second-order symmetric tensor on-the-fly.
 
@@
 
-Efficient and scalable matrix-vector products is not enough to obtain a method with linear scaling with respect to the number of unknown DoFs $N$.
-Efficiency of standard iterative solvers deteriorate for large systems of linear equations.
-It has been widely accepted that the geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can greatly improve the performance of iterative solvers by smoothing the defect on a hierarchy of levels.
-As a result the number of iterations for convergence becomes independent of the number of mesh refinements and thus leads to the $\mathcal O (N)$ solution method.
+Efficient and scalable matrix-vector products are not enough to obtain a method suitable for large computations.
+In particular, linear scaling with respect to the number of unknown DoFs $N$ is desirable.
+The efficiency of standard iterative solvers deteriorate for large systems of linear equations due to incresing iteration counts under mesh refinement.
+On the other hand, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts independent of the number of mesh refinements by smoothing the defect on a hierarchy of levels.
 For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
@@
 For this work we adopt the matrix-free level operators within the GMG and Chebyshev pre- and post-smoothing on each level.
-Given a generally heterogeneous non-linear material, definition of transfer operations (restriction and prolongation) and level operators is not trivial.
-One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1985, Hill1972,Hashin1983,Castaneda1997} to design specific transfer operators \cite{Miehe2007}\footnote{
+In general, given a heterogeneous non-linear material the definition of transfer operators (for restriction and prolongation) and level operators is not trivial.
+One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1985, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
   We note that this approach is limited to the case where for each quadrature point within an element the fourth order tangent operator is the same.
 }.
-In this approach the fine-scale displacement is decomposed into long-wave and short-wave contributions by additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
-However, for a homogeneous patch of elements, the transfer operators reduce to the standard geometric transfer.
+In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive splitting, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
+However, for a patch of elements consisting of the same material, the transfer operators reduce to the standard geometric transfer.
 Based on this observation we adopt the following approach.
 It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous fine-strain elastic material.
-In this case the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
+In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
 We define the level operators as follows.
-Recall that the fine scale tangent \eqref{eq:algebraic_tangent} is obtain by linearization around the current displacement field $\gz u^h$.
+Recall that the fine scale tangent \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$.
 We then restrict this displacement field to multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
 Essentially level operators correspond to the linearization around smoothed representation of displacement field.
-Note that by doing so the operator (tangent) $A^{l+1}$ on level $l+1$ is not directly related to the operator (tangent) $A^{l}$ on level $l$ via $A^{l+1}=I^{l+1}_{l}  A^l I^l_{l+1}$, where $I^l_{l+1}$ and $I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
-Finally, at the coarsest level we adopt CG solver which shall reduce residual by three orders of magnitude within at most hundred of steps.
-The classical multigrid V-cycle approach with two pre- and post-smoothing is adopted.
+Note that by doing so the (tangent) operator $A^{l+1}$ on level $l+1$ is not directly related to the (tangent) operator $A^{l}$ on level $l$ via $A^{l+1}=I^{l+1}_{l}  A^l I^l_{l+1}$, where $I^l_{l+1}$ and $I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
+Finally, at the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within at most one hundred steps.
+The multigrid algorithm is a V-cycle with two pre- and post-smoothing steps.
 For other approaches that combine homogenization and multigrid methods for small strain elastic materials, see \cite{Bayreuther2003,Fish1995}.
@@
 
-In this section we apply the proposed matrix-free operator algorithms for finite-strain hyper-elastic material as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}.
-Coarse mesh for 2D and 3D problems is depicted in Figure \ref{fig:miehe}.
-The 2D material consists of a matrix (depicted in blue), a hole and two spherical inclusions (depicted in red).
+In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyper-elastic material as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}.
+The coarse mesh for 2D and 3D problems is depicted in Figure \ref{fig:miehe}.
+The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red).
 The size of the square domain is $10^{-3} \rm{mm}$.
-The matrix material is taken to have Poisson's ratio of $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is take to be $100$ times harder.
+The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is take to be $100$ times harder.
 The 3D material is obtained by extrusion in the third direction.
-Note that a relatively coarse mesh is adopted at the coarsest multigrid level, however the geometry of inclusions is captured accurately by employing manifold description of the boundaries and interfaces.
-{\color{red}Daniel: do we have any paper referencing manifolds in deal.II?}
-Both domains are fixed at the bottom surface and the distributed load is applied at the top in $(1,0)$ and $(1,1,0)$ direction for 2D and 3D problems, respectively.
+Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
+{\color{red}Daniel: do we have any paper referencing manifolds in deal.II? No, I am not aware of any.}
+Both domains are fixed at the bottom surface and the distributed load is applied at the top in $(1,0)$ and $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
 The load of $0.5\,\rm{N}$ is applied in 5 steps.
-The displacement tolerance for Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for forces is $10^{-8}$.
+The displacement tolerance for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for 2the forces is $10^{-8}$.
 The relative convergence criteria for linear solver is $10^{-6}$.
@@
 We are interested in the following metrics:
-(i) walltime per DoF of matrix-vector multiplication;
-(ii) memory requirement per DoF;
-(iii) average number of CG iterations throughout the whole run (i.e. each Newton-Raphson iteration and each loading step);
-(iv) walltime per DoF of CG solver;
+\begin{itemize}
+\item walltime and memory requirement per DoF of matrix-vector multiplication;
+\item average number of CG iterations throughout the whole run (i.e. each Newton-Raphson iteration and each loading step);
+\item walltime per DoF of CG solver;
+\end{itemize}
 %
-The results of the matrix-free approach with geometric-multigrid preconditioner are compared to the matrix-based approach with algebraic multigrid (AMG) preconditioner using Trilinos \cite{Heroux2005} library, version 12.12.1.
+The results of the matrix-free approach with geometric-multigrid preconditioner are compared to the matrix-based approach with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
 The aggregation threshold for AMG is taken as $10^{-4}$.
@@
 The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
-Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly shows the influence of different caching strategies.
-Two conclusions can be drawn: scalar caching implementation is the most time consuming, at each step we additionally evaluate the gradient of displacement field in reference configuration, required to evaluate the Kirchhoff stress $\gz \tau$.
-On another hand there is little difference between the approach where the fourth order material tangent tensor is cached (``tensor4'') and the one where we only cache the second order Kirchhoff tensor and utilize the chosen hyper-elastic constitutive equation to efficiently implement the action of the material tangent on the second order symmetric tensor.
+Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies.
+Two conclusions can be drawn: The scalar caching implementation is the most time consuming, at each step we additionally evaluate the gradient of the displacement field in the reference configuration required to evaluate the Kirchhoff stress $\gz \tau$.
+On the other hand, there is little difference between the approach where the fourth order material tangent tensor is cached (``tensor4'') and the one where we only cache the second order Kirchhoff tensor and utilize the chosen hyper-elastic constitutive equation to efficiently implement the action of the material tangent on the second order symmetric tensor.
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
-Consequently we can apply matrix-free approach to any fine-strain material model by caching the fourth order material tangent in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
+Consequently, we can apply the matrix-free approach to any fine-strain material model by caching the fourth order material tangent in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
 
-As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind the matrix-free methods.
-Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} show that even with caching of one fourth order symmetric tensor and one second order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
+As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
+Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} show that even with caching one fourth order symmetric tensor and one second order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
 
 Next we evaluate the performance of the proposed geometric multigrid preconditioner. First we consider the number of CG iterations.
-Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_Emmy_cg3} show a very mild increase in the average number of CG iteration for different polynomial degrees.
-On another hand the black-box AGM preconditioner requires much more iterations for convergence.
+Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_Emmy_cg3} show a very mild increase in the average number of CG iterations for different polynomial degrees.
+On the other hand, the black-box AGM preconditioner requires much more iterations for convergence.
 Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution time for polynomial degrees up to $p=5$.
 For higher orders in 2D the walltime is comparable to the AMG preconditioner, see Figure \ref{fig:benchmark_miehe_Emmy_sol2}.
-The reason for that is the coarse level solver within the GMG preconditioner. Evidently the adopted here Chebyshev smoother with the CG solver have unsatisfactory performance for higher degrees.
+The reason for that is the coarse level solver within the GMG preconditioner. Evidently, the combination of the Chebyshev smoother with the CG solver has unsatisfactory performance for higher degrees.
 A possible solution is to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.
 This should also improve the performance of the preconditioner for lower degrees.
-An valid alternative is to employ matrix-based AMG at the coarsest level.
+A valid alternative is to use matrix-based AMG at the coarsest level.
 
@@
 
-In this contribution we proposed and numerically investigated several matrix-free implementations of tangent operators for fine-strain elasticity of heterogeneous materials.
+In this contribution, we proposed and numerically investigated several matrix-free implementations of tangent operators for fine-strain elasticity with heterogeneous materials.
 The implementation which caches the fourth order material tangent together with the second order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
-This gives hope that the matrix-free implementation of tangent operator can be applied to any constitutive model which operates with the material tangent on the quadrature point level.
+This gives hope that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material tangent on the quadrature point level.
 
 We also studied the performance of the GMG preconditioner with standard geometric transfer operations between each level.
-The GMG was applied to heterogeneous material assuming that the coarsest level can provide adequate discretization of a heterogeneity.
-The numerical studies indicate that the proposed preconditioner lead to the solution approach that is faster than the matrix-based algebraic multigrid preconditioner for moderate polynomial degrees.
-However the considerable amount of time is spent in the coarse level solver.
-Therefore our future work will be focused on extending the proposed here matrix-free multigrid solution approach to homogenization based transfer operators \cite{Miehe2007}.
-We are also interested  extending these methods to incompressible three-field formulation.
+The GMG was applied to heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
+The numerical studies indicate that the proposed preconditioner leads to a solution approach that is faster than the matrix-based algebraic multigrid preconditioner for moderate polynomial degrees.
+However, a considerable amount of time is spent in the coarse level solver.
+Therefore, our future work will be focused on extending the proposed matrix-free multigrid solution approach to homogenization based transfer operators \cite{Miehe2007}.
+We are also interested in extending these methods to incompressible three-field formulations.
 

### 2018-10-15 doc/large_strain.tex
split figures into 2, add subsections

@@
 \label{sec:mf}
-Classically in order to solve \eqref{eq:linear_system}, the matrix $\gz K$ and the force vector $\gz F$ corresponding to the discretization are assembled and
+Classically in order to solve \eqref{eq:linear_system}, the matrix $\gz A$ and the force vector $\gz F$ corresponding to the discretization are assembled and
 a direct or iterative solver is used to solve the linear system.
@@
   \SetKwInOut{Output}{Return}
-  \Input{RHS FE vector $\gz x$, current FE solution $\gz u$,
+  \Input{RHS FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
   cacheed $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
@@
   $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
-  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point}
+  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point,
+  evaluated based on $\overline{\gz u^h}$}
   \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
@@
   \Input{RHS FE vector $\gz x$,
-  cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point}
+  cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point,
+  evaluated based on $\overline{\gz u^h}$}
   \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
@@
 \label{sec:example}
-
-In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyper-elastic material as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}.
-The coarse mesh for 2D and 3D problems is depicted in Figure \ref{fig:miehe}.
-The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red).
-The size of the square domain is $10^{-3} \rm{mm}$.
-The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is take to be $100$ times harder.
-The 3D material is obtained by extrusion in the third direction.
-Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
-{\color{red}Daniel: do we have any paper referencing manifolds in deal.II? No, I am not aware of any.}
-Both domains are fixed at the bottom surface and the distributed load is applied at the top in $(1,0)$ and $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
-The load of $0.5\,\rm{N}$ is applied in 5 steps.
-The displacement tolerance for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for 2the forces is $10^{-8}$.
-The relative convergence criteria for linear solver is $10^{-6}$.
-%
-Figure \ref{fig:miehe_gmg} illustrates the multigrid mesh for the 2D problem after two global refinements.
-%
-The benchmark calculations are performed for 2D and 3D problems with various combination of polynomial degrees and number of global refinements, see Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
-%
-We are interested in the following metrics:
-\begin{itemize}
-\item walltime and memory requirement per DoF of matrix-vector multiplication;
-\item average number of CG iterations throughout the whole run (i.e. each Newton-Raphson iteration and each loading step);
-\item walltime per DoF of CG solver;
-\end{itemize}
-%
-The results of the matrix-free approach with geometric-multigrid preconditioner are compared to the matrix-based approach with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
-The aggregation threshold for AMG is taken as $10^{-4}$.
-Computations are done on a single node of two clusters.
-A node on ``Emmy'' cluster has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB Shared Cache per chip and 64 GB of RAM.
-Intel compilers version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
-The ``IWR'' results were obtained ona single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB Shared Cache per chip.
-For the simulation, GNU 8.1.0 with flags ``-O3 -march=native'' and OpenMPI 3.0.0 were used.
-
-Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the benchmarks as performed on the two clusters.
-As expected the matrix-vector multiplication becomes very expensive for higher order discretization for matrix-based approaches.
-The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
-Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies.
-Two conclusions can be drawn: The scalar caching implementation is the most time consuming, at each step we additionally evaluate the gradient of the displacement field in the reference configuration required to evaluate the Kirchhoff stress $\gz \tau$.
-On the other hand, there is little difference between the approach where the fourth order material tangent tensor is cached (``tensor4'') and the one where we only cache the second order Kirchhoff tensor and utilize the chosen hyper-elastic constitutive equation to efficiently implement the action of the material tangent on the second order symmetric tensor.
-This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
-Consequently, we can apply the matrix-free approach to any fine-strain material model by caching the fourth order material tangent in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
-
-As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
-Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} show that even with caching one fourth order symmetric tensor and one second order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
-
-Next we evaluate the performance of the proposed geometric multigrid preconditioner. First we consider the number of CG iterations.
-Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_Emmy_cg3} show a very mild increase in the average number of CG iterations for different polynomial degrees.
-On the other hand, the black-box AGM preconditioner requires much more iterations for convergence.
-Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution time for polynomial degrees up to $p=5$.
-For higher orders in 2D the walltime is comparable to the AMG preconditioner, see Figure \ref{fig:benchmark_miehe_Emmy_sol2}.
-The reason for that is the coarse level solver within the GMG preconditioner. Evidently, the combination of the Chebyshev smoother with the CG solver has unsatisfactory performance for higher degrees.
-A possible solution is to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.
-This should also improve the performance of the preconditioner for lower degrees.
-A valid alternative is to use matrix-based AMG at the coarsest level.
 
@@
 
+In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyper-elastic material as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}.
+The coarse mesh for 2D and 3D problems is depicted in Figure \ref{fig:miehe}.
+The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red).
+The size of the square domain is $10^{-3} \rm{mm}$.
+The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is take to be $100$ times harder.
+The 3D material is obtained by extrusion in the third direction.
+Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
+{\color{red}Daniel: do we have any paper referencing manifolds in deal.II? No, I am not aware of any.}
+Both domains are fixed at the bottom surface and the distributed load is applied at the top in $(1,0)$ and $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
+The load of $0.5\,\rm{N}$ is applied in 5 steps.
+The displacement tolerance for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for 2the forces is $10^{-8}$.
+The relative convergence criteria for linear solver is $10^{-6}$.
+%
+Figure \ref{fig:miehe_gmg} illustrates the multigrid mesh for the 2D problem after two global refinements.
+%
+The benchmark calculations are performed for 2D and 3D problems with various combination of polynomial degrees and number of global refinements, see Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
+%
+We are interested in the following metrics:
+\begin{itemize}
+\item walltime and memory requirement per DoF of matrix-vector multiplication;
+\item average number of CG iterations throughout the whole run (i.e. each Newton-Raphson iteration and each loading step);
+\item walltime per DoF of CG solver;
+\end{itemize}
+%
+The results of the matrix-free approach with geometric-multigrid preconditioner are compared to the matrix-based approach with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
+The aggregation threshold for AMG is taken as $10^{-4}$.
+Computations are done on a single node of two clusters.
+A node on ``Emmy'' cluster has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB Shared Cache per chip and 64 GB of RAM.
+Intel compilers version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
+The ``IWR'' results were obtained ona single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB Shared Cache per chip.
+For the simulation, GNU 8.1.0 with flags ``-O3 -march=native'' and OpenMPI 3.0.0 were used.
+
+\subsection{Matrix-vector multiplication}
+
 \begin{figure}[!ht]
   \centering
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.49\textwidth}
       \centering
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.49\textwidth}
     \centering
@@
   ~
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.49\textwidth}
       \centering
@@
   \end{subfigure}
+  \caption{Emmy cluster, RRZE, matrix-vector multiplication.}%
+  \label{fig:benchmark_miehe_Emmy}
+\end{figure}
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.49\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{IWR_timing2d.eps}
+      \caption{vmult 2D}
+      \label{fig:benchmark_miehe_IWR_vmult2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{IWR_timing3d.eps}
+    \caption{vmult 3D}
+    \label{fig:benchmark_miehe_IWR_vmult3}
+  \end{subfigure}
   ~
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.49\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{IWR_memory2d.eps}
+      \caption{memory consumption 2D}
+      \label{fig:benchmark_miehe_IWR_memory2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{IWR_memory3d.eps}
+    \caption{memory consumption 3D}
+    \label{fig:benchmark_miehe_IWR_memory3}
+  \end{subfigure}
+  \caption{IWR cluster, matrix-vector multiplication}%
+  \label{fig:benchmark_miehe_IWR}
+\end{figure}
+
+Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the benchmarks as performed on the two clusters.
+As expected the matrix-vector multiplication becomes very expensive for higher order discretization for matrix-based approaches.
+The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
+Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies.
+Two conclusions can be drawn: The scalar caching implementation is the most time consuming, at each step we additionally evaluate the gradient of the displacement field in the reference configuration required to evaluate the Kirchhoff stress $\gz \tau$.
+On the other hand, there is little difference between the approach where the fourth order material tangent tensor is cached (``tensor4'') and the one where we only cache the second order Kirchhoff tensor and utilize the chosen hyper-elastic constitutive equation to efficiently implement the action of the material tangent on the second order symmetric tensor.
+This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
+Consequently, we can apply the matrix-free approach to any fine-strain material model by caching the fourth order material tangent in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
+
+As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
+Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} show that even with caching one fourth order symmetric tensor and one second order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
+
+\subsection{Iterative solver}
+
+\begin{figure}[!ht]
+  \begin{subfigure}[b]{0.49\textwidth}
     \centering
@@
   ~
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.49\textwidth}
     \centering
@@
   \end{subfigure}
-  \caption{Emmy cluster, RRZE.}%
-  \label{fig:benchmark_miehe_Emmy}
+  \caption{Emmy cluster, RRZE, iterative solver.}%
+  \label{fig:benchmark_miehe_Emmy_cg}
 \end{figure}
@@
 \begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.45\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{IWR_timing2d.eps}
-      \caption{vmult 2D}
-      \label{fig:benchmark_miehe_IWR_vmult2}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.45\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{IWR_timing3d.eps}
-    \caption{vmult 3D}
-    \label{fig:benchmark_miehe_IWR_vmult3}
-  \end{subfigure}
-  ~
-  \begin{subfigure}[b]{0.45\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{IWR_memory2d.eps}
-      \caption{memory consumption 2D}
-      \label{fig:benchmark_miehe_IWR_memory2}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.45\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{IWR_memory3d.eps}
-    \caption{memory consumption 3D}
-    \label{fig:benchmark_miehe_IWR_memory3}
-  \end{subfigure}
-  ~
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.49\textwidth}
     \centering
@@
   \caption{IWR cluster.}%
-  \label{fig:benchmark_miehe_IWR}
+  \label{fig:benchmark_miehe_IWR_cg}
 \end{figure}
+
+Next we evaluate the performance of the proposed geometric multigrid preconditioner. First we consider the number of CG iterations.
+Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_Emmy_cg3} show a very mild increase in the average number of CG iterations for different polynomial degrees.
+On the other hand, the black-box AGM preconditioner requires much more iterations for convergence.
+Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution time for polynomial degrees up to $p=5$.
+For higher orders in 2D the walltime is comparable to the AMG preconditioner, see Figure \ref{fig:benchmark_miehe_Emmy_sol2}.
+The reason for that is the coarse level solver within the GMG preconditioner. Evidently, the combination of the Chebyshev smoother with the CG solver has unsatisfactory performance for higher degrees.
+A possible solution is to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.
+This should also improve the performance of the preconditioner for lower degrees.
+A valid alternative is to use matrix-based AMG at the coarsest level.
 

### 2018-10-16 doc/large_strain.tex

@@
     The performance of finite element solvers on modern computer architectures is typically memory bound.
-    The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
-    In order to improve the performance of the iterative solver, so-called matrix-free methods are
+    The main cause for this is that loading matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
+    In order to improve the performance of iterative solvers within the high-performance computing context, so-called matrix-free methods are
     widely adopted in the fluid mechanics community, where the matrix-vector products are computed on-the-fly.
 
-    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyper-elastic tangent operator numerically.
-    In order to improve the convergence behavior of iterative solvers, we also propose a way to construct level tangent operators
+    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyperelastic tangent operator numerically.
+    In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
     and employ them to define a geometric multigrid preconditioner.
-    Our implementation employs MPI {\color{red}and Intel Threading Building Blocks parallelization} and SIMD vectorization.
-    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions.
+    Our implementation employs MPI and Intel Threading Building Blocks parallelization, and SIMD vectorization.
+    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
   \end{abstract}
@@
 
-The performance of finite element (FEM) solvers on modern computer architectures is typically memory bound.
-The main cause for this is that loading matrix elements is significantly slower than performing the arithmetic operations when solving the problem.
-In order to improve the performance of the iterative solver, so-called matrix-free methods are
-widely adopted in fluid mechanics community, where the matrix-vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016}.
+The performance of finite element solvers on modern computer architectures is typically memory bound.
+The main cause for this is that loading pre-computed matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
+In order to improve the performance of iterative solvers, so-called matrix-free methods,
+where the matrix-vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016},
+are widely adopted in the fluid mechanics community.
 Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016, ljungkvist2017multigrid}.
@@
 To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
-We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a master thesis of a student of two of the authors \cite{Mentler2017}.
-A possible reason is that the tangent operators (coming from linearization of the nonlinear balance equations) are
-more elaborate as compared to the operators used in fluid mechanics and it is not obvious whether the matrix-free methods can be advantageous in this case.
+We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a masters thesis produced by a student of two of the authors \cite{Mentler2017}.
+One possible reason is that the tangent operators (stemming from linearization of the nonlinear balance equations) are
+more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether the matrix-free methods can be advantageous in this case.
 Another reason could be the use of lower order elements due to lacking smoothness of the underlying solution.
-FEM with linear or quadratic elements, that avoid locking, are often adopted in the solid mechanics community.
-Higher order elements are rarely employed.
-However, we note that large scale computations with $10^{12}$ unknowns even with linear FEM inevitably require matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
+Finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community.
+Even higher order elements are rarely employed.
+However, we note, even with linear FEM, that large scale computations with $10^{12}$ unknowns inevitably requires a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
 Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.
 
-In this work, we investigate different implementations of the finite-strain hyper-elastic tangent operator with respcet to their computational cost.
-In order to improve the convergence of iterative solvers, we also propose a way to construct level matrix-free tangent operators
+In this work, we investigate different implementations of the finite-strain hyperelastic tangent operator with respcet to their computational cost.
+In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
 and employ them to define a geometric multigrid preconditioner.
-Our implementation employs MPI and Intel Threading Building Blocks parallelization and SIMD vectorization.
+Our implementation employs MPI and Intel Threading Building Blocks parallelization, and SIMD vectorization.
 
-The paper is organized as follows.
-In Section \ref{sec:theory}, we briefly introduce the partial differential equations used in  finite-strain solid mechanics.
+The paper is organized as follows: In Section \ref{sec:theory} we briefly introduce the partial differential equations used in  finite-strain solid mechanics.
 Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the different matrix-free operator evaluation approaches.
-In Section \ref{sec:gmg}, we propose a geometric multigrid preconditioner suitable for finite-strain linear elasticity.
-The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of heterogeneous hyper-elastic material in two and three dimensions in Section \ref{sec:example}.
+In Section \ref{sec:gmg} we propose a geometric multigrid preconditioner suitable for finite-strain linear elasticity.
+The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}.
 The results are summarized in Section \ref{sec:summary}.
@@
 is defined via the mapping $\gz \varphi_t: \mcl B \rightarrow \mathbb E^3$, which places material points of $\mcl B$ in the Euclidean space $\mathbb E^3$.
-The location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
+The current location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
 The displacement field reads $\gz u = \gz x - \gz X$.
@@
 $\d \gz x = \gz F \cdot \d \gz X$,
-where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient.
-The mapping has to be one-to-one and must exclude self-penetration.
-Consequently, the Jacobian has to be positive $J = \det \gz F > 0$.
-The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
+where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient and $\gz I$ is the second order unit tensor.
+The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian $J = \det \gz F > 0$ has to be positive.
+% The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
 We shall denote the Piola transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
@@
 %
-In what follows, we characterize the material behavior using the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$, where $\gz C$ is the right Cauchy-Green tensor $\gz C := \gz F^T \cdot \gz F = \gz U^2$.
+In what follows, we parameterize the material behavior using the right Cacuhy-Green tensor $\gz C := \gz F^T \cdot \gz F$.
+We will also use the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$ and the left Cauchy-Green tensor $\gz b := \gz F \cdot \gz F^T$.
 Clearly $J = \sqrt{\det \gz C}$ and $\partial (\bullet)/\partial \gz E = 2\, \partial (\bullet)/\partial \gz C$.
-We will also use the left Cauchy-Green tensor $\gz b := \gz F \cdot \gz F^T = \gz V^2$.
-
-For conservative systems without volumetric load, the total potential energy functional $\mcl E$ is introduced as
+%
+For conservative systems without the body force, the total potential energy functional $\mcl E$ is introduced as
 \begin{align}
@@
 The first term in \eqref{eq:tangent} is related to the linearization of the Kirchhoff stress and therefore called material part of the tangent operator.
-The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the the linearization of the deformation gradient $\gz F$.
+The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the linearization of the inverse of the deformation gradient $\gz F^{-1}$.
 
@@
 
-We now introduce a FE triangulation $\mathcal{T}^h$ of $\mcl B_0$ and
+We now introduce a FE triangulation $\mathcal{B}^h_0$ of $\mcl B_0$ and
 the associated FE space of continuous piecewise elements with polynomial space of fixed degree. % : $V^h \subset H^1 (\mcl B_0)$.
-The displacement fields are given in a
+The displacement fields are given in
 a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support on a patch of elements):
@@
 \end{alignat}
-where the superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the sets of unknown degrees of freedom (DoF).
+where the superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the set of unknown degrees of freedom (DoF).
 %{\color{red} For the sake of this study, $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.}
 
-The Newton-Raphson solution approach is adopted for the here considered nonlinear problem.
-Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is to obtained as solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent})
+The Newton-Raphson solution approach is adopted for the nonlinear problem considered here.
+Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is obtained as the solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent}):
 \begin{align}
@@
 and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the bilinear form for the trial solution field $\overline{\gz u^h}$.
+Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the deformed configuration, whereas the integration is done in the reference (undeformed) configuration.
 
@@
 \label{sec:mf}
-Classically in order to solve \eqref{eq:linear_system}, the matrix $\gz A$ and the force vector $\gz F$ corresponding to the discretization are assembled and
+Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the force vector $\gz F$ corresponding to the discretization are assembled and
 a direct or iterative solver is used to solve the linear system.
-On modern computer architectures however, it turns out that loading data is significantly slower than performing the arithmetic operations when solving.
-For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations in the solver on-the-fly
+On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
+For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations within the solver on-the-fly
 rather than loading matrix elements from memory.
-For iterative solvers, this is possible since it is sufficient to compute matrix-vector products.
+For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
 %
@@
 \end{align*}
-where in the last step $\tilde{a}_K(\varphi_i,\varphi_j)(\gz x_q)$ is a representation for the evaluation of the bilinear in one quadrature point and a quadrature rule is used.
-For the sake of demonstration, let us assume $A$ represents a mass matrix. In this case, it holds
+where in the last step $\tilde{a}_K(\varphi_i,\varphi_j)(\gz x_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $q$.
+For the sake of demonstration, let us assume that $A$ represents a mass matrix. In this case, it holds that
 \begin{align*}
@@
 \end{align*}
-Further assuming that the number of the one-dimensional quadrature points $n$ matches the degree of the ansatz space plus one, the evaluation of all the shape functions ($n^d$)
-in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in case we consider a problem in $d$ space dimensions. For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$
-and this is the same as for a regular matrix-vector where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, we can actually do better
-than that. Indeed, the evaluation of all the shape functions in all the ansatz points can be rewritten as
+Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
+in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in that case that we consider a problem in $d$ space dimensions.
+For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$,
+and this is the same as for a regular matrix-vector product where the entries are already precomputed.
+Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently.
+Indeed, the evaluation of all the shape functions in all the ansatz points can be rewritten as
 \begin{lstlisting}
@@
 in three space dimensions. In general, this reordering reduces the arithmetic complexity to $\mathcal{O}(dn^{d+1})$ and is called sum factorization.
-Hence, we can expect that such a matrix-free approach is faster than a regular matrix-base especially for 3D and high polynomial degree.
-More information on techniques to increase the computational density can be found in \cite{kronbichler12,vos10}.
-We briefly note that the SIMD vectorization is applied at the element level, that is the operator is applied simultaneously on several elements (called ``block'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
+Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for 3D and high polynomial degree with the consistent quadrature rule.
+More information on techniques developed to increase the computational density can be found in \cite{kronbichler12,vos10}.
+We briefly note that the SIMD vectorization is applied at the finite element level, that is the operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
 
 Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
-Below, we propose three algorithms to implement tangent operator of finite-strain linear elasticity (see \eqref{eq:algebraic_tangent}) using matrix-free approach.
+Below, we propose three algorithms to implement tangent operator of finite-strain linear elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
 %
-In order to obtain gradients with respect to the deformed configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II}\cite{dealII90} library.
-In addition to the standard mapping from the reference (unit / isoparametric) element to the element in real space, this class computes the mapping to the deformed configuration given a displacement field.
+In order to obtain gradients with respect to the deformed configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
+In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the deformed configuration given a displacement field.
 %
-Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II} the integration is done over the deformed configuration.
+Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is performed over the deformed configuration.
 Therefore, we have to divide by the Jacobian $J$ of the mapping at a given quadrature point additionally to express the integral in the undeformed configuration.
@@
   \Input{RHS FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
-  cacheed $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
+  cached $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
   \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
@@
           evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
-          ``submit'' contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
+          queue $\boldsymbol{\mathcal{G}}$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
           evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
           evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
-          ``submit'' contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
+          queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
         }
+      evaluate queued contractions using sum factorization and distribute results to the destination vector
   }
@@
           evaluate $\boldsymbol{\mathcal{G}}\gz g = \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
-          ``submit'' contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
+          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
           evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
           evaluate $\boldsymbol{\mathcal{C}} \gz g_s = c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
-          ``submit'' contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
+          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
         }
+        evaluate queued contractions using sum factorization and distribute results to the destination vector
   }
@@
 In this case, we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
-This algorithm utilizes the chosen constitutive relationship.
-Therefore, the action of the material fourth order tangent $J \boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$ is cheap to evaluate, see \eqref{eq:simplified_action}.
+This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed in the operator itself.
+It is, however, somewhat more generic than Algorithm \ref{alg:mf_tensor1} since the dependence of $\gz \tau$ exists outside of the oeperator itself (although it must remain consistent with the action of $J \boldsymbol {\mathcal{C}}$).
+Therefore, the action of the material fourth order tangent $J \boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$, as defined in \eqref{eq:simplified_action}, is cheap to evaluate.
 To that end, we cache two scalars that depend on the Jacobian $J$.
@@
           evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
-          ``submit'' contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
+          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
           evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
           evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
-          ``submit'' contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
+          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
         }
+        evaluate queued contractions using sum factorization and distribute results to the destination vector
   }
@@
 Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent tensor.
-In this case, we have to cache the fourth order symmetric tensor $\boldsymbol{\mathcal{C}}$ for each element and quadrature point and perform the double contraction with the second-order symmetric tensor on-the-fly.
+In this case, we have to cache the fourth order symmetric tensor $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\gz g_s$ on-the-fly.
 
@@
 In particular, linear scaling with respect to the number of unknown DoFs $N$ is desirable.
-The efficiency of standard iterative solvers deteriorate for large systems of linear equations due to incresing iteration counts under mesh refinement.
-On the other hand, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts independent of the number of mesh refinements by smoothing the defect on a hierarchy of levels.
+The efficiency of standard iterative solvers and preconditioners deteriorate for large systems of linear equations due to increasing iteration counts under mesh refinement.
+In contrast to other preconditioners, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes.
 For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
 
-For this work we adopt the matrix-free level operators within the GMG and Chebyshev pre- and post-smoothing on each level.
+For this work we adopt the matrix-free level operators within the GMG and employ Chebyshev pre- and post-smoothing on each level.
 In general, given a heterogeneous non-linear material the definition of transfer operators (for restriction and prolongation) and level operators is not trivial.
 One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1985, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
-  We note that this approach is limited to the case where for each quadrature point within an element the fourth order tangent operator is the same.
+  We note that this approach is limited to the case where for each quadrature point within an element the fourth-order tangent operator is the same.
 }.
-In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive splitting, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
+In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
 However, for a patch of elements consisting of the same material, the transfer operators reduce to the standard geometric transfer.
-Based on this observation we adopt the following approach.
-It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous fine-strain elastic material.
+Based on this observation we adopt the following approach: It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous fine-strain elastic material.
 In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
-We define the level operators as follows.
-Recall that the fine scale tangent \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$.
-We then restrict this displacement field to multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
-Essentially level operators correspond to the linearization around smoothed representation of displacement field.
-Note that by doing so the (tangent) operator $A^{l+1}$ on level $l+1$ is not directly related to the (tangent) operator $A^{l}$ on level $l$ via $A^{l+1}=I^{l+1}_{l}  A^l I^l_{l+1}$, where $I^l_{l+1}$ and $I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
-Finally, at the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within at most one hundred steps.
+We define the level operators as follows: Recall that the fine scale tangent \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$.
+We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
+Essentially, level operators correspond to the linearization around smoothed representation of displacement field\footnote{
+Note that this differs from the classical approach where the (tangent) operator $A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $A^{l}$ on level $l$ via $A^{l+1}=I^{l+1}_{l}  A^l I^l_{l+1}$, where $I^l_{l+1}$ and $I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
+}
+Finally, at the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
 The multigrid algorithm is a V-cycle with two pre- and post-smoothing steps.
@@
 \label{sec:example}
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{coarse_mesh.png}
+      \caption{2D coarse mesh}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{coarse_mesh_3d.png}
+    \caption{3D coarse mesh}
+  \end{subfigure}
+  \caption{Discretization of the heterogeneous material at the coarsest mesh level.}%
+  \label{fig:miehe}
+\end{figure}
+
+In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic material as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}.
+The coarse mesh for the 2D and 3D problems is illustrated in Figure \ref{fig:miehe}.
+The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red).
+The size of the square domain is $10^{-3} \rm{mm}$.
+The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
+The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
+Note that, although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
+Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
+The distributed load of $0.5\,\rm{N}$ is applied in 5 steps.
+With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the forces is $10^{-8}$.
+The relative convergence criteria for linear solver is $10^{-6}$.
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.8\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{gmg_2d.png}
+    % \caption{2D multigrid mesh with 2 global refinements}
+  \end{subfigure}
+  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. Color indicate the process owning the element.}%
+  \label{fig:miehe_gmg}
+\end{figure}
+%
+Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.
 
@@
   \end{tabular}
-  \caption{Parameters for 2D benchmark.
+  \caption{Parameters for the 2D benchmark.
   }
@@
   \end{tabular}
-  \caption{Parameters for 3D benchmark.
+  \caption{Parameters for the 3D benchmark.
   }
@@
 
-\begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.45\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{coarse_mesh.png}
-      \caption{2D coarse mesh}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.45\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{coarse_mesh_3d.png}
-    \caption{3D coarse mesh}
-  \end{subfigure}
-  \caption{Heterogeneous material.}%
-  \label{fig:miehe}
-\end{figure}
+The benchmark calculations are performed for 2D and 3D problems with various combination of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
+Shared memory parallelization using Intel Threading Building Blocks was disabled for this study.
+%
 
-\begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.8\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{gmg_2d.png}
-    % \caption{2D multigrid mesh with 2 global refinements}
-  \end{subfigure}
-  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. Color indicate the process owning the element.}%
-  \label{fig:miehe_gmg}
-\end{figure}
-
-In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyper-elastic material as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}.
-The coarse mesh for 2D and 3D problems is depicted in Figure \ref{fig:miehe}.
-The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red).
-The size of the square domain is $10^{-3} \rm{mm}$.
-The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is take to be $100$ times harder.
-The 3D material is obtained by extrusion in the third direction.
-Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
-{\color{red}Daniel: do we have any paper referencing manifolds in deal.II? No, I am not aware of any.}
-Both domains are fixed at the bottom surface and the distributed load is applied at the top in $(1,0)$ and $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
-The load of $0.5\,\rm{N}$ is applied in 5 steps.
-The displacement tolerance for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for 2the forces is $10^{-8}$.
-The relative convergence criteria for linear solver is $10^{-6}$.
-%
-Figure \ref{fig:miehe_gmg} illustrates the multigrid mesh for the 2D problem after two global refinements.
-%
-The benchmark calculations are performed for 2D and 3D problems with various combination of polynomial degrees and number of global refinements, see Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
-%
 We are interested in the following metrics:
 \begin{itemize}
-\item walltime and memory requirement per DoF of matrix-vector multiplication;
-\item average number of CG iterations throughout the whole run (i.e. each Newton-Raphson iteration and each loading step);
-\item walltime per DoF of CG solver;
+\item the wall-clock time and memory requirement per DoF for matrix-vector multiplication;
+\item the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step);
+\item the wall-clock time per DoF for the CG solver;
 \end{itemize}
 %
-The results of the matrix-free approach with geometric-multigrid preconditioner are compared to the matrix-based approach with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
-The aggregation threshold for AMG is taken as $10^{-4}$.
-Computations are done on a single node of two clusters.
-A node on ``Emmy'' cluster has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB Shared Cache per chip and 64 GB of RAM.
-Intel compilers version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
-The ``IWR'' results were obtained ona single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB Shared Cache per chip.
-For the simulation, GNU 8.1.0 with flags ``-O3 -march=native'' and OpenMPI 3.0.0 were used.
+The results of the matrix-free approach with a geometric-multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
+The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
+Computations are done on a single node of two clusters:
+A node on ``Emmy'' cluster has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM.
+Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
+The ``IWR'' results were obtained on a single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip.
+For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
 
@@
       \includegraphics[width=\textwidth]{Emmy_RRZE_timing2d.eps}
-      \caption{vmult 2D}
+      \caption{vmult (2D)}
       \label{fig:benchmark_miehe_Emmy_vmult2}
@@
     \includegraphics[width=\textwidth]{Emmy_RRZE_timing3d.eps}
-    \caption{vmult 3D}
+    \caption{vmult (3D)}
     \label{fig:benchmark_miehe_Emmy_vmult3}
@@
       \includegraphics[width=\textwidth]{Emmy_RRZE_memory2d.eps}
-      \caption{memory consumption 2D}
+      \caption{memory consumption (2D)}
       \label{fig:benchmark_miehe_Emmy_memory2}
@@
     \includegraphics[width=\textwidth]{Emmy_RRZE_memory3d.eps}
-    \caption{memory consumption 3D}
+    \caption{memory consumption (3D)}
     \label{fig:benchmark_miehe_Emmy_memory3}
@@
       \includegraphics[width=\textwidth]{IWR_timing2d.eps}
-      \caption{vmult 2D}
+      \caption{vmult (2D)}
       \label{fig:benchmark_miehe_IWR_vmult2}
@@
     \includegraphics[width=\textwidth]{IWR_timing3d.eps}
-    \caption{vmult 3D}
+    \caption{vmult (3D)}
     \label{fig:benchmark_miehe_IWR_vmult3}
@@
       \includegraphics[width=\textwidth]{IWR_memory2d.eps}
-      \caption{memory consumption 2D}
+      \caption{memory consumption (2D)}
       \label{fig:benchmark_miehe_IWR_memory2}
@@
     \includegraphics[width=\textwidth]{IWR_memory3d.eps}
-    \caption{memory consumption 3D}
+    \caption{memory consumption (3D)}
     \label{fig:benchmark_miehe_IWR_memory3}
@@
 
-Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the benchmarks as performed on the two clusters.
-As expected the matrix-vector multiplication becomes very expensive for higher order discretization for matrix-based approaches.
+Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication benchmarks as performed on the two clusters.
+As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches.
 The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
 Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies.
-Two conclusions can be drawn: The scalar caching implementation is the most time consuming, at each step we additionally evaluate the gradient of the displacement field in the reference configuration required to evaluate the Kirchhoff stress $\gz \tau$.
-On the other hand, there is little difference between the approach where the fourth order material tangent tensor is cached (``tensor4'') and the one where we only cache the second order Kirchhoff tensor and utilize the chosen hyper-elastic constitutive equation to efficiently implement the action of the material tangent on the second order symmetric tensor.
+Two conclusions can be drawn from these results:
+The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the reference configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
+There is little difference between the approach where the fourth order material tangent tensor is cached (``tensor4'') and the one where we only cache the second order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material tangent on the second order symmetric tensor.
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
-Consequently, we can apply the matrix-free approach to any fine-strain material model by caching the fourth order material tangent in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
+Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the fourth order material tangent in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
+It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws.
 
 As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
-Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} show that even with caching one fourth order symmetric tensor and one second order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
+Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth order symmetric tensor and one second order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
 
-\subsection{Iterative solver}
+\subsection{Preconditioned iterative solver}
 
@@
     \includegraphics[width=\textwidth]{Emmy_RRZE_cg2d.eps}
-    \caption{CG iterations 2D}
+    \caption{CG iterations (2D)}
     \label{fig:benchmark_miehe_Emmy_cg2}
@@
     \includegraphics[width=\textwidth]{Emmy_RRZE_cg3d.eps}
-    \caption{CG iterations 3D}
+    \caption{CG iterations (3D)}
     \label{fig:benchmark_miehe_Emmy_cg3}
@@
     \includegraphics[width=\textwidth]{Emmy_RRZE_solver2d.eps}
-    \caption{CG solution time 2D}
+    \caption{CG solution time (2D)}
     \label{fig:benchmark_miehe_Emmy_sol2}
@@
     \includegraphics[width=\textwidth]{Emmy_RRZE_solver3d.eps}
-    \caption{CG solution time 3D}
+    \caption{CG solution time (3D)}
     \label{fig:benchmark_miehe_Emmy_sol3}
@@
     \includegraphics[width=\textwidth]{IWR_cg2d.eps}
-    \caption{CG iterations 2D}
+    \caption{CG iterations (2D)}
     \label{fig:benchmark_miehe_IWR_cg2}
@@
     \includegraphics[width=\textwidth]{IWR_cg3d.eps}
-    \caption{CG iterations 3D}
+    \caption{CG iterations (3D)}
     \label{fig:benchmark_miehe_IWR_cg3}
@@
     \includegraphics[width=\textwidth]{IWR_solver2d.eps}
-    \caption{CG solution time 2D}
+    \caption{CG solution time (2D)}
     \label{fig:benchmark_miehe_IWR_sol2}
@@
     \includegraphics[width=\textwidth]{IWR_solver3d.eps}
-    \caption{CG solution time 3D}
+    \caption{CG solution time (3D)}
     \label{fig:benchmark_miehe_IWR_sol3}
@@
 
-Next we evaluate the performance of the proposed geometric multigrid preconditioner. First we consider the number of CG iterations.
+Next we evaluate the performance of the proposed geometric multigrid preconditioner.
+First we consider the number of iterations in the conjugate gradient (CG) solver.
 Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_Emmy_cg3} show a very mild increase in the average number of CG iterations for different polynomial degrees.
-On the other hand, the black-box AGM preconditioner requires much more iterations for convergence.
-Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution time for polynomial degrees up to $p=5$.
-For higher orders in 2D the walltime is comparable to the AMG preconditioner, see Figure \ref{fig:benchmark_miehe_Emmy_sol2}.
-The reason for that is the coarse level solver within the GMG preconditioner. Evidently, the combination of the Chebyshev smoother with the CG solver has unsatisfactory performance for higher degrees.
-A possible solution is to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.
-This should also improve the performance of the preconditioner for lower degrees.
-A valid alternative is to use matrix-based AMG at the coarsest level.
+On the other hand, the black-box AMG preconditioner requires many more iterations for convergence.
+Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for polynomial degrees up to $p=5$.
+For higher orders in 2D the wall-clock time is comparable to the AMG preconditioner, see Figure \ref{fig:benchmark_miehe_Emmy_sol2}.
+The reason for this is the suboptimal coarse level solver within the GMG preconditioner.
+Evidently, the combination of the Chebyshev smoother with the CG solver at the coarsest mesh level has unsatisfactory performance for higher polynomial degrees.
+One possible solution to this issues would be to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.
+Although untested in this work, we hypothesize that this should also improve the performance of the preconditioner for lower degrees.
+A valid alternative would be to use matrix-based AMG at the coarsest level.
 
@@
 
-In this contribution, we proposed and numerically investigated several matrix-free implementations of tangent operators for fine-strain elasticity with heterogeneous materials.
-The implementation which caches the fourth order material tangent together with the second order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
+In this contribution, we proposed and numerically investigated several matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
+The implementation that caches the fourth order material tangent together with the second order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
 This gives hope that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material tangent on the quadrature point level.
@@
 The GMG was applied to heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
-The numerical studies indicate that the proposed preconditioner leads to a solution approach that is faster than the matrix-based algebraic multigrid preconditioner for moderate polynomial degrees.
-However, a considerable amount of time is spent in the coarse level solver.
-Therefore, our future work will be focused on extending the proposed matrix-free multigrid solution approach to homogenization based transfer operators \cite{Miehe2007}.
-We are also interested in extending these methods to incompressible three-field formulations.
+The numerical studies indicate that the proposed preconditioner leads to a solution approach that is faster than the matrix-based AMG for moderately high polynomial degrees (approximately up to order five).
+However, a considerable amount of time of GMG preconditioner is spent in the coarse level solver.
+Our future work will, therefore, be focused on extending the proposed matrix-free multigrid solution approach to homogenization based transfer operators \cite{Miehe2007}.
+We are also interested in extending these methods to near incompressible three-field formulations.
 
@@
 
-{\color{red}
-J-P.~Pelteret was supported by the European Research Council (ERC) through
-the Advanced Grant 289049 MOCOPOLY.(?)
-}
+% J-P.~Pelteret was supported by the European Research Council (ERC) through
+% the Advanced Grant 289049 MOCOPOLY
 

### 2018-10-18 doc/large_strain.tex
start with editing according to Prof. Steinmann comments

@@
 
-  \author[a]{Paul Steinmann}
+  \author[a,c]{Paul Steinmann}
   \ead{paul.steinmann@fau.de}
@@
 
+  \address[c]{Glasgow computational engineering center,
+      University of Glasgow, G12 8QQ Glasgow, United Kingdom
+  }
+
   \begin{abstract}
@@
     In order to improve the performance of iterative solvers within the high-performance computing context, so-called matrix-free methods are
-    widely adopted in the fluid mechanics community, where the matrix-vector products are computed on-the-fly.
+    widely adopted in the fluid mechanics community, where matrix-vector products are computed on-the-fly.
 
@@
 In order to improve the performance of iterative solvers, so-called matrix-free methods,
-where the matrix-vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016},
+where matrix-vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016},
 are widely adopted in the fluid mechanics community.
@@
 To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
-We are only aware of a single paper in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a masters thesis produced by a student of two of the authors \cite{Mentler2017}.
+We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent masters thesis produced by a student of the first two of the authors \cite{Mentler2017}.
 One possible reason is that the tangent operators (stemming from linearization of the nonlinear balance equations) are
@@
 Another reason could be the use of lower order elements due to lacking smoothness of the underlying solution.
-Finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community.
+The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community.
 Even higher order elements are rarely employed.
-However, we note, even with linear FEM, that large scale computations with $10^{12}$ unknowns inevitably requires a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
+However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
 Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.
 
-In this work, we investigate different implementations of the finite-strain hyperelastic tangent operator with respcet to their computational cost.
+In this work, we investigate different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost.
 In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
-and employ them to define a geometric multigrid preconditioner.
-Our implementation employs MPI and Intel Threading Building Blocks parallelization, and SIMD vectorization.
+and employ them to define a \mbox{geometric} multigrid preconditioner.
+Our implementation employs MPI and Intel Threading Building Blocks parallelization, together with SIMD vectorization.
 
@@
 Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the different matrix-free operator evaluation approaches.
-In Section \ref{sec:gmg} we propose a geometric multigrid preconditioner suitable for finite-strain linear elasticity.
+In Section \ref{sec:gmg} we propose a geometric multigrid preconditioner suitable for finite-strain hyperelasticity.
 The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}.
@@
 The deformation of a body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$ at time $t$
-is defined via the mapping $\gz \varphi_t: \mcl B \rightarrow \mathbb E^3$, which places material points of $\mcl B$ in the Euclidean space $\mathbb E^3$.
+is defined via the deformation map $\gz \varphi_t: \mcl B_0 \rightarrow \mcl B_t$, which places material points of $\mcl B$ into the Euclidean space $\mathbb E^3$.
 The current location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
@@
 %
-We assume that the deformation map is linear such that
+The tangent map is linear such that
 $\d \gz x = \gz F \cdot \d \gz X$,
@@
 % The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
-We shall denote the Piola transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
+We shall denote the push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
 \begin{align}
-  \chi\left( \mathbf{A} \right)_{ij}
+  \varphi\left( \mathbf{A} \right)_{ij}
   &= F_{iA} A_{AB} F_{jB} \\
-\chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
+\varphi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
   &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}.
@@
 %
-In what follows, we parameterize the material behavior using the right Cacuhy-Green tensor $\gz C := \gz F^T \cdot \gz F$.
+In what follows, we parameterize the material behavior using the right Cauchy-Green tensor $\gz C := \gz F^T \cdot \gz F$.
 We will also use the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$ and the left Cauchy-Green tensor $\gz b := \gz F \cdot \gz F^T$.
@@
 %
-For conservative systems without the body force, the total potential energy functional $\mcl E$ is introduced as
+For conservative systems without body forces, the total potential energy functional $\mcl E$ is introduced as
 \begin{align}
@@
 
-The principal of stationary potential energy requires for all  equilibrium states $\gz u$ that the directional derivative with respect to the displacement
+The principle of stationary potential energy at equilibrium requires that the directional derivative with respect to the displacement
 \begin{align}
@@
 \end{align}
-where $\gz P := \partial \mcl \psi / \partial \gz F$ is the first Piola-Kirchhoff stress tensor.
+where $\gz P := \partial \mcl \psi / \partial \gz F$ is the Piola stress tensor.
 The double contraction in the first term can be re-written in terms of the symmetric Kirchhoff stress tensor $\gz \tau := \gz P \cdot \gz F^T$ as
@@
 \gz P : \Grad \, \delta \gz u =
-(\gz \tau \cdot \gz F^{-T}) : \Grad \, \delta \gz u =
-\gz \tau : (\Grad \, \delta \gz u \cdot \gz F^{-1}) =
+\left[\gz \tau \cdot \gz F^{-T}\right] : \Grad \, \delta \gz u =
+\gz \tau : \left[\Grad \, \delta \gz u \cdot \gz F^{-1}\right] =
 \gz \tau : \grad \, \delta \gz u \, ,
@@
 where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$.
-Note that $\gz \tau$ is the Piola transformation of the second Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
-that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.
+Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
+that is $\gz \tau = \varphi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.
 
@@
 \end{align}
-Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
+Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \varphi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
 The first term in \eqref{eq:tangent} is related to the linearization of the Kirchhoff stress and therefore called material part of the tangent operator.
@@
   \equiv J \boldsymbol{\sigma}
-  = \chi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
+  = \varphi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
   = \mu \mathbf{b} - \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{I}
@@
 J \boldsymbol{\mathcal{C}}
-  = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
+  = \varphi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
   = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
@@
   A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
-  A_{ij} &=
+  A_{ij} \equiv a(\gz N_i, \gz N_j) &=
   \int_{\mcl B_0}
@@
   \right]
-  }_{\tilde{a}(\gz N_i, \gz N_j)}
+  }_{\displaystyle \tilde{a}(\gz N_i, \gz N_j)}
   \d V
@@
 \end{align}
-Here $\gz A$ is the discrete tangent operator, $\gz F$ is the discrete gradient
-and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the bilinear form for the trial solution field $\overline{\gz u^h}$.
+Here $\gz A$ is the discrete tangent matrix, $\gz F$ is the discrete gradient
+and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form for the trial solution field $\overline{\gz u^h}$.
 Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the deformed configuration, whereas the integration is done in the reference (undeformed) configuration.
@@
 %
-The matrix-vector product $A x$ can be expressed by
+The matrix-vector product $\gz A \gz x$ can be expressed by
 \begin{align*}
- (Ax)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
-        &= \sum_K\sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz x_q) x_j w_q
+ (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
+        &\approx \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz x_q) x_j W_q \\
+        &= \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi^K_q) x_j w_q
+  \label{eq:mf_vmult}
 \end{align*}
-where in the last step $\tilde{a}_K(\varphi_i,\varphi_j)(\gz x_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $q$.
-For the sake of demonstration, let us assume that $A$ represents a mass matrix. In this case, it holds that
+where in the last step $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi^K_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $q$.
+For the sake of demonstration, let us assume that $\gz A$ represents a mass matrix with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
 \begin{align*}
- \tilde{a}_K(N_i,N_j)(\gz x_q) = N_i(\gz x_q)N_j(\gz x_q).
+ \tilde{a}(N_i,N_j)(\gz \xi^K_q) = N_i(\gz \xi^K_q)N_j(\gz \xi^K_q).
 \end{align*}
@@
 Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently.
-Indeed, the evaluation of all the shape functions in all the ansatz points can be rewritten as
-\begin{lstlisting}
-template<int k, int q>
-void evaluate_dof_vector_in_all_quadrature_points
-  (const double u[k][k][k],
-   const double phi[k][q],
-   double return_value[q][q][q])
-{
-  double tmp1[k][k][q] {};
-  for (int i1=0; i1<k; ++i1)
-    for (int i2=0; i2<k; ++i2)
-      for (int q3=0; q3<q; ++q3)
-        for (int i3=0; i3<k; ++i3)
-          tmp1[i1][i2][q3] += u[i1][i2][i3]*phi[i3][q3];
-
-  double temp2[k][q][q] {};
-  for (int i1=0; i1<k; ++i1)
-    for (int q3=0; q3<q; ++q3)
-      for (int q2=0; q2<q; ++q2)
-        for (int i2=0; i2<k; ++i2)
-          tmp2[i1][q2][q3] += tmp1[i1][i2][q3]*phi[i2][q2];
-
-  for (int q2=0; q2<q; ++q2)
-    for (int q3=0; q3<q; ++q3)
-      for (int q1=0; q1<q; ++q1)
-        for (int i1=0; i1<k; ++i1)
-  return_value[q1][q2][q3] += tmp2[i1][q2][q3]*phi[i1][q1];
-}
-\end{lstlisting}
-in three space dimensions. In general, this reordering reduces the arithmetic complexity to $\mathcal{O}(dn^{d+1})$ and is called sum factorization.
+Let us illustrate this in 2D. Tensor-product quadrature points can be expressed as a combination of one dimensional quadrature points
+\begin{align}
+  \gz \xi_q = (\widetilde{\xi}_{q_1}, \widetilde{\xi}_{q_2}),
+\end{align}
+where for quadrature points $q$ we can associate a multi-index $(q_1,q_2)$.
+Weights associated with the quadrature formula $w_q$ can be expressed as a product of one-dimensional weights
+\begin{align}
+  w_q = \widetilde{w}_{q_1} \widetilde{w}_{q_2}
+\end{align}
+Shape functions can also be expressed as product of $d$ one dimensional basis functions
+\begin{align}
+  N_i(\gz \xi_q) =
+  \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
+  \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2})
+\end{align}
+where for each DoF index $i$ we can associate a multi-index $(i_1, i_2)$.
+%
+Therefore the result of application of the scalar-valued mass operator, represented as a vector $y_i$, on a single element $K$ can be written as
+\begin{align*}
+  y_{i} = y_{i_1\,i_2} & = \sum_{q_1} \sum_{q_2} \sum_{j_1} \sum_{j_2}
+  \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
+  \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2})
+  \widetilde{N}_{j_1}(\widetilde{\xi}_{q_1})
+  \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
+  x_{j_1\,j_2}
+  \widetilde{w}_{q_1} \widetilde{w}_{q_2}
+  \\
+  &=
+  \sum_{q_1} \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1}) \widetilde{w}_{q_1}
+  \sum_{q_2} \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2}) \widetilde{w}_{q_2}
+  \left[
+    \sum_{j_1}
+    \widetilde{N}_{j_1}(\widetilde{\xi}_{q_1})
+    \sum_{j_2}
+    \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
+    x_{j_1\,j_2}
+  \right]
+\end{align*}
+%
+{\color{red} Daniel: please adjust the discussion above, I think we need to add more on what happens with those 4 loops}
+In general, this reordering reduces the arithmetic complexity to $\mathcal{O}(dn^{d+1})$ and is called sum factorization.
 Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for 3D and high polynomial degree with the consistent quadrature rule.

### 2018-10-18 doc/large_strain.tex
more

@@
 $\d \gz x = \gz F \cdot \d \gz X$,
-where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient and $\gz I$ is the second order unit tensor.
+where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient and $\gz I$ is the second-order unit tensor.
 The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian $J = \det \gz F > 0$ has to be positive.
@@
 \end{align}
-Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \varphi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
+Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \varphi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
 The first term in \eqref{eq:tangent} is related to the linearization of the Kirchhoff stress and therefore called material part of the tangent operator.
@@
 In general, this reordering reduces the arithmetic complexity to $\mathcal{O}(dn^{d+1})$ and is called sum factorization.
-Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for 3D and high polynomial degree with the consistent quadrature rule.
-More information on techniques developed to increase the computational density can be found in \cite{kronbichler12,vos10}.
+Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for 3D and high polynomial degree with the corresponding quadrature rule.
+More information on matrix-free techniques can be found in \cite{kronbichler12,vos10}.
 We briefly note that the SIMD vectorization is applied at the finite element level, that is the operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
@@
 Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
-Below, we propose three algorithms to implement tangent operator of finite-strain linear elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
+Below, we propose three algorithms to implement the tangent matrix of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
 %
@@
 Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is performed over the deformed configuration.
-Therefore, we have to divide by the Jacobian $J$ of the mapping at a given quadrature point additionally to express the integral in the undeformed configuration.
+Therefore, we have to divide by the Jacobian $J$ of the deformation map at a given quadrature point additionally to express the integral in the undeformed configuration.
 
@@
 Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to additions and multiplications.
-As evident from Section \ref{sec:theory}, we need to reevaluate the Kirchhoff stress at each quadrature point in order to apply the operator.
+As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the operator.
 This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the undeformed configuration.
@@
   }
-  \caption{Matrix-free tangent cell operator, variant 1: cache second order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate refential quantities like $\gz F$ at runtime.}
+  \caption{Matrix-free tangent cell operator, variant 1: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate refential quantities like $\gz F$ at runtime.}
   \label{alg:mf_tensor2}
@@
 In this case, we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
-This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed in the operator itself.
-It is, however, somewhat more generic than Algorithm \ref{alg:mf_tensor1} since the dependence of $\gz \tau$ exists outside of the oeperator itself (although it must remain consistent with the action of $J \boldsymbol {\mathcal{C}}$).
-Therefore, the action of the material fourth order tangent $J \boldsymbol{\mathcal{C}}$ on the second order symmetric tensor $\gz g_s$, as defined in \eqref{eq:simplified_action}, is cheap to evaluate.
+This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
+Therefore, the action of the material fourth-order tangent $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\gz \nabla_{sym} \gz x$ is cheap to evaluate.
 To that end, we cache two scalars that depend on the Jacobian $J$.
@@
   }
-  \caption{Matrix-free tangent cell operator: cache fourth order material tangent $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
+  \caption{Matrix-free tangent cell operator: cache fourth-order material tangent $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
   \label{alg:mf_tensor4}
@@
 Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent tensor.
-In this case, we have to cache the fourth order symmetric tensor $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\gz g_s$ on-the-fly.
+In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\gz \nabla_{sym} \gz x$ on-the-fly.
 
@@
 For this work we adopt the matrix-free level operators within the GMG and employ Chebyshev pre- and post-smoothing on each level.
-In general, given a heterogeneous non-linear material the definition of transfer operators (for restriction and prolongation) and level operators is not trivial.
+In general, given a heterogeneous non-linear material, the definition of transfer operators (for restriction and prolongation) and level operators is not trivial.
 One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1985, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
@@
 However, for a patch of elements consisting of the same material, the transfer operators reduce to the standard geometric transfer.
-Based on this observation we adopt the following approach: It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous fine-strain elastic material.
+Based on this observation we adopt the following approach: It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material.
 In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
@@
 Essentially, level operators correspond to the linearization around smoothed representation of displacement field\footnote{
-Note that this differs from the classical approach where the (tangent) operator $A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $A^{l}$ on level $l$ via $A^{l+1}=I^{l+1}_{l}  A^l I^l_{l+1}$, where $I^l_{l+1}$ and $I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
+Note that this differs from the classical approach where the (tangent) operator $\gz A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $\gz A^{l}$ on level $l$ via $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, where $\gz I^l_{l+1}$ and $\gz I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
 }
@@
 The distributed load of $0.5\,\rm{N}$ is applied in 5 steps.
-With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the forces is $10^{-8}$.
-The relative convergence criteria for linear solver is $10^{-6}$.
+With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$.
+The relative convergence criteria for the linear solver is $10^{-6}$.
 
@@
   \end{subfigure}
-  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. Color indicate the process owning the element.}%
+  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. Color indicates the process owning the element.}%
   \label{fig:miehe_gmg}
@@
   \end{tabular}
-  \caption{Parameters for the 2D benchmark.
+  \caption{Parameters for the 2D benchmark: $p$ is the polynomial degree,
+  $q$ is the quadrature order, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
   }
@@
   \end{tabular}
-  \caption{Parameters for the 3D benchmark.
+  \caption{Parameters for the 3D benchmark: $p$ is the polynomial degree,
+  $q$ is the quadrature order, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
   }
@@
 Computations are done on a single node of two clusters:
-A node on ``Emmy'' cluster has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM.
+A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM.
 Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
@@
       \includegraphics[width=\textwidth]{Emmy_RRZE_timing2d.eps}
-      \caption{vmult (2D)}
+      \caption{matrix-vector product (2D)}
       \label{fig:benchmark_miehe_Emmy_vmult2}
@@
     \includegraphics[width=\textwidth]{Emmy_RRZE_timing3d.eps}
-    \caption{vmult (3D)}
+    \caption{matrix-vector product (3D)}
     \label{fig:benchmark_miehe_Emmy_vmult3}
@@
       \includegraphics[width=\textwidth]{IWR_timing2d.eps}
-      \caption{vmult (2D)}
+      \caption{matrix-vector product (2D)}
       \label{fig:benchmark_miehe_IWR_vmult2}
@@
     \includegraphics[width=\textwidth]{IWR_timing3d.eps}
-    \caption{vmult (3D)}
+    \caption{matrix-vector product (3D)}
     \label{fig:benchmark_miehe_IWR_vmult3}
@@
 The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the reference configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
-There is little difference between the approach where the fourth order material tangent tensor is cached (``tensor4'') and the one where we only cache the second order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material tangent on the second order symmetric tensor.
+There is little difference between the approach where the fourth-order material tangent tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material tangent on the second-order symmetric tensor.
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
-Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the fourth order material tangent in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
+Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the fourth-order material tangent in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
 It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws.
@@
 As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
-Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth order symmetric tensor and one second order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
+Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
 
@@
 In this contribution, we proposed and numerically investigated several matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
-The implementation that caches the fourth order material tangent together with the second order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
+The implementation that caches the fourth-order material tangent together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
 This gives hope that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material tangent on the quadrature point level.

### 2018-10-18 doc/large_strain.tex
minor

@@
 
-D.~Davydov acknowledge the financial support of the German Research Foundation (DFG), grant DA 1664/2-1
-and the Competence Network for Technical and Scientific High Performance Computing in Bavaria (KONWIHR).
+D.~Davydov acknowledge the financial support of the German Research Foundation (DFG), grant DA 1664/2-1.
 

### 2018-10-18 doc/large_strain.tex
minor
fix bib

@@
     x_{j_1\,j_2}
-  \right]
+  \right]\, \forall i_1 \, i_2
 \end{align*}
@@
 In general, given a heterogeneous non-linear material, the definition of transfer operators (for restriction and prolongation) and level operators is not trivial.
-One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1985, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
+One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
   We note that this approach is limited to the case where for each quadrature point within an element the fourth-order tangent operator is the same.

### 2018-10-18 doc/large_strain.tex
Improve description of sum factorization

@@
 and this is the same as for a regular matrix-vector product where the entries are already precomputed.
-Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently.
+Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.
 Let us illustrate this in 2D. Tensor-product quadrature points can be expressed as a combination of one dimensional quadrature points
@@
 %
-Therefore the result of application of the scalar-valued mass operator, represented as a vector $y_i$, on a single element $K$ can be written as
+Therefore, the result of application of the scalar-valued mass operator, represented as a vector $y_i$, on a single element $K$ can be written as
 \begin{align*}
@@
 %
-{\color{red} Daniel: please adjust the discussion above, I think we need to add more on what happens with those 4 loops}
-In general, this reordering reduces the arithmetic complexity to $\mathcal{O}(dn^{d+1})$ and is called sum factorization.
+These four loops all have the same structure.
+We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point and fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
+Since each of these $2\cdot d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, we end up with an arithmetic complexity of $\mathcal{O}(dn^{d+1})$.
 Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for 3D and high polynomial degree with the corresponding quadrature rule.

### 2018-10-19 doc/large_strain.tex
a few from Daniel

@@
 
-  \address[c]{Glasgow computational engineering center,
+  \address[c]{Glasgow Computational Engineering Center (GCEC),
       University of Glasgow, G12 8QQ Glasgow, United Kingdom
@@
   \end{subfigure}
-  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. Color indicates the process owning the element.}%
+  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
   \label{fig:miehe_gmg}
@@
 
-D.~Davydov acknowledge the financial support of the German Research Foundation (DFG), grant DA 1664/2-1.
+D.~Davydov acknowledges the financial support of the German Research Foundation (DFG), grant DA 1664/2-1.
 

### 2018-10-19 doc/large_strain.tex
add jacobian from isoparameter to real element in MF

@@
 Here $\gz A$ is the discrete tangent matrix, $\gz F$ is the discrete gradient
-and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form for the trial solution field $\overline{\gz u^h}$.
+and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
 Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the deformed configuration, whereas the integration is done in the reference (undeformed) configuration.
@@
  (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
-        &\approx \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz x_q) x_j W_q \\
-        &= \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi^K_q) x_j w_q
+        &\approx \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz x_q) x_j w_q \\
+        &= \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K
   \label{eq:mf_vmult}
 \end{align*}
-where in the last step $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi^K_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $q$.
+where in the last step $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point on isoparametric element $\gz \xi_q$ and $J^K$ is the Jacobian of the mapping from the isoparametric to the element $K$.
 For the sake of demonstration, let us assume that $\gz A$ represents a mass matrix with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
 \begin{align*}
- \tilde{a}(N_i,N_j)(\gz \xi^K_q) = N_i(\gz \xi^K_q)N_j(\gz \xi^K_q).
+ \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
 \end{align*}
@@
 \end{align}
-where for quadrature points $q$ we can associate a multi-index $(q_1,q_2)$.
+where for each quadrature points $q$ we can associate a multi-index $(q_1,q_2)$.
 Weights associated with the quadrature formula $w_q$ can be expressed as a product of one-dimensional weights
@@
 \end{align}
-Shape functions can also be expressed as product of $d$ one dimensional basis functions
+Shape functions can also be expressed as product of one dimensional basis functions
 \begin{align}
@@
 \begin{align*}
-  y_{i} = y_{i_1\,i_2} & = \sum_{q_1} \sum_{q_2} \sum_{j_1} \sum_{j_2}
+  y_{i} = y_{i_1\,i_2} & = J^K \sum_{q_1} \sum_{q_2} \sum_{j_1} \sum_{j_2}
   \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
@@
   \\
-  &=
+  &= J^K
   \sum_{q_1} \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1}) \widetilde{w}_{q_1}
@@
 These four loops all have the same structure.
-We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point and fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
-Since each of these $2\cdot d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, we end up with an arithmetic complexity of $\mathcal{O}(dn^{d+1})$.
+We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
+Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, we end up with an arithmetic complexity of $\mathcal{O}(dn^{d+1})$.
 Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for 3D and high polynomial degree with the corresponding quadrature rule.

### 2018-10-19 doc/large_strain.tex
improve MF algorithms

@@
   \ForEach{ element $K \in \Omega^h$ }{
+        evaluate the following tensors at each quadrature point using sum factorization \;
+        $\gz \nabla_X \gz u^h$ \tcp*{2nd order}
+        $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
+        $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{
-          evaluate $\gz \nabla_X \gz u^h$ \tcp*{2nd order}
           evaluate $\gz F = \gz I + \gz \nabla_X \gz u^h$ \tcp*{2nd order}
@@
           evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
-          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
           evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
-          evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
-          queue $\boldsymbol{\mathcal{G}}$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
-          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
-          evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
+          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
+          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
+          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
           queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
@@
   \ForEach{ element $K \in \Omega^h$ }{
+        evaluate the following tensors at each quadrature point using sum factorization \;
+        $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
+        $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{
-          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
-          evaluate $\boldsymbol{\mathcal{G}}\gz g = \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
+          evaluate $\boldsymbol{\mathcal{G}}\gz g := \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
           queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
-          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
-          evaluate $\boldsymbol{\mathcal{C}} \gz g_s = c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
+          evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
           queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
@@
   \ForEach{ element $K \in \Omega^h$ }{
+        evaluate the following tensors at each quadrature point using sum factorization \;
+        $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
+        $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{
-          evaluate $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
-          evaluate $\boldsymbol{\mathcal{G}} \gz g = \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
+          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
           queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
-          evaluate $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
-          evaluate $\boldsymbol{\mathcal{C}}\gz g_s = \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
+          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
           queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;

### 2018-10-19 Calculations/post_process.py
adjust font size in Python post-processor script

@@ -6,6 +6,7 @@ import matplotlib.pyplot as plt
 import matplotlib as mp
 import numpy as np
 import fileinput
+from matplotlib.ticker import MaxNLocator
 
 def remove_creation_date(file_name):
     for line in fileinput.input(file_name, inplace=True):
@@ -188,6 +189,10 @@ mem3d_t4 = [tup[3]/tup[1] for tup in mf3d_data_tensor4]
 # file location
 fig_prefix = os.path.join(os.getcwd(), '../doc/' + os.path.basename(os.path.normpath(prefix)) + '_')
 
+params = {'legend.fontsize': 20,
+          'font.size': 20}
+plt.rcParams.update(params)
+
 plt.plot(deg2d,time2d_tr, 'rs--', label='Trilinos')
 plt.plot(deg2d,time2d_sc, 'bo--', label='MF scalar')
 plt.plot(deg2d,time2d_t2, 'g^--', label='MF tensor2')
@@ -195,12 +200,15 @@ plt.plot(deg2d,time2d_t4, 'cv--', label='MF tensor4')
 plt.xlabel('degree')
 plt.ylabel('vmult wall time (s) / DoF')
 leg = plt.legend(loc='best', ncol=1)
+plt.tight_layout()
 plt.savefig(fig_prefix + 'timing2d.eps', format='eps')
 remove_creation_date(fig_prefix + 'timing2d.eps')
 
 # clear
 plt.clf()
 
+ax = plt.figure().gca()
+ax.xaxis.set_major_locator(MaxNLocator(integer=True))
 plt.plot(deg3d,time3d_tr, 'rs--', label='Trilinos')
 plt.plot(deg3d,time3d_sc, 'bo--', label='MF scalar')
 plt.plot(deg3d,time3d_t2, 'g^--', label='MF tensor2')
@@ -208,11 +216,13 @@ plt.plot(deg3d,time3d_t4, 'cv--', label='MF tensor4')
 plt.xlabel('degree')
 plt.ylabel('vmult wall time (s) / DoF')
 leg = plt.legend(loc='best', ncol=1)
+plt.tight_layout()
 plt.savefig(fig_prefix + 'timing3d.eps', format='eps')
 remove_creation_date(fig_prefix + 'timing3d.eps')
 
 # clear
 plt.clf()
[…]

### 2018-10-19 Calculations/pre_process.py
cleanp pre-process script and create two more parameter files for extra runs

@@ -74,15 +74,36 @@ end
 
 subsection Misc
   set Output folder = {7}
-  set Output points = 0,0{9}
+  set Output points = {9}
+  set Output solution = {13}
 end
 
 subsection Boundary conditions
-  set Dirichlet IDs and expressions = 1:0,0{10}
-  set Dirichlet IDs and component mask = 1:true,true{11}
-  set Neumann IDs and expressions = 11:(0.05*t)/(2.0e-6),0{12}
+  set Dirichlet IDs and expressions = {10}
+  set Dirichlet IDs and component mask = {11}
+  set Neumann IDs and expressions = {12}
 end
 """
+
+output_points  = {
+  2: '0,0',
+  3: '0,0,0.5e-3'
+}
+
+dirichlet_id   = {
+  2: '1:0,0',
+  3: '1:0,0,0;2:0,0,0'
+}
+
+dirichlet_mask = {
+  2: '1:true,true',
+  3: '1:true,true,true;2:false,false,true'
+}
+
+neumann_bc     = {
+  2: '11:(0.05*t)/(2.0e-6),0',
+  3: '11:(0.05*t)/(2.0e-6),0,(0.05*t)/(2.0e-6)'
+}
 
 base_prm = args.base_prm
 base_name = args.base_prm.split('.')[0]
@@ -114,12 +135,38 @@ for pqrd in poly_quad_ref_dim:
             s[2],
             out_dir + name,
             pqrd[3],
-            "" if pqrd[3]==2 else ",0.5e-3",
-            "" if pqrd[3]==2 else ",0;2:0,0,0",
-            "" if pqrd[3]==2 else ",true;2:false,false,true",
-            "" if pqrd[3]==2 else ",(0.05*t)/(2.0e-6)"
+            output_points[pqrd[3]],
+            dirichlet_id[pqrd[3]],
+            dirichlet_mask[pqrd[3]],
+            neumann_bc[pqrd[3]],
+            'false'
         ))
         filenames.append(name)
+
+# two more runs just to illustrate the deformed mesh
+for pqrd in [(2,3,2,2),(2,3,2,3)]:
+    for s in [('MF_CG', 'gmg', 'tensor4')]:
+        name = '__' + base_name + '_{6}d_p{0}q{1}r{2}_{3}_{4}_{5}'.format(pqrd[0],pqrd[1],pqrd[2],s[0],s[1],s[2],pqrd[3])
+        fname = name + '.prm'
+        print '{0}'.format(fname)
+        fout = open(fname, 'w')
[…]

### 2018-10-19 doc/large_strain.tex
one more correction from Prof

@@
 
-In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic material as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}.
-The coarse mesh for the 2D and 3D problems is illustrated in Figure \ref{fig:miehe}.
+In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}.
+The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}.
 The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red).

### 2018-10-19 doc/large_strain.tex
show deformed meshes
more loading BC more straight forward, also update the text

@@
 Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
-The distributed load of $0.5\,\rm{N}$ is applied in 5 steps.
+The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
 With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$.
 The relative convergence criteria for the linear solver is $10^{-6}$.
+Shared memory parallelization using Intel Threading Building Blocks was disabled for this study.
 
@@
 Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.
+Figure \ref{fig:miehe_deformed} shows deformed meshes at the final loading step with quadratic elements and two global mesh refinements.
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{deformed.png}
+      \caption{2D deformed mesh}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{deformed_3d.png}
+    \caption{3D deformed mesh}
+  \end{subfigure}
+  \caption{Deformed meshes at the final loading step with quadratic elements and two global mesh refinements.}%
+  \label{fig:miehe_deformed}
+\end{figure}
 
@@
 The benchmark calculations are performed for 2D and 3D problems with various combination of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
-Shared memory parallelization using Intel Threading Building Blocks was disabled for this study.
 %
-
 We are interested in the following metrics:

### 2018-10-19 Calculations/pre_process.py

@@ -101,8 +101,8 @@ dirichlet_mask = {
 }
 
 neumann_bc     = {
-  2: '11:(0.05*t)/(2.0e-6),0',
-  3: '11:(0.05*t)/(2.0e-6),0,(0.05*t)/(2.0e-6)'
+  2: '11:12.5e3*t,0',
+  3: '11:12.5e3*t,0,12.5e3*t'
 }
 
 base_prm = args.base_prm

### 2018-10-19 doc/large_strain.tex
denote push-forward via \chi for a good reason

@@
 more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether the matrix-free methods can be advantageous in this case.
-Another reason could be the use of lower order elements due to lacking smoothness of the underlying solution.
+Another reason could be the use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
 The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community.
@@
 % The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
-We shall denote the push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by
+We shall denote the push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by a linear transformation $\chi$ defined as
 \begin{align}
-  \varphi\left( \mathbf{A} \right)_{ij}
+  \chi\left( \mathbf{A} \right)_{ij}
   &= F_{iA} A_{AB} F_{jB} \\
-\varphi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
+  \chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
   &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}.
 \end{align}
+Note that $\d \gz x = \chi(\d \gz X)$.
 
@@
 Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
-that is $\gz \tau = \varphi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.
+that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.
 
@@
 \end{align}
-Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \varphi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
+Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
 The first term in \eqref{eq:tangent} is related to the linearization of the Kirchhoff stress and therefore called material part of the tangent operator.
@@
   \equiv J \boldsymbol{\sigma}
-  = \varphi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
+  = \chi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
   = \mu \mathbf{b} - \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{I}
@@
 J \boldsymbol{\mathcal{C}}
-  = \varphi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
+  = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
   = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
@@
 \begin{alignat}{2}
-       \gz u^h &=:  \sum_{i \in \mcl I_\varphi} u^i \gz N_i (\gz X) \quad \quad \quad
-\delta \gz u^h &&=: \sum_{i \in \mcl I_\varphi} \delta u^i \gz N_i (\gz X) \,,
+       \gz u^h &=:  \sum_{i \in \mcl I} u^i \gz N_i (\gz X) \quad \quad \quad
+\delta \gz u^h &&=: \sum_{i \in \mcl I} \delta u^i \gz N_i (\gz X) \,,
 \end{alignat}
-where the superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I_\varphi$ is the set of unknown degrees of freedom (DoF).
+where the superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I$ is the set of unknown degrees of freedom (DoF).
 %{\color{red} For the sake of this study, $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.}

### 2018-10-19 doc/large_strain.tex
typo

@@
 \end{align*}
-where in the last step $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point on isoparametric element $\gz \xi_q$ and $J^K$ is the Jacobian of the mapping from the isoparametric to the element $K$.
+where in the last step $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point on isoparametric element $\gz \xi_q$ and $J^K$ is the Jacobian of the mapping from the isoparametric element to the element $K$.
 For the sake of demonstration, let us assume that $\gz A$ represents a mass matrix with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that

### 2018-10-19 doc/large_strain.tex
provisional fix for linearization part

@@
 F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
-F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u).
+F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u.
 \end{align}
 For conservative traction boundary conditions, the directional derivative is given by
-\begin{align}
-\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u)
-=
-\int_{\mcl B_0} \D_{\Delta \gz u} \gz \tau : \overline{\grad^s} \, \delta \gz u \, \d V
-+
-\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot \D_{\Delta \gz u}  \gz F^{-1} \right] \d V.
-\end{align}
+\begin{equation}
+\begin{split}
+\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u
+&=
+\int_{\mcl B_0} \left[\D_{\Delta \gz u} \gz \tau \cdot \Delta \gz u\right] : \overline{\grad^s} \, \delta \gz u \, \d V
+\\
+& +
+\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot
+\left[\D_{\Delta \gz u}  \gz F^{-1} \cdot \Delta \gz u\right]
+\right] \d V.
+\end{split}
+\end{equation}
 Following \cite{Wriggers2008}, this can be simplified to
-\begin{align}
-\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u)
-  =
-  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V
-  +
+\begin{equation}
+  \begin{split}
+\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u
+  &=
+  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V \\
+  &+
   \int_{\mcl B_0}
@@
   \d V.
+\end{split}
 \label{eq:tangent}
-\end{align}
+\end{equation}
 Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.

### 2018-10-19 doc/large_strain.tex
drop superscripts from gradient
fix explanation of material and geometric parts

@@
 F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
-F(\overline{\gz u}, \delta \gz u) + \D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u.
+F(\overline{\gz u}, \delta \gz u) + \D F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u.
 \end{align}
@@
 \begin{split}
-\D_{\Delta \gz u}  F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u
+\D F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u
 &=
-\int_{\mcl B_0} \left[\D_{\Delta \gz u} \gz \tau \cdot \Delta \gz u\right] : \overline{\grad^s} \, \delta \gz u \, \d V
+\int_{\mcl B_0} \left[\D \left[\gz F \cdot \gz S \cdot \gz F^T\right] \cdot \Delta \gz u\right] : \overline{\grad^s} \, \delta \gz u \, \d V
 \\
@@
 \int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot
-\left[\D_{\Delta \gz u}  \gz F^{-1} \cdot \Delta \gz u\right]
+\left[\D \gz F^{-1} \cdot \Delta \gz u\right]
 \right] \d V.
@@
   \begin{split}
-\D_{\Delta \gz u} F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u
+\D F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u
   &=
@@
 Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
-The first term in \eqref{eq:tangent} is related to the linearization of the Kirchhoff stress and therefore called material part of the tangent operator.
-The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the linearization of the inverse of the deformation gradient $\gz F^{-1}$.
+The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and therefore called material part of the tangent operator.
+The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.
 

### 2018-10-23 doc/large_strain.tex
few fixes based on JPs comments

@@
 \begin{align}
-  w_q = \widetilde{w}_{q_1} \widetilde{w}_{q_2}
+  w_q = \widetilde{w}_{q_1} \widetilde{w}_{q_2} \, .
 \end{align}
@@
 We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
-Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, we end up with an arithmetic complexity of $\mathcal{O}(dn^{d+1})$.
-Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for 3D and high polynomial degree with the corresponding quadrature rule.
+Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, what results is an algorithm with arithmetic complexity of $\mathcal{O}(dn^{d+1})$.
+Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for the 3D case, and where the discretization and evaluation employs high polynomial degree basis functions with the corresponding quadrature rule.
 More information on matrix-free techniques can be found in \cite{kronbichler12,vos10}.
@@
   }
-  \caption{Matrix-free tangent cell operator, variant 1: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate refential quantities like $\gz F$ at runtime.}
+  \caption{Matrix-free tangent cell operator, variant 1: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
   \label{alg:mf_tensor2}

### 2018-10-23 doc/large_strain.tex
fix 3D force density

@@
 Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
-The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
+The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \sqrt{2} \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
 With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$.

### 2018-10-23 doc/large_strain.tex
Add further funding acknowledgement for PS

@@
 
-P.~Steinmann acknowledges the support of the Cluster of Excellence Engineering of Advanced Materials (EAM) which made this collaboration possible.
+P.~Steinmann acknowledges the support of the Cluster of Excellence Engineering of Advanced Materials (EAM) which made this collaboration possible, as well as funding by the EPSRC Strategic Support Package ``Engineering of Active Materials by Multiscale/Multiphysics Computational Mechanics''.
 

### 2018-10-23 Calculations/post_process.py
print more in post-processor debug output

@@ -88,7 +88,7 @@ for f in files:
             cg_iterations = int(re.findall(pattern,line)[0])
 
         if start_line in line:
-            print 'p={0} q={1} cells={2} dofs={3} tr_memory={4} mf_memory={5} file={6}'.format(p, q, cells, dofs, tr_memory, mf_memory, f)
+            print 'dim={0} p={1} q={2} cells={3} dofs={4} tr_memory={5} mf_memory={6} cg_it={7} file={8}'.format(dim, p, q, cells, dofs, tr_memory, mf_memory, cg_iterations, f)
             ready = True
 
         if ready:

### 2018-10-24 doc/large_strain.tex
minor update of fig title
update Linearization section based on discussion with Steinmann

@@
 \begin{align}
-\delta \mcl E =
-\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, .
+\D_{\displaystyle \delta \gz u} \mcl E :=
+\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, \qquad \forall \delta \gz u \, .
 \label{eq:stationary}
@@
 F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
-F(\overline{\gz u}, \delta \gz u) + \D F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u.
+F(\overline{\gz u}, \delta \gz u) + \D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u) ,
 \end{align}
+where $\D_{\displaystyle \Delta \gz u}(\bullet)$ denotes the directional derivative in the direction $\Delta \gz u$.
 For conservative traction boundary conditions, the directional derivative is given by
@@
 \begin{split}
-\D F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u
+\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
 &=
-\int_{\mcl B_0} \left[\D \left[\gz F \cdot \gz S \cdot \gz F^T\right] \cdot \Delta \gz u\right] : \overline{\grad^s} \, \delta \gz u \, \d V
+\int_{\mcl B_0}
+\D_{\displaystyle \Delta \gz u} \left(\gz F \cdot \gz S \cdot \gz F^T\right)  :
+\overline{\grad^s} \, \delta \gz u
+\, \d V
 \\
 & +
-\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot
-\left[\D \gz F^{-1} \cdot \Delta \gz u\right]
+\int_{\mcl B_0}
+\overline{\gz \tau} :
+\left[
+  \Grad \, \delta \gz u \cdot
+  \D_{\displaystyle \Delta \gz u} \gz F^{-1}
 \right] \d V.
 \end{split}
+\label{eq:tangent_pre}
 \end{equation}
-Following \cite{Wriggers2008}, this can be simplified to
+Following \cite{Wriggers2008}, this can be simplified to\footnote{To that end
+$\D_{\displaystyle \Delta \gz u} \gz F = \Grad \Delta \gz u$,
+$\;\D_{\displaystyle \Delta \gz u} \gz F^{-1} = - \gz F^{-1} \cdot \D_{\displaystyle \Delta \gz u} \gz F \cdot \gz F^{-1}$ and
+$\D_{\displaystyle \Delta \gz u} \gz E = \frac{1}{2}\left[\D_{\displaystyle \Delta \gz u} \gz F^T \cdot \gz F + \gz F^T \cdot \D_{\displaystyle \Delta \gz u} \gz F\right]$
+are employed together with
+$\D_{\displaystyle \Delta \gz u} \gz S = 2 \partial \gz S / \partial \gz C : \D_{\displaystyle \Delta \gz u} \gz E$.
+}
 \begin{equation}
   \begin{split}
-\D F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u
+\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
   &=
@@
   \end{subfigure}
-  \caption{Discretization of the heterogeneous material at the coarsest mesh level.}%
+  \caption{Discretization of the heterogeneous material at the coarsest mesh level and the prescribed boundary conditions.}%
   \label{fig:miehe}

### 2018-11-01 doc/large_strain.tex
do not use \nabla in algorithms (we don't use it in theoretical part)

@@
 \begin{alignat}{2}
-       \gz u^h &=:  \sum_{i \in \mcl I} u^i \gz N_i (\gz X) \quad \quad \quad
-\delta \gz u^h &&=: \sum_{i \in \mcl I} \delta u^i \gz N_i (\gz X) \,,
+       \gz u^h &=:  \sum_{i \in \mcl I} u_i \gz N_i (\gz X) \quad \quad \quad
+\delta \gz u^h &&=: \sum_{i \in \mcl I} \delta u_i \gz N_i (\gz X) \,,
 \end{alignat}
@@
   \SetKwInOut{Output}{Return}
-  \Input{RHS FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
+  \Input{Source FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
   cached $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
@@
         evaluate the following tensors at each quadrature point using sum factorization \;
-        $\gz \nabla_X \gz u^h$ \tcp*{2nd order}
-        $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
-        $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
+        $\Grad \overline{\gz u^h}$ \tcp*{2nd order}
+        $\gz g_s := \overline{\grad^s} \gz x$ \tcp*{2nd order symmetric}
+        $\gz g := \overline{\grad} \gz x$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{
-          evaluate $\gz F = \gz I + \gz \nabla_X \gz u^h$ \tcp*{2nd order}
+          evaluate $\gz F = \gz I + \Grad \overline{\gz u^h}$ \tcp*{2nd order}
           evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
@@
           evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
-          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
+          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
           evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
-          queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
+          queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
         }
@@
   \SetKwInOut{Output}{Return}
-  \Input{RHS FE vector $\gz x$,
+  \Input{Source FE vector $\gz x$,
   cached
@@
         evaluate the following tensors at each quadrature point using sum factorization \;
-        $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
-        $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
+        $\gz g_s := \overline{\grad^s} \gz x$ \tcp*{2nd order symmetric}
+        $\gz g := \overline{\grad} \gz x$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{
           evaluate $\boldsymbol{\mathcal{G}}\gz g := \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
-          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
+          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
           evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
-          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
+          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
         }
@@
 This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
-Therefore, the action of the material fourth-order tangent $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\gz \nabla_{sym} \gz x$ is cheap to evaluate.
+Therefore, the action of the material fourth-order tangent $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
 To that end, we cache two scalars that depend on the Jacobian $J$.
@@
   \SetKwInOut{Output}{Return}
-  \Input{RHS FE vector $\gz x$,
+  \Input{Source FE vector $\gz x$,
   cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point,
@@
         evaluate the following tensors at each quadrature point using sum factorization \;
-        $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
-        $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
+        $\gz g_s := \overline{\grad^s} \gz x$ \tcp*{2nd order symmetric}
+        $\gz g := \overline{\grad} \gz x$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{
           evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
-          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
+          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
           evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
-          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
+          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
         }
@@
 Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent tensor.
-In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\gz \nabla_{sym} \gz x$ on-the-fly.
+In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\overline{\grad^s} \gz x$ on-the-fly.
 

### 2018-11-01 doc/large_strain.tex
make J_q and fix complexity of MF mass matrix

@@
 The matrix-vector product $\gz A \gz x$ can be expressed by
-\begin{align*}
+\begin{align}
+  \begin{split}
  (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
-        &\approx \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz x_q) x_j w_q \\
-        &= \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K
+        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
+  \end{split}
   \label{eq:mf_vmult}
-\end{align*}
+\end{align}
 where in the last step $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point on isoparametric element $\gz \xi_q$ and $J^K$ is the Jacobian of the mapping from the isoparametric element to the element $K$.
@@
 \begin{align*}
-  y_{i} = y_{i_1\,i_2} & = J^K \sum_{q_1} \sum_{q_2} \sum_{j_1} \sum_{j_2}
+  y_{i} = y_{i_1\,i_2} & = \sum_{q_1} \sum_{q_2} \sum_{j_1} \sum_{j_2}
   \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
@@
   x_{j_1\,j_2}
-  \widetilde{w}_{q_1} \widetilde{w}_{q_2}
+  \widetilde{w}_{q_1} \widetilde{w}_{q_2} J^K_{q1\,q2}
   \\
-  &= J^K
+  &=
   \sum_{q_1} \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1}) \widetilde{w}_{q_1}
   \sum_{q_2} \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2}) \widetilde{w}_{q_2}
+  J^K_{q1\,q2}
   \left[
@@
 We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
-Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, what results is an algorithm with arithmetic complexity of $\mathcal{O}(dn^{d+1})$.
+Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, what results is an algorithm with arithmetic complexity of $\mathcal{O}(2dn^{d+1})$.
 Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for the 3D case, and where the discretization and evaluation employs high polynomial degree basis functions with the corresponding quadrature rule.

### 2018-11-01 doc/large_strain.tex
illustrate multi-index for nodes and quadrature

@@
 Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{nodes.png}
+      \caption{Nodes of cubic Lagrange element}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{quad_points.png}
+    \caption{Gaussian quadrature points}
+  \end{subfigure}
+  \caption{Illustration of tensorial indexing for nodes and quadrature points. In this sketch a node/quadrature point with the number $7$ can be indexed with a multi-index $\{3,2\}$.}%
+  \label{fig:multi_index}
+\end{figure}
+
 Let us illustrate this in 2D. Tensor-product quadrature points can be expressed as a combination of one dimensional quadrature points
@@
 where for each DoF index $i$ we can associate a multi-index $(i_1, i_2)$.
+See Figure \ref{fig:multi_index} for illustration.
 %

### 2018-11-01 doc/large_strain.tex
explain no-difference between x_j and X_{j1 j2}
new paragraph for SIMD comment and TBB
improvements based on Daniel's comments

@@
 \end{align}
-where in the last step $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point on isoparametric element $\gz \xi_q$ and $J^K$ is the Jacobian of the mapping from the isoparametric element to the element $K$.
+where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point on isoparametric element $\gz \xi_q$ and $J^K$ is the Jacobian of the mapping from the isoparametric element to the element $K$.
 For the sake of demonstration, let us assume that $\gz A$ represents a mass matrix with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
@@
 %
-Therefore, the result of application of the scalar-valued mass operator, represented as a vector $y_i$, on a single element $K$ can be written as
+Therefore, the result of the application of the scalar-valued mass operator, represented as a vector $y_i$, on a single element $K$ can be written as
 \begin{align*}
-  y_{i} = y_{i_1\,i_2} & = \sum_{q_1} \sum_{q_2} \sum_{j_1} \sum_{j_2}
+  y_{i} \equiv Y_{i_1\,i_2} & = \sum_{q_1} \sum_{q_2} \sum_{j_1} \sum_{j_2}
   \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
@@
   \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
-  x_{j_1\,j_2}
+  X_{j_1\,j_2}
   \widetilde{w}_{q_1} \widetilde{w}_{q_2} J^K_{q1\,q2}
@@
     \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
-    x_{j_1\,j_2}
+    X_{j_1\,j_2}
   \right]\, \forall i_1 \, i_2
@@
 %
+Note that $x_j \equiv X_{j_1\,j_2}$; the former access the element source vector using a linear index $j$, whereas the latter uses the associated multi-index accesses $\{j_1\,j_2\}$.
+%
 These four loops all have the same structure.
 We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
-Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, what results is an algorithm with arithmetic complexity of $\mathcal{O}(2dn^{d+1})$.
+Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, this results in an algorithm with an arithmetic complexity of $\mathcal{O}(2dn^{d+1})$.
 Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for the 3D case, and where the discretization and evaluation employs high polynomial degree basis functions with the corresponding quadrature rule.
 More information on matrix-free techniques can be found in \cite{kronbichler12,vos10}.
+
 We briefly note that the SIMD vectorization is applied at the finite element level, that is the operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
@@
         evaluate the following tensors at each quadrature point using sum factorization \;
-        $\Grad \overline{\gz u^h}$ \tcp*{2nd order}
+        $\Grad \, \overline{\gz u^h}$ \tcp*{2nd order}
         $\gz g_s := \overline{\grad^s} \gz x$ \tcp*{2nd order symmetric}

### 2018-11-01 doc/large_strain.tex
one more space

@@
         \ForEach{quadrature point $q$ on $K$}{
-          evaluate $\gz F = \gz I + \Grad \overline{\gz u^h}$ \tcp*{2nd order}
+          evaluate $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ \tcp*{2nd order}
           evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}

### 2018-11-01 doc/large_strain.tex
few mixxing subscripts
add a note in intro related to linear and quadratic elemetns and MF. Color text changes for ease of review

@@
 Even higher order elements are rarely employed.
+{\color{red}
+This can potentially reduce advantage of matrix-free methods, that are generally more competitive for higher order elements.
+}
 However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
@@
 where for each DoF index $i$ we can associate a multi-index $(i_1, i_2)$.
+{\color{red}
 See Figure \ref{fig:multi_index} for illustration.
+}
 %
@@
   X_{j_1\,j_2}
-  \widetilde{w}_{q_1} \widetilde{w}_{q_2} J^K_{q1\,q2}
+  \widetilde{w}_{q_1} \widetilde{w}_{q_2} J^K_{q_1\,q_2}
   \\
@@
   \sum_{q_2} \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2}) \widetilde{w}_{q_2}
-  J^K_{q1\,q2}
+  J^K_{q_1\,q_2}
   \left[
@@
 %
+{\color{red}
 Note that $x_j \equiv X_{j_1\,j_2}$; the former access the element source vector using a linear index $j$, whereas the latter uses the associated multi-index accesses $\{j_1\,j_2\}$.
+}
 %

### 2018-11-01 doc/large_strain.tex
use tangent tensors where we speak about 4th order tesnors and reserve operators for FE spaces only

@@
 
-    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyperelastic tangent operator numerically.
+    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyperelastic {\color{red}finite element} tangent operator numerically.
     In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
@@
 We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent masters thesis produced by a student of the first two of the authors \cite{Mentler2017}.
-One possible reason is that the tangent operators (stemming from linearization of the nonlinear balance equations) are
+One possible reason is that the tangent operators\footnote{{\color{red}Here and
+below by operators we mean linear operators on vector space $\mathbb R^N$ (or  endomorphisms of $\mathbb R^N$), where $N$ is the number of unknown degrees of freedom in the FE discretization.}} (stemming from linearization of the nonlinear balance equations) are
 more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether the matrix-free methods can be advantageous in this case.
@@
 \end{equation}
-Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
-The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and therefore called material part of the tangent operator.
+Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order {\color{red}spatial} tangent {\color{red}stiffness} tensor is the push forward of the material tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
+The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and therefore called material part of the tangent.
 The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.
@@
 \end{align}
-The Kirchhoff stress and its associated material tangent are
+The Kirchhoff stress and its associated {\color{red}fourth-order material tangent tensor} are
 \begin{gather}
@@
 \end{align}
-Here $\gz A$ is the discrete tangent matrix, $\gz F$ is the discrete gradient
+Here $\gz A$ is the discrete tangent {\color{red}operator}, $\gz F$ is the discrete gradient
 and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
@@
 Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
-Below, we propose three algorithms to implement the tangent matrix of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
+Below, we propose three algorithms to implement the tangent {\color{red}operator} of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
 %
@@
   cached $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
-  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
+  \Output{action of the FE tangent {\color{red}operator} \eqref{eq:algebraic_tangent} on $\gz x$}
   \ForEach{ element $K \in \Omega^h$ }{
@@
   }
-  \caption{Matrix-free tangent cell operator: scalar cache}
+  \caption{Matrix-free tangent operator: {\color{red}cache scalar quantities only}}
   \label{alg:mf_scalar}
@@
   evaluated based on $\overline{\gz u^h}$}
-  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
+  \Output{action of the FE tangent {\color{red}operator} \eqref{eq:algebraic_tangent} on $\gz x$}
   \ForEach{ element $K \in \Omega^h$ }{
@@
   }
-  \caption{Matrix-free tangent cell operator, variant 1: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
+  \caption{Matrix-free tangent operator: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
   \label{alg:mf_tensor2}
@@
 This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
-Therefore, the action of the material fourth-order tangent $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
+Therefore, the action of the material fourth-order tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
 To that end, we cache two scalars that depend on the Jacobian $J$.
@@
   }
-  \caption{Matrix-free tangent cell operator: cache fourth-order material tangent $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
+  \caption{Matrix-free tangent operator: cache fourth-order material tangent {\color{red}stiffness tensor} $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
   \label{alg:mf_tensor4}
@@
 
-Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent tensor.
+Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent {\color{red}stiffness} tensor.
 In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\overline{\grad^s} \gz x$ on-the-fly.
@@
 One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
-  We note that this approach is limited to the case where for each quadrature point within an element the fourth-order tangent operator is the same.
+  We note that this approach is limited to the case where for each quadrature point within an element the fourth-order tangent {\color{red}stiffness tensor} is the same.
 }.
@@
 In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
-We define the level operators as follows: Recall that the fine scale tangent \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$.
+We define the level operators as follows: Recall that the fine scale tangent {\color{red}operator} \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$.
 We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
@@
 The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the reference configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
-There is little difference between the approach where the fourth-order material tangent tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material tangent on the second-order symmetric tensor.
+There is little difference between the approach where the fourth-order material tangent {\color{red}stiffness} tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material tangent {\color{red}stiffness tensor} on the second-order symmetric tensor.
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
-Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the fourth-order material tangent in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
+Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the fourth-order material tangent {\color{red}stiffness tensor} in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
 It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws.
@@
 In this contribution, we proposed and numerically investigated several matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
-The implementation that caches the fourth-order material tangent together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
-This gives hope that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material tangent on the quadrature point level.
+The implementation that caches the fourth-order material tangent {\color{red}stiffness tensor} together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
+This gives hope that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material tangent {\color{red}stiffness tensor} on the quadrature point level.
 

### 2018-11-01 doc/large_strain.tex
referential/spatial configuration

@@
 %
-The deformation of a body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$ at time $t$
+The deformation of a body $\mcl B$ from the {\color{red}referential} configuration $\mcl B_0$ to the {\color{red}spatial} configuration $\mcl B_t$ at time $t$
 is defined via the deformation map $\gz \varphi_t: \mcl B_0 \rightarrow \mcl B_t$, which places material points of $\mcl B$ into the Euclidean space $\mathbb E^3$.
@@
 \end{align}
-where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
+where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the {\color{red}referential} configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
 
@@
 and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
-Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the deformed configuration, whereas the integration is done in the reference (undeformed) configuration.
+Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the {\color{red}spatial} configuration, whereas the integration is done in the {\color{red}referential} configuration.
 
@@
 %
-In order to obtain gradients with respect to the deformed configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
-In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the deformed configuration given a displacement field.
+In order to obtain gradients with respect to the {\color{red}spatial} configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
+In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the {\color{red}spatial} configuration given a displacement field.
 %
-Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is performed over the deformed configuration.
-Therefore, we have to divide by the Jacobian $J$ of the deformation map at a given quadrature point additionally to express the integral in the undeformed configuration.
+Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is performed over the {\color{red}spatial} configuration.
+Therefore, we have to divide by the Jacobian $J$ of the deformation map at a given quadrature point additionally to express the integral in the {\color{red}referential} configuration.
 
@@
 As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the operator.
-This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the undeformed configuration.
+This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the {\color{red}referential} configuration.
 
@@
 In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point.
-In this case, we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
+In this case, we also avoid the need to evaluate gradients of displacement field with respect to the {\color{red}referential} configuration.
 This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
@@
 Two conclusions can be drawn from these results:
-The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the reference configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
+The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the {\color{red}referential} configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
 There is little difference between the approach where the fourth-order material tangent {\color{red}stiffness} tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material tangent {\color{red}stiffness tensor} on the second-order symmetric tensor.

### 2018-11-01 doc/large_strain.tex
minor

@@
 One possible reason is that the tangent operators\footnote{{\color{red}Here and
-below by operators we mean linear operators on vector space $\mathbb R^N$ (or  endomorphisms of $\mathbb R^N$), where $N$ is the number of unknown degrees of freedom in the FE discretization.}} (stemming from linearization of the nonlinear balance equations) are
+below by operators we mean finite-dimensional linear operators on vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.}} (stemming from linearization of the nonlinear balance equations) are
 more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether the matrix-free methods can be advantageous in this case.

### 2018-11-01 doc/large_strain.tex
add a note on WHY we stick with operator term

@@
 One possible reason is that the tangent operators\footnote{{\color{red}Here and
-below by operators we mean finite-dimensional linear operators on vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.}} (stemming from linearization of the nonlinear balance equations) are
+below by operators we mean finite-dimensional linear operators on vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
+Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
+}} (stemming from linearization of the nonlinear balance equations) are
 more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether the matrix-free methods can be advantageous in this case.

### 2018-11-02 doc/large_strain.tex
few line breaks
switch to IJNME template

@@
 %\documentclass[12pt]{article}
-\documentclass[preprint,12pt,times]{elsarticle}
+% \documentclass[preprint,12pt,times]{elsarticle}
+\documentclass[times,doublespace]{nmeauth}
 
-\biboptions{sort&compress,comma,square}
+%%% ELSOPTIONS:
+% \biboptions{sort&compress,comma,square}
+% \usepackage{amsmath}
 
-% \usepackage{url}
+%%%
 
-% \usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}
-%\usepackage{amssymb}
-\usepackage{amsmath}
-%\usepackage{amsfonts}
 \usepackage{bm}
-%\usepackage{bbm}
-%\usepackage{caption}
 \usepackage{xcolor}
-%\usepackage{geometry}
-%\usepackage{graphicx}
-%\usepackage[colorlinks,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
 \usepackage{listings}
@@
 }
-%\usepackage{layouts}
-%\usepackage{mathrsfs}
-%\usepackage{pgfkeys,pgfmath,pgfcore}
-%\usepackage{srcltx}
-%\usepackage[labelformat=simple]{subcaption}
-%\usepackage{xspace}
 \usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}
@@
 
-%\renewcommand{\thetable}{\arabic{table}}
-%\captionsetup[table]{labelformat=simple, labelsep=colon}
-
-%\pgfkeys{/pgf/number format/.cd,1000 sep={}}
-
-%% see http://tex.stackexchange.com/questions/2441/how-to-add-a-forced-line-break-inside-a-table-cell
-%\newcommand{\specialcell}[2][c]{%
-%  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
-
-% put your own definitions here:
-% \renewcommand\thesubfigure{(\alph{subfigure})}
+%
+% Definitions:
+%
 \def\gz  #1{           \mbox{$\boldsymbol{#1}$}}
-% \DeclareMathAlphabet{\bsf}{OT1}{cmss}{bx}{n}
-% \def\msf  #1{           \mbox{\!\!      $\sf #1$}}
 \def\Grad #1 {{\rm Grad} #1}
@@
 \def\D {\mbox{D}}
-% \newcommand{\norm}[1]{\left\lvert #1 \right\rvert}
-%\newcommand\normDouble[1]{\left\lVert#1\right\rVert}
-%\newcommand{\normXY}[1]{\lvert #1 \rvert_{2d}}
-%\newcommand{\dbracket}[1]{\left[\!\!\left[ #1 \right]\!\!\right]}
-
 \def\mcl  #1{               {\cal #1}}
-%\def\bcl  #1{\mbox{\boldmath$\cal #1$}}
-
-%\DeclareMathOperator{\det}{det}
 \DeclareMathOperator{\trace}{tr}
@@
 
-\begin{frontmatter}
+\runningheads{D. Davydov et al.}{A matrix-free approach for finite-strain hyperelastic problems}
+
 \title{
@@
 
-  \author[a]{Denis Davydov\corref{cor}}
-  \ead{denis.davydov@fau.de}
+  \author{
+    Denis Davydov\affil{1} \corrauth ,
+    Jean-Paul Pelteret\affil{1},
+    Daniel Arndt\affil{2},
+    Paul Steinmann\affil{1,3}
+  }
 
-  \author[a]{Jean-Paul Pelteret}
-  \ead{jean-paul.pelteret@fau.de}
+  \address{
+    \affilnum{1}Chair of Applied Mechanics,
+    Friedrich-Alexander-Universit\"{a}t Erlangen-N\"{u}rnberg,
+    Egerlandstr.\ 5, 91058 Erlangen, Germany \break
+    \affilnum{2}Interdisciplinary Center for Scientific Computing (IWR),
+    Heidelberg University,
+    Im Neuenheimer Feld 205, 69120 Heidelberg, Germany \break
+    \affilnum{3}Glasgow Computational Engineering Center (GCEC),
+        University of Glasgow, G12 8QQ Glasgow, United Kingdom \break
+    }
 
-  \author[b]{Daniel Arndt}
-  \ead{daniel.arndt@iwr.uni-heidelberg.de}
+  \corraddr{E-mail: denis.davydov@fau.de}
 
-  \author[a,c]{Paul Steinmann}
-  \ead{paul.steinmann@fau.de}
+  % \author[a]{Denis Davydov\corref{cor}}
+  % \ead{denis.davydov@fau.de}
 
-  \cortext[cor]{Corresponding author.}
+  % \author[a]{Jean-Paul Pelteret}
+  % \ead{jean-paul.pelteret@fau.de}
 
-  \address[a]{Chair of Applied Mechanics,
-  Friedrich-Alexander-Universit\"{a}t Erlangen-N\"{u}rnberg,
-  Egerlandstr.\ 5, 91058 Erlangen, Germany}
+  % \author[b]{Daniel Arndt}
+  % \ead{daniel.arndt@iwr.uni-heidelberg.de}
 
-  \address[b]{Interdisciplinary Center for Scientific Computing (IWR),
-      Heidelberg University,
-      Im Neuenheimer Feld 205,
-      69120 Heidelberg,
-      Germany}
+  % \author[a,c]{Paul Steinmann}
+  % \ead{paul.steinmann@fau.de}
 
-  \address[c]{Glasgow Computational Engineering Center (GCEC),
-      University of Glasgow, G12 8QQ Glasgow, United Kingdom
-  }
+  % \cortext[cor]{Corresponding author.}
+
+  % \address[a]{Chair of Applied Mechanics,
+  % Friedrich-Alexander-Universit\"{a}t Erlangen-N\"{u}rnberg,
+  % Egerlandstr.\ 5, 91058 Erlangen, Germany}
+
+  % \address[b]{Interdisciplinary Center for Scientific Computing (IWR),
+  %     Heidelberg University,
+  %     Im Neuenheimer Feld 205,
+  %     69120 Heidelberg,
+  %     Germany}
+
+  % \address[c]{Glasgow Computational Engineering Center (GCEC),
+  %     University of Glasgow, G12 8QQ Glasgow, United Kingdom
+  % }
 
@@
 
-  \begin{keyword}
-      adaptive finite element method \sep
-      geometric multigrid \sep
-      finite-strain \sep
-      matrix-free \sep
+  \keywords{
+  % \begin{keyword}
+      adaptive finite element method ; %\sep
+      geometric multigrid ; %\sep
+      finite-strain ; %\sep
+      matrix-free ; %\sep
       hyperelasticity
-  \end{keyword}
+  % \end{keyword}
+  }
 
-  \end{frontmatter}
+  \maketitle
+  % \end{frontmatter}
 

### 2018-11-02 doc/large_strain.tex
most of Prof. Steinmann's comments

@@
 is defined via the deformation map $\gz \varphi_t: \mcl B_0 \rightarrow \mcl B_t$, which places material points of $\mcl B$ into the Euclidean space $\mathbb E^3$.
-The current location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
+The {\color{red}spatial} location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
 The displacement field reads $\gz u = \gz x - \gz X$.
@@
 $\d \gz x = \gz F \cdot \d \gz X$,
-where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient and $\gz I$ is the second-order unit tensor.
+where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called {\color{red}the} deformation gradient and $\gz I$ is the second-order unit tensor.
 The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian $J = \det \gz F > 0$ has to be positive.
@@
 %
-This leads to the following scalar non-linear equation
+This leads to the following scalar{\color{red}-valued} non-linear equation
 \begin{align}
@@
 \end{align}
-where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$.
+where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$
+{\color{red}is involved due to the symmetry of $\gz \tau$}.
 Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
@@
 \end{equation}
-Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order {\color{red}spatial} tangent {\color{red}stiffness} tensor is the push forward of the material tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
-The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and therefore called material part of the tangent.
-The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.
+Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order {\color{red}spatial} tangent {\color{red}stiffness} tensor is the push forward of the material {\color{red}part of the referential} tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
+The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and {\color{red}is} therefore called material part of the tangent.
+The second term in \eqref{eq:tangent} is called {\color{red}the} geometric part of the tangent since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.
 
@@
 \end{align}
-The Kirchhoff stress and its associated {\color{red}fourth-order material tangent tensor} are
+The Kirchhoff stress and its associated {\color{red}fourth-order material part of spatial tangent tensor} are
 \begin{gather}
@@
 \begin{align}
-  A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
+  \sum_{j \in \mcl I} A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
   A_{ij} \equiv a(\gz N_i, \gz N_j) &=
@@
 \end{align}
-Here $\gz A$ is the discrete tangent {\color{red}operator}, $\gz F$ is the discrete gradient
+Here $\gz A$ is the discrete tangent {\color{red}operator}, $\gz F$ is the discrete gradient {\color{red}of the potential energy}
 and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
@@
 \label{sec:mf}
-Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the force vector $\gz F$ corresponding to the discretization are assembled and
+Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the {\color{red}residual} force vector $\gz F$ corresponding to the discretization are assembled and
 a direct or iterative solver is used to solve the linear system.
@@
 \end{align}
-where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point on isoparametric element $\gz \xi_q$ and $J^K$ is the Jacobian of the mapping from the isoparametric element to the element $K$.
+where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ {\color{red}and
+$w_q$ is the quadrature weight}.
 For the sake of demonstration, let us assume that $\gz A$ represents a mass matrix with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
@@
 Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
-in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in that case that we consider a problem in $d$ space dimensions.
+in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ {\color{red}in $d$ space dimensions}.
 For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$,
@@
 
-We briefly note that the SIMD vectorization is applied at the finite element level, that is the operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
+We briefly note that the SIMD vectorization is applied at the finite element level, {\color{red}i.e.} the {\color{red}matrix-free} operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
 
@@
 %
-Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is performed over the {\color{red}spatial} configuration.
-Therefore, we have to divide by the Jacobian $J$ of the deformation map at a given quadrature point additionally to express the integral in the {\color{red}referential} configuration.
+Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is {\color{red}eventually} performed over the {\color{red}spatial} configuration.
+Therefore, we have to {\color{red}additionally} divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the {\color{red}referential} configuration.
 
@@
 Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to additions and multiplications.
-As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the operator.
+As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the {\color{red}tangent} operator.
 This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the {\color{red}referential} configuration.
@@
 This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
-Therefore, the action of the material fourth-order tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
+Therefore, the action of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
 To that end, we cache two scalars that depend on the Jacobian $J$.
@@
 
-Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent {\color{red}stiffness} tensor.
+Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor}.
 In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\overline{\grad^s} \gz x$ on-the-fly.
@@
 Efficient and scalable matrix-vector products are not enough to obtain a method suitable for large computations.
-In particular, linear scaling with respect to the number of unknown DoFs $N$ is desirable.
-The efficiency of standard iterative solvers and preconditioners deteriorate for large systems of linear equations due to increasing iteration counts under mesh refinement.
+In particular, linear scaling with respect to the number $N$ of unknown DoFs is desirable.
+The efficiency of standard iterative solvers and preconditioners deteriorates for large systems of linear equations due to increasing iteration counts under mesh refinement.
 In contrast to other preconditioners, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes.
@@
 For this work we adopt the matrix-free level operators within the GMG and employ Chebyshev pre- and post-smoothing on each level.
-In general, given a heterogeneous non-linear material, the definition of transfer operators (for restriction and prolongation) and level operators is not trivial.
+In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{{\color{red}Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.}} (for restriction and prolongation) and level operators is not trivial.
 One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
@@
 The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
-Note that, although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
+Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
 Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
@@
   \caption{Parameters for the 2D benchmark: $p$ is the polynomial degree,
-  $q$ is the quadrature order, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
+  $q$ is the {\color{red}number of quadrature points in 1D}, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
   }
@@
   \caption{Parameters for the 3D benchmark: $p$ is the polynomial degree,
-  $q$ is the quadrature order, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
+  $q$ is the {\color{red}number of quadrature points in 1D}, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
   }
@@
 The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the {\color{red}referential} configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
-There is little difference between the approach where the fourth-order material tangent {\color{red}stiffness} tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material tangent {\color{red}stiffness tensor} on the second-order symmetric tensor.
+There is little difference between the approach where the fourth-order material tangent {\color{red}stiffness} tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} on the second-order symmetric tensor.
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
@@
 In this contribution, we proposed and numerically investigated several matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
-The implementation that caches the fourth-order material tangent {\color{red}stiffness tensor} together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
-This gives hope that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material tangent {\color{red}stiffness tensor} on the quadrature point level.
+The implementation that caches the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
+This gives hope that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} on the quadrature point level.
 

### 2018-11-02 doc/large_strain.tex
unify writing about material tangent

@@
 \end{equation}
-Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order {\color{red}spatial} tangent {\color{red}stiffness} tensor is the push forward of the material {\color{red}part of the referential} tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
-The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and {\color{red}is} therefore called material part of the tangent.
-The second term in \eqref{eq:tangent} is called {\color{red}the} geometric part of the tangent since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.
+Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order {\color{red}material part of the spatial} tangent {\color{red}stiffness} tensor is the push forward of the material {\color{red}part of the referential tangent stiffness tensor} $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
+The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and {\color{red}is} therefore called material part of the {\color{red}directional derivative}.
+The second term in \eqref{eq:tangent} is called {\color{red}the} geometric part of the {\color{red}directional derivative} since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.
 
@@
   }
-  \caption{Matrix-free tangent operator: cache fourth-order material tangent {\color{red}stiffness tensor} $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
+  \caption{Matrix-free tangent operator: cache {\color{red}material part of the fourth-order spatial tangent stiffness tensor} $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
   \label{alg:mf_tensor4}
@@
 The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the {\color{red}referential} configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
-There is little difference between the approach where the fourth-order material tangent {\color{red}stiffness} tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} on the second-order symmetric tensor.
+There is little difference between the approach where the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} on the second-order symmetric tensor.
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
-Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the fourth-order material tangent {\color{red}stiffness tensor} in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
+Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
 It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws.

### 2018-11-02 Calculations/post_process.py
change xlabel: degree -> polynomial degree

@@ -197,7 +197,7 @@ plt.plot(deg2d,time2d_tr, 'rs--', label='Trilinos')
 plt.plot(deg2d,time2d_sc, 'bo--', label='MF scalar')
 plt.plot(deg2d,time2d_t2, 'g^--', label='MF tensor2')
 plt.plot(deg2d,time2d_t4, 'cv--', label='MF tensor4')
-plt.xlabel('degree')
+plt.xlabel('polynomial degree')
 plt.ylabel('vmult wall time (s) / DoF')
 leg = plt.legend(loc='best', ncol=1)
 plt.tight_layout()
@@ -213,7 +213,7 @@ plt.plot(deg3d,time3d_tr, 'rs--', label='Trilinos')
 plt.plot(deg3d,time3d_sc, 'bo--', label='MF scalar')
 plt.plot(deg3d,time3d_t2, 'g^--', label='MF tensor2')
 plt.plot(deg3d,time3d_t4, 'cv--', label='MF tensor4')
-plt.xlabel('degree')
+plt.xlabel('polynomial degree')
 plt.ylabel('vmult wall time (s) / DoF')
 leg = plt.legend(loc='best', ncol=1)
 plt.tight_layout()
@@ -228,7 +228,7 @@ plt.plot(deg2d,mem2d_tr, 'rs--', label='Trilinos')
 plt.plot(deg2d,mem2d_sc, 'bo--', label='MF scalar')
 plt.plot(deg2d,mem2d_t2, 'g^--', label='MF tensor2')
 plt.plot(deg2d,mem2d_t4, 'cv--', label='MF tensor4')
-plt.xlabel('degree')
+plt.xlabel('polynomial degree')
 plt.ylabel('memory (Mb) / DoF')
 leg = plt.legend(loc='best', ncol=1)
 plt.tight_layout()
@@ -245,7 +245,7 @@ plt.plot(deg3d,mem3d_tr, 'rs--', label='Trilinos')
 plt.plot(deg3d,mem3d_sc, 'bo--', label='MF scalar')
 plt.plot(deg3d,mem3d_t2, 'g^--', label='MF tensor2')
 plt.plot(deg3d,mem3d_t4, 'cv--', label='MF tensor4')
-plt.xlabel('degree')
+plt.xlabel('polynomial degree')
 plt.ylabel('memory (Mb) / DoF')
 leg = plt.legend(loc='best', ncol=1)
 plt.tight_layout()
[…]

### 2018-11-02 doc/large_strain.tex
fix few more comments

@@
 }} (stemming from linearization of the nonlinear balance equations) are
-more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether the matrix-free methods can be advantageous in this case.
-Another reason could be the use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
-The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community.
-Even higher order elements are rarely employed.
+more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
+Another reason could be the {\color{red}frequent} use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
+The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
+{\color{red}whereas} higher order elements are rarely employed.
 {\color{red}
@@
 \end{gather}
-where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor.
+where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor
+{\color{red}with $S_{ijkl}=\frac{1}{2}\left[\delta_{ik}\delta_{jl}+\delta_{il}\delta_{jk}\right]$}.
 The action that $J \boldsymbol{\mathcal{C}}$ performs when contracted with an arbitrary rank-2 symmetric tensor is therefore

### 2018-11-02 doc/large_strain.tex
clarify benchmarks

@@
 
-Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication benchmarks as performed on the two clusters.
+Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication {\color{red}for the considered finite-strain hyperelastic benchmark problem} as performed on the two clusters.
 As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches.

### 2018-11-02 doc/large_strain.tex
minor changes related to matrices and operators

@@
 %
-The matrix-vector product $\gz A \gz x$ can be expressed by
+The matrix-vector product $\gz A \gz x$ {\color{red}(i.e. the action of operator $\gz A$ on a vector $\gz x$)} can be expressed by
 \begin{align}
@@
 $w_q$ is the quadrature weight}.
-For the sake of demonstration, let us assume that $\gz A$ represents a mass matrix with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
+For the sake of demonstration, let us assume that $\gz A$ represents a mass {\color{red}operator} with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
 \begin{align*}

### 2018-11-05 doc/large_strain.tex
minor fixes to fit IJNME template
Beautify footnote

@@
 \end{equation}
-Following \cite{Wriggers2008}, this can be simplified to\footnote{To that end
-$\D_{\displaystyle \Delta \gz u} \gz F = \Grad \Delta \gz u$,
-$\;\D_{\displaystyle \Delta \gz u} \gz F^{-1} = - \gz F^{-1} \cdot \D_{\displaystyle \Delta \gz u} \gz F \cdot \gz F^{-1}$ and
-$\D_{\displaystyle \Delta \gz u} \gz E = \frac{1}{2}\left[\D_{\displaystyle \Delta \gz u} \gz F^T \cdot \gz F + \gz F^T \cdot \D_{\displaystyle \Delta \gz u} \gz F\right]$
+Following \cite{Wriggers2008}, this can be simplified to\footnote{To that end,
+$\!\D_{\displaystyle \Delta \gz u} \gz E = \frac{1}{2}\left[\D_{\displaystyle \Delta \gz u} \gz F^T \cdot \gz F + \gz F^T \cdot \D_{\displaystyle \Delta \gz u} \gz F\right]$, $\D_{\displaystyle \Delta \gz u} \gz F = \Grad \Delta \gz u\;$  and
+$\!\D_{\displaystyle \Delta \gz u} \gz F^{-1} = - \gz F^{-1} \cdot \D_{\displaystyle \Delta \gz u} \gz F \cdot \gz F^{-1}$
 are employed together with
@@
 
-\section*{Acknowledgements}
+\acks
+% \section*{Acknowledgements}
 
@@
 
-\section*{Bibliography}
+% \section*{Bibliography}
 
-\bibliographystyle{elsarticle-num}
+\bibliographystyle{wileyj}
+% \bibliographystyle{elsarticle-num}
+
 \bibliography{bibliography}

### 2018-11-05 doc/large_strain.tex
Fix some more typos

@@
 One possible reason is that the tangent operators\footnote{{\color{red}Here and
-below by operators we mean finite-dimensional linear operators on vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
+below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
 Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
@@
 {\color{red}
-This can potentially reduce advantage of matrix-free methods, that are generally more competitive for higher order elements.
+This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements.
 }
@@
 \end{align}
-The Kirchhoff stress and its associated {\color{red}fourth-order material part of spatial tangent tensor} are
+The Kirchhoff stress and its associated {\color{red}fourth-order material part of the spatial tangent tensor} are
 \begin{gather}
@@
 \end{align}
-Here $\gz A$ is the discrete tangent {\color{red}operator}, $\gz F$ is the discrete gradient {\color{red}of the potential energy}
+Here, $\gz A$ is the discrete tangent {\color{red}operator}, $\gz F$ is the discrete gradient {\color{red}of the potential energy}
 and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
@@
 {\color{red}
-See Figure \ref{fig:multi_index} for illustration.
+See Figure \ref{fig:multi_index} for an illustration.
 }
@@
 {\color{red}
-Note that $x_j \equiv X_{j_1\,j_2}$; the former access the element source vector using a linear index $j$, whereas the latter uses the associated multi-index accesses $\{j_1\,j_2\}$.
+Note that $x_j \equiv X_{j_1\,j_2}$; the former accesses the element source vector using a linear index $j$, whereas the latter uses the associated multi-index accesses $\{j_1\,j_2\}$.
 }

### 2018-11-08 doc/large_strain.tex
one more citation

@@
 The multigrid algorithm is a V-cycle with two pre- and post-smoothing steps.
-For other approaches that combine homogenization and multigrid methods for small strain elastic materials, see \cite{Bayreuther2003,Fish1995}.
+For other approaches that combine homogenization and multigrid methods for small strain elastic materials {\color{red}and fracture}, see \cite{Bayreuther2003,Fish1995} {\color{red}and \cite{Kaczmarczyk2010}, respectively}.
 

### 2018-11-08 doc/large_strain.tex
add switch between IJNME and CMAME, enable CMAME for now

@@
 \RequirePackage[l2tabu, orthodox]{nag}
-%\documentclass[12pt]{article}
-% \documentclass[preprint,12pt,times]{elsarticle}
+
+\newif\ifijnme %Declaration of ijnme
+% \ijnmetrue %Sets ijnme to true
+\ijnmefalse %Sets ijnme to false
+
+\ifijnme
 \documentclass[times,doublespace]{nmeauth}
-
-%%% ELSOPTIONS:
-% \biboptions{sort&compress,comma,square}
-% \usepackage{amsmath}
-
-%%%
+\else
+\documentclass[preprint,12pt,times]{elsarticle}
+\biboptions{sort&compress,comma,square}
+\usepackage{amsmath}
+\fi
 
@@
 
+\ifijnme
 \runningheads{D. Davydov et al.}{A matrix-free approach for finite-strain hyperelastic problems}
+\else
+\journal{CMAME}
+\begin{frontmatter}
+\fi
 
@@
 
+\ifijnme
   \author{
@@
 
-  % \author[a]{Denis Davydov\corref{cor}}
-  % \ead{denis.davydov@fau.de}
+\else
+  \author[a]{Denis Davydov\corref{cor}}
+  \ead{denis.davydov@fau.de}
 
-  % \author[a]{Jean-Paul Pelteret}
-  % \ead{jean-paul.pelteret@fau.de}
+  \author[a]{Jean-Paul Pelteret}
+  \ead{jean-paul.pelteret@fau.de}
 
-  % \author[b]{Daniel Arndt}
-  % \ead{daniel.arndt@iwr.uni-heidelberg.de}
+  \author[b]{Daniel Arndt}
+  \ead{daniel.arndt@iwr.uni-heidelberg.de}
 
-  % \author[a,c]{Paul Steinmann}
-  % \ead{paul.steinmann@fau.de}
+  \author[a,c]{Paul Steinmann}
+  \ead{paul.steinmann@fau.de}
 
-  % \cortext[cor]{Corresponding author.}
+  \cortext[cor]{Corresponding author.}
 
-  % \address[a]{Chair of Applied Mechanics,
-  % Friedrich-Alexander-Universit\"{a}t Erlangen-N\"{u}rnberg,
-  % Egerlandstr.\ 5, 91058 Erlangen, Germany}
+  \address[a]{Chair of Applied Mechanics,
+  Friedrich-Alexander-Universit\"{a}t Erlangen-N\"{u}rnberg,
+  Egerlandstr.\ 5, 91058 Erlangen, Germany}
 
-  % \address[b]{Interdisciplinary Center for Scientific Computing (IWR),
-  %     Heidelberg University,
-  %     Im Neuenheimer Feld 205,
-  %     69120 Heidelberg,
-  %     Germany}
+  \address[b]{Interdisciplinary Center for Scientific Computing (IWR),
+      Heidelberg University,
+      Im Neuenheimer Feld 205,
+      69120 Heidelberg,
+      Germany}
 
-  % \address[c]{Glasgow Computational Engineering Center (GCEC),
-  %     University of Glasgow, G12 8QQ Glasgow, United Kingdom
-  % }
+  \address[c]{Glasgow Computational Engineering Center (GCEC),
+      University of Glasgow, G12 8QQ Glasgow, United Kingdom
+  }
+\fi
 
@@
 
+\ifijnme
+
   \keywords{
-  % \begin{keyword}
-      adaptive finite element method ; %\sep
-      geometric multigrid ; %\sep
-      finite-strain ; %\sep
-      matrix-free ; %\sep
+      adaptive finite element method ;
+      geometric multigrid ;
+      finite-strain ;
+      matrix-free ;
       hyperelasticity
-  % \end{keyword}
   }
@@
   \maketitle
-  % \end{frontmatter}
+
+\else
+    \begin{keyword}
+        adaptive finite element method \sep
+        geometric multigrid \sep
+        finite-strain \sep
+        matrix-free \sep
+        hyperelasticity
+    \end{keyword}
+
+\end{frontmatter}
+\fi
 
@@
 
+\ifijnme
 \acks
-% \section*{Acknowledgements}
+\else
+\section*{Acknowledgements}
+\fi
 
@@
 
-% \section*{Bibliography}
-
+\ifijnme
 \bibliographystyle{wileyj}
-% \bibliographystyle{elsarticle-num}
+\else
+\section*{Bibliography}
+\bibliographystyle{elsarticle-num}
+\fi
 

### 2018-11-09 doc/large_strain.tex
fix Lukasz's citation

@@
 The multigrid algorithm is a V-cycle with two pre- and post-smoothing steps.
-For other approaches that combine homogenization and multigrid methods for small strain elastic materials {\color{red}and fracture}, see \cite{Bayreuther2003,Fish1995} {\color{red}and \cite{Kaczmarczyk2010}, respectively}.
+{\color{red}For examples of other approaches that combine homogenization and multigrid methods, see \cite{Bayreuther2003,Fish1995} for applications using small strain elastic materials and \cite{Kaczmarczyk2010} with applications in fracture.}
 

### 2018-11-12 doc/large_strain.tex
Fix more of PS's comments.

@@
 
-For the current study, we use the Neo-Hookean model
+For the current study, we use the compressible Neo-Hookean model
 \begin{gather}
@@
 \label{sec:mf}
-Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the {\color{red}residual} force vector $\gz F$ corresponding to the discretization are assembled and
-a direct or iterative solver is used to solve the linear system.
+Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the {\color{red}residual} force vector $\gz F$ corresponding to the discretization are assembled (\textcolor{red}{that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored}) and a direct or iterative solver is used to solve the linear system.
+\textcolor{red}{In this case the the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.}
 On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
-For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations within the solver on-the-fly
-rather than loading matrix elements from memory.
+For this reason, recent implementations often focus on so-called matrix-free approaches \textcolor{red}{where the matrix is not assembled but rather the result of its operation on a vector is directly calculated}.
+The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
 For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.

### 2018-11-16 doc/large_strain.tex
add article and full stop

@@
 We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
-Essentially, level operators correspond to the linearization around smoothed representation of displacement field\footnote{
+Essentially, level operators correspond to the linearization around the smoothed representation of displacement field\footnote{
 Note that this differs from the classical approach where the (tangent) operator $\gz A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $\gz A^{l}$ on level $l$ via $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, where $\gz I^l_{l+1}$ and $\gz I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
-}
+}.
 Finally, at the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.

### 2018-11-16 doc/large_strain.tex

@@
 
-    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyperelastic {\color{red}finite element} tangent operator numerically.
+    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyperelastic finite element tangent operator numerically.
     In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
@@
 We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent masters thesis produced by a student of the first two of the authors \cite{Mentler2017}.
-One possible reason is that the tangent operators\footnote{{\color{red}Here and
+One possible reason is that the tangent operators\footnote{Here and
 below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
 Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
-}} (stemming from linearization of the nonlinear balance equations) are
+} (stemming from linearization of the nonlinear balance equations) are
 more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
-Another reason could be the {\color{red}frequent} use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
+Another reason could be the frequent use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
 The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
-{\color{red}whereas} higher order elements are rarely employed.
-{\color{red}
+whereas higher order elements are rarely employed.
 This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements.
-}
 However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
@@
 %
-The deformation of a body $\mcl B$ from the {\color{red}referential} configuration $\mcl B_0$ to the {\color{red}spatial} configuration $\mcl B_t$ at time $t$
+The deformation of a body $\mcl B$ from the referential configuration $\mcl B_0$ to the spatial configuration $\mcl B_t$ at time $t$
 is defined via the deformation map $\gz \varphi_t: \mcl B_0 \rightarrow \mcl B_t$, which places material points of $\mcl B$ into the Euclidean space $\mathbb E^3$.
-The {\color{red}spatial} location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
+The spatial location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
 The displacement field reads $\gz u = \gz x - \gz X$.
@@
 $\d \gz x = \gz F \cdot \d \gz X$,
-where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called {\color{red}the} deformation gradient and $\gz I$ is the second-order unit tensor.
+where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called the deformation gradient and $\gz I$ is the second-order unit tensor.
 The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian $J = \det \gz F > 0$ has to be positive.
@@
 \end{align}
-where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the {\color{red}referential} configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
+where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the referential configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
 
@@
 %
-This leads to the following scalar{\color{red}-valued} non-linear equation
+This leads to the following scalar-valued non-linear equation
 \begin{align}
@@
 where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$
-{\color{red}is involved due to the symmetry of $\gz \tau$}.
+is involved due to the symmetry of $\gz \tau$.
 Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
@@
 \end{equation}
-Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order {\color{red}material part of the spatial} tangent {\color{red}stiffness} tensor is the push forward of the material {\color{red}part of the referential tangent stiffness tensor} $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
-The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and {\color{red}is} therefore called material part of the {\color{red}directional derivative}.
-The second term in \eqref{eq:tangent} is called {\color{red}the} geometric part of the {\color{red}directional derivative} since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.
+Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order material part of the spatial tangent stiffness tensor is the push forward of the material part of the referential tangent stiffness tensor $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
+The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and is therefore called material part of the directional derivative.
+The second term in \eqref{eq:tangent} is called the geometric part of the directional derivative since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.
 
@@
 \end{align}
-The Kirchhoff stress and its associated {\color{red}fourth-order material part of the spatial tangent tensor} are
+The Kirchhoff stress and its associated fourth-order material part of the spatial tangent tensor are
 \begin{gather}
@@
 where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor
-{\color{red}with $S_{ijkl}=\frac{1}{2}\left[\delta_{ik}\delta_{jl}+\delta_{il}\delta_{jk}\right]$}.
+with $S_{ijkl}=\frac{1}{2}\left[\delta_{ik}\delta_{jl}+\delta_{il}\delta_{jk}\right]$.
 The action that $J \boldsymbol{\mathcal{C}}$ performs when contracted with an arbitrary rank-2 symmetric tensor is therefore
@@
 where the superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I$ is the set of unknown degrees of freedom (DoF).
-%{\color{red} For the sake of this study, $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.}
 
@@
 \end{align}
-Here, $\gz A$ is the discrete tangent {\color{red}operator}, $\gz F$ is the discrete gradient {\color{red}of the potential energy}
+Here, $\gz A$ is the discrete tangent operator, $\gz F$ is the discrete gradient of the potential energy
 and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
-Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the {\color{red}spatial} configuration, whereas the integration is done in the {\color{red}referential} configuration.
+Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the spatial configuration, whereas the integration is done in the referential configuration.
 
@@
 \label{sec:mf}
-Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the {\color{red}residual} force vector $\gz F$ corresponding to the discretization are assembled (\textcolor{red}{that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored}) and a direct or iterative solver is used to solve the linear system.
-\textcolor{red}{In this case the the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.}
+Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
+In this case the the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
 On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
-For this reason, recent implementations often focus on so-called matrix-free approaches \textcolor{red}{where the matrix is not assembled but rather the result of its operation on a vector is directly calculated}.
+For this reason, recent implementations often focus on so-called matrix-free approaches where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
 The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
@@
 %
-The matrix-vector product $\gz A \gz x$ {\color{red}(i.e. the action of operator $\gz A$ on a vector $\gz x$)} can be expressed by
+The matrix-vector product $\gz A \gz x$ (i.e. the action of operator $\gz A$ on a vector $\gz x$) can be expressed by
 \begin{align}
@@
 \end{align}
-where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ {\color{red}and
-$w_q$ is the quadrature weight}.
-For the sake of demonstration, let us assume that $\gz A$ represents a mass {\color{red}operator} with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
+where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
+$w_q$ is the quadrature weight.
+For the sake of demonstration, let us assume that $\gz A$ represents a mass operator with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
 \begin{align*}
@@
 Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
-in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ {\color{red}in $d$ space dimensions}.
+in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
 For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$,
@@
 where for each DoF index $i$ we can associate a multi-index $(i_1, i_2)$.
-{\color{red}
 See Figure \ref{fig:multi_index} for an illustration.
-}
 %
@@
 %
-{\color{red}
 Note that $x_j \equiv X_{j_1\,j_2}$; the former accesses the element source vector using a linear index $j$, whereas the latter uses the associated multi-index accesses $\{j_1\,j_2\}$.
-}
 %
@@
 
-We briefly note that the SIMD vectorization is applied at the finite element level, {\color{red}i.e.} the {\color{red}matrix-free} operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
+We briefly note that the SIMD vectorization is applied at the finite element level, i.e.
+the matrix-free operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
 
 Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
-Below, we propose three algorithms to implement the tangent {\color{red}operator} of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
+Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
 %
-In order to obtain gradients with respect to the {\color{red}spatial} configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
-In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the {\color{red}spatial} configuration given a displacement field.
+In order to obtain gradients with respect to the spatial configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
+In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
 %
-Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is {\color{red}eventually} performed over the {\color{red}spatial} configuration.
-Therefore, we have to {\color{red}additionally} divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the {\color{red}referential} configuration.
+Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the spatial configuration.
+Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.
 
@@
   cached $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
-  \Output{action of the FE tangent {\color{red}operator} \eqref{eq:algebraic_tangent} on $\gz x$}
+  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
   \ForEach{ element $K \in \Omega^h$ }{
@@
   }
-  \caption{Matrix-free tangent operator: {\color{red}cache scalar quantities only}}
+  \caption{Matrix-free tangent operator: cache scalar quantities only}
   \label{alg:mf_scalar}
@@
 Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to additions and multiplications.
-As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the {\color{red}tangent} operator.
-This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the {\color{red}referential} configuration.
+As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the tangent operator.
+This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the referential configuration.
 
@@
   evaluated based on $\overline{\gz u^h}$}
-  \Output{action of the FE tangent {\color{red}operator} \eqref{eq:algebraic_tangent} on $\gz x$}
+  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
   \ForEach{ element $K \in \Omega^h$ }{
@@
 In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point.
-In this case, we also avoid the need to evaluate gradients of displacement field with respect to the {\color{red}referential} configuration.
+In this case, we also avoid the need to evaluate gradients of displacement field with respect to the referential configuration.
 This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
-Therefore, the action of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
+Therefore, the action of the material part of the fourth-order spatial tangent stiffness tensor $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
 To that end, we cache two scalars that depend on the Jacobian $J$.
@@
   }
-  \caption{Matrix-free tangent operator: cache {\color{red}material part of the fourth-order spatial tangent stiffness tensor} $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
+  \caption{Matrix-free tangent operator: cache material part of the fourth-order spatial tangent stiffness tensor $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
   \label{alg:mf_tensor4}
@@
 
-Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor}.
+Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material part of the fourth-order spatial tangent stiffness tensor.
 In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\overline{\grad^s} \gz x$ on-the-fly.
@@
 For this work we adopt the matrix-free level operators within the GMG and employ Chebyshev pre- and post-smoothing on each level.
-In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{{\color{red}Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.}} (for restriction and prolongation) and level operators is not trivial.
+In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial.
 One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
-  We note that this approach is limited to the case where for each quadrature point within an element the fourth-order tangent {\color{red}stiffness tensor} is the same.
+  We note that this approach is limited to the case where for each quadrature point within an element the fourth-order tangent stiffness tensor is the same.
 }.
@@
 In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
-We define the level operators as follows: Recall that the fine scale tangent {\color{red}operator} \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$.
+We define the level operators as follows: Recall that the fine scale tangent operator \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$.
 We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
@@
 The multigrid algorithm is a V-cycle with two pre- and post-smoothing steps.
-{\color{red}For examples of other approaches that combine homogenization and multigrid methods, see \cite{Bayreuther2003,Fish1995} for applications using small strain elastic materials and \cite{Kaczmarczyk2010} with applications in fracture.}
+For examples of other approaches that combine homogenization and multigrid methods, see \cite{Bayreuther2003,Fish1995} for applications using small strain elastic materials and \cite{Kaczmarczyk2010} with applications in fracture.
 
@@
   \caption{Parameters for the 2D benchmark: $p$ is the polynomial degree,
-  $q$ is the {\color{red}number of quadrature points in 1D}, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
+  $q$ is the number of quadrature points in 1D, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
   }
@@
   \caption{Parameters for the 3D benchmark: $p$ is the polynomial degree,
-  $q$ is the {\color{red}number of quadrature points in 1D}, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
+  $q$ is the number of quadrature points in 1D, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
   }
@@
 
-Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication {\color{red}for the considered finite-strain hyperelastic benchmark problem} as performed on the two clusters.
+Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
 As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches.
@@
 Two conclusions can be drawn from these results:
-The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the {\color{red}referential} configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
-There is little difference between the approach where the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} on the second-order symmetric tensor.
+The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
+There is little difference between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor.
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
-Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
+Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
 It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws.
@@
 In this contribution, we proposed and numerically investigated several matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
-The implementation that caches the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
-This gives hope that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} on the quadrature point level.
+The implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
+This gives hope that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
 

### 2019-01-08 doc/large_strain.tex
remove constants from operation counts

@@
 in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
-For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$,
+For the total matrix-vector product we get $\mathcal{O}(n^{2d})$,
 and this is the same as for a regular matrix-vector product where the entries are already precomputed.
@@
 We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
-Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, this results in an algorithm with an arithmetic complexity of $\mathcal{O}(2dn^{d+1})$.
+Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, this results in an algorithm with an arithmetic complexity of $\mathcal{O}(n^{d+1})$.
 Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for the 3D case, and where the discretization and evaluation employs high polynomial degree basis functions with the corresponding quadrature rule.

### 2019-01-08 doc/large_strain.tex
address minor point 4 of Reviewer 2

@@
 Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
-Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
 %
-In order to obtain gradients with respect to the spatial configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
+In order to obtain gradients with respect to the spatial configuration,
+{\color{red}
+that appear in the tangent operator \eqref{eq:algebraic_tangent} of finite-strain elasticity,
+}
+we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
 In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
@@
 Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.
+
+{\color{red}
+Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
+}
 
@@
 The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
-Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies.
+Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
+{\color{red}
+described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.
+}
 Two conclusions can be drawn from these results:

### 2019-01-08 doc/large_strain.tex
Rev2 minor point 6

@@
 In this contribution, we proposed and numerically investigated several matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
-The implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic in 3D.
+The implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements} in 3D.
 This gives hope that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.

### 2019-01-08 doc/large_strain.tex
clarify what's noverl in GMG section and what's used from deal.II

@@
 The multigrid algorithm is a V-cycle with two pre- and post-smoothing steps.
+
+{\color{red}
+To summarize, in this section we propose a method by which to construct level tangent operators for finite-strain elasticity to be employed within the GMG preconditioner.
+The other parts of the GMG preconditioner such as Chebyshev smoother, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
+We note that at the time of writing, $p-$ version of the GMG is not available within the \texttt{deal.II} library.
+}
+
 For examples of other approaches that combine homogenization and multigrid methods, see \cite{Bayreuther2003,Fish1995} for applications using small strain elastic materials and \cite{Kaczmarczyk2010} with applications in fracture.

### 2019-01-08 doc/large_strain.tex
remove TBB, expand on how SIMD is done

@@
     and employ them to define a geometric multigrid preconditioner.
-    Our implementation employs MPI and Intel Threading Building Blocks parallelization, and SIMD vectorization.
     The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
@@
 and employ them to define a \mbox{geometric} multigrid preconditioner.
-Our implementation employs MPI and Intel Threading Building Blocks parallelization, together with SIMD vectorization.
+MPI parallelization together with SIMD vectorization
+{\color{red}
+are provided by the \texttt{deal.II} \cite{dealII90} finite element library, used for this study.
+}
 
@@
 
-We briefly note that the SIMD vectorization is applied at the finite element level, i.e.
-the matrix-free operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.
-
 Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
@@
 
+Note that the SIMD vectorization
+{\color{red}
+in \texttt{deal.II} \cite{dealII90}
+}
+is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
+{\color{red}
+For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements.
+This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
+\textit{FEEvaluation} class already provides all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
+Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
+For more information about the implementation details and performance analysis (????), see \cite{kronbichler12}.
+}
+
 \section{Geometric multigrid preconditioning}
@@
 The relative convergence criteria for the linear solver is $10^{-6}$.
-Shared memory parallelization using Intel Threading Building Blocks was disabled for this study.
 

### 2019-01-08 doc/large_strain.tex
mention MPI details and add more info on SIMD

@@
 For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements.
+For example, deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
 This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
-\textit{FEEvaluation} class already provides all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
+In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type.
+The core of the sum factorization algorithms is provided by
+\textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
 Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
-For more information about the implementation details and performance analysis (????), see \cite{kronbichler12}.
+Distributed memory MPI parallelization of matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so called locally owned elements.
+For more information about the implementation details of sum factorization techniques in deal.II and their performance analysis (????), see \cite{kronbichler12}.
 }

### 2019-01-09 doc/large_strain.tex
mention that we won't study GMG in details (not the focus)

@@
 The other parts of the GMG preconditioner such as Chebyshev smoother, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
-We note that at the time of writing, $p-$ version of the GMG is not available within the \texttt{deal.II} library.
+The efficiency of the preconditioner, from the perspective of linear algebra, with the so defined level operators will be assessed by numerical examples in the next section.
+We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
+Finally, we note that at the time of writing, $p-$ version of the GMG is not available within the \texttt{deal.II} library.
 }

### 2019-01-09 doc/large_strain.tex
more details on MPI

@@
 The relative convergence criteria for the linear solver is $10^{-6}$.
-
-\begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.8\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{gmg_2d.png}
-    % \caption{2D multigrid mesh with 2 global refinements}
-  \end{subfigure}
-  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
-  \label{fig:miehe_gmg}
-\end{figure}
-%
-Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.
 Figure \ref{fig:miehe_deformed} shows deformed meshes at the final loading step with quadratic elements and two global mesh refinements.
@@
 \end{figure}
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.8\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{fine_level_2d.png}
+    % \caption{2D multigrid mesh with 2 global refinements}
+  \end{subfigure}
+  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
+  \label{fig:miehe_fine_level}
+\end{figure}
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.8\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{gmg_2d.png}
+    % \caption{2D multigrid mesh with 2 global refinements}
+  \end{subfigure}
+  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
+  \label{fig:miehe_gmg}
+\end{figure}
+%
+%
+{\color{red}
+MPI parallelization is achieved using the standard domain decomposition, where each MPI process ``owns'' only a subset of elements.
+To that end, \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library is used.
+Figure \ref{fig:miehe_fine_level} illustrates such decomposition on the fine mesh level for the 2D problem.
+Same approach is adopted at the multigrid levels.}
+Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.
+{\color{red}
+Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.
+}
 

### 2019-01-09 doc/large_strain.tex
more comments on GMG results

@@
 Next we evaluate the performance of the proposed geometric multigrid preconditioner.
+{\color{red}
+Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective.
+}
 First we consider the number of iterations in the conjugate gradient (CG) solver.
@@
 On the other hand, the black-box AMG preconditioner requires many more iterations for convergence.
+{\color{red}
+Consequently we can concluded that, although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still results in a well behaving multigrid preconditioner for the here considered finite strain elasticity problem.
+}
+
+{\color{red}
+Additionally, we also compare solution times of the preconditioned iterative solvers.
+}
 Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for polynomial degrees up to $p=5$.
@@
 Evidently, the combination of the Chebyshev smoother with the CG solver at the coarsest mesh level has unsatisfactory performance for higher polynomial degrees.
+{\color{red}
+Consequently, the coarse level solver is the bottleneck of the adopted here geometric multigrid preconditioner.
+Its performance from the perspective of linear algebra shall be improved first and only then numerical implementational details of the complete preconditioner shall be studied.
+}
 One possible solution to this issues would be to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.

### 2019-01-09 doc/large_strain.tex
better stress novelty in conclusions

@@
 
-In this contribution, we proposed and numerically investigated several matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
-The implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements} in 3D.
-This gives hope that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
-
-We also studied the performance of the GMG preconditioner with standard geometric transfer operations between each level.
+In this contribution, we proposed and numerically investigated {\color{red}three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
+{\color{red}
+To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to such materials.}
+The implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
+{\color{red}
+Given that in this case we did not utilize the specific form of the material part of the stiffness tensor, we conclude}
+that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
+
+{\color{red}
+We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner
+}
+with standard geometric transfer operations between each level.
 The GMG was applied to heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
-The numerical studies indicate that the proposed preconditioner leads to a solution approach that is faster than the matrix-based AMG for moderately high polynomial degrees (approximately up to order five).
-However, a considerable amount of time of GMG preconditioner is spent in the coarse level solver.
+The numerical studies indicate that the proposed preconditioner is efficient from the perspective of linear algebra.
+{\color{red}
+Namely, the number of iterations in the iterative solver is almost independent on the polynomial degree of the finite element basis.
+The multigrid matrix-free preconditioner also}
+leads to a solution approach that is faster than the matrix-based AMG for moderately high polynomial degrees (approximately up to order five).
+However, a considerable amount of time of GMG preconditioner is spent in the coarse level solver,
+{\color{red}
+which is the main bottleneck of the here considered iterative solver.}
 Our future work will, therefore, be focused on extending the proposed matrix-free multigrid solution approach to homogenization based transfer operators \cite{Miehe2007}.

### 2019-01-09 doc/large_strain.tex
add details about Chebyshev smoother, move that to Numerical examples part

@@
 
-For this work we adopt the matrix-free level operators within the GMG and employ Chebyshev pre- and post-smoothing on each level.
+For this work we adopt the matrix-free level operators within the GMG.
 In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial.
@@
 }.
-Finally, at the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
-The multigrid algorithm is a V-cycle with two pre- and post-smoothing steps.
 
@@
 To summarize, in this section we propose a method by which to construct level tangent operators for finite-strain elasticity to be employed within the GMG preconditioner.
-The other parts of the GMG preconditioner such as Chebyshev smoother, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
+The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
 The efficiency of the preconditioner, from the perspective of linear algebra, with the so defined level operators will be assessed by numerical examples in the next section.
@@
 The relative convergence criteria for the linear solver is $10^{-6}$.
+{\color{red}
+The geometric multigrid algorithm is a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
+by the \texttt{deal.II} library via \textit{PreconditionChebyshev} class.
+The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
+is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
+At the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
+}
+
 Figure \ref{fig:miehe_deformed} shows deformed meshes at the final loading step with quadratic elements and two global mesh refinements.

### 2019-01-09 include/mf_elasticity.h
add LIKWID markers around vmult's

@@ -80,10 +80,15 @@ static const unsigned int debug_level = 0;
 #include <deal.II/physics/elasticity/kinematics.h>
 #include <deal.II/physics/elasticity/standard_tensors.h>
 
+#include <config.h>
 #include <material.h>
 #include <mf_ad_nh_operator.h>
 #include <mf_nh_operator.h>
 #include <sys/stat.h>
+
+#ifdef WITH_LIKWID
+#include <likwid.h>
+#endif
 
 #include <fstream>
 #include <iostream>
@@ -1089,6 +1094,9 @@ namespace Cook_Membrane
   void
   Solid<dim, degree, n_q_points_1d, NumberType>::run()
   {
+#ifdef WITH_LIKWID
+    LIKWID_MARKER_INIT;
+#endif
     make_grid();
     system_setup();
     output_results();
@@ -1121,6 +1129,10 @@ namespace Cook_Membrane
 
     // for post-processing, print average CG iterations over the whole run:
     timer_out << std::endl << "Average CG iter = " << (total_n_cg_iterations/total_n_cg_solve) << std::endl;
+#ifdef WITH_LIKWID
+    LIKWID_MARKER_CLOSE;
+#endif
+
   }
 
 
@@ -2041,14 +2053,26 @@ namespace Cook_Membrane
           for (unsigned int i = 0; i < n_times; ++i)
             {
               TimerOutput::Scope t(timer, "vmult (Trilinos)");
+#ifdef WITH_LIKWID
+              LIKWID_MARKER_START("vmult_Trilinos");
+#endif
               tangent_matrix.vmult(dst_mb, src_trilinos);
+#ifdef WITH_LIKWID
+              LIKWID_MARKER_STOP("vmult_Trilinos");
+#endif
             }
 
           MPI_Barrier(mpi_communicator);
           for (unsigned int i = 0; i < n_times; ++i)
             {
               TimerOutput::Scope t(timer, "vmult (MF)");
+#ifdef WITH_LIKWID
+              LIKWID_MARKER_START("vmult_MF");
+#endif
               mf_nh_operator.vmult(dst_mf, src);
+#ifdef WITH_LIKWID
+              LIKWID_MARKER_STOP("vmult_MF");
+#endif
             }
 
 #ifdef DEBUG

### 2019-01-09 doc/large_strain.tex
mention where the Roofline would go
more details on LIKWID

@@
 
+% likwid-mpirun -np 10 -nperdomain S:10 -g MEM_DP -m $EXC
+{\color{red}
+Finally, we analyze the node-level performance of the two approaches using the Roofline model.
+Memory bandwidth (MBytes/s) and the number of floating point operations per second (MFLOP/s) are measured using the \texttt{MEM\_DP} group of the LIKWID \cite{likwid} tool, version 4.2.1.
+The measurements are done on a single socket of RRZE cluster by pinning all 10 MPI processes to this socket.
+[DD]TODO: plot and discuss.
+}
+
 \subsection{Preconditioned iterative solver}

### 2019-01-09 Calculations/likwid_emmy_benchmark_intel.sh
adjust python script to prepare LIKWID runs

@@ -1,0 +1,14 @@
+#!/bin/bash
+#   allocate 1 nodes (with likwid) with 40 CPU per node for 2 hours:
+#PBS -l nodes=1:ppn=40:likwid,walltime=2:00:00
+#   job name
+#PBS -N dealii_mf_cm_likwid
+#   stdout and stderr files:
+#PBS -o /home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/_build/likwid_bench.txt -e /home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/_build/likwid_bench_error.txt
+#   first non-empty non-comment line ends PBS options
+
+# submit with: qsub <name>.sh
+cd /home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/Calculations/
+module load intel64/18.0up03 likwid/4.2.1
+
+source likwid_run.sh

### 2019-01-09 Calculations/pre_process.py

@@ -7,16 +7,17 @@ import argparse
 parser = argparse.ArgumentParser(
     description='Prepare input files.',
     formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-parser.add_argument('base_prm', metavar='base_prm', default='holes.prm', nargs='?',
+parser.add_argument('--base_prm', metavar='base_prm', default='holes.prm',
                     help='Base parameter file to use')
-parser.add_argument('dir', metavar='dir', default='Emmy_RRZE', nargs='?',
+parser.add_argument('--dir', metavar='dir', default='Emmy_RRZE',
                     help='Subdirectory to store calculations')
-parser.add_argument('prefix', metavar='prefix', default='/home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/_build/', nargs='?',
+parser.add_argument('--prefix', metavar='prefix', default='/home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/_build/',
                     help='Build directory with executable `main`')
-parser.add_argument('calc', metavar='calc', default='/home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/Calculations/', nargs='?',
+parser.add_argument('--calc', metavar='calc', default='/home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/Calculations/',
                     help='Directory with calculations where .prm files will be generated and `base_prm` is located')
-parser.add_argument('mpirun', metavar='mpirun', default='mpirun -np 20', nargs='?',
+parser.add_argument('--mpirun', metavar='mpirun', default='mpirun -np 20',
                     help='mpi run command with cores')
+parser.add_argument('--likwid', help='Prepare LIKWID run', action="store_true")
 args = parser.parse_args()
 
 # parameters (list of tuples):
@@ -37,6 +38,12 @@ poly_quad_ref_dim = [
     (4,5,2,3),
 ]
 
+poly_quad_ref_dim_likwid = [
+    (2,3,6,2),
+    (4,5,5,2),
+    (6,7,4,2),
+]
+
 # Solvers (type, preconditioner and caching)
 solvers = [
     ('MF_CG', 'gmg', 'scalar'),
@@ -45,9 +52,22 @@ solvers = [
     ('CG',    'amg', 'scalar'),
 ]
 
[…]

### 2019-01-09 Calculations/pre_process.py
store likwid runs in a separate directory

@@ -135,6 +135,7 @@ if out_dir and not out_dir.endswith('/'):
 # add prefix to the base name
 if args.likwid:
   base_name = 'likwid_' + base_name
+  out_dir = 'LIKWID_' + out_dir
 
 print 'base parameter file: {0}'.format(base_prm)
 print 'output directory:    {0}'.format(out_dir)

### 2019-01-10 doc/large_strain.tex
node data on memory bandwidth and performance
minor improvements throughout (i.e. mention node-level everywhere)

@@
     and employ them to define a geometric multigrid preconditioner.
-    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
+    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner {\color{red} on a single node of an higher performance computing cluster} for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
   \end{abstract}
@@
 To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
-We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent masters thesis produced by a student of the first two of the authors \cite{Mentler2017}.
+We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent masters thesis produced by a student of the first two of the authors \cite{Mentler2017},
+{\color{red} which is also limited to the small strain linear elasticity.}
 One possible reason is that the tangent operators\footnote{Here and
@@
 
-In this work, we investigate different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost.
+In this work, we investigate different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost
+and performance {\color{red} on a single node of an higher performance computing cluster.}
 In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
@@
 
+{\color{red}In this study we limit ourselves to comparison on a node level, which should already indicate the differences between the  matrix-free and matrix-based approaches.}
 The benchmark calculations are performed for 2D and 3D problems with various combination of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
-%
-We are interested in the following metrics:
-\begin{itemize}
-\item the wall-clock time and memory requirement per DoF for matrix-vector multiplication;
-\item the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step);
-\item the wall-clock time per DoF for the CG solver;
-\end{itemize}
+{\color{red}
+We chose to have around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix based approach.
+On another hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we should observe that the matrix-based approach is memory bound.
+}
 %
@@
 For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
+{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. fully occupied node.}
 
@@
 Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
-As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches.
+{\color{red}
+We are interested in the wall-clock time per DoF as a first metric.}
+As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
+{\color{red}as the matrix becomes more and more dense.}
 The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
@@
 {\color{red}
-described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.
-}
+described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.}
 Two conclusions can be drawn from these results:
 The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
-There is little difference between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor.
+There is little difference {\color{red}in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor.
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
 Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
-It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws.
+It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
+{\color{red}so long as the resulting tangent operator is expressed via the \eqref{eq:algebraic_tangent}.}
 
@@
 Memory bandwidth (MBytes/s) and the number of floating point operations per second (MFLOP/s) are measured using the \texttt{MEM\_DP} group of the LIKWID \cite{likwid} tool, version 4.2.1.
-The measurements are done on a single socket of RRZE cluster by pinning all 10 MPI processes to this socket.
+The measurements are done on a single socket of RRZE cluster by pinning all 10 MPI processes to this socket, whose
+memory bandwidth and peak performance are $B=50 \,\rm{GB/s}$ and $P=176\, \rm{GFLOP/s}$, respectively.
 [DD]TODO: plot and discuss.
@@
 {\color{red}
-Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective.
-}
-First we consider the number of iterations in the conjugate gradient (CG) solver.
-Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_Emmy_cg3} show a very mild increase in the average number of CG iterations for different polynomial degrees.
+Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end}
+we consider {\color{red}the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
+Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_Emmy_cg3} show a very mild increase in the average number of CG iterations for different polynomial degrees
+{\color{red} for the GMG preconditioner.}
 On the other hand, the black-box AMG preconditioner requires many more iterations for convergence.
@@
 Additionally, we also compare solution times of the preconditioned iterative solvers.
-}
+Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner,
+as well as to identify some bottlenecks within the former method that shall guide our further research in this direction.}
 Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for polynomial degrees up to $p=5$.
-For higher orders in 2D the wall-clock time is comparable to the AMG preconditioner, see Figure \ref{fig:benchmark_miehe_Emmy_sol2}.
+For higher orders in 2D the wall-clock time {\color{red}per DoF} is comparable to the AMG preconditioner, see Figure \ref{fig:benchmark_miehe_Emmy_sol2}.
 The reason for this is the suboptimal coarse level solver within the GMG preconditioner.
@@
 Consequently, the coarse level solver is the bottleneck of the adopted here geometric multigrid preconditioner.
-Its performance from the perspective of linear algebra shall be improved first and only then numerical implementational details of the complete preconditioner shall be studied.
-}
+Its performance from the perspective of linear algebra shall be improved first and only then numerical implementational details of the complete preconditioner shall be studied.}
 One possible solution to this issues would be to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.
@@
 
-In this contribution, we proposed and numerically investigated {\color{red}three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
+In this contribution, we proposed and numerically investigated {\color{red}on the node level three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
 {\color{red}
 To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to such materials.}
-The implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
+{\color{red}Among the three algorithms, }
+the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
 {\color{red}
@@
 that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
+{\color{red}
+Roofline model indicated the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
+Further studies are required to identify how to further improve the intensity and reach the compute bound case.}
 

### 2019-01-10 Calculations/likwid_emmy_benchmark_intel.sh
add tensor2 to LIKWID script, increase to 3 hours

@@ -1,6 +1,6 @@
 #!/bin/bash
-#   allocate 1 nodes (with likwid) with 40 CPU per node for 2 hours:
-#PBS -l nodes=1:ppn=40:likwid,walltime=2:00:00
+#   allocate 1 nodes (with likwid) with 40 CPU per node for 3 hours:
+#PBS -l nodes=1:ppn=40:likwid,walltime=3:00:00
 #   job name
 #PBS -N dealii_mf_cm_likwid
 #   stdout and stderr files:

### 2019-01-10 Calculations/pre_process.py

@@ -53,6 +53,7 @@ solvers = [
 ]
 
 solvers_likwid = [
+    ('MF_CG', 'gmg', 'tensor2'),
     ('MF_CG', 'gmg', 'tensor4'),
     ('CG',    'amg', 'scalar'),
 ]

### 2019-01-10 doc/large_strain.tex
more info on what Martin did already

@@
 {\color{red}
-For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements.
+The main reason is that the operations are typically the same on all elements.
+For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{For the heterogeneous materials considered here,
+the elements are first according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
+This functionality is provided by the \texttt{deal.II} library within the \textit{MatrixFree} class.
+}.
 For example, deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
 This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
-In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type.
+In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type
+and overloads basic arithmetic operations based on compiler intrinsics.
 The core of the sum factorization algorithms is provided by
@@
 Distributed memory MPI parallelization of matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so called locally owned elements.
-For more information about the implementation details of sum factorization techniques in deal.II and their performance analysis (????), see \cite{kronbichler12}.
+For more information about the implementation details and data structures for sum factorization approaches in deal.II (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and shared memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
 }
@@
 
+{\color{red}
+To summarize, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order bases to two factors.
+First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly a memory
+bound, see Figure XXXX.
+Second, is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
+Section \ref{sec:mf} and in the reference \cite{kronbichler12}.
+}
+
 \subsection{Preconditioned iterative solver}

### 2019-01-10 doc/large_strain.tex
minor comparision to Martin's results

@@
 {\color{red}
-We are interested in the wall-clock time per DoF as a first metric.}
+We are interested in the wall-clock time (average over 10 runs) per DoF as a first metric.}
 As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
-{\color{red}as the matrix becomes more and more dense.}
-The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
+{\color{red}as the matrix becomes more and more dense.
+The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for Laplace operator in \cite{kronbichler12}.
+Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces to capture the geometry as well as
+the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration),
+which increases the memory consumption makes element operations more expensive.
+Nonetheless }
+the matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
 Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,

### 2019-01-10 doc/large_strain.tex
better distinguish between read/write operations and sum factorization

@@
   \ForEach{ element $K \in \Omega^h$ }{
+        {\color{red}extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ } \;
         evaluate the following tensors at each quadrature point using sum factorization \;
-        $\Grad \, \overline{\gz u^h}$ \tcp*{2nd order}
-        $\gz g_s := \overline{\grad^s} \gz x$ \tcp*{2nd order symmetric}
-        $\gz g := \overline{\grad} \gz x$ \tcp*{2nd order}
+        $\Grad \, \overline{\gz u^h}_K$ \tcp*{2nd order}
+        $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
+        $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{
@@
         }
-      evaluate queued contractions using sum factorization and distribute results to the destination vector
+      evaluate queued contractions using sum factorization \;
+      distribute results to the destination vector
   }
@@
   \ForEach{ element $K \in \Omega^h$ }{
+        {\color{red}extract local vector values on this element $\gz x_K$ } \;
         evaluate the following tensors at each quadrature point using sum factorization \;
-        $\gz g_s := \overline{\grad^s} \gz x$ \tcp*{2nd order symmetric}
-        $\gz g := \overline{\grad} \gz x$ \tcp*{2nd order}
+        $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
+        $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{
@@
         }
-        evaluate queued contractions using sum factorization and distribute results to the destination vector
+        evaluate queued contractions using sum factorization \;
+        distribute results to the destination vector
   }

### 2019-01-10 Calculations/post_process_likwid.py
minor improvements to LIKWID post-processing script

@@ -152,6 +152,9 @@ degree_to_points = {
     8:'v'
 }
 
+mf_perf = []
+tr_perf = []
+
 # Now plot each measured point:
 for d in likwid_data:
   f = d[5][0]
@@ -161,11 +164,18 @@ for d in likwid_data:
   style = d[4] + degree_to_points[d[1]]
   label = '{0} (p={1})'.format(d[3],d[1])
   plt.loglog(x,y, style, label=label)
+  if 'MF' in label:
+    mf_perf.append(y[0])
+  else:
+    tr_perf.append(y[0])
 
 plt.xlabel('intensity (Flop/byte)')
 plt.ylabel('performance (GFlop/s)')
+plt.ylim(top=9000)
+plt.text(1,200,'P={0} GFlop/s'.format(P))
+plt.text(0.02, 5, 'B={0} GB/s'.format(B), rotation=30.)
 
-leg = plt.legend(loc='upper left', ncol=1, labelspacing=0.2)
+leg = plt.legend(loc='upper left', ncol=1, labelspacing=0.1)
 
 
 # file location
@@ -177,6 +187,11 @@ print 'Saving figure in: {0}'.format(fig_file)
 
 plt.tight_layout()
 plt.savefig(fig_file, format='pdf')
+
+# Finally report average performance for all MF and Trilinos runs:
+print 'Average performance:'
+print '  MF:       {0}'.format(np.mean(mf_perf))
+print '  Trilinos: {0}'.format(np.mean(tr_perf))
 
 # plt.savefig(fig_prefix + 'timing2d.eps', format='eps')
 # remove_creation_date(fig_prefix + 'timing2d.eps')

### 2019-01-10 doc/large_strain.tex
first discussion on Roofline

@@
 memory bandwidth and peak performance are $B=50 \,\rm{GB/s}$ and $P=176\, \rm{GFLOP/s}$, respectively.
-[DD]TODO: plot and discuss.
+}
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.8\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline.pdf}
+    % \caption{2D multigrid mesh with 2 global refinements}
+  \end{subfigure}
+  \caption{Roofline model for a selection of polynomial degrees on a 2D test example.}%
+  \label{fig:roofline}
+\end{figure}
+
+{\color{red}
+Figure \ref{fig:roofline} shows the Roofline performance model.
+As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$.
+Matrix-free implementation of fine-strain elastic tangent operators (on average) leads to a much better performance of
+$22.6\,\rm{GFlop/s}$, as well as about an order of magnitude higher computational intensity, but still stays in the memory bound regime.
+The achieved performance is about 13\% of the peak arithmetic performance, which is considerably lower than the
+70\% reported for the Laplace operation in \cite{kronbichler12}.
+Further detailed studies are necessary to explain this behavior.
 }

### 2019-01-10 doc/large_strain.tex
a bit more on LIKWID

@@
 70\% reported for the Laplace operation in \cite{kronbichler12}.
+When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
+we observe very marginal increase in computational intensity and almost unchanged performance.
+This indicates the origin for suboptimal results of matrix-free approach is most likely not related to the algorithms studied here.
 Further detailed studies are necessary to explain this behavior.
@@
 To summarize, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order bases to two factors.
-First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly a memory
-bound, see Figure XXXX.
+First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
+bound, see Figure \ref{fig:roofline}.
 Second, is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
-Section \ref{sec:mf} and in the reference \cite{kronbichler12}.
+Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
 }
@@
 Roofline model indicated the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
-Further studies are required to identify how to further improve the intensity and reach the compute bound case.}
+Further studies are required to identify means to improve the computational intensity and performance of the matrix-free implementation.}
 

### 2019-01-10 Calculations/post_process_likwid.py
sort by label and degree

@@ -123,7 +123,7 @@ for f in files:
 
 # now we have lists of tuples ready
 # first, sort by label:
-likwid_data.sort(key=lambda tup: tup[3])
+likwid_data.sort(key=lambda tup: (tup[3], tup[1]))
 
 #####################
 #       PLOT        #

### 2019-01-10 doc/large_strain.tex
minor

@@
 Figure \ref{fig:roofline} shows the Roofline performance model.
-As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$.
+As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$ (4.5\% of the peak performance).
 Matrix-free implementation of fine-strain elastic tangent operators (on average) leads to a much better performance of
@@
 we observe very marginal increase in computational intensity and almost unchanged performance.
-This indicates the origin for suboptimal results of matrix-free approach is most likely not related to the algorithms studied here.
+This indicates that the origin for suboptimal results of matrix-free approach is most likely not related to the algorithms numerically investigated here.
 Further detailed studies are necessary to explain this behavior.

### 2019-01-10 include/mf_nh_operator.h
add MatrixFree object(s) and diagonal to NeoHookOperator::memory_consumption()

@@ -184,7 +184,12 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::memory_consumption()
   return cached_scalar.memory_consumption() +
          cached_second_scalar.memory_consumption() +
          cached_tensor2.memory_consumption() +
-         cached_tensor4.memory_consumption();
+         cached_tensor4.memory_consumption() +
+         // matrix-free data:
+         data_current->memory_consumption() +
+         (mf_caching == "scalar" ? data_reference->memory_consumption() : 0) +
+         // diagonal (we actually need only one vector)
+         inverse_diagonal_entries->memory_consumption();
 }
 
 

### 2019-01-11 include/mf_nh_operator.h
MF: do not include diagonals into memory measurements (that also broke the code, but I don't have time to investigate what's wrong in deal.II)

@@ -187,9 +187,9 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::memory_consumption()
          cached_tensor4.memory_consumption() +
          // matrix-free data:
          data_current->memory_consumption() +
-         (mf_caching == "scalar" ? data_reference->memory_consumption() : 0) +
-         // diagonal (we actually need only one vector)
-         inverse_diagonal_entries->memory_consumption();
+         (mf_caching == "scalar" ? data_reference->memory_consumption() : 0);
+  // note: do not include diagonals, we want to measure only memory needed for vmult
+  // for performance analysis.
 }
 
 

### 2019-01-11 Calculations/likwid_emmy_benchmark_intel.sh
python: add 3D and scalar strategy to LIKWID runs, request 7 hours

@@ -1,6 +1,6 @@
 #!/bin/bash
-#   allocate 1 nodes (with likwid) with 40 CPU per node for 3 hours:
-#PBS -l nodes=1:ppn=40:likwid,walltime=3:00:00
+#   allocate 1 nodes (with likwid) with 40 CPU per node for 7 hours:
+#PBS -l nodes=1:ppn=40:likwid,walltime=7:00:00
 #   job name
 #PBS -N dealii_mf_cm_likwid
 #   stdout and stderr files:

### 2019-01-11 Calculations/pre_process.py

@@ -42,6 +42,8 @@ poly_quad_ref_dim_likwid = [
     (2,3,6,2),
     (4,5,5,2),
     (6,7,4,2),
+    (2,3,3,3),
+    (4,5,2,3),
 ]
 
 # Solvers (type, preconditioner and caching)
@@ -53,6 +55,7 @@ solvers = [
 ]
 
 solvers_likwid = [
+    ('MF_CG', 'gmg', 'scalar'),
     ('MF_CG', 'gmg', 'tensor2'),
     ('MF_CG', 'gmg', 'tensor4'),
     ('CG',    'amg', 'scalar'),

### 2019-01-11 doc/large_strain.tex
improvements based on Daniel's comments

@@
     and employ them to define a geometric multigrid preconditioner.
-    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner {\color{red} on a single node of an higher performance computing cluster} for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
+    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner {\color{red} on a single node} for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
   \end{abstract}
@@
 {\color{red}
-that appear in the tangent operator \eqref{eq:algebraic_tangent} of finite-strain elasticity,
-}
+that appear in the tangent operator \eqref{eq:algebraic_tangent} of finite-strain elasticity,}
 we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
@@
 For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{For the heterogeneous materials considered here,
-the elements are first according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
+the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
 This functionality is provided by the \texttt{deal.II} library within the \textit{MatrixFree} class.
 }.
-For example, deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
+For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
 This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
@@
 {\color{red}
-The geometric multigrid algorithm is a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
+The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
 by the \texttt{deal.II} library via \textit{PreconditionChebyshev} class.
@@
 At the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
+Note that for the matrix-free approach we are limited to smoothers that use only diagonal elements of the operator.
 }
@@
 We chose to have around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix based approach.
-On another hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we should observe that the matrix-based approach is memory bound.
+On another hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we can observe that the matrix-based approach is memory bound.
 }
@@
 Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces to capture the geometry as well as
-the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration),
-which increases the memory consumption makes element operations more expensive.
+the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
+Both factors increase the memory consumption and make element operations more expensive.
 Nonetheless }
@@
 The achieved performance is about 13\% of the peak arithmetic performance, which is considerably lower than the
-70\% reported for the Laplace operation in \cite{kronbichler12}.
+70\% reported for the Laplace operation on Cartesian meshes in \cite{kronbichler12}.
 When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),

### 2019-01-11 Calculations/post_process_likwid.py
likwid: plot ceilings in Roofline, confirm peak performance calculation

@@ -46,9 +46,9 @@ sections = [
 
 # NODE data:
 # memory bandwidth
-B=50   # 50 GB/s single socket
+B=50.0   # 50 GB/s single socket
 # peak performance
-P=176  # 176 Gflop/s single socket
+P=176.0  # 176 Gflop/s single socket
 
 table_names = [
     'Event',
@@ -136,12 +136,23 @@ plt.rcParams.update(params)
 # Roofline model
 # p = min (P, b I)
 # where I is measured intensity (Flops/byte)
-x = [0.05*i for i in range(200)]
-y = [min(P,B*i) for i in x]
+def Roofline(I,P,B):
+    return np.array([min(P,B*i) for i in I])
 
-plt.grid(True, which="both",color='grey', linestyle=':')
+x = np.linspace(1./B, 10., num=500)
+base = np.array([1./B for i in x])
+roofline_style = 'b-'
+peak = Roofline(x,P,B)
+# see https://github.com/matplotlib/matplotlib/issues/8623#issuecomment-304892552
+plt.loglog(x,peak, roofline_style, label=None, nonposx='clip', nonposy='clip')
 
-plt.loglog(x,y, 'r-', label='Roofline')
+# various ceilings (w/o FMA, w/o FMA and vectorization):
+for p_ in [P/2, P/2/4]:
+    plt.plot(x,Roofline(x,p_,B), roofline_style, label=None)
+
+plt.fill_between(x, base, peak, where=peak>base, interpolate=True, zorder=1, color='aqua')
+
+plt.grid(True, which="both",color='grey', linestyle=':', zorder=5)
 
 
 # map degrees to point labels:
@@ -171,9 +182,11 @@ for d in likwid_data:
 
 plt.xlabel('intensity (Flop/byte)')
 plt.ylabel('performance (GFlop/s)')
-plt.ylim(top=9000)
-plt.text(1,200,'P={0} GFlop/s'.format(P))
-plt.text(0.02, 5, 'B={0} GB/s'.format(B), rotation=30.)
+plt.ylim(top=9000,bottom=1)
+plt.text(0.04, 15, 'B={0} GB/s'.format(B), rotation=30.)
+plt.text(3,200,'Peak DP')  #  2.2 GHz
+plt.text(3,100,'w/o FMA')
+plt.text(3,25, 'w/o SIMD')
 
 leg = plt.legend(loc='upper left', ncol=1, labelspacing=0.1)
 
[…]

### 2019-01-11 doc/large_strain.tex
Update doc/large_strain.tex
Co-Authored-By: davydden <davydden@gmail.com>
update on what we measure as memory, att todo's

@@
 
+\usepackage{todonotes}
 \usepackage{bm}
@@
 We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
-Finally, we note that at the time of writing, $p-$ version of the GMG is not available within the \texttt{deal.II} library.
+Finally, we note that at the time of writing, $p-$ version of the GMG is not available within the \texttt{deal.II} library and therefore not considered here.
 }
@@
 {\color{red}
-We are interested in the wall-clock time (average over 10 runs) per DoF as a first metric.}
+We are interested in the wall-clock time (average over 10 runs per each linear solver step) per DoF as a first metric.}
 As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
@@
 Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
+{\color{red}
+In those graphs we report the memory needed to cache additional quantities required for the tangent operator here considered,
+as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
+from unit to real cell \cite{kronbichler12}.
+Note that for the Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
+configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{rerun RRZE and IWR}\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}
 

### 2019-01-11 doc/large_strain.tex
say how memory bandwidth is measured

@@
 The measurements are done on a single socket of RRZE cluster by pinning all 10 MPI processes to this socket, whose
-memory bandwidth and peak performance are $B=50 \,\rm{GB/s}$ and $P=176\, \rm{GFLOP/s}$, respectively.
+memory bandwidth\footnote{As measured with the \textit{load\_avx} benchmark of \textit{likwid-bench} \cite{likwid}.} and peak performance are $B=47.2 \,\rm{GB/s}$ and $P=176\, \rm{GFlop/s}$, respectively.
 }

### 2019-01-11 Calculations/post_process_likwid.py

@@ -45,10 +45,20 @@ sections = [
 ]
 
 # NODE data:
+clock_speed = 2.2  # GHz
 # memory bandwidth
-B=50.0   # 50 GB/s single socket
-# peak performance
-P=176.0  # 176 Gflop/s single socket
+# likwid-bench -t load_avx -w S0:1GB:10:1:2
+B=47.16855 # 50 GB/s single socket
+
+print 'Peak performance calculations:'
+P = clock_speed
+print '  Clock speed:  {0}'.format(clock_speed)
+P = P * 2                             # add+mult per cycle
+print '  w FMA:        {0}'.format(P)
+P = P * 4                             # vectorization over 4 doubles = 256 bits (AVX), VECTORIZATION_LEVEL=2
+print '  w FMA w SIMD: {0}'.format(P)
+P = P * 10                            # 10 cores
+print '  peak:         {0}'.format(P)
 
 table_names = [
     'Event',
@@ -183,7 +193,7 @@ for d in likwid_data:
 plt.xlabel('intensity (Flop/byte)')
 plt.ylabel('performance (GFlop/s)')
 plt.ylim(top=9000,bottom=1)
-plt.text(0.04, 15, 'B={0} GB/s'.format(B), rotation=30.)
+plt.text(0.04, 15, 'B={:.1f} GB/s'.format(B), rotation=30.)
 plt.text(3,200,'Peak DP')  #  2.2 GHz
 plt.text(3,100,'w/o FMA')
 plt.text(3,25, 'w/o SIMD')
@@ -208,14 +218,3 @@ print '  Trilinos: {0}'.format(np.mean(tr_perf))
 
 # plt.savefig(fig_prefix + 'timing2d.eps', format='eps')
 # remove_creation_date(fig_prefix + 'timing2d.eps')
-
-print 'Peak performance calculations:'
-clock_speed = 2.2  # GHz
-performance = clock_speed
-print '  Clock speed:  {0}'.format(clock_speed)
-performance = performance * 2   # add+mult per cycle
-print '  w FMA:        {0}'.format(performance)
-performance = performance * 4   # vectorization over 4 doubles = 256 bits (AVX), VECTORIZATION_LEVEL=2
-print '  w FMA w SIMD: {0}'.format(performance)
-performance = performance * 10  # 10 cores
-print '  peak:         {0}'.format(performance)

### 2019-01-11 Calculations/post_process_likwid.py
tweak LIKWID plot

@@ -139,7 +139,7 @@ likwid_data.sort(key=lambda tup: (tup[3], tup[1]))
 #       PLOT        #
 #####################
 
-params = {'legend.fontsize': 20,
+params = {'legend.fontsize': 14,
           'font.size': 20}
 plt.rcParams.update(params)
 
@@ -160,7 +160,7 @@ plt.loglog(x,peak, roofline_style, label=None, nonposx='clip', nonposy='clip')
 for p_ in [P/2, P/2/4]:
     plt.plot(x,Roofline(x,p_,B), roofline_style, label=None)
 
-plt.fill_between(x, base, peak, where=peak>base, interpolate=True, zorder=1, color='aqua')
+plt.fill_between(x, base, peak, where=peak>base, interpolate=True, zorder=1, color='aqua', alpha=0.1)
 
 plt.grid(True, which="both",color='grey', linestyle=':', zorder=5)
 
@@ -192,11 +192,11 @@ for d in likwid_data:
 
 plt.xlabel('intensity (Flop/byte)')
 plt.ylabel('performance (GFlop/s)')
-plt.ylim(top=9000,bottom=1)
-plt.text(0.04, 15, 'B={:.1f} GB/s'.format(B), rotation=30.)
-plt.text(3,200,'Peak DP')  #  2.2 GHz
-plt.text(3,100,'w/o FMA')
-plt.text(3,25, 'w/o SIMD')
+plt.ylim(top=300,bottom=1)
+plt.text(0.02, 5, 'B={:.1f} GB/s'.format(B), rotation=42)
+plt.text(3.7,200,'Peak DP', fontsize=14)
+plt.text(3.7,100,'w/o FMA', fontsize=14)
+plt.text(3.7,25, 'w/o SIMD', fontsize=14)
 
 leg = plt.legend(loc='upper left', ncol=1, labelspacing=0.1)
 

### 2019-01-11 Calculations/post_process_likwid.py
LIWKID: extend python script to distinguish between 2d and 3d plots

@@ -18,8 +18,11 @@ def remove_creation_date(file_name):
 parser = argparse.ArgumentParser(
     description='Post-Process timing/memory info and plot figures.',
     formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-parser.add_argument('prefix', metavar='prefix', default='LIKWID_Emmy_RRZE', nargs='?',
+parser.add_argument('--prefix', metavar='prefix', default='LIKWID_Emmy_RRZE',
                     help='A folder to look for benchmark results')
+parser.add_argument('--dim', metavar='dim', default=2, type=int,
+                    help='Dimension (2 or 3)')
+
 args = parser.parse_args()
 
 prefix = args.prefix if args.prefix.startswith('/') else os.path.join(os.getcwd(), args.prefix)
@@ -178,17 +181,18 @@ tr_perf = []
 
 # Now plot each measured point:
 for d in likwid_data:
-  f = d[5][0]
-  b = d[5][1]
-  x = [f/b]
-  y = [f/1000]  # MFLOP->GFLOP
-  style = d[4] + degree_to_points[d[1]]
-  label = '{0} (p={1})'.format(d[3],d[1])
-  plt.loglog(x,y, style, label=label)
-  if 'MF' in label:
-    mf_perf.append(y[0])
-  else:
-    tr_perf.append(y[0])
+  if d[0] == args.dim:
+    f = d[5][0]
+    b = d[5][1]
+    x = [f/b]
+    y = [f/1000]  # MFLOP->GFLOP
+    style = d[4] + degree_to_points[d[1]]
+    label = '{0} (p={1})'.format(d[3],d[1])
+    plt.loglog(x,y, style, label=label)
+    if 'MF' in label:
+      mf_perf.append(y[0])
+    else:
+      tr_perf.append(y[0])
 
 plt.xlabel('intensity (Flop/byte)')
 plt.ylabel('performance (GFlop/s)')
@@ -203,18 +207,15 @@ leg = plt.legend(loc='upper left', ncol=1, labelspacing=0.1)
 
 # file location
 fig_prefix = os.path.join(os.getcwd(), '../doc/' + os.path.basename(os.path.normpath(prefix)) + '_')
-
-fig_file = fig_prefix + 'roofline.pdf'
+fig_file = fig_prefix + 'roofline_{0}d.pdf'.format(args.dim)
 
 print 'Saving figure in: {0}'.format(fig_file)
 
 plt.tight_layout()
[…]

### 2019-01-11 doc/large_strain.tex
extend Roofline discussion, add 2d figure and 3d (incomplete)

@@
   \centering
-  \begin{subfigure}[b]{0.8\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline.pdf}
-    % \caption{2D multigrid mesh with 2 global refinements}
-  \end{subfigure}
-  \caption{Roofline model for a selection of polynomial degrees on a 2D test example.}%
+  \begin{subfigure}[b]{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline_2d.pdf}
+    \caption{2D}
+    \label{fig:roofline_2d}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline_3d.pdf}
+    \caption{3D}
+    \label{fig:roofline_3d}
+  \end{subfigure}
+  \caption{Roofline model for a selection of polynomial degrees. `Peak DP' denotes the peak double precision performance ($P$),
+  `w/o FMA' represents a ceiling without Fused Multiply-Add ($P/2$), finally `w/o SIMD' is a ceiling where both FMA
+  and SIMD vectorization are discarded ($P/8$).}%
   \label{fig:roofline}
@@
 {\color{red}
-Figure \ref{fig:roofline} shows the Roofline performance model.
-As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$ (4.5\% of the peak performance).
+Figure \ref{fig:roofline} shows results of the Roofline performance model.
+As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
 Matrix-free implementation of fine-strain elastic tangent operators (on average) leads to a much better performance of
-$22.6\,\rm{GFlop/s}$, as well as about an order of magnitude higher computational intensity, but still stays in the memory bound regime.
-The achieved performance is about 13\% of the peak arithmetic performance, which is considerably lower than the
-70\% reported for the Laplace operation on Cartesian meshes in \cite{kronbichler12}.
+$22.6\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory bound regime.
+The achieved performance is about 13\% of the peak arithmetic performance. This is much lower than the
+70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.
+Unfortunately, the performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.
+
 When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
-we observe very marginal increase in computational intensity and almost unchanged performance.
-This indicates that the origin for suboptimal results of matrix-free approach is most likely not related to the algorithms numerically investigated here.
+we observe a marginal increase in computational intensity and almost unchanged performance.
+This coincides with our expectation that the later algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
+On another hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
+Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
+second order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
+As a result both algorithms will have comparable memory requirements (also confirmed by Figure ....),
+however the ``scalar'' algorithm performs a large number of repetitive operations that are
+independent on the right hand side vector (steps 8--11 in Algorithm \ref{alg:mf_scalar}), which makes it the worst option among the three.
+
+Finally we note that all three algorithms are far from the theoretical peak performance which
+indicates that the origin for suboptimal results of matrix-free approach is most likely not related to the various flavours of algorithms numerically investigated here.
 Further detailed studies are necessary to explain this behavior.
-}
-
-{\color{red}
-To summarize, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order bases to two factors.
+
+To summarize this section, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order bases to two factors.
 First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
@@
 Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
+Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach with relatively high computational intensity.
 }

### 2019-01-11 Calculations/likwid_emmy_benchmark_intel.sh
ask 15 instead 7 hours for LIKWID runs

@@ -1,6 +1,6 @@
 #!/bin/bash
-#   allocate 1 nodes (with likwid) with 40 CPU per node for 7 hours:
-#PBS -l nodes=1:ppn=40:likwid,walltime=7:00:00
+#   allocate 1 nodes (with likwid) with 40 CPU per node for 15 hours:
+#PBS -l nodes=1:ppn=40:likwid,walltime=15:00:00
 #   job name
 #PBS -N dealii_mf_cm_likwid
 #   stdout and stderr files:

### 2019-01-11 include/mf_nh_operator.h
mf: reinit phi_reference nad read displacements only for scalar caching

@@ -516,12 +516,16 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::local_apply_cell(
       // initialize on this cell
       phi_current.reinit(cell);
       phi_current_s.reinit(cell);
-      phi_reference.reinit(cell);
 
       // read-in total displacement and src vector and evaluate gradients
-      phi_reference.read_dof_values_plain(*displacement);
       phi_current.read_dof_values(src);
       phi_current_s.read_dof_values(src);
+
+      if (mf_caching == "scalar")
+        {
+          phi_reference.reinit(cell);
+          phi_reference.read_dof_values_plain(*displacement);
+        }
 
       do_operation_on_cell(phi_current, phi_current_s, phi_reference, cell);
 

### 2019-01-11 doc/large_strain.tex
Fix some typos and improve some formulations

@@
   \begin{abstract}
-    The performance of finite element solvers on modern computer architectures is typically memory bound.
+    The performance of finite element solvers on modern computer architectures is typically memory bound {\color{red} for sufficiently large problems}.
     The main cause for this is that loading matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
@@
 
-The performance of finite element solvers on modern computer architectures is typically memory bound.
+The performance of finite element solvers on modern computer architectures is typically memory bound {\color{red} for sufficiently large problems}.
 The main cause for this is that loading pre-computed matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
@@
 To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
-We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent masters thesis produced by a student of the first two of the authors \cite{Mentler2017},
-{\color{red} which is also limited to the small strain linear elasticity.}
+We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent master's thesis produced by a student of the first two of the authors \cite{Mentler2017},
+{\color{red} which is also limited to small strain linear elasticity.omit?}
 One possible reason is that the tangent operators\footnote{Here and
@@
 whereas higher order elements are rarely employed.
-This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements.
+This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements ({\color{red} reference?}).
 However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
@@
 
-Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
+{\color{red}For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.}
 %
@@
 {\color{red}
-that appear in the tangent operator \eqref{eq:algebraic_tangent} of finite-strain elasticity,}
+that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem,}
 we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
@@
 In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point.
-In this case, we also avoid the need to evaluate gradients of displacement field with respect to the referential configuration.
+In this case, we also avoid the need to evaluate gradients of the displacement field with respect to the referential configuration.
 This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
@@
 the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
-This functionality is provided by the \texttt{deal.II} library within the \textit{MatrixFree} class.
+This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
 }.
@@
 and overloads basic arithmetic operations based on compiler intrinsics.
-The core of the sum factorization algorithms is provided by
+The core of the sum factorization algorithms is provided by the
 \textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
 Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
-Distributed memory MPI parallelization of matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so called locally owned elements.
+Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
 For more information about the implementation details and data structures for sum factorization approaches in deal.II (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and shared memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
@@
 The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
-The efficiency of the preconditioner, from the perspective of linear algebra, with the so defined level operators will be assessed by numerical examples in the next section.
+The efficiency of the preconditioner with the considered level operators will be assessed by numerical examples in the next section.
 We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
-Finally, we note that at the time of writing, $p-$ version of the GMG is not available within the \texttt{deal.II} library and therefore not considered here.
+Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
 }
@@
 The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
-by the \texttt{deal.II} library via \textit{PreconditionChebyshev} class.
+by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
 The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
 is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
-At the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
-Note that for the matrix-free approach we are limited to smoothers that use only diagonal elements of the operator.
-}
+On the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
+Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while must default smoothers rely on an explicit  matrix form.}
 
@@
 Figure \ref{fig:miehe_fine_level} illustrates such decomposition on the fine mesh level for the 2D problem.
-Same approach is adopted at the multigrid levels.}
+The same approach is adopted on the multigrid levels.}
 Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.
@@
 
-{\color{red}In this study we limit ourselves to comparison on a node level, which should already indicate the differences between the  matrix-free and matrix-based approaches.}
-The benchmark calculations are performed for 2D and 3D problems with various combination of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
-{\color{red}
-We chose to have around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix based approach.
-On another hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we can observe that the matrix-based approach is memory bound.
+{\color{red}In this study, we limit ourselves to a comparison on a node level, which should already indicate the differences between the matrix-free and matrix-based approaches.}
+The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
+{\color{red}
+We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach.
+On the other hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we can observe that the matrix-based approach is memory bound.
 }
@@
 {\color{red}as the matrix becomes more and more dense.
-The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for Laplace operator in \cite{kronbichler12}.
+The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
 Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces to capture the geometry as well as
@@
 Both factors increase the memory consumption and make element operations more expensive.
-Nonetheless }
+Nonetheless, }
 the matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
@@
 The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
-There is little difference {\color{red}in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor.
+There is little difference {\color{red}in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'').
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
@@
 It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
-{\color{red}so long as the resulting tangent operator is expressed via the \eqref{eq:algebraic_tangent}.}
+{\color{red}as long as the resulting tangent operator is expressed via \eqref{eq:algebraic_tangent}.}
 
@@
 from unit to real cell \cite{kronbichler12}.
-Note that for the Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
+Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
 configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{rerun RRZE and IWR}\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}
@@
   `w/o FMA' represents a ceiling without Fused Multiply-Add ($P/2$), finally `w/o SIMD' is a ceiling where both FMA
-  and SIMD vectorization are discarded ($P/8$).}%
+    and SIMD vectorization are discarded ($P/8$).{\color{red} TODO Improve design}}%
   \label{fig:roofline}
@@
 As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
-Matrix-free implementation of fine-strain elastic tangent operators (on average) leads to a much better performance of
-$22.6\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory bound regime.
+The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
+$22.6\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
 The achieved performance is about 13\% of the peak arithmetic performance. This is much lower than the
 70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.
-Unfortunately, the performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.
+Unfortunately, performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.
 
@@
 
-To summarize this section, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order bases to two factors.
+{\color{red}
+To summarize this section, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order elements to two factors.
 First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
@@
 we consider {\color{red}the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
-Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_Emmy_cg3} show a very mild increase in the average number of CG iterations for different polynomial degrees
+Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a very mild increase in the average number of CG iterations for different polynomial degrees
 {\color{red} for the GMG preconditioner.}
@@
 {\color{red}
-Consequently we can concluded that, although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still results in a well behaving multigrid preconditioner for the here considered finite strain elasticity problem.
-}
-
-{\color{red}
-Additionally, we also compare solution times of the preconditioned iterative solvers.
+Consequently, we can conclude that, although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a well behaving multigrid preconditioner for the here considered finite strain elasticity problem and that the considered smoother is suitable.
+}
+
+{\color{red}
+Additionally, we compare solution times of the preconditioned iterative solvers.
 Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner,
@@
 {\color{red}
-Consequently, the coarse level solver is the bottleneck of the adopted here geometric multigrid preconditioner.
-Its performance from the perspective of linear algebra shall be improved first and only then numerical implementational details of the complete preconditioner shall be studied.}
-One possible solution to this issues would be to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.
+Consequently, the coarse level solver is the bottleneck of the here described geometric multigrid preconditioner.
+Its performance shall first be improved from a linear algebra perspective and only then implementational details of the complete preconditioner shall be studied.}
+One possible solution to this issue would be to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.
 Although untested in this work, we hypothesize that this should also improve the performance of the preconditioner for lower degrees.
@@
 
-In this contribution, we proposed and numerically investigated {\color{red}on the node level three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
+In this contribution, we proposed and numerically investigated {\color{red}three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
 {\color{red}
 To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to such materials.}
-{\color{red}Among the three algorithms, }
+{\color{red}Among the three algorithms,}
 the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
@@
 that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
-{\color{red}
-Roofline model indicated the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
+
+{\color{red}
+The Roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
 Further studies are required to identify means to improve the computational intensity and performance of the matrix-free implementation.}
@@
 leads to a solution approach that is faster than the matrix-based AMG for moderately high polynomial degrees (approximately up to order five).
-However, a considerable amount of time of GMG preconditioner is spent in the coarse level solver,
+However, a considerable amount of time is spent in the coarse level solver of the GMG preconditioner,
 {\color{red}

### 2019-01-11 doc/large_strain.tex
Fix a bracket
use \todo package to do TODO

@@
 We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent master's thesis produced by a student of the first two of the authors \cite{Mentler2017},
-{\color{red} which is also limited to small strain linear elasticity.omit?}
+{\color{red} which is also limited to small strain linear elasticity.}\todo{omit?}
 One possible reason is that the tangent operators\footnote{Here and
@@
 whereas higher order elements are rarely employed.
-This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements ({\color{red} reference?}).
+This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements\todo{reference?}.
 However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
@@
 On the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
-Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while must default smoothers rely on an explicit  matrix form.}
+Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}
 
@@
   `w/o FMA' represents a ceiling without Fused Multiply-Add ($P/2$), finally `w/o SIMD' is a ceiling where both FMA
-    and SIMD vectorization are discarded ($P/8$).{\color{red} TODO Improve design}}%
+    and SIMD vectorization are discarded ($P/8$).}%
   \label{fig:roofline}
@@
 {\color{red}
-Figure \ref{fig:roofline} shows results of the Roofline performance model.
+Figure\todo{Improve design} \ref{fig:roofline} shows results of the Roofline performance model.
 As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
@@
 indicates that the origin for suboptimal results of matrix-free approach is most likely not related to the various flavours of algorithms numerically investigated here.
-Further detailed studies are necessary to explain this behavior.
+Further detailed studies are necessary to explain this behavior.}
 

### 2019-01-12 Calculations/likwid_emmy_benchmark_intel.sh
LIKWID: run vmult() on first load step and first NR and exit gracefully

@@ -1,6 +1,6 @@
 #!/bin/bash
-#   allocate 1 nodes (with likwid) with 40 CPU per node for 15 hours:
-#PBS -l nodes=1:ppn=40:likwid,walltime=15:00:00
+#   allocate 1 nodes (with likwid) with 40 CPU per node for 1 hour:
+#PBS -l nodes=1:ppn=40:likwid,walltime=1:00:00
 #   job name
 #PBS -N dealii_mf_cm_likwid
 #   stdout and stderr files:

### 2019-01-12 Calculations/pre_process.py

@@ -107,6 +107,10 @@ subsection Boundary conditions
   set Dirichlet IDs and component mask = {11}
   set Neumann IDs and expressions = {12}
 end
+
+subsection Time
+  set End time  = {14}
+end
 """
 
 output_points  = {
@@ -168,7 +172,8 @@ for pqrd in poly_quad_ref_dim:
             dirichlet_id[pqrd[3]],
             dirichlet_mask[pqrd[3]],
             neumann_bc[pqrd[3]],
-            'false'
+            'false',
+            0.2 if args.likwid else 1.0  # make sure we do 1 fake solution step
         ))
         filenames.append(name)
 
@@ -194,7 +199,8 @@ if not args.likwid:
               dirichlet_id[pqrd[3]],
               dirichlet_mask[pqrd[3]],
               neumann_bc[pqrd[3]],
-              'true'
+              'true',
+              1.0
           ))
 
 

### 2019-01-12 include/mf_elasticity.h

@@ -1095,6 +1095,7 @@ namespace Cook_Membrane
   Solid<dim, degree, n_q_points_1d, NumberType>::run()
   {
 #ifdef WITH_LIKWID
+    pcout << "LIKWID_MARKER_INIT" << std::endl;
     LIKWID_MARKER_INIT;
 #endif
     make_grid();
@@ -1130,6 +1131,7 @@ namespace Cook_Membrane
     // for post-processing, print average CG iterations over the whole run:
     timer_out << std::endl << "Average CG iter = " << (total_n_cg_iterations/total_n_cg_solve) << std::endl;
 #ifdef WITH_LIKWID
+    pcout << "LIKWID_MARKER_CLOSE" << std::endl;
     LIKWID_MARKER_CLOSE;
 #endif
 
@@ -2076,6 +2078,17 @@ namespace Cook_Membrane
               LIKWID_MARKER_STOP("vmult_MF");
 #endif
             }
+
+#ifdef WITH_LIKWID
+          // save HPC time and simply break out of NR loop
+          AssertThrow (parameters.delta_t == parameters.end_time,
+                      ExcMessage("LIKWID runs are fake and should have 1 timestep only"));
+          pcout << " FAKE CONVERGED!" << std::endl;
+          print_conv_footer();
+          total_n_cg_solve = 1;  // we later divide by that, so terminate gracefully
+          break;
+#endif
+
 
 #ifdef DEBUG
           if (!parameters.skip_tangent_assembly)

### 2019-01-12 doc/large_strain.tex
update Roffline (full 3D, quicker runs), minor related changes in the text

@@
 The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
-$22.6\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
+$22.1\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
 The achieved performance is about 13\% of the peak arithmetic performance. This is much lower than the
@@
 second order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
-As a result both algorithms will have comparable memory requirements (also confirmed by Figure ....),
+As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} \todo{rerun and check}),
 however the ``scalar'' algorithm performs a large number of repetitive operations that are

### 2019-01-14 Calculations/post_process_likwid.py
hack LIKWID post-processing script to plot Roofline for breakdown of vmult into various stages

@@ -22,10 +22,13 @@ parser.add_argument('--prefix', metavar='prefix', default='LIKWID_Emmy_RRZE',
                     help='A folder to look for benchmark results')
 parser.add_argument('--dim', metavar='dim', default=2, type=int,
                     help='Dimension (2 or 3)')
+parser.add_argument('--breakdown', help='Post process breakdown results of vmult', action="store_true")
 
 args = parser.parse_args()
 
 prefix = args.prefix if args.prefix.startswith('/') else os.path.join(os.getcwd(), args.prefix)
+if args.breakdown:
+    prefix = prefix + '_breakdown'
 
 files = []
 for root, dirs, files_ in os.walk(prefix):
@@ -44,7 +47,8 @@ pattern = r'[+\-]?(?:[0-9]\d*)(?:\.\d*)?(?:[eE][+\-]?\d+)?'
 sections = [
     'MFLOP/s STAT',
     'Memory bandwidth [MBytes/s] STAT',
-    'Operational intensity STAT'
+    'Operational intensity STAT',
+    'Runtime (RDTSC) [s] STAT'
 ]
 
 # NODE data:
@@ -70,6 +74,26 @@ table_names = [
 
 likwid_data = []
 
+regions = []
+if args.breakdown:
+    regions = [
+        'vmult_sum_factorization',
+        'vmult_reinit_read_write',
+        'vmult_quadrature_loop'
+    ]
+
+# labels/colors to be used for regions in breakdown run:
+region_labels = [
+    'sum factorization',
+    'read / write',
+    'quadrature loop'
+]
+region_colors = [
+    'b',
+    'g',
+    'c'
+]
+
 for f in files:
     # guess parameters from the file name:
     fname = os.path.basename(f)
@@ -77,7 +101,10 @@ for f in files:
     dim = int(re.findall(pattern,strings[2])[0])
     p   = int(re.findall(pattern,strings[3])[0])
     q   = int(re.findall(pattern,strings[3])[1])
-    region = 'vmult_MF' if 'MF_CG' in f else 'vmult_Trilinos'
+    if not args.breakdown:
+        regions = [
+            'vmult_MF' if 'MF_CG' in f else 'vmult_Trilinos'
+        ]
     label = ''
     color = ''  # use colors consistent with post_process.py
     if 'MF_CG' in fname:
@@ -94,19 +121,24 @@ for f in files:
         label = 'Trilinos'
         color = 'r'
 
[…]

### 2019-01-15 Calculations/post_process_likwid.py
python: plot stacked barplot for timing

@@ -132,7 +132,11 @@ for f in files:
     found_region = False
     found_table = False
     r_idx = np.nan
+    n_cells = np.nan
     for line in fin:
+        if 'Number of active cells' in line:
+            n_cells = int(re.findall(pattern,line)[0])
+
         # Check if we found one of the regions:
         if 'Region:' in line:
             found_region = False
@@ -175,7 +179,7 @@ for f in files:
         if args.breakdown:
             label = region_labels[r_idx]
             color = region_colors[r_idx]
-        tp = tuple((dim,p,q,label,color,t))
+        tp = tuple((dim,p,q,label,color,t,r_idx,n_cells))
         likwid_data.append(tp)
 
 # now we have lists of tuples ready
@@ -272,3 +276,54 @@ if not args.breakdown:
   print 'Average performance:'
   print '  MF:       {0}'.format(np.mean(mf_perf))
   print '  Trilinos: {0}'.format(np.mean(tr_perf))
+else:
+  # clear
+  plt.clf()
+  ax = plt.figure().gca()
+  ax.xaxis.set_major_locator(MaxNLocator(integer=True))
+  plt.ticklabel_format(style='sci', axis='y', scilimits=(0,0))
+  # FIXME: hard-code degrees
+  ind = [2, 4, 6] if args.dim==2 else [2, 4]
+  ind_str = ['{0}'.format(s) for s in ind]
+  ind_plt = range(len(ind))
+  # plot stack bar graph for time
+  bar_data = [[ np.nan for i in range(len(ind))] for i in range(len(region_labels))]
+
+  for d in likwid_data:
+    if d[0] == args.dim:
+      r_idx = d[6] # region index
+      t = d[5][3] / d[7] # runtime unhalted per cell
+      p = d[1]
+      for idx, s in enumerate(ind):
+        if s == p:
+          bar_data[r_idx][idx] = t
+
+  # setup "bottom" for bar data
+  bar_data_bottom = [[ 0 for i in range(len(ind))] for i in range(len(region_labels))]
+  for i in range(len(region_labels)):
+    if i > 0:
+      for j in range(len(ind)):
+        for k in range(i-1):
+          bar_data_bottom[i][j] = bar_data_bottom[i][j] + bar_data[k][j]
+
+  width = 0.5
+  bars = [i for i in range(len(region_labels))]
+  for i in range(len(region_labels)):
+    if i==0:
[…]

### 2019-01-15 doc/large_strain.tex
Remove TODO from Roofline model
preliminary figures and text for breakdown of ocmputing times and Roofline for tensor4

@@
 {\color{red}
-Figure\todo{Improve design} \ref{fig:roofline} shows results of the Roofline performance model.
+Figure \ref{fig:roofline} shows results of the Roofline performance model.
 As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
@@
 
-Finally we note that all three algorithms are far from the theoretical peak performance which
-indicates that the origin for suboptimal results of matrix-free approach is most likely not related to the various flavours of algorithms numerically investigated here.
-Further detailed studies are necessary to explain this behavior.}
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.49\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_2d.pdf}
+      \caption{Roofline (2D)}
+      \label{fig:roofline_2d_tensor4}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_3d.pdf}
+    \caption{Roofline (3D)}
+    \label{fig:roofline_3d_tensor4}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[b]{0.49\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_stackedbar_2d.pdf}
+      \caption{Computing times (2D)}
+      \label{fig:breakdown_stackedbar_2d_tensor4}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_stackedbar_3d.pdf}
+    \caption{Computing times (2D)}
+    \label{fig:breakdown_stackedbar_3d_tensor4}
+  \end{subfigure}
+  \caption{Breakdown of computing times and Roofline analysis of various steps of Algorithm \ref{alg:mf_tensor4}.}%
+  \label{fig:breakdown}
+\end{figure}
+
+In order to have a better understanding of the behavior of the most promising ``tensor4'' implementation,
+we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
+(i) `read / write' denotes reading and writing from/to a global vector (step 1 and 13);
+(ii) `sum factorization' denotes application of sum factorization techniques (steps 4,5 and 12);
+(iii) `quadrature loop' denotes operations performed on each quadrature (steps 6 -- 11).
+It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that
+`read / write' steps takes negligible amount of time spent on each element.
+For higher order bases the majority of time is spent within the quadrature loop.
+When analyzing the Roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
+we observe that this part of the algorithm demonstrates rather low performance, which in our opinion is the source of the
+suboptimal performance of the whole matrix-free matrix-vector product (see Figure \ref{fig:roofline}).
+In this case the quadrature loop consists of, essentially, a
+single contraction between two second order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
+double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
+Further optimization of \texttt{Tensor} classes in \textit{deal.II} are required to
+improve the performance.
+Such studies, however, are beyond the scope of the current work.
+}
 
@@
 
+{\color{red}
+Based on the results presented in this work, we can conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with
+the geometric multigrid preconditioner is a very competitive solution strategy, as compared to more traditional matrix-based approach.}
+
 \ifijnme

### 2019-01-15 Calculations/post_process_likwid.py

@@ -261,7 +261,7 @@ leg = plt.legend(loc='upper left', ncol=1, labelspacing=0.1)
 # file location
 fig_prefix = os.path.join(os.getcwd(), '../doc/' + os.path.basename(os.path.normpath(prefix)) + '_')
 
-name = 'roofline_breakdown_{0}d.pdf' if args.breakdown else 'roofline_{0}d.pdf'
+name = 'roofline_{0}d.pdf'
 
 fig_file = fig_prefix + name.format(args.dim)
 
@@ -318,11 +318,11 @@ else:
 
   plt.legend(bars, region_labels, loc='upper left')
 
-  plt.ylabel('Walltime / number of elements')
+  plt.ylabel('wall time (s) / number of elements')
   plt.xticks(ind_plt, ind_str)
-  plt.xlabel("polynomial degree")
+  plt.xlabel('polynomial degree')
 
-  name = 'roofline_breakdown_stackedbar_{0}d.pdf'
+  name = 'stackedbar_{0}d.pdf'
   fig_file = fig_prefix + name.format(args.dim)
   print 'Saving figure in: {0}'.format(fig_file)
   plt.tight_layout()

### 2019-01-17 doc/large_strain.tex
clarify coarse level solver
minor fixes to the text

@@
 is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
-On the coarsest level we use a CG solver to reduce the residual by three orders of magnitude within (at most) one hundred steps.
+On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga1999}} to reduce the residual by three orders of magnitude.
 Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}
@@
     \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_stackedbar_3d.pdf}
-    \caption{Computing times (2D)}
+    \caption{Computing times (3D)}
     \label{fig:breakdown_stackedbar_3d_tensor4}
@@
 (i) `read / write' denotes reading and writing from/to a global vector (step 1 and 13);
-(ii) `sum factorization' denotes application of sum factorization techniques (steps 4,5 and 12);
+(ii) `sum factorization' denotes application of sum factorization techniques (steps 4, 5 and 12);
 (iii) `quadrature loop' denotes operations performed on each quadrature (steps 6 -- 11).

### 2019-01-17 doc/large_strain.tex
clarify current estimate of eigenvalues for coarse level

@@
 On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga1999}} to reduce the residual by three orders of magnitude.
+{\color{red}Maximum and minimum eigenvalues at this level are estimated from the CG solver set to reduce the residual by $10$ orders of magnitude within (at most) as many
+iterations as the coarse level problem size.\todo{remove if we do 30}}
 Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}

### 2019-01-17 include/mf_elasticity.h
add parameters for coarse level solver

@@ -362,6 +362,9 @@ namespace Cook_Membrane
       double       preconditioner_aggregation_threshold = 1e-4;
       unsigned int cond_number_cg_iterations            = 20;
       std::string  mf_caching                           = "scalar";
+      bool mf_coarse_chebyshev                          = true;
+      bool mf_coarse_chebyshev_accurate_eigenval        = true;
+      unsigned int mf_chebyshev_n_cg_iterations         = 30;
 
       void
       add_parameters(ParameterHandler &prm);
@@ -413,6 +416,23 @@ namespace Cook_Membrane
                           mf_caching,
                           "Type of caching for matrix-free operator",
                           Patterns::Selection("scalar|tensor2|tensor4"));
+
+        prm.add_parameter(
+          "MF Chebyshev number CG iterations",
+          mf_chebyshev_n_cg_iterations,
+          "Number of CG iterations to estiamte condition number "
+          "for Chebyshev smoother",
+          Patterns::Integer(2));
+
+        prm.add_parameter("MF Chebyshev coarse",
+                          mf_coarse_chebyshev,
+                          "Use Chebyshev smoother as coarse level solver");
+
+        prm.add_parameter(
+          "MF Chebyshev coarse accurate eigenvalues",
+          mf_coarse_chebyshev_accurate_eigenval,
+          "Accurately estimate eigenvalues for coarse level Chebyshev"
+          "solver");
       }
       prm.leave_subsection();
     }
@@ -1857,7 +1877,7 @@ namespace Cook_Membrane
       }
 
     // setup GMG preconditioner
-    const bool cheb_coarse = true;
+    const bool cheb_coarse = parameters.mf_coarse_chebyshev;
     {
       MGLevelObject<typename SmootherChebyshev::AdditionalData> smoother_data;
       smoother_data.resize(0, triangulation.n_global_levels() - 1);
@@ -1871,7 +1891,9 @@ namespace Cook_Membrane
               smoother_data[level].degree =
                 numbers::invalid_unsigned_int; // use as a solver
               smoother_data[level].eig_cg_n_iterations =
-                mg_mf_nh_operator[level].m();
+                (parameters.mf_coarse_chebyshev_accurate_eigenval ?
+                   mg_mf_nh_operator[level].m() :
+                   parameters.mf_chebyshev_n_cg_iterations);
             }
           else
             {
@@ -1881,7 +1903,8 @@ namespace Cook_Membrane
               // parameter is retrieved
[…]

### 2019-01-17 include/mf_elasticity.h
add configure option to do breakdown measurements with LIKWID within a cell

@@ -2076,7 +2076,11 @@ namespace Cook_Membrane
           constraints.set_zero(src_trilinos);
           copy_trilinos(src, src_trilinos);
 
+#ifdef WITH_BREAKDOWN
+          const unsigned int n_times = 1;
+#else
           const unsigned int n_times = 10;
+#endif
           MPI_Barrier(mpi_communicator);
           if (!parameters.skip_tangent_assembly)
             for (unsigned int i = 0; i < n_times; ++i)
@@ -2095,11 +2099,11 @@ namespace Cook_Membrane
           for (unsigned int i = 0; i < n_times; ++i)
             {
               TimerOutput::Scope t(timer, "vmult (MF)");
-#ifdef WITH_LIKWID
+#if defined(WITH_LIKWID) && !defined(WITH_BREAKDOWN)
               LIKWID_MARKER_START("vmult_MF");
 #endif
               mf_nh_operator.vmult(dst_mf, src);
-#ifdef WITH_LIKWID
+#if defined(WITH_LIKWID) && !defined(WITH_BREAKDOWN)
               LIKWID_MARKER_STOP("vmult_MF");
 #endif
             }

### 2019-01-17 include/mf_nh_operator.h

@@ -12,7 +12,12 @@
 
 #include <deal.II/multigrid/mg_constrained_dofs.h>
 
+#include <config.h>
 #include <material.h>
+
+#ifdef WITH_LIKWID
+#  include <likwid.h>
+#endif
 
 using namespace dealii;
 
@@ -514,6 +519,9 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::local_apply_cell(
   for (unsigned int cell = cell_range.first; cell < cell_range.second; ++cell)
     {
       // initialize on this cell
+#if defined(WITH_LIKWID) && defined(WITH_BREAKDOWN)
+      LIKWID_MARKER_START("vmult_reinit_read_write");
+#endif
       phi_current.reinit(cell);
       phi_current_s.reinit(cell);
 
@@ -527,10 +535,23 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::local_apply_cell(
           phi_reference.read_dof_values_plain(*displacement);
         }
 
+#if defined(WITH_LIKWID) && defined(WITH_BREAKDOWN)
+      LIKWID_MARKER_STOP("vmult_reinit_read_write");
+#endif
+
       do_operation_on_cell(phi_current, phi_current_s, phi_reference, cell);
+
+#if defined(WITH_LIKWID) && defined(WITH_BREAKDOWN)
+      LIKWID_MARKER_START("vmult_reinit_read_write");
+#endif
 
       phi_current.distribute_local_to_global(dst);
       phi_current_s.distribute_local_to_global(dst);
+
+#if defined(WITH_LIKWID) && defined(WITH_BREAKDOWN)
+      LIKWID_MARKER_STOP("vmult_reinit_read_write");
+#endif
+
     }
 }
 
@@ -668,11 +689,23 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::do_operation_on_cell(
   const number            mu           = cell_mat->mu;
   const number            lambda       = cell_mat->lambda;
 
+#if defined(WITH_LIKWID) && defined(WITH_BREAKDOWN)
+  LIKWID_MARKER_START("vmult_sum_factorization");
+#endif
+
   if (mf_caching == "scalar")
     phi_reference.evaluate(false, true, false);
 
   phi_current.evaluate(false, true, false);
   phi_current_s.evaluate(false, true, false);
+
[…]

### 2019-01-17 doc/large_strain.tex
fix bibtex reference to Varga

--- 2019-01-17 doc/large_strain.tex
@@
 is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
-On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga1999}} to reduce the residual by three orders of magnitude.
+On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
 Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}

### 2019-01-17 doc/large_strain.tex
update discussions around MF GMG to fit new results

@@
 we consider {\color{red}the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
-Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a very mild increase in the average number of CG iterations for different polynomial degrees
-{\color{red} for the GMG preconditioner.}
-On the other hand, the black-box AMG preconditioner requires many more iterations for convergence.
-{\color{red}
-Consequently, we can conclude that, although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a well behaving multigrid preconditioner for the here considered finite strain elasticity problem and that the considered smoother is suitable.
+Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
+{\color{red} noticeable} increase in the average number of CG iterations for different polynomial degrees
+{\color{red} for the GMG preconditioner.
+Although not reported in this section, we identified that
+the coarse level solver setup will greatly affect the performance of the preconditioner.
+In particular, we can improve the number of iterations by choosing a more accurate coarse level solver,
+however this leads to an overall more expensive linear solver.}
+The black-box AMG preconditioner
+{\color{red} demonstrate the same behavior of the number of CG iterations, however it}
+requires many more iterations for convergence.
+{\color{red}
+Consequently, we can conclude that, although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid for the here considered finite strain elasticity problem and that the considered smoother is suitable.
 }
@@
 Additionally, we compare solution times of the preconditioned iterative solvers.
-Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner,
-as well as to identify some bottlenecks within the former method that shall guide our further research in this direction.}
-Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for polynomial degrees up to $p=5$.
-For higher orders in 2D the wall-clock time {\color{red}per DoF} is comparable to the AMG preconditioner, see Figure \ref{fig:benchmark_miehe_Emmy_sol2}.
-The reason for this is the suboptimal coarse level solver within the GMG preconditioner.
-Evidently, the combination of the Chebyshev smoother with the CG solver at the coarsest mesh level has unsatisfactory performance for higher polynomial degrees.
-{\color{red}
-Consequently, the coarse level solver is the bottleneck of the here described geometric multigrid preconditioner.
-Its performance shall first be improved from a linear algebra perspective and only then implementational details of the complete preconditioner shall be studied.}
-One possible solution to this issue would be to perform further coarsening combined with the homogenization-based transfer operators \cite{Miehe2007}.
-Although untested in this work, we hypothesize that this should also improve the performance of the preconditioner for lower degrees.
-A valid alternative would be to use matrix-based AMG at the coarsest level.
+Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.}
+Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
+{\color{red}Importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
+On another hand, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
+Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
+Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for engendering problems
+with finite strain material models.
+Extension of this study from the node level to the cluster level is beyond the scope of this work,
+however we hypothesize that the matrix-free and multigrid frameworks provided by \texttt{deal.II} library would provide a competitive solution for that case as well.}
 
@@
 {\color{red}
-To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to such materials.}
+To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to partial differential equations arising in this case.}
 {\color{red}Among the three algorithms,}
@@
 The Roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
-Further studies are required to identify means to improve the computational intensity and performance of the matrix-free implementation.}
+We identified that one of the bottlenecks is related to the quadrature loop on each element where double contractions of fourth order and second order symmetric tensor
+as well as single contraction of two second order tensors is performed.
+Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by the \texttt{deal.II}.}
 
@@
 The GMG was applied to heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
-The numerical studies indicate that the proposed preconditioner is efficient from the perspective of linear algebra.
-{\color{red}
-Namely, the number of iterations in the iterative solver is almost independent on the polynomial degree of the finite element basis.
+The numerical studies indicate that the proposed preconditioner
+{\color{red}
+leads to much fewer iterations of the iterative solver as compared to the algebraic multigrid preconditioner.
 The multigrid matrix-free preconditioner also}
-leads to a solution approach that is faster than the matrix-based AMG for moderately high polynomial degrees (approximately up to order five).
-However, a considerable amount of time is spent in the coarse level solver of the GMG preconditioner,
-{\color{red}
-which is the main bottleneck of the here considered iterative solver.}
-Our future work will, therefore, be focused on extending the proposed matrix-free multigrid solution approach to homogenization based transfer operators \cite{Miehe2007}.
-We are also interested in extending these methods to near incompressible three-field formulations.
-
-{\color{red}
-Based on the results presented in this work, we can conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with
-the geometric multigrid preconditioner is a very competitive solution strategy, as compared to more traditional matrix-based approach.}
+leads to a solution approach that is faster than the matrix-based AMG {\color{red}for all polynomial degrees studied here.
+Importantly the solution wall time per degree of freedom is close to being constant.
+On the contrary, the matrix-based implementation with AMG becomes prohibitively expensive for higher order bases.
+We conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with
+the geometric multigrid preconditioner is a very competitive solution strategy, as compared to more traditional matrix-based approach.
+Our future work will be focused on adopting the proposed matrix-free multigrid solution approach to $FE^2$ homogenization
+as well studying its behavior on the cluster level as opposed to node level, considered here.}
 
@@
 
-D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for a fruitful discussion on matrix-free operator evaluation approaches.
+D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to {\color{red}Martin Kronbichler (TU Munich),} Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for a fruitful discussion on matrix-free operator evaluation approaches.
 

### 2019-01-17 Calculations/pre_process.py
python: add breakdown option to preprocessing script

@@ -18,6 +18,7 @@ parser.add_argument('--calc', metavar='calc', default='/home/woody/iwtm/iwtm108/
 parser.add_argument('--mpirun', metavar='mpirun', default='mpirun -np 20',
                     help='mpi run command with cores')
 parser.add_argument('--likwid', help='Prepare LIKWID run', action="store_true")
+parser.add_argument('--breakdown', help='LIKWID run with breakdown of costs', action="store_true")
 args = parser.parse_args()
 
 # parameters (list of tuples):
@@ -46,6 +47,14 @@ poly_quad_ref_dim_likwid = [
     (4,5,2,3),
 ]
 
+poly_quad_ref_dim_likwid_breakdown = [
+    (2,3,1,2),
+    (4,5,1,2),
+    (6,7,1,2),
+    (2,3,0,3),
+    (4,5,0,3),
+]
+
 # Solvers (type, preconditioner and caching)
 solvers = [
     ('MF_CG', 'gmg', 'scalar'),
@@ -61,6 +70,11 @@ solvers_likwid = [
     ('CG',    'amg', 'scalar'),
 ]
 
+solvers_likwid_breakdown = [
+    ('MF_CG', 'gmg', 'tensor4')
+]
+
+
 # MPI run command (override if use LIKWID)
 mpirun_args = args.mpirun
 if args.likwid:
@@ -72,6 +86,9 @@ mpicmd = mpirun_args + ' ' + args.prefix + 'main ' + args.calc + '{0}.prm 2>&1 |
 if args.likwid:
   solvers = solvers_likwid
   poly_quad_ref_dim = poly_quad_ref_dim_likwid
+  if args.breakdown:
+    solvers = solvers_likwid_breakdown
+    poly_quad_ref_dim = poly_quad_ref_dim_likwid_breakdown
 
 #
 # from here on the actual preprocessing:
@@ -146,6 +163,8 @@ if out_dir and not out_dir.endswith('/'):
 if args.likwid:
   base_name = 'likwid_' + base_name
   out_dir = 'LIKWID_' + out_dir
+  if args.breakdown:
+    out_dir = out_dir + '_breakdown'
 
 print 'base parameter file: {0}'.format(base_prm)
 print 'output directory:    {0}'.format(out_dir)

### 2019-01-18 Calculations/pre_process.py
python: fix output dir for breakdown preprocessing

@@ -156,15 +156,15 @@ base_prm = args.base_prm
 base_name = args.base_prm.split('.')[0]
 out_dir = args.dir
 
-if out_dir and not out_dir.endswith('/'):
-    out_dir = out_dir + '/'
-
 # add prefix to the base name
 if args.likwid:
   base_name = 'likwid_' + base_name
   out_dir = 'LIKWID_' + out_dir
   if args.breakdown:
     out_dir = out_dir + '_breakdown'
+
+if out_dir and not out_dir.endswith('/'):
+    out_dir = out_dir + '/'
 
 print 'base parameter file: {0}'.format(base_prm)
 print 'output directory:    {0}'.format(out_dir)

### 2019-01-18 Calculations/post_process_likwid.py
python: switch vmult() Roofline to log2

@@ -194,18 +194,26 @@ params = {'legend.fontsize': 14,
           'font.size': 20}
 plt.rcParams.update(params)
 
+plt.xscale('log', basex=2)
+plt.yscale('log', basey=2)
+
 # Roofline model
 # p = min (P, b I)
 # where I is measured intensity (Flops/byte)
 def Roofline(I,P,B):
     return np.array([min(P,B*i) for i in I])
 
-x = np.linspace(1./B, 10. if not args.breakdown else 100., num=500 if not args.breakdown else 5000)
-base = np.array([1./B for i in x])
+# log10:
+#x = np.linspace(1./B, 10. if not args.breakdown else 100., num=500 if not args.breakdown else 5000)
+#base = np.array([1./B for i in x])
+
+x = np.linspace(2**(-4), 2**4, num=500)
+base = np.array([x[0] for i in x])
+
 roofline_style = 'b-'
 peak = Roofline(x,P,B)
 # see https://github.com/matplotlib/matplotlib/issues/8623#issuecomment-304892552
-plt.loglog(x,peak, roofline_style, label=None, nonposx='clip', nonposy='clip')
+plt.plot(x,peak, roofline_style, label=None)
 
 # various ceilings (w/o FMA, w/o FMA and vectorization):
 for p_ in [P/2, P/2/4]:
@@ -236,7 +244,7 @@ for d in likwid_data:
     y = [f/1000]  # MFLOP->GFLOP
     style = d[4] + degree_to_points[d[1]]
     label = '{0} (p={1})'.format(d[3],d[1])
-    plt.loglog(x,y, style, label=label)
+    plt.plot(x,y, style, label=label)
     if 'MF' in label:
       mf_perf.append(y[0])
     else:
@@ -244,13 +252,18 @@ for d in likwid_data:
 
 plt.xlabel('intensity (Flop/byte)')
 plt.ylabel('performance (GFlop/s)')
-plt.ylim(top=300,bottom=1)
+#plt.ylim(top=300,bottom=1)
+plt.ylim(top=2**8,bottom=2**2)
+plt.xlim(right=2**4,left=(2**(-4)+0.01))
+plt.axes().set_aspect('equal', adjustable=None) #'datalim')
+plt.axes().yaxis.set_major_formatter(mp.ticker.ScalarFormatter())
[…]

### 2019-01-18 doc/large_strain.tex
python: update post-processing script to fit breakdown results with log2

@@
 Figure \ref{fig:roofline} shows results of the Roofline performance model.
-As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
+As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.4\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
 The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
-$22.1\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
+$22.0\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
 The achieved performance is about 13\% of the peak arithmetic performance. This is much lower than the

### 2019-01-18 Calculations/post_process_likwid.py

@@ -207,7 +207,7 @@ def Roofline(I,P,B):
 #x = np.linspace(1./B, 10. if not args.breakdown else 100., num=500 if not args.breakdown else 5000)
 #base = np.array([1./B for i in x])
 
-x = np.linspace(2**(-4), 2**4, num=500)
+x = np.linspace(2**(-5), 2**6+10, num=2000)
 base = np.array([x[0] for i in x])
 
 roofline_style = 'b-'
@@ -253,17 +253,20 @@ for d in likwid_data:
 plt.xlabel('intensity (Flop/byte)')
 plt.ylabel('performance (GFlop/s)')
 #plt.ylim(top=300,bottom=1)
-plt.ylim(top=2**8,bottom=2**2)
-plt.xlim(right=2**4,left=(2**(-4)+0.01))
+ymin = 2**2 if not args.breakdown else 2**(0) - 0.1
+plt.ylim(top=2**8,bottom=ymin)
+xmax = 2**6 + 10 if args.breakdown else 2**4
+xmin = 2**(-4)+0.01 if not args.breakdown else 2**(-5) + 1./80
+plt.xlim(right=xmax,left=xmin)
 plt.axes().set_aspect('equal', adjustable=None) #'datalim')
-plt.axes().yaxis.set_major_formatter(mp.ticker.ScalarFormatter())
-plt.axes().xaxis.set_major_formatter(mp.ticker.FuncFormatter(lambda x, pos: '{0}'.format(x) if x < 1.0 else '{0}'.format(int(round(x))) ))
+plt.axes().yaxis.set_major_formatter(mp.ticker.FuncFormatter(lambda x, pos: '{0}'.format(int(round(x))) ))
+plt.axes().xaxis.set_major_formatter(mp.ticker.FuncFormatter(lambda x, pos: '1/{0}'.format(int(round(1./x))) if x < 1.0 else '{0}'.format(int(round(x))) ))
 
-ang = 45 if not args.breakdown else 50
-y_pos = 13 if not args.breakdown else 7
-plt.text(2**(-4)+0.01, y_pos, 'B={:.1f} GB/s'.format(B), rotation=ang)
+ang = 45
+y_pos = 13
+plt.text(xmin, y_pos, 'B={:.1f} GB/s'.format(B), rotation=ang)
 
-x_pos = 4 if not args.breakdown else 25
+x_pos = 4 if not args.breakdown else 16
[…]

### 2019-01-18 Calculations/post_process_likwid.py
minor comment in python script

@@ -308,7 +308,7 @@ else:
   for d in likwid_data:
     if d[0] == args.dim:
       r_idx = d[6] # region index
-      t = d[5][3] / d[7] # runtime unhalted per cell
+      t = d[5][3] / d[7] # 3rd element in Sections - Runtime unhalted, report per cell
       p = d[1]
       for idx, s in enumerate(ind):
         if s == p:

### 2019-01-18 doc/large_strain.tex
mention that breakdown is done on smaller meshes
remove a couple of todo

@@
 Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
-configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{rerun RRZE and IWR}\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}
+configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{rerun IWR}\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}
 
@@
 second order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
-As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} \todo{rerun and check}),
+As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}),
 however the ``scalar'' algorithm performs a large number of repetitive operations that are
@@
 (iii) `quadrature loop' denotes operations performed on each quadrature (steps 6 -- 11).
+Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one
+global refinement in 2D and no global mesh refinement in 3D.
 It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that

### 2019-01-28 doc/large_strain.tex

@@
 (ii) `sum factorization' denotes application of sum factorization techniques (steps 4, 5 and 12);
-(iii) `quadrature loop' denotes operations performed on each quadrature (steps 6 -- 11).
+(iii) `quadrature loop' denotes operations performed on each quadrature (steps \mbox{6 -- 11}).
 Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one

### 2019-01-31 Calculations/__init__.py
python: wrote a parser for LIKWID

@@ -1,0 +1,1 @@
+# dummy file to let Python import from this directory

### 2019-01-31 Calculations/utilities.py

@@ -1,0 +1,118 @@
+import os
+import fileinput
+import re
+
+def remove_creation_date(file_name):
+    '''remove creationg data from .eps file in place'''
+    for line in fileinput.input(file_name, inplace=True):
+        if not 'CreationDate' in line:
+            print line,
+
+def collection_toutput_files(files, prefix):
+    '''starting from prefix, collection full path to all files that
+    end with .toutput'''
+    for root, dirs, files_ in os.walk(prefix):
+        for f in files_:
+            if f.endswith(".toutput"):
+                files.append(os.path.join(root, f))
+
+def get_regex_pattern():
+    return r'[+\-]?(?:[0-9]\d*)(?:\.\d*)?(?:[eE][+\-]?\d+)?'
+
+
+def parse_likwid_file(filename, last_line = ''):
+    '''Parse terminal output of LIKWID
+
+    Arguments:
+    filename -- the path to the file
+    last_line -- the last line of non-LIKWID output. Starting from the line after this
+        the parser will be active
+
+    Return:
+    parsed data stored as:
+        dictionary (region)
+        dictionary (table name i.e. Event, Event_Sum, Metric, Metric_Sum or alike
+        dictionary of rows. That is, one can get the row as
+        result['vmult']['Metric_Sum']['MFLOP/s']
+    '''
+    result = {}
+    fin = open(filename, 'r')
+
+    row_separator = '---------'
+
+    found_start = (last_line == "")
+    region = ''
+    separator_counter = 0
+    table_name = ''
+    table_ind = 0
+
+    for line in fin:
+        line = line.strip()
+
+        # skip empty lines
+        if line == "":
+            continue
+
+        if found_start:
+            #
+            # Main logic to parse
+            #
+
+            # Check if we found one of the regions:
+            if 'Region:' in line:
+                region = line[8:]
+                print '-- Region: {0}'.format(region)
+                # regions should be unique
+                assert region not in result
+                result[region] = {}
+                separator_counter = 0
+                table_name = ''
+                continue
+
+            if 'Group:' in line:
+                # FIXME: read in groups as well?
+                continue
+
+            # If we are in LIKWID part, we should have some region always around
+            assert region != ''
+
+            # At this point we have only 3 options: we are on one of the separators,
+            # inside the header or inside the core of the table.
+            if row_separator in line:
+                separator_counter = separator_counter + 1
+                # reset the counter if we are at the end of the current table
+                if separator_counter == 3:
+                    separator_counter = 0
[…]

### 2019-01-31 Calculations/post_process_likwid_simd.py
python: start writing a script to get data for SIMD/MPI table

@@ -1,0 +1,72 @@
+import argparse
+import numpy as np
+
+# own functions
+from utilities import *
+
+# define command line arguments
+parser = argparse.ArgumentParser(
+    description='Post-Process timing/memory info and create a table.',
+    formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+parser.add_argument('--prefix', metavar='prefix', default='LIKWID_Emmy_RRZE',
+                    help='A folder prefix to look for benchmark results')
+parser.add_argument('--dim', metavar='dim', default=2, type=int,
+                    help='Dimension (2 or 3)')
+
+args = parser.parse_args()
+
+
+# serial | MPI | SIMD | MPI+SIMD
+suffixes = [
+    '_1proc_novec',
+    '_novec',
+    '_1proc',
+    ''
+]
+
+prefix = args.prefix if args.prefix.startswith('/') else os.path.join(os.getcwd(), args.prefix)
+
+# We will generate a table similar to Table 1 in Kronbichler 2012
+table_data = []
+
+# Go through all the suffixes and parse the data
+for idx, s in enumerate(suffixes):
+    this_prefix = prefix + s
+    files = collection_toutput_files(prefix + s)
+
+    print 'Gather data from {0}'.format(this_prefix)
+    print 'found {0} files'.format(len(files))
+
+    # no go through the files:
+    for f in files:
+        # guess parameters from the file name:
+        fname = os.path.basename(f)
+        strings = fname.split('_')
+        dim = int(re.findall(get_regex_pattern(),strings[2])[0])
+        p   = int(re.findall(get_regex_pattern(),strings[3])[0])
+        q   = int(re.findall(get_regex_pattern(),strings[3])[1])
+
+        # we are only interested in tensor2 (Algorithm 3) for the currently chosen dimension
+        if not (dim == args.dim and '_tensor4' in fname):
+            continue
+
+        print 'dim={0} p={1} q={2} file={3}'.format(dim,p,q,fname)
+
+        # these files contain a single region
+        result = parse_likwid_file(f,'LIKWID_MARKER_CLOSE')['vmult_MF']
+
+        # depending on the run, the LIKWID output may not have Sum table
+        if 'Metric_Sum' not in result:
+            data = result['Metric']
+            flops = data['MFLOP/s'][0]
+            runtime = data['Runtime (RDTSC) [s]'][0]
+        else:
+            data = result['Metric_Sum']
[…]

### 2019-01-31 Calculations/utilities.py

@@ -8,13 +8,16 @@ def remove_creation_date(file_name):
         if not 'CreationDate' in line:
             print line,
 
-def collection_toutput_files(files, prefix):
+def collection_toutput_files(prefix):
     '''starting from prefix, collection full path to all files that
     end with .toutput'''
+    files = []
     for root, dirs, files_ in os.walk(prefix):
         for f in files_:
             if f.endswith(".toutput"):
                 files.append(os.path.join(root, f))
+
+    return files
 
 def get_regex_pattern():
     return r'[+\-]?(?:[0-9]\d*)(?:\.\d*)?(?:[eE][+\-]?\d+)?'
@@ -37,6 +40,7 @@ def parse_likwid_file(filename, last_line = ''):
     '''
     result = {}
     fin = open(filename, 'r')
+    debug_output = False
 
     row_separator = '---------'
 
@@ -61,7 +65,8 @@ def parse_likwid_file(filename, last_line = ''):
             # Check if we found one of the regions:
             if 'Region:' in line:
                 region = line[8:]
-                print '-- Region: {0}'.format(region)
+                if debug_output:
+                    print '-- Region: {0}'.format(region)
                 # regions should be unique
                 assert region not in result
                 result[region] = {}
@@ -92,10 +97,12 @@ def parse_likwid_file(filename, last_line = ''):
                 table_name = columns[0]
                 if 'Sum' in line and 'Min' in line and 'Max' in line and 'Avg' in line:
                     table_name = table_name + '_Sum'
-                print '   Name  : {0}'.format(table_name)
+
                 # table names for a given region should be unique
                 assert table_name not in result[region]
                 result[region][table_name] = {}
+                if debug_output:
+                    print '   Name  : {0}'.format(table_name)
             else:
                 # we should have table name around already
                 assert table_name != ''
@@ -107,8 +114,9 @@ def parse_likwid_file(filename, last_line = ''):
                 val = columns[1:]
 
                 # finally put the data
-                print '      {0}'.format(key)
                 result[region][table_name][key] = val
+                if debug_output:
+                    print '      {0}'.format(key)
         else:
             # If we have not found LIKWID part yet
             if last_line in line:

### 2019-01-31 Calculations/post_process_likwid_simd.py
working table with MPI/SIMD speedups

@@ -24,10 +24,17 @@ suffixes = [
     ''
 ]
 
+# FIXME: hard-code polynomial orders
+poly_to_row = {
+    2 : 0,
+    4 : 1,
+    6 : 2
+}
+
 prefix = args.prefix if args.prefix.startswith('/') else os.path.join(os.getcwd(), args.prefix)
 
 # We will generate a table similar to Table 1 in Kronbichler 2012
-table_data = []
+table_data = [ [ np.nan for i in range(12)] for i in range(3)]
 
 # Go through all the suffixes and parse the data
 for idx, s in enumerate(suffixes):
@@ -65,8 +72,55 @@ for idx, s in enumerate(suffixes):
             flops = data['MFLOP/s STAT'][0]                # take Sum
             runtime = data['Runtime (RDTSC) [s] STAT'][2]  # take Max
 
+        runtime = float(runtime)
+        flops = float(flops) / 1000  # MFLOP/s -> GFLOP/s
+
         print '  {0}'.format(flops)
         print '  {0}'.format(runtime)
 
+        row = poly_to_row[p]
+        # first column
+        col = 1 if idx==0 else 3 * idx
+        # fill in the table
+        table_data[row][0] = p
+        table_data[row][col] = runtime
+        table_data[row][col+1] = flops
+        if idx > 0:
+            # FIXME: assume that fully serial runs are already in the table
+            table_data[row][col+2] = table_data[row][1] / runtime
 
 
+# Finally write out a latex multicolumn table
+file_name = os.path.join(os.getcwd(), '../doc/parallelization.tex')
+print 'Saving table in ' + file_name
+
+with open(file_name, 'w') as f:
+    # start with the header
+    f.write("""\
+\\begin{table}
+\centering
+\\begin{tabular}{c|cc|ccc|ccc|ccc}
+\hline
+              & \multicolumn{2}{c|}{serial} & \multicolumn{3}{c|}{MPI} & \multicolumn{3}{c|}{SIMD} & \multicolumn{3}{c|}{MPI+SIMD}  \\\\
+\hline
+p             & time  & GFlop/s              & time & GFlop/s & speedup & time & GFlop/s & speedup & time & GFlop/s & speedup \\\\
+\hline
+""")
+
+    # now print the gathered data:
+    for row in table_data:
+        line = '{0}'.format(int(row[0]))
+        for i in range(1, len(row)):
[…]

### 2019-01-31 Calculations/post_process_likwid_simd.py
plot 3D speedup table

@@ -34,7 +34,8 @@ poly_to_row = {
 prefix = args.prefix if args.prefix.startswith('/') else os.path.join(os.getcwd(), args.prefix)
 
 # We will generate a table similar to Table 1 in Kronbichler 2012
-table_data = [ [ np.nan for i in range(12)] for i in range(3)]
+n_rows = 3 if args.dim == 2 else 2
+table_data = [ [ np.nan for i in range(12)] for i in range(n_rows)]
 
 # Go through all the suffixes and parse the data
 for idx, s in enumerate(suffixes):
@@ -89,9 +90,8 @@ for idx, s in enumerate(suffixes):
             # FIXME: assume that fully serial runs are already in the table
             table_data[row][col+2] = table_data[row][1] / runtime
 
-
 # Finally write out a latex multicolumn table
-file_name = os.path.join(os.getcwd(), '../doc/parallelization.tex')
+file_name = os.path.join(os.getcwd(), '../doc/parallelization_{0}d.tex'.format(args.dim))
 print 'Saving table in ' + file_name
 
 with open(file_name, 'w') as f:
@@ -119,8 +119,8 @@ p             & time  & GFlop/s              & time & GFlop/s & speedup & time &
     f.write("""\
 \hline
 \end{tabular}
-\caption{Wall-clock time in seconds and performance in GFlops of tensor4 algorithm for various combinations of polynomial degrees,
+\caption{Wall-clock time in seconds and performance in GFlops of Algorithm \\ref{alg:mf_tensor4} in 2D for various combinations of polynomial degrees,
 vectorization and parallelization.}
-\label{tab:numbers}
+\label{tab:numbers_2d}
 \end{table}
 """)

### 2019-01-31 Calculations/post_process_likwid_simd.py
divide time by 10 (we do 10 vmults)

@@ -73,7 +73,7 @@ for idx, s in enumerate(suffixes):
             flops = data['MFLOP/s STAT'][0]                # take Sum
             runtime = data['Runtime (RDTSC) [s] STAT'][2]  # take Max
 
-        runtime = float(runtime)
+        runtime = float(runtime) / 10  # FIXME: hard-code that we run 10 times
         flops = float(flops) / 1000  # MFLOP/s -> GFLOP/s
 
         print '  {0}'.format(flops)

### 2019-01-31 Calculations/post_process_likwid_simd.py
cleanup MPI/SIMD tables

@@ -99,7 +99,8 @@ with open(file_name, 'w') as f:
     f.write("""\
 \\begin{table}
 \centering
-\\begin{tabular}{c|cc|ccc|ccc|ccc}
+\\resizebox{\\textwidth}{!}{
+\\begin{tabular}{|c|cc|ccc|ccc|ccc|}
 \hline
               & \multicolumn{2}{c|}{serial} & \multicolumn{3}{c|}{MPI} & \multicolumn{3}{c|}{SIMD} & \multicolumn{3}{c|}{MPI+SIMD}  \\\\
 \hline
@@ -119,6 +120,7 @@ p             & time  & GFlop/s              & time & GFlop/s & speedup & time &
     f.write("""\
 \hline
 \end{tabular}
+}
 \caption{Wall-clock time in seconds and performance in GFlops of Algorithm \\ref{alg:mf_tensor4} in 2D for various combinations of polynomial degrees,
 vectorization and parallelization.}
 \label{tab:numbers_2d}

### 2019-01-31 doc/large_strain.tex
make GMG MPI picture consistent with fine level one (use 4 procs in both)
working table with MPI/SIMD speedups
plot 3D speedup table
cleanup MPI/SIMD tables
fixed precision=3 for tables
remove obsolete todo

@@
 
+\usepackage{pgfkeys,pgfmath,pgfcore}
+% \pgfkeys{/pgf/number format/.cd,1000 sep={}}
+\pgfkeys{/pgf/number format/fixed,/pgf/number format/precision=3}
+
 %
@@
     \centering
-    \includegraphics[width=\textwidth]{gmg_2d.png}
+    \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
     % \caption{2D multigrid mesh with 2 global refinements}
   \end{subfigure}
-  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
+  \caption{2D multigrid mesh after two global refinements distributed into four MPI processes. The color indicates the process owning the element.}%
   \label{fig:miehe_gmg}
@@
 Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
-configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{rerun IWR}\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}
+configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}
 
@@
 \end{figure}
+
+\todo{Discuss MPI/SIMD}
+\input{parallelization_2d.tex}
+\input{parallelization_3d.tex}
 

### 2019-01-31 doc/large_strain.tex
Temp commit

@@
 \newcommand{\diff}{\mathop{}\!\mathrm{d}}
+
+\newcommand{\changeJP}[1]{\textcolor{blue}{#1}}
 
@@
 
-\subsection{Constitutive modelling}
+\subsection{Constitutive modelling
+\label{sec: Constitutive modelling}
+}
 
@@
 
+\section{\changeJP{Description of numerical framework}}
+
+The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
+This library offers a number of paradigms by which to parallelize a finite-element code.
+In this work we employ \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
+%
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.8\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{fine_level_2d.png}
+    % \caption{2D multigrid mesh with 2 global refinements}
+  \end{subfigure}
+  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
+  \label{fig:miehe_fine_level}
+\end{figure}
+%
+%
+Figure \ref{fig:miehe_fine_level} illustrates a distributed MPI decomposition on the fine mesh level for the 2D problem that will be introduced later in Section \ref{sec:example}.
+
+In practise this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of discussion in this manuscript, has been achieved using MPI.
+
+The results of the matrix-free approach with a geometric-multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
+
 \section{Matrix-free operator evaluation}
@@
 Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
-For more information about the implementation details and data structures for sum factorization approaches in deal.II (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and shared memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
+For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
 }
@@
     \centering
-    \includegraphics[width=\textwidth]{fine_level_2d.png}
-    % \caption{2D multigrid mesh with 2 global refinements}
-  \end{subfigure}
-  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
-  \label{fig:miehe_fine_level}
-\end{figure}
-
-\begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.8\textwidth}
-    \centering
     \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
@@
 %
-{\color{red}
-MPI parallelization is achieved using the standard domain decomposition, where each MPI process ``owns'' only a subset of elements.
-To that end, \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library is used.
-Figure \ref{fig:miehe_fine_level} illustrates such decomposition on the fine mesh level for the 2D problem.
-The same approach is adopted on the multigrid levels.}
-Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.
-{\color{red}
-Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.
-}
-
+\changeJP{The same distributed approach to mesh decomposition (as most evident on the finest level) is adopted on the multigrid levels.}
+Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements \changeJP{for the mesh introduced in figure \ref{fig:miehe_fine_level}}.
+{\color{red}
+Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.}
 \begin{table}
@@
 %
-The results of the matrix-free approach with a geometric-multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
+The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library.
 The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
-Computations are done on a single node of two clusters:
+
+Computations \changeJP{were performed} on a single node of two clusters:
 A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM.

### 2019-01-31 doc/large_strain.tex
Temp commit

@@
 memory bandwidth\footnote{As measured with the \textit{load\_avx} benchmark of \textit{likwid-bench} \cite{likwid}.} and peak performance are $B=47.2 \,\rm{GB/s}$ and $P=176\, \rm{GFlop/s}$, respectively.
+Turbo mode of CPU was disabled.
 }
@@
 Figure \ref{fig:roofline} shows results of the Roofline performance model.
-As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.4\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
+As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $\pgfmathprintnumber{8.2093245}\,\rm{GFlop/s}$ (4.6\% of the peak performance) in 2D.
 The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
-$22.0\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
-The achieved performance is about 13\% of the peak arithmetic performance. This is much lower than the
+$\pgfmathprintnumber{18.8047083222}\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
+The achieved performance is about 10.6\% of the peak arithmetic performance. This is much lower than the
 70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.

### 2019-01-31 doc/large_strain.tex
Finish basic outline of sec 3.

@@
 
-In practise this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of discussion in this manuscript, has been achieved using MPI.
-
-The results of the matrix-free approach with a geometric-multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
+In practise this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
+For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
+For the matrix-based approach, the Trilinos \cite{Heroux2005} library, version 12.12.1 was leveraged; 
+the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the package ML package \cite{Gee2006a}.
+The second implementation, which is discussed in detail in the following Sections, utilized a matrix-free approach in conjunction with a geometric-multigrid preconditioner.
 

### 2019-01-31 doc/large_strain.tex
Finish basic outline of sec 3.

@@
 memory bandwidth\footnote{As measured with the \textit{load\_avx} benchmark of \textit{likwid-bench} \cite{likwid}.} and peak performance are $B=47.2 \,\rm{GB/s}$ and $P=176\, \rm{GFlop/s}$, respectively.
-Turbo mode of CPU was disabled.
 }
@@
 Figure \ref{fig:roofline} shows results of the Roofline performance model.
-As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $\pgfmathprintnumber{8.2093245}\,\rm{GFlop/s}$ (4.6\% of the peak performance) in 2D.
+As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.4\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
 The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
-$\pgfmathprintnumber{18.8047083222}\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
-The achieved performance is about 10.6\% of the peak arithmetic performance. This is much lower than the
+$22.0\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
+The achieved performance is about 13\% of the peak arithmetic performance. This is much lower than the
 70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.

### 2019-02-01 doc/large_strain.tex
Update abstract

@@
 
-    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyperelastic finite element tangent operator numerically.
+    \changeJP{To date, there have been few assessments into the applicability of the matrix-free approach to problems in solid mechanics.
+    In this work, we perform an initial investigation on the application of the matrix-free approach to problems in quasi-static finite-strain hyperelasticity to determine whether it is viable for further extension.
+    Specifically, we study different numerical implementations of the finite element tangent operator, and determine whether generalized methods of incorporating complex constitutive behavior might be feasible.
+    }
     In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
@@
     The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner {\color{red} on a single node} for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
+    \changeJP{We conclude that the application of matrix-free methods to finite-strain solid mechanics is promising, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
   \end{abstract}

### 2019-02-01 doc/large_strain.tex
Add todo note for JP
Touch up abstract
Add more details in introduction

@@
 \newcommand{\changeJP}[1]{\textcolor{blue}{#1}}
+\newcommand{\todojp}[1]{\todo[inline,color=red!40,size=\small]{TODO (J-P): #1}}
 
@@
 
-    \changeJP{To date, there have been few assessments into the applicability of the matrix-free approach to problems in solid mechanics.
+    \changeJP{To date, there have been few (if any) assessments into the applicability of the matrix-free approach to problems in solid mechanics.
     In this work, we perform an initial investigation on the application of the matrix-free approach to problems in quasi-static finite-strain hyperelasticity to determine whether it is viable for further extension.
@@
 To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
-We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent master's thesis produced by a student of the first two of the authors \cite{Mentler2017},
-{\color{red} which is also limited to small strain linear elasticity.}\todo{omit?}
+We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018}.
 One possible reason is that the tangent operators\footnote{Here and
@@
 whereas higher order elements are rarely employed.
-This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements\todo{reference?}.
+This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements\todo{reference? Martin's papers?}.
 However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
@@
 
-In this work, we investigate different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost
-and performance {\color{red} on a single node of an higher performance computing cluster.}
+In this work, we perform \changeJP{a preliminary investigation of} different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost
+and performance {\color{red} on a single node of a high performance computing cluster.}
+\changeJP{
+The \texttt{deal.II} \cite{dealII90} finite element library, which provides MPI parallelization together with a \changeJP{generalized matrix-free framework that incorporates SIMD vectorization} and an interface to high-performance, MPI distributed matrix-based libraries, is used for this study.
+}
 In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
 and employ them to define a \mbox{geometric} multigrid preconditioner.
-MPI parallelization together with SIMD vectorization
-{\color{red}
-are provided by the \texttt{deal.II} \cite{dealII90} finite element library, used for this study.
-}
 
 The paper is organized as follows: In Section \ref{sec:theory} we briefly introduce the partial differential equations used in  finite-strain solid mechanics. 
-Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the different matrix-free operator evaluation approaches.
-In Section \ref{sec:gmg} we propose a geometric multigrid preconditioner suitable for finite-strain hyperelasticity.
+\changeJP{
+Sections \ref{sec:fe} and \ref{sec: Description of numerical framework} cover the finite element discretization and provide details of the numerical framework with which this study is implemented.
+In \ref{sec:mf} we describe the different matrix-free operator evaluation approaches that are evaluated in this study, and in section \ref{sec:gmg} we propose a geometric multigrid preconditioner that is suitable for finite-strain hyperelasticity.
+}
 The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}.
-The results are summarized in Section \ref{sec:summary}.
+\changeJP{Lastly, the} results are summarized in Section \ref{sec:summary}.
 
@@
 
-\section{\changeJP{Description of numerical framework}}
+\section{\changeJP{Description of numerical framework}
+\label{sec: Description of numerical framework}
+}
 

### 2019-02-01 doc/large_strain.tex
Add extra detail to section on implementation
Add indentations and formatting to algorithms

@@
 The second implementation, which is discussed in detail in the following Sections, utilized a matrix-free approach in conjunction with a geometric-multigrid preconditioner.
+As will be detailed later, the pre-existing matrix-free framework that was used offers further opportunity for low-level parallelism that may be exploited.
 
@@
         {\color{red}extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ } \;
-        evaluate the following tensors at each quadrature point using sum factorization \;
-        $\Grad \, \overline{\gz u^h}_K$ \tcp*{2nd order}
-        $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
-        $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
+        evaluate the following tensors at each quadrature point using sum factorization :\\
+        \enskip $\Grad \, \overline{\gz u^h}_K$ \tcp*{2nd order}
+        \enskip $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
+        \enskip $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{
@@
         {\color{red}extract local vector values on this element $\gz x_K$ } \;
-        evaluate the following tensors at each quadrature point using sum factorization \;
-        $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
-        $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
+        evaluate the following tensors at each quadrature point using sum factorization :\\
+        \enskip$\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
+        \enskip$\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{

### 2019-02-01 benchmark/tensors.cc
add a benchmark to do tensor products

@@ -1,0 +1,117 @@
+
+// benchmark double dot product between symmetric tensors and single dot
+// between two second order tensors
+
+#include <deal.II/base/symmetric_tensor.h>
+#include <deal.II/base/table.h>
+#include <deal.II/base/tensor.h>
+#include <deal.II/base/timer.h>
+#include <deal.II/base/vectorization.h>
+
+#include <deal.II/physics/elasticity/standard_tensors.h>
+
+#include <config.h>
+
+#ifdef WITH_LIKWID
+#  include <likwid.h>
+#else
+// if build without LIKWID, keep dummy definitions to avoid writing #ifdef...
+// all over the place
+#  define LIKWID_MARKER_INIT
+#  define LIKWID_MARKER_THREADINIT
+#  define LIKWID_MARKER_SWITCH
+#  define LIKWID_MARKER_REGISTER(regionTag)
+#  define LIKWID_MARKER_START(regionTag)
+#  define LIKWID_MARKER_STOP(regionTag)
+#  define LIKWID_MARKER_CLOSE
+#  define LIKWID_MARKER_GET(regionTag, nevents, events, time, count)
+#endif
+
+using namespace dealii;
+
+
+template <int dim = 2, int degree = 4, typename number = double>
+void
+test(const unsigned int n_cells = 22528) // 90112 active cells in 2D for p=4 =>
+                                         // num of cell batches ~ 90112/4
+{
+  constexpr int n_q_points = Utilities::pow(degree + 1, dim);
+  std::cout << "dim                = " << dim << std::endl
+            << "degree             = " << degree << std::endl
+            << "SIMD width         = "
+            << VectorizedArray<number>::n_array_elements << std::endl
+            << "n_cell_batches     = " << n_cells << std::endl
+            << "n_q_points         = " << n_q_points << std::endl
+            << "number of products = " << n_cells * n_q_points << std::endl;
+
+  Table<2, Tensor<2, dim, VectorizedArray<number>>>          cached_tensor2;
+  Table<2, SymmetricTensor<4, dim, VectorizedArray<number>>> cached_tensor4;
+
+  cached_tensor2.reinit(n_cells, n_q_points);
+  cached_tensor4.reinit(n_cells, n_q_points);
+
+  // set tensors to something (does not matter)
+  const VectorizedArray<number> one = make_vectorized_array(1.);
[…]

### 2019-02-01 doc/large_strain.tex
Changes to MF section

@@
 {\color{red}
-that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem,}
+that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem \changeJP{derived with a Lagrangian formulation},}
 we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
@@
 {\color{red}
-Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
+Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach \changeJP{, each of which have a different level of abstraction, algorithmic complexity, and memory requirements}.
+\changeJP{They are introduced in order of lowest memory footprint to largest:}
 }
@@
 {\color{red}
-The main reason is that the operations are typically the same on all elements.
-For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{For the heterogeneous materials considered here,
+The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements\footnote{%
+\changeJP{In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
+This motivates the caching strategy employed in Algorithm 3, as the pre-caching of the chosen data ensures a context in which SIMD parallelization of subsequent operations is guaranteed.
+Algorithms 1 and 2 sacrifice some of this generality for decreased memory requirements.}
+}.
+For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{%
+For the heterogeneous materials considered here,
 the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.

### 2019-02-01 benchmark/tensors.cc
output one more convinience scalar

@@ -42,7 +42,11 @@ test(const unsigned int n_cells = 22528) // 90112 active cells in 2D for p=4 =>
             << VectorizedArray<number>::n_array_elements << std::endl
             << "n_cell_batches     = " << n_cells << std::endl
             << "n_q_points         = " << n_q_points << std::endl
-            << "number of products = " << n_cells * n_q_points << std::endl;
+            << "number of products = " << n_cells * n_q_points << std::endl
+            << "multiplier         = " // Runtime * MFLOP/s * multiplier = FLOP for a single operation
+            << 1000000. / (n_cells * n_q_points *
+                           VectorizedArray<number>::n_array_elements)
+            << std::endl;
 
   Table<2, Tensor<2, dim, VectorizedArray<number>>>          cached_tensor2;
   Table<2, SymmetricTensor<4, dim, VectorizedArray<number>>> cached_tensor4;

### 2019-02-01 benchmark/tensors.cc
implement various tensor2 contractions

@@ -43,23 +43,30 @@ test(const unsigned int n_cells = 22528) // 90112 active cells in 2D for p=4 =>
             << "n_cell_batches     = " << n_cells << std::endl
             << "n_q_points         = " << n_q_points << std::endl
             << "number of products = " << n_cells * n_q_points << std::endl
-            << "multiplier         = " // Runtime * MFLOP/s * multiplier = FLOP for a single operation
-            << 1000000. / (n_cells * n_q_points *
-                           VectorizedArray<number>::n_array_elements)
+            << "multiplier         = "
+            << 1000000. /
+                 (n_cells * n_q_points) // Runtime * MFLOP/s * multiplier = FLOP
+                                        // for a single operation
             << std::endl;
 
   Table<2, Tensor<2, dim, VectorizedArray<number>>>          cached_tensor2;
+  Table<2, SymmetricTensor<2, dim, VectorizedArray<number>>> cached_tensor2sym;
   Table<2, SymmetricTensor<4, dim, VectorizedArray<number>>> cached_tensor4;
 
   cached_tensor2.reinit(n_cells, n_q_points);
   cached_tensor4.reinit(n_cells, n_q_points);
+  cached_tensor2sym.reinit(n_cells, n_q_points);
 
   // set tensors to something (does not matter)
   const VectorizedArray<number> one = make_vectorized_array(1.);
+  const VectorizedArray<number> zero = make_vectorized_array(0.);
+
   for (unsigned int cell = 0; cell < n_cells; ++cell)
     for (unsigned int q = 0; q < n_q_points; ++q)
       {
         cached_tensor2(cell, q) =
+          one * Physics::Elasticity::StandardTensors<dim>::I;
+        cached_tensor2sym(cell, q) =
           one * Physics::Elasticity::StandardTensors<dim>::I;
         cached_tensor4(cell, q) =
           one * Physics::Elasticity::StandardTensors<dim>::S +
@@ -95,6 +102,8 @@ test(const unsigned int n_cells = 22528) // 90112 active cells in 2D for p=4 =>
   timer.leave_subsection();
 
   // now test 2-nd order tensors
+
+  // OPTION1 : operator*
   timer.enter_subsection("tensor2");
   LIKWID_MARKER_START("tensor2");
[…]

### 2019-02-01 doc/large_strain.tex
Move comment related to implementation details of GMG
Fiddle with intro to numerical examples; expanded on some points

@@
 For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
+\changeJP{Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.}
 
@@
 The relative convergence criteria for the linear solver is $10^{-6}$.
-{\color{red}
-The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
-by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
-The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
-is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
-On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
-Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}
-
+%
 Figure \ref{fig:miehe_deformed} shows deformed meshes at the final loading step with quadratic elements and two global mesh refinements.
-
+%
 \begin{figure}[!ht]
@@
 \end{figure}
-
+%
 \begin{figure}[!ht]
@@
 Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements \changeJP{for the mesh introduced in figure \ref{fig:miehe_fine_level}}.
-{\color{red}
-Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.}
+%
 \begin{table}
@@
 
-{\color{red}In this study, we limit ourselves to a comparison on a node level, which should already indicate the differences between the matrix-free and matrix-based approaches.}
+\changeJP{%
+Aligned with the approach taken in similar previous investigations \cite{}\todo{Cite Martin's papers?}, in this study we limit ourselves to comparisons made on a node level.
+Examinations performed in this manner should already indicate the differences between the matrix-free and matrix-based approaches, and give some preliminary understanding of the scalability potential of the numerical implementation without confounding factors related to data exchange though interconnects via MPI.
+}
 The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
 {\color{red}
-We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach.
-On the other hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we can observe that the matrix-based approach is memory bound.
-}
+We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach \changeJP{with a high polynomial order finite element}.
+\changeJP{Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound}.
+}
+
+%
+% Below are are implementational details that might affect the comparative result
 %
@@
 The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
+{\color{red}
+The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
+by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
+The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
+is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
+On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
+Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}
 
@@
 For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
-{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. fully occupied node.}
+{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. \changeJP{one} fully occupied node.}
 

### 2019-02-01 benchmark/tensors.cc
print version info

@@ -2,6 +2,9 @@
 // benchmark double dot product between symmetric tensors and single dot
 // between two second order tensors
 
+#include <deal.II/base/config.h>
+
+#include <deal.II/base/revision.h>
 #include <deal.II/base/symmetric_tensor.h>
 #include <deal.II/base/table.h>
 #include <deal.II/base/tensor.h>
@@ -11,6 +14,7 @@
 #include <deal.II/physics/elasticity/standard_tensors.h>
 
 #include <config.h>
+#include <version.h>
 
 #ifdef WITH_LIKWID
 #  include <likwid.h>
@@ -28,7 +32,7 @@
 #endif
 
 using namespace dealii;
-
+using namespace Cook_Membrane;
 
 template <int dim = 2, int degree = 4, typename number = double>
 void
@@ -36,6 +40,32 @@ test(const unsigned int n_cells = 22528) // 90112 active cells in 2D for p=4 =>
                                          // num of cell batches ~ 90112/4
 {
   constexpr int n_q_points = Utilities::pow(degree + 1, dim);
+
+  std::cout
+    << "-----------------------------------------------------------------------------"
+    << std::endl
+#ifdef DEBUG
+    << "--     . running in DEBUG mode" << std::endl;
+#else
+    << "--     . running in OPTIMIZED mode" << std::endl;
+#endif
+
+  std::cout << "--     . vectorization over "
+            << VectorizedArray<number>::n_array_elements
+            << " elements, VECTORIZATION_LEVEL="
+            << DEAL_II_COMPILER_VECTORIZATION_LEVEL << std::endl;
+
+  std::cout << "--     . version " << GIT_TAG << " (revision " << GIT_SHORTREV
+            << " on branch " << GIT_BRANCH << ")" << std::endl;
+  std::cout << "--     . deal.II " << DEAL_II_PACKAGE_VERSION << " (revision "
+            << DEAL_II_GIT_SHORTREV << " on branch " << DEAL_II_GIT_BRANCH
+            << ")" << std::endl;
+
+  std::cout
+    << "-----------------------------------------------------------------------------"
+    << std::endl
+    << std::endl;
+
   std::cout << "dim                = " << dim << std::endl
             << "degree             = " << degree << std::endl
             << "SIMD width         = "

### 2019-02-01 benchmark/tensors.cc
same number of total cells for SIMD non-SIMD runs

@@ -36,8 +36,7 @@ using namespace Cook_Membrane;
 
 template <int dim = 2, int degree = 4, typename number = double>
 void
-test(const unsigned int n_cells = 22528) // 90112 active cells in 2D for p=4 =>
-                                         // num of cell batches ~ 90112/4
+test(const unsigned int n_cells = 90112/VectorizedArray<number>::n_array_elements)
 {
   constexpr int n_q_points = Utilities::pow(degree + 1, dim);
 

### 2019-02-01 doc/large_strain.tex
Touch ups to mtrx-vec multiplication section; fix some spelling errors

@@
 more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
-Another reason could be the frequent use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
+Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
 The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
@@
 }.
-For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
+For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second-order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
 This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
@@
 The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
-Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces to capture the geometry as well as
+Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces \cite{Arndt2017a} to capture the geometry as well as
 the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
@@
 Nonetheless, }
-the matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
+the matrix-free implementation is faster than the matrix-based counterpart already for \changeJP{bi-cubic Lagrangian basis in 2D and tri-quadratic in 3D}.
 Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
@@
 As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
-Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
-{\color{red}
-In those graphs we report the memory needed to cache additional quantities required for the tangent operator here considered,
+Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point \changeJP{(``tensor4'')}, the matrix-free approach takes much less memory than its matrix-based counterpart.
+{\color{red}
+In those graphs we report the \changeJP{overall} memory needed to cache additional quantities required for the tangent operator here considered,
 as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
 from unit to real cell \cite{kronbichler12}.
-Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
+Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both \changeJP{deformed} and undeformed
 configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}
@@
 {\color{red}
-Finally, we analyze the node-level performance of the two approaches using the Roofline model.
+Finally, we analyze the node-level performance of the two approaches using the roofline model.
 Memory bandwidth (MBytes/s) and the number of floating point operations per second (MFLOP/s) are measured using the \texttt{MEM\_DP} group of the LIKWID \cite{likwid} tool, version 4.2.1.
@@
 {\color{red}
-Figure \ref{fig:roofline} shows results of the Roofline performance model.
+Figure \ref{fig:roofline} shows results of the roofline performance model.
 As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $\pgfmathprintnumber{8.2093245}\,\rm{GFlop/s}$ (4.6\% of the peak performance) in 2D.
@@
 Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
-second order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
+second-order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
 As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}),
@@
   \end{subfigure}
-  \caption{Breakdown of computing times and Roofline analysis of various steps of Algorithm \ref{alg:mf_tensor4}.}%
+  \caption{Breakdown of computing times and roofline analysis of various steps of Algorithm \ref{alg:mf_tensor4}.}%
   \label{fig:breakdown}
@@
 we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
-(i) `read / write' denotes reading and writing from/to a global vector (step 1 and 13);
-(ii) `sum factorization' denotes application of sum factorization techniques (steps 4, 5 and 12);
-(iii) `quadrature loop' denotes operations performed on each quadrature (steps \mbox{6 -- 11}).
+(i) `read / write' denotes reading and writing from/to a global vector (\changeJP{lines 2 and 13});
+(ii) `sum factorization' denotes application of sum factorization techniques (\changeJP{lines} 4, 5 and 12);
+(iii) `quadrature loop' denotes operations performed on each quadrature (\changeJP{lines \mbox{6 -- 11}}).
 Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one
@@
 `read / write' steps takes negligible amount of time spent on each element.
-For higher order bases the majority of time is spent within the quadrature loop.
-When analyzing the Roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
+For higher-order bases the majority of time is spent within the quadrature loop.
+When analyzing the roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
 we observe that this part of the algorithm demonstrates rather low performance, which in our opinion is the source of the
@@
 In this case the quadrature loop consists of, essentially, a
-single contraction between two second order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
-double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
-Further optimization of \texttt{Tensor} classes in \textit{deal.II} are required to
+single contraction between two second-order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
+double contraction between a symmetric second-order tensor and a fourth-order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
+Further optimization of \texttt{Tensor} \changeJP{and \texttt{SymmetricTensor}} classes in \textit{deal.II} are required to
 improve the performance.
@@
 {\color{red}
-To summarize this section, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order elements to two factors.
+To summarize this section, we can attribute the speed-up of the matrix-free finite-strain tangent operators for higher-order elements to two factors.
 First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
@@
 {\color{red}Importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
-On another hand, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
+\changeJP{In comparison}, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
 Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
@@
 {\color{red}
-The Roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
-We identified that one of the bottlenecks is related to the quadrature loop on each element where double contractions of fourth order and second order symmetric tensor
-as well as single contraction of two second order tensors is performed.
+The roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
+We identified that one of the bottlenecks is related to the quadrature loop on each element where double contractions of fourth-order and second-order symmetric tensor
+as well as single contraction of two second-order tensors is performed.
 Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by the \texttt{deal.II}.}
@@
 Importantly the solution wall time per degree of freedom is close to being constant.
-On the contrary, the matrix-based implementation with AMG becomes prohibitively expensive for higher order bases.
+On the contrary, the matrix-based implementation with AMG becomes prohibitively expensive for higher-order bases.
 We conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with

### 2019-02-01 doc/large_strain.tex
Touch ups to mtrx-vec multiplication section; fix some spelling errors

@@
 Figure \ref{fig:roofline} shows results of the roofline performance model.
-As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $\pgfmathprintnumber{8.2093245}\,\rm{GFlop/s}$ (4.6\% of the peak performance) in 2D.
+As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.4\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
 The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
-$\pgfmathprintnumber{18.8047083222}\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
-The achieved performance is about 10.6\% of the peak arithmetic performance. This is much lower than the
+$22.0\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
+The achieved performance is about 13\% of the peak arithmetic performance. This is much lower than the
 70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.

### 2019-02-01 doc/large_strain.tex
Changes to results, conclusions, acknowledgements

@@
 %
+\changeJP{We conduct two performance studies using the described discretization, the first being for only matrix-vector multiplication, and the second for the preconditioned iterative linear solver.}
 The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library.
@@
 however this leads to an overall more expensive linear solver.}
+\changeJP{However, the settings used in this study are optimized for the considered problem in terms of the wall-time spent in the linear solver.}
 The black-box AMG preconditioner
-{\color{red} demonstrate the same behavior of the number of CG iterations, however it}
+{\color{red} demonstrates the same \changeJP{trend in terms of} the number of CG iterations, however it}
 requires many more iterations for convergence.
-{\color{red}
-Consequently, we can conclude that, although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid for the here considered finite strain elasticity problem and that the considered smoother is suitable.
+\changeJP{
+Consequently, we can conclude that for the here considered finite strain elasticity problem (i) the selected smoother is suitable, and that (ii) although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid.
 }
@@
 Additionally, we compare solution times of the preconditioned iterative solvers.
-Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.}
+Our goal is to \changeJP{determine} whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.
+}
+\changeJP{We note that although the well-established implementation of the AMG preconditioner \cite{Gee2006a} is the best available choice for the matrix-based implementation of the boundary value problem, the current implementation of the matrix-free GMG preconditioner is not yet optimized and can be improved further.}
 Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
@@
 Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
-Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for engendering problems
-with finite strain material models.
+Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for \changeJP{engineering problems
+with compressible hyperelastic} finite strain material models.
 Extension of this study from the node level to the cluster level is beyond the scope of this work,
@@
 To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to partial differential equations arising in this case.}
-{\color{red}Among the three algorithms,}
+{\color{red}Among the three \changeJP{examined} algorithms,}
 the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
@@
 
-D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to {\color{red}Martin Kronbichler (TU Munich),} Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for a fruitful discussion on matrix-free operator evaluation approaches.
+D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to {\color{red}Martin Kronbichler (TU Munich),} Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for \changeJP{fruitful discussions} on matrix-free operator evaluation approaches.
 
@@
 P.~Steinmann acknowledges the support of the Cluster of Excellence Engineering of Advanced Materials (EAM) which made this collaboration possible, as well as funding by the EPSRC Strategic Support Package ``Engineering of Active Materials by Multiscale/Multiphysics Computational Mechanics''.
+
+\todo{Add back of the envelope calculations by Veselin Dobrev? See {https://github.com/CEED/Forum/issues/1}.}
 

### 2019-02-01 doc/large_strain.tex
Changes to results, conclusions, acknowledgements

@@
 Figure \ref{fig:roofline} shows results of the roofline performance model.
-As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.4\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
+As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $\pgfmathprintnumber{8.2093245}\,\rm{GFlop/s}$ (4.6\% of the peak performance) in 2D.
 The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
-$22.0\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
-The achieved performance is about 13\% of the peak arithmetic performance. This is much lower than the
+$\pgfmathprintnumber{18.8047083222}\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
+The achieved performance is about 10.6\% of the peak arithmetic performance. This is much lower than the
 70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.

### 2019-02-01 doc/large_strain.tex
Add blurb on the origin and use of the NH model

--- 2019-02-01 doc/large_strain.tex
@@
 \end{gather}
-where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively. It can be shown that the first and second derivatives of the strain energy function are given by
+where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively.
+\changeJP{
+Derived using statistical thermodynamics applied to cross-linked polymers, the Neo-Hookean model \cite{Treloar1975a,Treloar1976a} is commonly used to model rubber-like materials in the finite-strain regime.
+}
+It can be shown that the first and second derivatives of the strain energy function are given by
 \begin{align}

### 2019-02-05 doc/large_strain.tex
Improve some wordings and fox typos

--- 2019-02-01 doc/large_strain.tex
@@
 
-\usepackage{pgfkeys,pgfmath,pgfcore}
-% \pgfkeys{/pgf/number format/.cd,1000 sep={}}
-\pgfkeys{/pgf/number format/fixed,/pgf/number format/precision=3}
-
 %
@@
 
-\newcommand{\changeJP}[1]{\textcolor{blue}{#1}}
-\newcommand{\todojp}[1]{\todo[inline,color=red!40,size=\small]{TODO (J-P): #1}}
-
 \begin{document}
@@
 
-    \changeJP{To date, there have been few (if any) assessments into the applicability of the matrix-free approach to problems in solid mechanics.
-    In this work, we perform an initial investigation on the application of the matrix-free approach to problems in quasi-static finite-strain hyperelasticity to determine whether it is viable for further extension.
-    Specifically, we study different numerical implementations of the finite element tangent operator, and determine whether generalized methods of incorporating complex constitutive behavior might be feasible.
-    }
+    In this work, we extend the application of matrix-free approaches to problems in solid mechanics and investigate different implementations of the finite-strain hyperelastic finite element tangent operator numerically.
     In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
@@
     The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner {\color{red} on a single node} for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
-    \changeJP{We conclude that the application of matrix-free methods to finite-strain solid mechanics is promising, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
   \end{abstract}
@@
 To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
-We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018}.
+We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent master's thesis produced by a student of the first two of the authors \cite{Mentler2017},
+{\color{red} which is also limited to small strain linear elasticity.}\todo{omit?}
 One possible reason is that the tangent operators\footnote{Here and
@@
 more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
-Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
+Another reason could be the frequent use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
 The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
 whereas higher order elements are rarely employed.
-This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements\todo{reference? Martin's papers?}.
+This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements\todo{reference?}.
 However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
@@
 
-In this work, we perform \changeJP{a preliminary investigation of} different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost
-and performance {\color{red} on a single node of a high performance computing cluster.}
-\changeJP{
-The \texttt{deal.II} \cite{dealII90} finite element library, which provides MPI parallelization together with a \changeJP{generalized matrix-free framework that incorporates SIMD vectorization} and an interface to high-performance, MPI distributed matrix-based libraries, is used for this study.
-}
+In this work, we investigate different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost
+and performance {\color{red} on a single node of an higher performance computing cluster.}
 In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
 and employ them to define a \mbox{geometric} multigrid preconditioner.
+MPI parallelization together with SIMD vectorization
+{\color{red}
+are provided by the \texttt{deal.II} \cite{dealII90} finite element library, used for this study.
+}
 
 The paper is organized as follows: In Section \ref{sec:theory} we briefly introduce the partial differential equations used in  finite-strain solid mechanics.
-\changeJP{
-Sections \ref{sec:fe} and \ref{sec: Description of numerical framework} cover the finite element discretization and provide details of the numerical framework with which this study is implemented.
-In \ref{sec:mf} we describe the different matrix-free operator evaluation approaches that are evaluated in this study, and in section \ref{sec:gmg} we propose a geometric multigrid preconditioner that is suitable for finite-strain hyperelasticity.
-}
+Sections \ref{sec:fe} and \ref{sec:mf} cover the finite element discretization and the different matrix-free operator evaluation approaches.
+In Section \ref{sec:gmg} we propose a geometric multigrid preconditioner suitable for finite-strain hyperelasticity.
 The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}.
-\changeJP{Lastly, the} results are summarized in Section \ref{sec:summary}.
+The results are summarized in Section \ref{sec:summary}.
 
@@
 
-\subsection{Constitutive modelling
-\label{sec: Constitutive modelling}
-}
+\subsection{Constitutive modelling}
 
@@
 \end{gather}
-where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively.
-\changeJP{
-Derived using statistical thermodynamics applied to cross-linked polymers, the Neo-Hookean model \cite{Treloar1975a,Treloar1976a} is commonly used to model rubber-like materials in the finite-strain regime.
-}
-It can be shown that the first and second derivatives of the strain energy function are given by
+where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively. It can be shown that the first and second derivatives of the strain energy function are given by
 \begin{align}
@@
 
-\section{\changeJP{Description of numerical framework}
-\label{sec: Description of numerical framework}
-}
-
-The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
-This library offers a number of paradigms by which to parallelize a finite-element code.
-In this work we employ \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
-%
-\begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.8\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{fine_level_2d.png}
-    % \caption{2D multigrid mesh with 2 global refinements}
-  \end{subfigure}
-  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
-  \label{fig:miehe_fine_level}
-\end{figure}
-%
-%
-Figure \ref{fig:miehe_fine_level} illustrates a distributed MPI decomposition on the fine mesh level for the 2D problem that will be introduced later in Section \ref{sec:example}.
-
-In practise this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
-For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
-For the matrix-based approach, the Trilinos \cite{Heroux2005} library, version 12.12.1 was leveraged; 
-the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the package ML package \cite{Gee2006a}.
-The second implementation, which is discussed in detail in the following Sections, utilized a matrix-free approach in conjunction with a geometric-multigrid preconditioner.
-As will be detailed later, the pre-existing matrix-free framework that was used offers further opportunity for low-level parallelism that may be exploited.
-
 \section{Matrix-free operator evaluation}
@@
 Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
-In this case the the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
-On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
+In this case, the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
+On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
 For this reason, recent implementations often focus on so-called matrix-free approaches where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
@@
 {\color{red}
-that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem \changeJP{derived with a Lagrangian formulation},}
+that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem,}
 we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
@@
 {\color{red}
-Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach \changeJP{, each of which have a different level of abstraction, algorithmic complexity, and memory requirements}.
-\changeJP{They are introduced in order of lowest memory footprint to largest:}
+Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
 }
@@
         {\color{red}extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ } \;
-        evaluate the following tensors at each quadrature point using sum factorization :\\
-        \enskip $\Grad \, \overline{\gz u^h}_K$ \tcp*{2nd order}
-        \enskip $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
-        \enskip $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
+        evaluate the following tensors at each quadrature point using sum factorization \;
+        $\Grad \, \overline{\gz u^h}_K$ \tcp*{2nd order}
+        $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
+        $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{
@@
         {\color{red}extract local vector values on this element $\gz x_K$ } \;
-        evaluate the following tensors at each quadrature point using sum factorization :\\
-        \enskip$\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
-        \enskip$\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
+        evaluate the following tensors at each quadrature point using sum factorization \;
+        $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
+        $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
         \ForEach{quadrature point $q$ on $K$}{
@@
 {\color{red}
-The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements\footnote{%
-\changeJP{In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
-This motivates the caching strategy employed in Algorithm 3, as the pre-caching of the chosen data ensures a context in which SIMD parallelization of subsequent operations is guaranteed.
-Algorithms 1 and 2 sacrifice some of this generality for decreased memory requirements.}
-}.
-For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{%
-For the heterogeneous materials considered here,
+The main reason is that the operations are typically the same on all elements.
+For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{For the heterogeneous materials considered here,
 the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
@@
 }.
-For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second-order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
+For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
 This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
@@
 \textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
-Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
+Thanks to the object-oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
 Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
-For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
+For more information about the implementation details and data structures for sum factorization approaches in deal.II (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and shared memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
 }
@@
 For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
-\changeJP{Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.}
 
@@
 We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
-Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
+% Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
 }
@@
 The relative convergence criteria for the linear solver is $10^{-6}$.
-%
+{\color{red}
+The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
+by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
+The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
+is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
+On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
+Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}
+
 Figure \ref{fig:miehe_deformed} shows deformed meshes at the final loading step with quadratic elements and two global mesh refinements.
-%
+
 \begin{figure}[!ht]
@@
 \end{figure}
-%
+
 \begin{figure}[!ht]
@@
     \centering
-    \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
+    \includegraphics[width=\textwidth]{fine_level_2d.png}
     % \caption{2D multigrid mesh with 2 global refinements}
   \end{subfigure}
-  \caption{2D multigrid mesh after two global refinements distributed into four MPI processes. The color indicates the process owning the element.}%
+  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
+  \label{fig:miehe_fine_level}
+\end{figure}
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.8\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{gmg_2d.png}
+    % \caption{2D multigrid mesh with 2 global refinements}
+  \end{subfigure}
+  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
   \label{fig:miehe_gmg}
@@
 %
-\changeJP{The same distributed approach to mesh decomposition (as most evident on the finest level) is adopted on the multigrid levels.}
-Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements \changeJP{for the mesh introduced in figure \ref{fig:miehe_fine_level}}.
-%
+{\color{red}
+MPI parallelization is achieved using the standard domain decomposition, where each MPI process ``owns'' only a subset of elements.
+To that end, \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library is used.
+Figure \ref{fig:miehe_fine_level} illustrates such decomposition on the fine mesh level for the 2D problem.
+The same approach is adopted on the multigrid levels.}
+Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.
+{\color{red}
+Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.
+}
+
 \begin{table}
@@
 
-\changeJP{%
-Aligned with the approach taken in similar previous investigations \cite{}\todo{Cite Martin's papers?}, in this study we limit ourselves to comparisons made on a node level.
-Examinations performed in this manner should already indicate the differences between the matrix-free and matrix-based approaches, and give some preliminary understanding of the scalability potential of the numerical implementation without confounding factors related to data exchange though interconnects via MPI.
-}
+{\color{red}In this study, we limit ourselves to a comparison on a node level, which should already indicate the differences between the matrix-free and matrix-based approaches.}
 The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
 {\color{red}
-We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach \changeJP{with a high polynomial order finite element}.
-\changeJP{Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound}.
-}
-
-%
-% Below are are implementational details that might affect the comparative result
-%
-\changeJP{We conduct two performance studies using the described discretization, the first being for only matrix-vector multiplication, and the second for the preconditioned iterative linear solver.}
-The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library.
+We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach.
+On the other hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we can observe that the matrix-based approach is memory bound.
+}
+%
+The results of the matrix-free approach with a geometric-multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
 The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
-{\color{red}
-The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
-by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
-The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
-is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
-On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
-Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}
-
-Computations \changeJP{were performed} on a single node of two clusters:
+Computations are done on a single node of two clusters:
 A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM.
@@
 For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
-{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. \changeJP{one} fully occupied node.}
+{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. fully occupied node.}
 
@@
 The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
-Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces \cite{Arndt2017a} to capture the geometry as well as
+Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces to capture the geometry as well as
 the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
@@
 Nonetheless, }
-the matrix-free implementation is faster than the matrix-based counterpart already for \changeJP{bi-cubic Lagrangian basis in 2D and tri-quadratic in 3D}.
+the matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
 Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
@@
 As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
-Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point \changeJP{(``tensor4'')}, the matrix-free approach takes much less memory than its matrix-based counterpart.
-{\color{red}
-In those graphs we report the \changeJP{overall} memory needed to cache additional quantities required for the tangent operator here considered,
+Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
+{\color{red}
+In those graphs we report the memory needed to cache additional quantities required for the tangent operator here considered,
 as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
 from unit to real cell \cite{kronbichler12}.
-Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both \changeJP{deformed} and undeformed
-configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}
+Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deformed and undeformed
+configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{this is roughly main memory transfer, do we elaborate on this/use it? I don't think so}}
 
@@
 {\color{red}
-Finally, we analyze the node-level performance of the two approaches using the roofline model.
+Finally, we analyze the node-level performance of the two approaches using the Roofline model.
 Memory bandwidth (MBytes/s) and the number of floating point operations per second (MFLOP/s) are measured using the \texttt{MEM\_DP} group of the LIKWID \cite{likwid} tool, version 4.2.1.
@@
 {\color{red}
-Figure \ref{fig:roofline} shows results of the roofline performance model.
+Figure \ref{fig:roofline} shows results of the Roofline performance model.
 As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.4\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
@@
 
-When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
+When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}) caching strategy,
 we observe a marginal increase in computational intensity and almost unchanged performance.
-This coincides with our expectation that the later algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
-On another hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
+This coincides with our expectation that the latter algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
+On the other hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
 Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
-second-order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
+second order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
 As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}),
@@
   \end{subfigure}
-  \caption{Breakdown of computing times and roofline analysis of various steps of Algorithm \ref{alg:mf_tensor4}.}%
+  \caption{Breakdown of computing times and Roofline analysis of various steps of Algorithm \ref{alg:mf_tensor4}.}%
   \label{fig:breakdown}
@@
 
-\todo{Discuss MPI/SIMD}
-\input{parallelization_2d.tex}
-\input{parallelization_3d.tex}
-
 In order to have a better understanding of the behavior of the most promising ``tensor4'' implementation,
 we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
-(i) `read / write' denotes reading and writing from/to a global vector (\changeJP{lines 2 and 13});
-(ii) `sum factorization' denotes application of sum factorization techniques (\changeJP{lines} 4, 5 and 12);
-(iii) `quadrature loop' denotes operations performed on each quadrature (\changeJP{lines \mbox{6 -- 11}}).
+(i) `read / write' denotes reading and writing from/to a global vector (step 1 and 13);
+(ii) `sum factorization' denotes application of sum factorization techniques (steps 4, 5 and 12);
+(iii) `quadrature loop' denotes operations performed on each quadrature (steps \mbox{6 -- 11}).
 Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one
 global refinement in 2D and no global mesh refinement in 3D.
-It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that
-`read / write' steps takes negligible amount of time spent on each element.
-For higher-order bases the majority of time is spent within the quadrature loop.
-When analyzing the roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
+It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that the
+`read / write' steps takes a negligible amount of time on each element.
+For higher order bases, the majority of time is spent within the quadrature loop.
+When analyzing the Roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
 we observe that this part of the algorithm demonstrates rather low performance, which in our opinion is the source of the
 suboptimal performance of the whole matrix-free matrix-vector product (see Figure \ref{fig:roofline}).
-In this case the quadrature loop consists of, essentially, a
-single contraction between two second-order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
-double contraction between a symmetric second-order tensor and a fourth-order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
-Further optimization of \texttt{Tensor} \changeJP{and \texttt{SymmetricTensor}} classes in \textit{deal.II} are required to
-improve the performance.
-Such studies, however, are beyond the scope of the current work.
-}
-
-{\color{red}
-To summarize this section, we can attribute the speed-up of the matrix-free finite-strain tangent operators for higher-order elements to two factors.
-First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
+In this case, the quadrature loop consists of, essentially, a
+single contraction between two second order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
+double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
+% Further optimization of \texttt{Tensor} classes in \textit{deal.II} are required to
+% improve the performance.
+% Such studies, however, are beyond the scope of the current work.
+}
+
+{\color{red}
+To summarize this section, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order elements to two factors.
+The first one is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
 bound, see Figure \ref{fig:roofline}.
-Second, is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
+The second one is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
 Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
@@
 
-Next we evaluate the performance of the proposed geometric multigrid preconditioner.
-{\color{red}
-Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end}
+Next, we evaluate the performance of the proposed geometric multigrid preconditioner.
+{\color{red}
+Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end,}
 we consider {\color{red}the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
@@
 however this leads to an overall more expensive linear solver.}
-\changeJP{However, the settings used in this study are optimized for the considered problem in terms of the wall-time spent in the linear solver.}
 The black-box AMG preconditioner
-{\color{red} demonstrates the same \changeJP{trend in terms of} the number of CG iterations, however it}
+{\color{red} demonstrate the same behavior of the number of CG iterations, but}
 requires many more iterations for convergence.
-\changeJP{
-Consequently, we can conclude that for the here considered finite strain elasticity problem (i) the selected smoother is suitable, and that (ii) although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid.
+{\color{red}
+Consequently, we can conclude that, although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid for the here considered finite strain elasticity problem and that the considered smoother is suitable.
 }
@@
 Additionally, we compare solution times of the preconditioned iterative solvers.
-Our goal is to \changeJP{determine} whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.
-}
-\changeJP{We note that although the well-established implementation of the AMG preconditioner \cite{Gee2006a} is the best available choice for the matrix-based implementation of the boundary value problem, the current implementation of the matrix-free GMG preconditioner is not yet optimized and can be improved further.}
+Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.}
 Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
-{\color{red}Importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
-\changeJP{In comparison}, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
+{\color{red}Most importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
+On the other hand, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
 Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
-Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for \changeJP{engineering problems
-with compressible hyperelastic} finite strain material models.
+Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for solid mechanics problems
+with finite strain material models.
 Extension of this study from the node level to the cluster level is beyond the scope of this work,
@@
 To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to partial differential equations arising in this case.}
-{\color{red}Among the three \changeJP{examined} algorithms,}
-the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
-{\color{red}
-Given that in this case we did not utilize the specific form of the material part of the stiffness tensor, we conclude}
+{\color{red}Among the three algorithms,}
+the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for polynomial degrees higher than one in 3D and polynomial degrees higher than two {\color{red} in 2D}.
+{\color{red}
+Given that we did not utilize the specific form of the material part of the stiffness tensor  in this case, we conclude}
 that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
@@
 {\color{red}
-The roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
-We identified that one of the bottlenecks is related to the quadrature loop on each element where double contractions of fourth-order and second-order symmetric tensor
-as well as single contraction of two second-order tensors is performed.
-Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by the \texttt{deal.II}.}
-
-{\color{red}
-We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner
-}
+The Roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
+We identified that one of the bottlenecks is related to the quadrature loop on each element where double contractions of fourth order and second order symmetric tensor
+as well as single contraction of two second order tensors is performed.
+% Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by the \texttt{deal.II}.
+}
+
+{\color{red}
+We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner}
 with standard geometric transfer operations between each level.
@@
 leads to a solution approach that is faster than the matrix-based AMG {\color{red}for all polynomial degrees studied here.
-Importantly the solution wall time per degree of freedom is close to being constant.
-On the contrary, the matrix-based implementation with AMG becomes prohibitively expensive for higher-order bases.
+Most importantly, the wall time per degree of freedom for solving the problem is close to being constant.
+On the other hand, the matrix-based implementation with AMG becomes prohibitively expensive for higher order bases.
 We conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with
@@
 Our future work will be focused on adopting the proposed matrix-free multigrid solution approach to $FE^2$ homogenization
-as well studying its behavior on the cluster level as opposed to node level, considered here.}
+as well studying its behavior on the cluster level
+% as opposed to node level considered here.
+}
 
@@
 
-D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to {\color{red}Martin Kronbichler (TU Munich),} Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for \changeJP{fruitful discussions} on matrix-free operator evaluation approaches.
+D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to {\color{red}Martin Kronbichler (TU Munich),} Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for a fruitful discussion on matrix-free operator evaluation approaches.
 
@@
 P.~Steinmann acknowledges the support of the Cluster of Excellence Engineering of Advanced Materials (EAM) which made this collaboration possible, as well as funding by the EPSRC Strategic Support Package ``Engineering of Active Materials by Multiscale/Multiphysics Computational Mechanics''.
-
-\todo{Add back of the envelope calculations by Veselin Dobrev? See {https://github.com/CEED/Forum/issues/1}.}
 

### 2019-02-06 doc/large_strain.tex

--- 2019-02-01 doc/large_strain.tex
@@
 \renewcommand\thesubfigure{(\alph{subfigure})}
-
-\usepackage{pgfkeys,pgfmath,pgfcore}
-% \pgfkeys{/pgf/number format/.cd,1000 sep={}}
-\pgfkeys{/pgf/number format/fixed,/pgf/number format/precision=3}
 
@@
 
-In practise this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
+In practice this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
 For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
@@
 \textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
-Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
+Thanks to the object-oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
 Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
@@
 
-For this work we adopt the matrix-free level operators within the GMG.
+For this work, we adopt the matrix-free level operators within the GMG.
 In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial.
@@
     \centering
-    \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
+    \includegraphics[width=\textwidth]{gmg_2d.png}
     % \caption{2D multigrid mesh with 2 global refinements}
   \end{subfigure}
-  \caption{2D multigrid mesh after two global refinements distributed into four MPI processes. The color indicates the process owning the element.}%
+  \caption{2D multigrid mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
   \label{fig:miehe_gmg}
@@
 Aligned with the approach taken in similar previous investigations \cite{}\todo{Cite Martin's papers?}, in this study we limit ourselves to comparisons made on a node level.
-Examinations performed in this manner should already indicate the differences between the matrix-free and matrix-based approaches, and give some preliminary understanding of the scalability potential of the numerical implementation without confounding factors related to data exchange though interconnects via MPI.
+Examinations performed in this manner should already indicate the differences between the matrix-free and matrix-based approaches, and give some preliminary understanding of the scalability potential of the numerical implementation without confounding factors related to data exchange through interconnects via MPI.
 }
@@
 is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
-On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
+On the coarsest level, we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
 Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}
@@
 Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both \changeJP{deformed} and undeformed
-configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}
+configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{this is roughly main memory transfer, do we elaborate on this/use it? I don't think so}}
 
@@
 
-When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
+When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}) caching strategy,
 we observe a marginal increase in computational intensity and almost unchanged performance.
-This coincides with our expectation that the later algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
-On another hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
+This coincides with our expectation that the latter algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
+On the other hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
 Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
@@
 \end{figure}
-
-\todo{Discuss MPI/SIMD}
-\input{parallelization_2d.tex}
-\input{parallelization_3d.tex}
 
@@
 global refinement in 2D and no global mesh refinement in 3D.
-It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that
-`read / write' steps takes negligible amount of time spent on each element.
-For higher-order bases the majority of time is spent within the quadrature loop.
+It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that the
+`read / write' steps take a negligible amount of time on each element.
+For higher order bases, the majority of time is spent within the quadrature loop.
 When analyzing the roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
@@
 suboptimal performance of the whole matrix-free matrix-vector product (see Figure \ref{fig:roofline}).
-In this case the quadrature loop consists of, essentially, a
-single contraction between two second-order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
-double contraction between a symmetric second-order tensor and a fourth-order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
-Further optimization of \texttt{Tensor} \changeJP{and \texttt{SymmetricTensor}} classes in \textit{deal.II} are required to
-improve the performance.
-Such studies, however, are beyond the scope of the current work.
+In this case the, quadrature loop consists of, essentially, a
+single contraction between two second order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
+double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
+% Further optimization of \texttt{Tensor} \changeJP{and \texttt{SymmetricTensor}}classes in \textit{deal.II} are required to
+% improve the performance.
+% Such studies, however, are beyond the scope of the current work.
 }
@@
 To summarize this section, we can attribute the speed-up of the matrix-free finite-strain tangent operators for higher-order elements to two factors.
-First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
+The first one is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
 bound, see Figure \ref{fig:roofline}.
-Second, is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
+The second one is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
 Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
@@
 The black-box AMG preconditioner
-{\color{red} demonstrates the same \changeJP{trend in terms of} the number of CG iterations, however it}
+{\color{red} demonstrates the same \changeJP{trend in terms of} the number of CG iterations, but}
 requires many more iterations for convergence.
@@
 Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
-{\color{red}Importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
+{\color{red}Most importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
 \changeJP{In comparison}, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
@@
 {\color{red}Among the three \changeJP{examined} algorithms,}
-the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
-{\color{red}
-Given that in this case we did not utilize the specific form of the material part of the stiffness tensor, we conclude}
+the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for polynomial degrees higher than one in 3D and polynomial degrees higher than two {\color{red} in 2D}.
+{\color{red}
+Given that we did not utilize the specific form of the material part of the stiffness tensor in this case, we conclude}
 that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
@@
 as well as single contraction of two second-order tensors is performed.
-Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by the \texttt{deal.II}.}
-
-{\color{red}
-We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner
-}
+% Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by \texttt{deal.II}.}
+
+{\color{red}
+We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner}
 with standard geometric transfer operations between each level.
@@
 leads to a solution approach that is faster than the matrix-based AMG {\color{red}for all polynomial degrees studied here.
-Importantly the solution wall time per degree of freedom is close to being constant.
-On the contrary, the matrix-based implementation with AMG becomes prohibitively expensive for higher-order bases.
+Most importantly, the wall time per degree of freedom for solving the problem is close to being constant.
+On the other hand, the matrix-based implementation with AMG becomes prohibitively expensive for higher-order bases.
 We conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with

### 2019-02-06 doc/large_strain.tex
remove todo and white spaces

@@
 Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both \changeJP{deformed} and undeformed
-configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{this is roughly main memory transfer, do we elaborate on this/use it? I don't think so}}
+configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.}
 

### 2019-02-10 doc/large_strain.tex
Add some references and some discussion of explicit SIMD speedups

@@
 whereas higher order elements are rarely employed.
-This can potentially reduce the advantage of matrix-free methods, that are generally more competitive for higher order elements\todo{reference? Martin's papers?}.
+This can potentially reduce the advantage of matrix-free methods using sum-factorization techniques, that are generally more competitive for higher order elements \cite{kronbichler12,kronbichler2017fast,muthing2017high}.
 However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
@@
 
-In practice this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
+In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
 For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
@@
 \changeJP{%
-Aligned with the approach taken in similar previous investigations \cite{}\todo{Cite Martin's papers?}, in this study we limit ourselves to comparisons made on a node level.
+Aligned with the approach taken in similar previous investigations, compare e.g. \cite{kronbichler2017fast}, 
+we limit ourselves to comparisons made on a node level in this study.
 Examinations performed in this manner should already indicate the differences between the matrix-free and matrix-based approaches, and give some preliminary understanding of the scalability potential of the numerical implementation without confounding factors related to data exchange through interconnects via MPI.
@@
 {\color{red}
-In those graphs we report the \changeJP{overall} memory needed to cache additional quantities required for the tangent operator here considered,
+In those graphs, we report the \changeJP{overall} memory needed to cache additional quantities required for the tangent operator here considered,
 as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
@@
 
+\input{parallelization_2d.tex}
+\input{parallelization_3d.tex}
+Table \ref{tab:numbers_2d} and Table \ref{tab:numbers_3d} show the speed-up obtained due to explicit SIMD vectorization intrinsics and due to MPI.
+For these rather small count of MPI processes, the results confirm that the performance gained is independent.
+Furthermore, the improvement due to explicit intrinsics for the data used in the quadrature loop, is comparable to the ones in \cite[Figure 6]{kronbichler2017fastcomputer} where the speedup factor is $1.2-1.8$ for a similar architecture.
+
 {\color{red}
@@
 the geometric multigrid preconditioner is a very competitive solution strategy, as compared to more traditional matrix-based approach.
-Our future work will be focused on adopting the proposed matrix-free multigrid solution approach to $FE^2$ homogenization
+Our future work will be focused on adopting the proposed matrix-free multigrid solution approach to $\operatorname{FE}^2$ homogenization
 as well studying its behavior on the cluster level
@@
 
-\todo{Add back of the envelope calculations by Veselin Dobrev? See {https://github.com/CEED/Forum/issues/1}.}
+% \todo{Add back of the envelope calculations by Veselin Dobrev? See {https://github.com/CEED/Forum/issues/1}.}
 

### 2019-02-10 doc/large_strain.tex
Relativize the 70%

@@
 The achieved performance is about 10.6\% of the peak arithmetic performance. This is much lower than the
-70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.
+70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}, but similar to what was obtained in \cite[Figure 14]{kronbichler2017fast} when not optimizing for Cartesian meshes.
 Unfortunately, performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.

### 2019-02-12 Calculations/post_process_likwid.py
python: add asserts on correct clockspeed during post-processing

@@ -49,7 +49,8 @@ sections = [
     'Memory bandwidth [MBytes/s] STAT',
     'Operational intensity STAT',
     'Runtime unhalted [s] STAT',
-    'Runtime (RDTSC) [s] STAT' # spoke with Georg Haager, this shall be closest to the walltime.
+    'Runtime (RDTSC) [s] STAT', # spoke with Georg Haager, this shall be closest to the walltime.
+    'Clock [MHz] STAT'
 ]
 
 # NODE data:
@@ -161,13 +162,18 @@ for f in files:
             if len(columns) > 1:
                 for idx, s in enumerate(sections):
                     if s == columns[1]:
-                        if 'Runtime' in s:
+                        if ('Runtime' in s) or ('Clock' in s):
                             # Take "Max" (third number)
                             val = float(columns[4])
                         else:
                             # Take "Sum" (first number)
                             val = float(columns[2])
                         print '   {0} {1}'.format(s,val)
+                        # make sure we run with clockspeed we use for Roofline:
+                        if 'Clock' in s:
+                            # allow 0.1% variation
+                            assert abs(val - clock_speed * 1000) < clock_speed
+
                         # we should get here only once for each region:
                         assert np.isnan(timing[r_idx][idx])
                         timing[r_idx][idx] = val

### 2019-02-12 Calculations/post_process_likwid_simd.py

@@ -15,10 +15,12 @@ parser.add_argument('--dim', metavar='dim', default=2, type=int,
 
 args = parser.parse_args()
 
+# expected blockspeed:
+clock_speed = 2.2  # GHz
 
 # serial | MPI | SIMD | MPI+SIMD
 suffixes = [
-    '_1proc_novec',
+    '_novec_1proc',
     '_novec',
     '_1proc',
     ''
@@ -68,10 +70,12 @@ for idx, s in enumerate(suffixes):
             data = result['Metric']
             flops = data['MFLOP/s'][0]
             runtime = data['Runtime (RDTSC) [s]'][0]
+            assert abs(float(data['Clock [MHz]'][0]) - clock_speed * 1000) < clock_speed
         else:
             data = result['Metric_Sum']
             flops = data['MFLOP/s STAT'][0]                # take Sum
             runtime = data['Runtime (RDTSC) [s] STAT'][2]  # take Max
+            assert abs(float(data['Clock [MHz] STAT'][2]) - clock_speed * 1000) < clock_speed
 
         runtime = float(runtime) / 10  # FIXME: hard-code that we run 10 times
         flops = float(flops) / 1000  # MFLOP/s -> GFLOP/s

### 2019-02-12 doc/large_strain.tex
fix closing bracket
update figures and reported in the text 2D average performance

--- 2019-02-06 doc/large_strain.tex
@@
 
-\section{\changeJP{Description of numerical framework}
-\label{sec: Description of numerical framework}
-}
-
-The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
-This library offers a number of paradigms by which to parallelize a finite-element code.
-In this work we employ \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
-%
-\begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.8\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{fine_level_2d.png}
-    % \caption{2D multigrid mesh with 2 global refinements}
-  \end{subfigure}
-  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
-  \label{fig:miehe_fine_level}
-\end{figure}
-%
-%
-Figure \ref{fig:miehe_fine_level} illustrates a distributed MPI decomposition on the fine mesh level for the 2D problem that will be introduced later in Section \ref{sec:example}.
-
-In practice this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
-For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
-For the matrix-based approach, the Trilinos \cite{Heroux2005} library, version 12.12.1 was leveraged;
-the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the package ML package \cite{Gee2006a}.
-The second implementation, which is discussed in detail in the following Sections, utilized a matrix-free approach in conjunction with a geometric-multigrid preconditioner.
-As will be detailed later, the pre-existing matrix-free framework that was used offers further opportunity for low-level parallelism that may be exploited.
-
 \section{Matrix-free operator evaluation}
@@
 Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
-In this case, the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
-On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
+In this case the the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
+On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
 For this reason, recent implementations often focus on so-called matrix-free approaches where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
@@
 \textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
-Thanks to the object-oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
+Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
 Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
-For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
+For more information about the implementation details and data structures for sum factorization approaches in deal.II (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and shared memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
 }
@@
 For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
-\changeJP{Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.}
-
-For this work, we adopt the matrix-free level operators within the GMG.
+
+For this work we adopt the matrix-free level operators within the GMG.
 In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial.
@@
 We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
-% Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
+Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
 }
@@
 \end{figure}
-%
+
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.8\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{fine_level_2d.png}
+    % \caption{2D multigrid mesh with 2 global refinements}
+  \end{subfigure}
+  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
+  \label{fig:miehe_fine_level}
+\end{figure}
+
 \begin{figure}[!ht]
@@
 
-\changeJP{%
-Aligned with the approach taken in similar previous investigations \cite{}\todo{Cite Martin's papers?}, in this study we limit ourselves to comparisons made on a node level.
-Examinations performed in this manner should already indicate the differences between the matrix-free and matrix-based approaches, and give some preliminary understanding of the scalability potential of the numerical implementation without confounding factors related to data exchange through interconnects via MPI.
-}
+{\color{red}In this study, we limit ourselves to a comparison on a node level, which should already indicate the differences between the matrix-free and matrix-based approaches.}
 The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
 {\color{red}
-We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach \changeJP{with a high polynomial order finite element}.
-\changeJP{Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound}.
-}
-
-%
-% Below are are implementational details that might affect the comparative result
-%
-\changeJP{We conduct two performance studies using the described discretization, the first being for only matrix-vector multiplication, and the second for the preconditioned iterative linear solver.}
-The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library.
+We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach.
+On the other hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we can observe that the matrix-based approach is memory bound.
+}
+%
+The results of the matrix-free approach with a geometric-multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
 The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
-{\color{red}
-The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
-by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
-The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
-is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
-On the coarsest level, we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
-Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}
-
-Computations \changeJP{were performed} on a single node of two clusters:
+Computations are done on a single node of two clusters:
 A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM.
@@
 As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
-Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point \changeJP{(``tensor4'')}, the matrix-free approach takes much less memory than its matrix-based counterpart.
-{\color{red}
-In those graphs we report the \changeJP{overall} memory needed to cache additional quantities required for the tangent operator here considered,
+Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
+{\color{red}
+In those graphs we report the memory needed to cache additional quantities required for the tangent operator here considered,
 as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
 from unit to real cell \cite{kronbichler12}.
-Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both \changeJP{deformed} and undeformed
-configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.}
+Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
+configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}
 
@@
 {\color{red}
-Figure \ref{fig:roofline} shows results of the roofline performance model.
+Figure \ref{fig:roofline} shows results of the Roofline performance model.
 As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $\pgfmathprintnumber{8.2093245}\,\rm{GFlop/s}$ (4.6\% of the peak performance) in 2D.
@@
 
-When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}) caching strategy,
+When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
 we observe a marginal increase in computational intensity and almost unchanged performance.
-This coincides with our expectation that the latter algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
-On the other hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
+This coincides with our expectation that the later algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
+On another hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
 Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
-second-order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
+second order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
 As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}),
@@
 we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
-(i) `read / write' denotes reading and writing from/to a global vector (\changeJP{lines 2 and 13});
-(ii) `sum factorization' denotes application of sum factorization techniques (\changeJP{lines} 4, 5 and 12);
-(iii) `quadrature loop' denotes operations performed on each quadrature (\changeJP{lines \mbox{6 -- 11}}).
+(i) `read / write' denotes reading and writing from/to a global vector (step 1 and 13);
+(ii) `sum factorization' denotes application of sum factorization techniques (steps 4, 5 and 12);
+(iii) `quadrature loop' denotes operations performed on each quadrature (steps \mbox{6 -- 11}).
 Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one
 global refinement in 2D and no global mesh refinement in 3D.
-It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that the
-`read / write' steps take a negligible amount of time on each element.
-For higher order bases, the majority of time is spent within the quadrature loop.
-When analyzing the roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
+It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that
+`read / write' steps takes negligible amount of time spent on each element.
+For higher order bases the majority of time is spent within the quadrature loop.
+When analyzing the Roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
 we observe that this part of the algorithm demonstrates rather low performance, which in our opinion is the source of the
 suboptimal performance of the whole matrix-free matrix-vector product (see Figure \ref{fig:roofline}).
-In this case the, quadrature loop consists of, essentially, a
+In this case the quadrature loop consists of, essentially, a
 single contraction between two second order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
 double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
-% Further optimization of \texttt{Tensor} \changeJP{and \texttt{SymmetricTensor}}classes in \textit{deal.II} are required to
-% improve the performance.
-% Such studies, however, are beyond the scope of the current work.
-}
-
-{\color{red}
-To summarize this section, we can attribute the speed-up of the matrix-free finite-strain tangent operators for higher-order elements to two factors.
-The first one is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
+Further optimization of \texttt{Tensor} classes in \textit{deal.II} are required to
+improve the performance.
+Such studies, however, are beyond the scope of the current work.
+}
+
+{\color{red}
+To summarize this section, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order elements to two factors.
+First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
 bound, see Figure \ref{fig:roofline}.
-The second one is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
+Second, is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
 Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
@@
 
-Next, we evaluate the performance of the proposed geometric multigrid preconditioner.
-{\color{red}
-Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end,}
+Next we evaluate the performance of the proposed geometric multigrid preconditioner.
+{\color{red}
+Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end}
 we consider {\color{red}the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
@@
 however this leads to an overall more expensive linear solver.}
-\changeJP{However, the settings used in this study are optimized for the considered problem in terms of the wall-time spent in the linear solver.}
 The black-box AMG preconditioner
-{\color{red} demonstrates the same \changeJP{trend in terms of} the number of CG iterations, but}
+{\color{red} demonstrate the same behavior of the number of CG iterations, however it}
 requires many more iterations for convergence.
-\changeJP{
-Consequently, we can conclude that for the here considered finite strain elasticity problem (i) the selected smoother is suitable, and that (ii) although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid.
+{\color{red}
+Consequently, we can conclude that, although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid for the here considered finite strain elasticity problem and that the considered smoother is suitable.
 }
@@
 Additionally, we compare solution times of the preconditioned iterative solvers.
-Our goal is to \changeJP{determine} whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.
-}
-\changeJP{We note that although the well-established implementation of the AMG preconditioner \cite{Gee2006a} is the best available choice for the matrix-based implementation of the boundary value problem, the current implementation of the matrix-free GMG preconditioner is not yet optimized and can be improved further.}
+Our goal is to see whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.}
 Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
-{\color{red}Most importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
-\changeJP{In comparison}, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
+{\color{red}Importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
+On another hand, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
 Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
-Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for \changeJP{engineering problems
-with compressible hyperelastic} finite strain material models.
+Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for engendering problems
+with finite strain material models.
 Extension of this study from the node level to the cluster level is beyond the scope of this work,
@@
 To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to partial differential equations arising in this case.}
-{\color{red}Among the three \changeJP{examined} algorithms,}
-the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for polynomial degrees higher than one in 3D and polynomial degrees higher than two {\color{red} in 2D}.
-{\color{red}
-Given that we did not utilize the specific form of the material part of the stiffness tensor in this case, we conclude}
+{\color{red}Among the three algorithms,}
+the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for quadratic elements in 3D and cubic {\color{red}elements in 2D}.
+{\color{red}
+Given that in this case we did not utilize the specific form of the material part of the stiffness tensor, we conclude}
 that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
@@
 {\color{red}
-The roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
-We identified that one of the bottlenecks is related to the quadrature loop on each element where double contractions of fourth-order and second-order symmetric tensor
-as well as single contraction of two second-order tensors is performed.
-% Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by \texttt{deal.II}.}
-
-{\color{red}
-We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner}
+The Roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
+We identified that one of the bottlenecks is related to the quadrature loop on each element where double contractions of fourth order and second order symmetric tensor
+as well as single contraction of two second order tensors is performed.
+Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by the \texttt{deal.II}.}
+
+{\color{red}
+We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner
+}
 with standard geometric transfer operations between each level.
@@
 leads to a solution approach that is faster than the matrix-based AMG {\color{red}for all polynomial degrees studied here.
-Most importantly, the wall time per degree of freedom for solving the problem is close to being constant.
-On the other hand, the matrix-based implementation with AMG becomes prohibitively expensive for higher-order bases.
+Importantly the solution wall time per degree of freedom is close to being constant.
+On the contrary, the matrix-based implementation with AMG becomes prohibitively expensive for higher order bases.
 We conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with
@@
 Our future work will be focused on adopting the proposed matrix-free multigrid solution approach to $FE^2$ homogenization
-as well studying its behavior on the cluster level
-% as opposed to node level considered here.
-}
+as well studying its behavior on the cluster level as opposed to node level, considered here.}
 

### 2019-02-13 doc/large_strain.tex
minor changes in intro and coloring of the new section

--- 2019-02-10 doc/large_strain.tex
@@
 Sections \ref{sec:fe} and \ref{sec: Description of numerical framework} cover the finite element discretization and provide details of the numerical framework with which this study is implemented.
-In \ref{sec:mf} we describe the different matrix-free operator evaluation approaches that are evaluated in this study, and in section \ref{sec:gmg} we propose a geometric multigrid preconditioner that is suitable for finite-strain hyperelasticity.
+In \ref{sec:mf} we describe the different matrix-free operator implementation approaches that are evaluated in this study, and in section \ref{sec:gmg} we propose a geometric multigrid preconditioner that is suitable for matrix-free implementation of finite-strain hyperelasticity.
 }
@@
 
+\changeJP{
 The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
@@
 Figure \ref{fig:miehe_fine_level} illustrates a distributed MPI decomposition on the fine mesh level for the 2D problem that will be introduced later in Section \ref{sec:example}.
-
+}
+
+\changeJP{
 In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
@@
 As will be detailed later, the pre-existing matrix-free framework that was used offers further opportunity for low-level parallelism that may be exploited.
+}
 
@@
 as well as single contraction of two second-order tensors is performed.
-% Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by \texttt{deal.II}.}
+% Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by \texttt{deal.II}.
+}
 

### 2019-02-13 doc/large_strain.tex
minor in JPs section

@@
 For the matrix-based approach, the Trilinos \cite{Heroux2005} library, version 12.12.1 was leveraged;
-the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the package ML package \cite{Gee2006a}.
+the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
 The second implementation, which is discussed in detail in the following Sections, utilized a matrix-free approach in conjunction with a geometric-multigrid preconditioner.
-As will be detailed later, the pre-existing matrix-free framework that was used offers further opportunity for low-level parallelism that may be exploited.
+As will be detailed later, the pre-existing matrix-free framework that was used offers further opportunity for low-level parallelism that we exploit.
 }

### 2019-02-13 doc/large_strain.tex
color footnote

@@
 For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{%
+{\color{red}
 For the heterogeneous materials considered here,
@@
 This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
+}
 }.

### 2019-02-13 doc/large_strain.tex
Changes: Denis, JP discussions

@@
 Sections \ref{sec:fe} and \ref{sec: Description of numerical framework} cover the finite element discretization and provide details of the numerical framework with which this study is implemented.
-In \ref{sec:mf} we describe the different matrix-free operator implementation approaches that are evaluated in this study, and in section \ref{sec:gmg} we propose a geometric multigrid preconditioner that is suitable for matrix-free implementation of finite-strain hyperelasticity.
+In \ref{sec:mf} we describe the different matrix-free operator implementations that are evaluated in this study, and in section \ref{sec:gmg} we propose a geometric multigrid preconditioner that is suitable for the matrix-free implementation of finite-strain hyperelasticity.
 }
@@
 
-\section{\changeJP{Description of numerical framework}
+\section{\changeJP{Description of the utilized numerical framework}
 \label{sec: Description of numerical framework}
@@
 The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
-This library offers a number of paradigms by which to parallelize a finite-element code.
+This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aims to support exascale computing.
 In this work we employ \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
@@
 
-\changeJP{
+\changeJP{%
 In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
 For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
-For the matrix-based approach, the Trilinos \cite{Heroux2005} library, version 12.12.1 was leveraged;
+}
+
+\changeJP{%
+The discrete tangent operator described by \eqref{eq:algebraic_tangent} can be implemented in one of two ways, namely through a classical matrix-based approach or, alternatively, a matrix-free approach.
+For the matrix-based approach adopted in this study, the Trilinos \cite{Heroux2005} library was leveraged;
 the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
+This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is possible in  \texttt{deal.II}.
+}
+
+\changeJP{%
 The second implementation, which is discussed in detail in the following Sections, utilized a matrix-free approach in conjunction with a geometric-multigrid preconditioner.
@@
 \changeJP{We conduct two performance studies using the described discretization, the first being for only matrix-vector multiplication, and the second for the preconditioned iterative linear solver.}
-The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library.
+The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library version 12.12.1.
 The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
@@
 memory bandwidth\footnote{As measured with the \textit{load\_avx} benchmark of \textit{likwid-bench} \cite{likwid}.} and peak performance are $B=47.2 \,\rm{GB/s}$ and $P=176\, \rm{GFlop/s}$, respectively.
-Turbo mode of CPU was disabled.
+Turbo mode of the CPU was disabled.
 }
@@
 
-{\color{red}
+%{\color{red}
+%Additionally, we compare solution times of the preconditioned iterative solvers.
+%Our goal is to \changeJP{determine} whether the matrix-free approach with a GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.
+%}
+\changeJP{%
 Additionally, we compare solution times of the preconditioned iterative solvers.
-Our goal is to \changeJP{determine} whether the matrix-free approach with GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.
-}
-\changeJP{We note that although the well-established implementation of the AMG preconditioner \cite{Gee2006a} is the best available choice for the matrix-based implementation of the boundary value problem, the current implementation of the matrix-free GMG preconditioner is not yet optimized and can be improved further.}
+Our goal is to \changeJP{determine} whether the matrix-free approach with a GMG preconditioner can be faster than the matrix-based approach with a well-established implementation of an AMG preconditioner.
+}
+%\changeJP{We note that although the well-established implementation of the AMG preconditioner \cite{Gee2006a} is the best available choice for the matrix-based implementation of the boundary value problem, the current implementation of the matrix-free GMG preconditioner is not yet optimized and can be improved further.}
 Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.

### 2019-02-13 doc/large_strain.tex
Reinstate comment on lack of p-refinement for GMG
Add note on GMG framework / level operator comparison

@@
 Our goal is to \changeJP{determine} whether the matrix-free approach with a GMG preconditioner can be faster than the matrix-based approach with a well-established implementation of an AMG preconditioner.
+We note specifically that the comparison is restricted to the choice of level operators used with the pre-existing GMG framework, and not the framework itself. 
 }

### 2019-02-16 Calculations/post_process_likwid.py
adjust y-range for Roofline plots for breakdown (one point was out). change bandwidth fontisze to be consistent with roofs

@@ -259,7 +259,7 @@ for d in likwid_data:
 plt.xlabel('intensity (Flop/byte)')
 plt.ylabel('performance (GFlop/s)')
 #plt.ylim(top=300,bottom=1)
-ymin = 2**2 if not args.breakdown else 2**(0) - 0.1
+ymin = 2**2 if not args.breakdown else 2**(0) - 0.3
 plt.ylim(top=2**8,bottom=ymin)
 xmax = 2**6 + 10 if args.breakdown else 2**4
 xmin = 2**(-4)+0.01 if not args.breakdown else 2**(-5) + 1./80
@@ -269,8 +269,8 @@ plt.axes().yaxis.set_major_formatter(mp.ticker.FuncFormatter(lambda x, pos: '{0}
 plt.axes().xaxis.set_major_formatter(mp.ticker.FuncFormatter(lambda x, pos: '1/{0}'.format(int(round(1./x))) if x < 1.0 else '{0}'.format(int(round(x))) ))
 
 ang = 45
-y_pos = 13
-plt.text(xmin, y_pos, 'B={:.1f} GB/s'.format(B), rotation=ang)
+y_pos = 7.5 if args.breakdown else 8.5
+plt.text(xmin, y_pos, 'B={:.1f} GB/s'.format(B), rotation=ang, fontsize=14)
 
 x_pos = 4 if not args.breakdown else 16
 plt.text(x_pos,200,'Peak DP', fontsize=14)

### 2019-02-19 Calculations/emmy_benchmark_intel_64n.sh
update weak scaling results with :noturbo

@@ -1,6 +1,6 @@
 #!/bin/bash
 #   allocate 64 nodes with 40 CPU per node for 3 hours:
-#PBS -l nodes=64:ppn=40,walltime=3:00:00
+#PBS -l nodes=64:ppn=40:noturbo,walltime=3:00:00
 #   job name
 #PBS -N dealii_mf_cm64
 #   stdout and stderr files:
@@ -12,3 +12,4 @@ cd /home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/Calculations/
 module load intel64/18.0up03
 
 mpirun -np 1280 /home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/_build/main /home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/Calculations/__holes_3d_p2q3r5_MF_CG_gmg_tensor4_64node.prm 2>&1 | tee __holes_3d_p2q3r5_MF_CG_gmg_tensor4_64node.toutput
+mv __holes_3d_p2q3r5_MF_CG_gmg_tensor4_64node.toutput Emmy_RRZE_scaling/holes_3d_p2q3r5_MF_CG_gmg_tensor4/holes_3d_p2q3r5_MF_CG_gmg_tensor4_64node.toutput

### 2019-02-19 Calculations/emmy_benchmark_intel_8n.sh

@@ -1,6 +1,6 @@
 #!/bin/bash
 #   allocate 8 nodes with 40 CPU per node for 3 hours:
-#PBS -l nodes=8:ppn=40,walltime=3:00:00
+#PBS -l nodes=8:ppn=40:noturbo,walltime=3:00:00
 #   job name
 #PBS -N dealii_mf_cm8
 #   stdout and stderr files:
@@ -12,3 +12,4 @@ cd /home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/Calculations/
 module load intel64/18.0up03
 
 mpirun -np 160 /home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/_build/main /home/woody/iwtm/iwtm108/deal.ii-mf-elasticity/Calculations/__holes_3d_p2q3r4_MF_CG_gmg_tensor4_8node.prm 2>&1 | tee __holes_3d_p2q3r4_MF_CG_gmg_tensor4_8node.toutput
+mv __holes_3d_p2q3r4_MF_CG_gmg_tensor4_8node.toutput Emmy_RRZE_scaling/holes_3d_p2q3r4_MF_CG_gmg_tensor4/holes_3d_p2q3r4_MF_CG_gmg_tensor4_8node.toutput

### 2019-02-19 doc/large_strain.tex
whitespace
add weak scaling table

@@
 \input{parallelization_3d.tex}
+{\color{red}
 Table \ref{tab:numbers_2d} and Table \ref{tab:numbers_3d} show the speed-up obtained due to explicit SIMD vectorization intrinsics and due to MPI.
@@
 Furthermore, the improvement due to explicit intrinsics for the data used in the quadrature loop, is comparable to the ones in \cite[Figure 6]{kronbichler2017fastcomputer} where the speedup factor is $1.2-1.8$ for a similar architecture.
+}
 
@@
 Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for \changeJP{engineering problems
-with compressible hyperelastic} finite strain material models.
-Extension of this study from the node level to the cluster level is beyond the scope of this work,
-however we hypothesize that the matrix-free and multigrid frameworks provided by \texttt{deal.II} library would provide a competitive solution for that case as well.}
+with compressible hyperelastic} finite strain material models.}
+
+\changeJP{Extension of this study from the node level to the cluster level is beyond the scope of this work,
+however we provide a preliminary weak scaling results that confirm that the matrix-free and multigrid frameworks provided by \texttt{deal.II} library can provide a competitive solution for that case as well.
+Table \ref{tab:weak_3d} demonstrates a very good weak scaling up to 1280 processes (64 nodes, the maximum we could request on the local HPC cluster) of the wall time for a single matrix-vector product and a single iteration in the CG solver for the 3D problem with quadratic basis.
+}
+
+\input{weak_scaling_3d.tex}
 

### 2019-02-19 Calculations/post_process_weak.py

@@ -10,15 +10,12 @@ parser = argparse.ArgumentParser(
     formatter_class=argparse.ArgumentDefaultsHelpFormatter)
 parser.add_argument('--prefix', metavar='prefix', default='Emmy_RRZE_scaling',
                     help='A folder prefix to look for benchmark results')
-parser.add_argument('--dim', metavar='dim', default=2, type=int,
+parser.add_argument('--dim', metavar='dim', default=3, type=int,
                     help='Dimension (2 or 3)')
 
 args = parser.parse_args()
 
 prefix = args.prefix if args.prefix.startswith('/') else os.path.join(os.getcwd(), args.prefix)
-
-# We will generate a table similar to Table 1 in Kronbichler 2012
-# table_data = [ [ np.nan for i in range(12)] for i in range(n_rows)]
 
 # Go through all the suffixes and parse the data
 files = collection_toutput_files(prefix)
@@ -27,84 +24,55 @@ files = collect_timing_files(prefix)
 print 'Gather data from {0}'.format(prefix)
 print 'found {0} files'.format(len(files))
 
+table_data = []
 # no go through the files:
+gref = 3
 for f in files:
     parsed_tup = parse_timing_file(f)
     table = parsed_tup[4]
-    cores = parsed_tup[6]
+    tot_cg = parsed_tup[6]
+    cores = parsed_tup[7]
     dofs = parsed_tup[1]
 
     vmult = table['vmult (MF)']
+    cg = table['Linear solver']
 
     t = float(vmult[1])/int(vmult[0])
+    tcg = float(cg[1])/int(tot_cg)
 
-    print 'Dofs={0} cores={1} vmult={2}'.format(dofs,cores,t)
+    print 'Dofs={0} cores={1} vmult={2} cg_it={3}'.format(dofs,cores,t, tcg)
 
-    # # these files contain a single region
-    # result = parse_likwid_file(f,'LIKWID_MARKER_CLOSE')['vmult_MF']
-
-    # # depending on the run, the LIKWID output may not have Sum table
-    # if 'Metric_Sum' not in result:
-    #     data = result['Metric']
-    #     flops = data['MFLOP/s'][0]
-    #     runtime = data['Runtime (RDTSC) [s]'][0]
-    #     assert abs(float(data['Clock [MHz]'][0]) - clock_speed * 1000) < clock_speed
-    # else:
-    #     data = result['Metric_Sum']
[…]

### 2019-02-19 Calculations/utilities.py

@@ -45,6 +45,7 @@ def parse_timing_file(f):
     timing = {}
     # reset CG iterations in case AMG did not have enough memory
     cg_iterations = np.nan
+    tot_cg_iterations = np.nan
     for line in fin:
         if 'running with' in line:
             cores = int(re.findall(pattern,line)[0])
@@ -72,6 +73,8 @@ def parse_timing_file(f):
 
         elif 'Average CG iter =' in line:
             cg_iterations = int(re.findall(pattern,line)[0])
+        elif 'Total CG iter = ' in line:
+            tot_cg_iterations = int(re.findall(pattern,line)[0])
 
         if ready:
             if dividers in line:
@@ -95,7 +98,7 @@ def parse_timing_file(f):
             ready = True
 
     # return data as a tuple
-    return tuple((p, dofs, tr_memory, mf_memory, timing, cg_iterations, cores))
+    return tuple((p, dofs, tr_memory, mf_memory, timing, cg_iterations, tot_cg_iterations, cores))
 
 
 def parse_likwid_file(filename, last_line = ''):

### 2019-02-19 doc/large_strain.tex
minor addition to the roofline interpretation of algorithms

@@
 we observe a marginal increase in computational intensity and almost unchanged performance.
-This coincides with our expectation that the latter algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
+This coincides with our expectation that the latter algorithm requires transferring fewer bytes to perform operations on a single quadrature point
+(compare line 9 in Algorithm \ref{alg:mf_tensor4} that loads a symmetric rank-4 tensor to the line 9 in Algorithm \ref{alg:mf_tensor2} that evaluates the
+same contraction by loading two scalars $c_1$ and $c_2$).
 On the other hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
+This is not surprising as at each quadrature point we need to re-evaluate the Kirchoff stress $\gz \tau$.
 Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional

### 2019-02-20 doc/large_strain.tex

@@
 \changeJP{Extension of this study from the node level to the cluster level is beyond the scope of this work,
-however we provide a preliminary weak scaling results that confirm that the matrix-free and multigrid frameworks provided by \texttt{deal.II} library can provide a competitive solution for that case as well.
-Table \ref{tab:weak_3d} demonstrates a very good weak scaling up to 1280 processes (64 nodes, the maximum we could request on the local HPC cluster) of the wall time for a single matrix-vector product and a single iteration in the CG solver for the 3D problem with quadratic basis.
+however we show preliminary weak scaling results that confirm that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for that case as well.
+Table \ref{tab:weak_3d} demonstrates a good weak scaling up to 1280 processes (64 nodes, the maximum we could request on the local HPC cluster) of the wall time for a single matrix-vector product and a single iteration in the CG solver for the 3D problem with quadratic basis.
 }

### 2019-02-22 doc/large_strain.tex
Adjust discussion on weak scaling

@@
 
-\changeJP{Extension of this study from the node level to the cluster level is beyond the scope of this work,
-however we show preliminary weak scaling results that confirm that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for that case as well.
-Table \ref{tab:weak_3d} demonstrates a good weak scaling up to 1280 processes (64 nodes, the maximum we could request on the local HPC cluster) of the wall time for a single matrix-vector product and a single iteration in the CG solver for the 3D problem with quadratic basis.
+\changeJP{Extension of this study from the node level to the cluster level is beyond the scope of this work.
+However, as a step in this direction we have performed a preliminary study of the weak scalability of the implementation.
+Table \ref{tab:weak_3d} demonstrates a good weak scaling up to 1280 processes\footnote{%
+This corresponds to 64 nodes, which is the maximum we could request on the RRZE Emmy HPC cluster.
+} of the wall time per matrix-vector product and per iteration in the CG solver for the 3D problem with a tri-quadratic basis.
+This confirms that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for cluster-sized problems as well.
 }

### 2019-02-22 doc/large_strain.tex
Change colour of J-P's suggestions
Add a reference to one of the largest deal.II runs

@@
 
-\newcommand{\changeJP}[1]{\textcolor{blue}{#1}}
+\newcommand{\changeJP}[1]{\textcolor{red}{#1}}
 \newcommand{\todojp}[1]{\todo[inline,color=red!40,size=\small]{TODO (J-P): #1}}
@@
 } of the wall time per matrix-vector product and per iteration in the CG solver for the 3D problem with a tri-quadratic basis.
-This confirms that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for cluster-sized problems as well.
+This confirms that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for cluster-sized problems as well,
+also compare \cite[Figure 7]{Krank2017} for scaling results for up to 147'456 CPU cores and 34.4 billion degrees of freedom for an incompressible flow problem.
 }

### 2019-02-22 doc/large_strain.tex
Fix difit separator

--- 2019-02-22 doc/large_strain.tex
@@
 } of the wall time per matrix-vector product and per iteration in the CG solver for the 3D problem with a tri-quadratic basis.
-This confirms that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for cluster-sized problems as well.
+This confirms that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for cluster-sized problems as well,
+also compare \cite[Figure 7]{Krank2017} for scaling results for up to 147,456 CPU cores and 34.4 billion degrees of freedom for an incompressible flow problem.
 }

### 2019-04-29 doc/large_strain.tex
remove all colors (can put back easily if needed)

@@
 
-\newcommand{\changeJP}[1]{\textcolor{red}{#1}}
+%\newcommand{\changeJP}[1]{\textcolor{red}{#1}}
+%\newcommand{\changeDD}[1]{\textcolor{red}{#1}}
+\newcommand{\changeDD}[1]{#1}
+\newcommand{\changeJP}[1]{#1}
+
 \newcommand{\todojp}[1]{\todo[inline,color=red!40,size=\small]{TODO (J-P): #1}}
@@
   \begin{abstract}
-    The performance of finite element solvers on modern computer architectures is typically memory bound {\color{red} for sufficiently large problems}.
+    The performance of finite element solvers on modern computer architectures is typically memory bound \changeDD{ for sufficiently large problems}.
     The main cause for this is that loading matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
@@
     and employ them to define a geometric multigrid preconditioner.
-    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner {\color{red} on a single node} for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
+    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner \changeDD{ on a single node} for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
     \changeJP{We conclude that the application of matrix-free methods to finite-strain solid mechanics is promising, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
@@
 
-The performance of finite element solvers on modern computer architectures is typically memory bound {\color{red} for sufficiently large problems}.
+The performance of finite element solvers on modern computer architectures is typically memory bound \changeDD{ for sufficiently large problems}.
 The main cause for this is that loading pre-computed matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
@@
 In this work, we perform \changeJP{a preliminary investigation of} different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost
-and performance {\color{red} on a single node of a high performance computing cluster.}
+and performance \changeDD{ on a single node of a high performance computing cluster.}
 \changeJP{
@@
 
-{\color{red}For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.}
+\changeDD{For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.}
 %
 In order to obtain gradients with respect to the spatial configuration,
-{\color{red}
+\changeDD{
 that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem \changeJP{derived with a Lagrangian formulation},}
@@
 
-{\color{red}
+\changeDD{
 Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach \changeJP{, each of which have a different level of abstraction, algorithmic complexity, and memory requirements}.
@@
   \ForEach{ element $K \in \Omega^h$ }{
-        {\color{red}extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ } \;
+        \changeDD{extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ } \;
         evaluate the following tensors at each quadrature point using sum factorization :\\
@@
   \ForEach{ element $K \in \Omega^h$ }{
-        {\color{red}extract local vector values on this element $\gz x_K$ } \;
+        \changeDD{extract local vector values on this element $\gz x_K$ } \;
         evaluate the following tensors at each quadrature point using sum factorization :\\
@@
 Note that the SIMD vectorization
-{\color{red}
+\changeDD{
 in \texttt{deal.II} \cite{dealII90}
@@
 is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
-{\color{red}
+\changeDD{
 The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements\footnote{%
@@
 For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{%
-{\color{red}
+\changeDD{
 For the heterogeneous materials considered here,
@@
 }.
-For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second-order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
+For example, \changeDD{the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second-order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
 This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
@@
 
-{\color{red}
+\changeDD{
 To summarize, in this section we propose a method by which to construct level tangent operators for finite-strain elasticity to be employed within the GMG preconditioner.
@@
 The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
-{\color{red}
+\changeDD{
 We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach \changeJP{with a high polynomial order finite element}.
@@
 The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
-{\color{red}
+\changeDD{
 The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
@@
 is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
-On the coarsest level, we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
+On the coarsest level, we use \changeDD{the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
 Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}
@@
 For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
-{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. \changeJP{one} fully occupied node.}
+\changeDD{Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. \changeJP{one} fully occupied node.}
 
@@
 Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
-{\color{red}
+\changeDD{
 We are interested in the wall-clock time (average over 10 runs per each linear solver step) per DoF as a first metric.}
 As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
-{\color{red}as the matrix becomes more and more dense.
+\changeDD{as the matrix becomes more and more dense.
 The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
@@
 Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
-{\color{red}
+\changeDD{
 described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.}
@@
 The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
-There is little difference {\color{red}in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'').
+There is little difference \changeDD{in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'').
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
@@
 It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
-{\color{red}as long as the resulting tangent operator is expressed via \eqref{eq:algebraic_tangent}.}
+\changeDD{as long as the resulting tangent operator is expressed via \eqref{eq:algebraic_tangent}.}
 
@@
 Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point \changeJP{(``tensor4'')}, the matrix-free approach takes much less memory than its matrix-based counterpart.
-{\color{red}
+\changeDD{
 In those graphs, we report the \changeJP{overall} memory needed to cache additional quantities required for the tangent operator here considered,
@@
 % likwid-mpirun -np 10 -nperdomain S:10 -g MEM_DP -m $EXC
-{\color{red}
+\changeDD{
 Finally, we analyze the node-level performance of the two approaches using the roofline model.
@@
 
-{\color{red}
+\changeDD{
 Figure \ref{fig:roofline} shows results of the roofline performance model.
@@
 Unfortunately, performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.
-
+}
+
+\changeDD{
 When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}) caching strategy,
@@
 independent on the right hand side vector (steps 8--11 in Algorithm \ref{alg:mf_scalar}), which makes it the worst option among the three.
+}
 
@@
 
+\changeDD{
 In order to have a better understanding of the behavior of the most promising ``tensor4'' implementation,
@@
 \input{parallelization_3d.tex}
-{\color{red}
+\changeDD{
 Table \ref{tab:numbers_2d} and Table \ref{tab:numbers_3d} show the speed-up obtained due to explicit SIMD vectorization intrinsics and due to MPI.
@@
 
-{\color{red}
+\changeDD{
 To summarize this section, we can attribute the speed-up of the matrix-free finite-strain tangent operators for higher-order elements to two factors.
@@
 Next, we evaluate the performance of the proposed geometric multigrid preconditioner.
-{\color{red}
+\changeDD{
 Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end,}
-we consider {\color{red}the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
+we consider \changeDD{the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
 Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
-{\color{red} noticeable} increase in the average number of CG iterations for different polynomial degrees
-{\color{red} for the GMG preconditioner.
+\changeDD{ noticeable} increase in the average number of CG iterations for different polynomial degrees
+\changeDD{ for the GMG preconditioner.
 Although not reported in this section, we identified that
@@
 The black-box AMG preconditioner
-{\color{red} demonstrates the same \changeJP{trend in terms of} the number of CG iterations, but}
+\changeDD{ demonstrates the same \changeJP{trend in terms of} the number of CG iterations, but}
 requires many more iterations for convergence.
@@
 
-%{\color{red}
+%\changeDD{
 %Additionally, we compare solution times of the preconditioned iterative solvers.
@@
 %\changeJP{We note that although the well-established implementation of the AMG preconditioner \cite{Gee2006a} is the best available choice for the matrix-based implementation of the boundary value problem, the current implementation of the matrix-free GMG preconditioner is not yet optimized and can be improved further.}
-Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
-{\color{red}Most importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
+Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for \changeDD{all} polynomial degrees.
+\changeDD{Most importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
 \changeJP{In comparison}, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
@@
 
-In this contribution, we proposed and numerically investigated {\color{red}three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
-{\color{red}
+In this contribution, we proposed and numerically investigated \changeDD{three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
+\changeDD{
 To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to partial differential equations arising in this case.}
-{\color{red}Among the three \changeJP{examined} algorithms,}
-the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for polynomial degrees higher than one in 3D and polynomial degrees higher than two {\color{red} in 2D}.
-{\color{red}
+\changeDD{Among the three \changeJP{examined} algorithms,}
+the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for polynomial degrees higher than one in 3D and polynomial degrees higher than two \changeDD{ in 2D}.
+\changeDD{
 Given that we did not utilize the specific form of the material part of the stiffness tensor in this case, we conclude}
@@
 
-{\color{red}
+\changeDD{
 The roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
@@
 
-{\color{red}
+\changeDD{
 We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner}
@@
 The numerical studies indicate that the proposed preconditioner
-{\color{red}
+\changeDD{
 leads to much fewer iterations of the iterative solver as compared to the algebraic multigrid preconditioner.
 The multigrid matrix-free preconditioner also}
-leads to a solution approach that is faster than the matrix-based AMG {\color{red}for all polynomial degrees studied here.
+leads to a solution approach that is faster than the matrix-based AMG \changeDD{for all polynomial degrees studied here.
 Most importantly, the wall time per degree of freedom for solving the problem is close to being constant.
@@
 
-D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to {\color{red}Martin Kronbichler (TU Munich),} Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for \changeJP{fruitful discussions} on matrix-free operator evaluation approaches.
+D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to \changeDD{Martin Kronbichler (TU Munich),} Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for \changeJP{fruitful discussions} on matrix-free operator evaluation approaches.
 

### 2019-04-30 doc/large_strain.tex
use mathcal instead of cal
use rm for d and D instead of mbox, somehow makes arXiv happier
remove changeJP/changeDD/todojp and other commented text

--- 2019-02-22 doc/large_strain.tex
@@
 
-%
-% Definitions:
-%
 \def\gz  #1{           \mbox{$\boldsymbol{#1}$}}
@@
 \def\div {\mbox{div\,}}
-\def\d {\mbox{d}}
-\def\D {\mbox{D}}
-\def\mcl  #1{               {\cal #1}}
+\def\d {{\rm d}}
+\def\D {{\rm D}}
+\def\mcl  #1{               {\mathcal #1}}
 \DeclareMathOperator{\trace}{tr}
 \newcommand{\diff}{\mathop{}\!\mathrm{d}}
-
-\newcommand{\changeJP}[1]{\textcolor{red}{#1}}
-\newcommand{\todojp}[1]{\todo[inline,color=red!40,size=\small]{TODO (J-P): #1}}
 
@@
   \begin{abstract}
-    The performance of finite element solvers on modern computer architectures is typically memory bound {\color{red} for sufficiently large problems}.
+    The performance of finite element solvers on modern computer architectures is typically memory bound for sufficiently large problems.
     The main cause for this is that loading matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
@@
 
-    \changeJP{To date, there have been few (if any) assessments into the applicability of the matrix-free approach to problems in solid mechanics.
+    To date, there have been few (if any) assessments into the applicability of the matrix-free approach to problems in solid mechanics.
     In this work, we perform an initial investigation on the application of the matrix-free approach to problems in quasi-static finite-strain hyperelasticity to determine whether it is viable for further extension.
     Specifically, we study different numerical implementations of the finite element tangent operator, and determine whether generalized methods of incorporating complex constitutive behavior might be feasible.
-    }
     In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
     and employ them to define a geometric multigrid preconditioner.
-    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner {\color{red} on a single node} for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
-    \changeJP{We conclude that the application of matrix-free methods to finite-strain solid mechanics is promising, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
+    The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner on a single node for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
+    We conclude that the application of matrix-free methods to finite-strain solid mechanics is promising, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.
   \end{abstract}
@@
 
-The performance of finite element solvers on modern computer architectures is typically memory bound {\color{red} for sufficiently large problems}.
+The performance of finite element solvers on modern computer architectures is typically memory bound for sufficiently large problems.
 The main cause for this is that loading pre-computed matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
@@
 
-In this work, we perform \changeJP{a preliminary investigation of} different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost
-and performance {\color{red} on a single node of a high performance computing cluster.}
-\changeJP{
-The \texttt{deal.II} \cite{dealII90} finite element library, which provides MPI parallelization together with a \changeJP{generalized matrix-free framework that incorporates SIMD vectorization} and an interface to high-performance, MPI distributed matrix-based libraries, is used for this study.
-}
+In this work, we perform a preliminary investigation of different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost
+and performance on a single node of a high performance computing cluster.
+The \texttt{deal.II} \cite{dealII90} finite element library, which provides MPI parallelization together with a generalized matrix-free framework that incorporates SIMD vectorization and an interface to high-performance, MPI distributed matrix-based libraries, is used for this study.
 In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
@@
 The paper is organized as follows: In Section \ref{sec:theory} we briefly introduce the partial differential equations used in  finite-strain solid mechanics.
-\changeJP{
 Sections \ref{sec:fe} and \ref{sec: Description of numerical framework} cover the finite element discretization and provide details of the numerical framework with which this study is implemented.
 In \ref{sec:mf} we describe the different matrix-free operator implementations that are evaluated in this study, and in section \ref{sec:gmg} we propose a geometric multigrid preconditioner that is suitable for the matrix-free implementation of finite-strain hyperelasticity.
-}
 The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}.
-\changeJP{Lastly, the} results are summarized in Section \ref{sec:summary}.
+Lastly, the results are summarized in Section \ref{sec:summary}.
 
@@
 \subsection{Weak form}
-%
-% Kinematics
-%
 The deformation of a body $\mcl B$ from the referential configuration $\mcl B_0$ to the spatial configuration $\mcl B_t$ at time $t$
@@
 
-%
-% Deformation gradient
-%
 The tangent map is linear such that
@@
 The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian $J = \det \gz F > 0$ has to be positive.
-% The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
 We shall denote the push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by a linear transformation $\chi$ defined as
@@
 
-%
-% strain measures
-%
 In what follows, we parameterize the material behavior using the right Cauchy-Green tensor $\gz C := \gz F^T \cdot \gz F$.
@@
 where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively.
-\changeJP{
 Derived using statistical thermodynamics applied to cross-linked polymers, the Neo-Hookean model \cite{Treloar1975a,Treloar1976a} is commonly used to model rubber-like materials in the finite-strain regime.
-}
 It can be shown that the first and second derivatives of the strain energy function are given by
@@
 
-\section{\changeJP{Description of the utilized numerical framework}
+\section{Description of the utilized numerical framework
 \label{sec: Description of numerical framework}
@@
 
-\changeJP{
 The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
@@
 Figure \ref{fig:miehe_fine_level} illustrates a distributed MPI decomposition on the fine mesh level for the 2D problem that will be introduced later in Section \ref{sec:example}.
-}
-
-\changeJP{%
+
 In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
 For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
-}
-
-\changeJP{%
+
 The discrete tangent operator described by \eqref{eq:algebraic_tangent} can be implemented in one of two ways, namely through a classical matrix-based approach or, alternatively, a matrix-free approach.
@@
 This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is possible in  \texttt{deal.II}.
-}
-
-\changeJP{%
+
 The second implementation, which is discussed in detail in the following Sections, utilized a matrix-free approach in conjunction with a geometric-multigrid preconditioner.
 As will be detailed later, the pre-existing matrix-free framework that was used offers further opportunity for low-level parallelism that we exploit.
-}
 
@@
 
-{\color{red}For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.}
+For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.
 %
 In order to obtain gradients with respect to the spatial configuration,
-{\color{red}
-that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem \changeJP{derived with a Lagrangian formulation},}
+that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem derived with a Lagrangian formulation,
 we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
@@
 
-{\color{red}
-Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach \changeJP{, each of which have a different level of abstraction, algorithmic complexity, and memory requirements}.
-\changeJP{They are introduced in order of lowest memory footprint to largest:}
-}
+Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach, each of which have a different level of abstraction, algorithmic complexity, and memory requirements.
+They are introduced in order of lowest memory footprint to largest:
 
@@
   \ForEach{ element $K \in \Omega^h$ }{
-        {\color{red}extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ } \;
+        extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ \;
         evaluate the following tensors at each quadrature point using sum factorization :\\
@@
   \ForEach{ element $K \in \Omega^h$ }{
-        {\color{red}extract local vector values on this element $\gz x_K$ } \;
+        extract local vector values on this element $\gz x_K$ \;
         evaluate the following tensors at each quadrature point using sum factorization :\\
@@
 Note that the SIMD vectorization
-{\color{red}
 in \texttt{deal.II} \cite{dealII90}
-}
 is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
-{\color{red}
 The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements\footnote{%
-\changeJP{In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
+In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
 This motivates the caching strategy employed in Algorithm 3, as the pre-caching of the chosen data ensures a context in which SIMD parallelization of subsequent operations is guaranteed.
-Algorithms 1 and 2 sacrifice some of this generality for decreased memory requirements.}
+Algorithms 1 and 2 sacrifice some of this generality for decreased memory requirements.
 }.
 For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{%
-{\color{red}
 For the heterogeneous materials considered here,
@@
 This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
-}
 }.
-For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second-order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
+For example, the deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second-order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
 This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
@@
 For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
-}
 
@@
 For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
-\changeJP{Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.}
+Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.
 
@@
 
-{\color{red}
 To summarize, in this section we propose a method by which to construct level tangent operators for finite-strain elasticity to be employed within the GMG preconditioner.
@@
 Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
-}
 
@@
 %
-\changeJP{The same distributed approach to mesh decomposition (as most evident on the finest level) is adopted on the multigrid levels.}
-Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements \changeJP{for the mesh introduced in figure \ref{fig:miehe_fine_level}}.
+The same distributed approach to mesh decomposition (as most evident on the finest level) is adopted on the multigrid levels.
+Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements for the mesh introduced in figure \ref{fig:miehe_fine_level}.
 %
@@
 
-\changeJP{%
 Aligned with the approach taken in similar previous investigations, compare e.g. \cite{kronbichler2017fast},
@@
 Examinations performed in this manner should already indicate the differences between the matrix-free and matrix-based approaches, and give some preliminary understanding of the scalability potential of the numerical implementation without confounding factors related to data exchange through interconnects via MPI.
-}
 The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
-{\color{red}
-We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach \changeJP{with a high polynomial order finite element}.
-\changeJP{Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound}.
-}
-
-%
-% Below are are implementational details that might affect the comparative result
-%
-\changeJP{We conduct two performance studies using the described discretization, the first being for only matrix-vector multiplication, and the second for the preconditioned iterative linear solver.}
+We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach with a high polynomial order finite element.
+Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound.
+
+We conduct two performance studies using the described discretization, the first being for only matrix-vector multiplication, and the second for the preconditioned iterative linear solver.
 The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library version 12.12.1.
 The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
-{\color{red}
 The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
@@
 is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
-On the coarsest level, we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
-Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}
-
-Computations \changeJP{were performed} on a single node of two clusters:
+On the coarsest level, we use the Chebyshev preconditioner as a solver \cite{Varga2009} to reduce the residual by three orders of magnitude.
+Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.
+
+Computations were performed on a single node of two clusters:
 A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM.
@@
 For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
-{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. \changeJP{one} fully occupied node.}
+Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. one fully occupied node.
 
@@
 Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
-{\color{red}
-We are interested in the wall-clock time (average over 10 runs per each linear solver step) per DoF as a first metric.}
+We are interested in the wall-clock time (average over 10 runs per each linear solver step) per DoF as a first metric.
 As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
-{\color{red}as the matrix becomes more and more dense.
+as the matrix becomes more and more dense.
 The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
@@
 Both factors increase the memory consumption and make element operations more expensive.
-Nonetheless, }
-the matrix-free implementation is faster than the matrix-based counterpart already for \changeJP{bi-cubic Lagrangian basis in 2D and tri-quadratic in 3D}.
+Nonetheless,
+the matrix-free implementation is faster than the matrix-based counterpart already for bi-cubic Lagrangian basis in 2D and tri-quadratic in 3D.
 Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
-{\color{red}
-described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.}
+described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.
 Two conclusions can be drawn from these results:
 The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
-There is little difference {\color{red}in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'').
+There is little difference in terms of the wall-clock time per DoF between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'').
 This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
@@
 It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
-{\color{red}as long as the resulting tangent operator is expressed via \eqref{eq:algebraic_tangent}.}
+as long as the resulting tangent operator is expressed via \eqref{eq:algebraic_tangent}.
 
 As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
-Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point \changeJP{(``tensor4'')}, the matrix-free approach takes much less memory than its matrix-based counterpart.
-{\color{red}
-In those graphs, we report the \changeJP{overall} memory needed to cache additional quantities required for the tangent operator here considered,
+Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point (``tensor4''), the matrix-free approach takes much less memory than its matrix-based counterpart.
+In those graphs, we report the overall memory needed to cache additional quantities required for the tangent operator here considered,
 as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
 from unit to real cell \cite{kronbichler12}.
-Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both \changeJP{deformed} and undeformed
-configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.}
-
-% likwid-mpirun -np 10 -nperdomain S:10 -g MEM_DP -m $EXC
-{\color{red}
+Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deformed and undeformed
+configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.
+
 Finally, we analyze the node-level performance of the two approaches using the roofline model.
@@
 Turbo mode of the CPU was disabled.
-}
 
@@
 
-{\color{red}
 Figure \ref{fig:roofline} shows results of the roofline performance model.
@@
 we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
-(i) `read / write' denotes reading and writing from/to a global vector (\changeJP{lines 2 and 13});
-(ii) `sum factorization' denotes application of sum factorization techniques (\changeJP{lines} 4, 5 and 12);
-(iii) `quadrature loop' denotes operations performed on each quadrature (\changeJP{lines \mbox{6 -- 11}}).
+(i) `read / write' denotes reading and writing from/to a global vector (lines 2 and 13);
+(ii) `sum factorization' denotes application of sum factorization techniques (lines 4, 5 and 12);
+(iii) `quadrature loop' denotes operations performed on each quadrature (lines \mbox{6 -- 11}).
 Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one
@@
 double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
-% Further optimization of \texttt{Tensor} \changeJP{and \texttt{SymmetricTensor}}classes in \textit{deal.II} are required to
-% improve the performance.
-% Such studies, however, are beyond the scope of the current work.
-}
 
@@
 \input{parallelization_3d.tex}
-{\color{red}
+
 Table \ref{tab:numbers_2d} and Table \ref{tab:numbers_3d} show the speed-up obtained due to explicit SIMD vectorization intrinsics and due to MPI.
@@
 Furthermore, the improvement due to explicit intrinsics for the data used in the quadrature loop, is comparable to the ones in \cite[Figure 6]{kronbichler2017fastcomputer} where the speedup factor is $1.2-1.8$ for a similar architecture.
-}
-
-{\color{red}
+
 To summarize this section, we can attribute the speed-up of the matrix-free finite-strain tangent operators for higher-order elements to two factors.
@@
 Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach with relatively high computational intensity.
-}
 
@@
 Next, we evaluate the performance of the proposed geometric multigrid preconditioner.
-{\color{red}
-Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end,}
-we consider {\color{red}the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).}
+Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end,
+we consider the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).
 Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
-{\color{red} noticeable} increase in the average number of CG iterations for different polynomial degrees
-{\color{red} for the GMG preconditioner.
+ noticeable increase in the average number of CG iterations for different polynomial degrees
+ for the GMG preconditioner.
 Although not reported in this section, we identified that
@@
 In particular, we can improve the number of iterations by choosing a more accurate coarse level solver,
-however this leads to an overall more expensive linear solver.}
-\changeJP{However, the settings used in this study are optimized for the considered problem in terms of the wall-time spent in the linear solver.}
+however this leads to an overall more expensive linear solver.
+However, the settings used in this study are optimized for the considered problem in terms of the wall-time spent in the linear solver.
 The black-box AMG preconditioner
-{\color{red} demonstrates the same \changeJP{trend in terms of} the number of CG iterations, but}
+demonstrates the same trend in terms of the number of CG iterations, but
 requires many more iterations for convergence.
-\changeJP{
 Consequently, we can conclude that for the here considered finite strain elasticity problem (i) the selected smoother is suitable, and that (ii) although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid.
-}
-
-%{\color{red}
-%Additionally, we compare solution times of the preconditioned iterative solvers.
-%Our goal is to \changeJP{determine} whether the matrix-free approach with a GMG preconditioner can be faster than the matrix-based approach with an AMG preconditioner.
-%}
-\changeJP{%
+
 Additionally, we compare solution times of the preconditioned iterative solvers.
-Our goal is to \changeJP{determine} whether the matrix-free approach with a GMG preconditioner can be faster than the matrix-based approach with a well-established implementation of an AMG preconditioner.
+Our goal is to determine whether the matrix-free approach with a GMG preconditioner can be faster than the matrix-based approach with a well-established implementation of an AMG preconditioner.
 We note specifically that the comparison is restricted to the choice of level operators used with the pre-existing GMG framework, and not the framework itself.
-}
-%\changeJP{We note that although the well-established implementation of the AMG preconditioner \cite{Gee2006a} is the best available choice for the matrix-based implementation of the boundary value problem, the current implementation of the matrix-free GMG preconditioner is not yet optimized and can be improved further.}
-Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for {\color{red}all} polynomial degrees.
-{\color{red}Most importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
-\changeJP{In comparison}, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
+Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for all polynomial degrees.
+Most importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
+In comparison, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
 Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
-Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for \changeJP{engineering problems
-with compressible hyperelastic} finite strain material models.}
-
-\changeJP{Extension of this study from the node level to the cluster level is beyond the scope of this work.
+Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for engineering problems
+with compressible hyperelastic finite strain material models.
+
+Extension of this study from the node level to the cluster level is beyond the scope of this work.
 However, as a step in this direction we have performed a preliminary study of the weak scalability of the implementation.
@@
 also compare \cite[Figure 7]{Krank2017} for scaling results for up to 147,456 CPU cores and 34.4 billion degrees of freedom for an incompressible flow problem.
-}
 
@@
 
-In this contribution, we proposed and numerically investigated {\color{red}three different} matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
-{\color{red}
-To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to partial differential equations arising in this case.}
-{\color{red}Among the three \changeJP{examined} algorithms,}
-the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for polynomial degrees higher than one in 3D and polynomial degrees higher than two {\color{red} in 2D}.
-{\color{red}
-Given that we did not utilize the specific form of the material part of the stiffness tensor in this case, we conclude}
+In this contribution, we proposed and numerically investigated three different matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
+To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to partial differential equations arising in this case.
+Among the three examined algorithms,
+the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for polynomial degrees higher than one in 3D and polynomial degrees higher than two in 2D.
+Given that we did not utilize the specific form of the material part of the stiffness tensor in this case, we conclude
 that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
 
-{\color{red}
 The roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
@@
 as well as single contraction of two second-order tensors is performed.
-% Further studies are required to identify means to improve the computational intensity and performance of the matrix-free framework provided by \texttt{deal.II}.
-}
-
-{\color{red}
-We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner}
+
+We also propose a method by which to construct level tangent operators and employ them to define a geometric multigrid preconditioner
 with standard geometric transfer operations between each level.
@@
 The numerical studies indicate that the proposed preconditioner
-{\color{red}
 leads to much fewer iterations of the iterative solver as compared to the algebraic multigrid preconditioner.
-The multigrid matrix-free preconditioner also}
-leads to a solution approach that is faster than the matrix-based AMG {\color{red}for all polynomial degrees studied here.
+The multigrid matrix-free preconditioner also
+leads to a solution approach that is faster than the matrix-based AMG for all polynomial degrees studied here.
 Most importantly, the wall time per degree of freedom for solving the problem is close to being constant.
@@
 Our future work will be focused on adopting the proposed matrix-free multigrid solution approach to $\operatorname{FE}^2$ homogenization
-as well studying its behavior on the cluster level
-% as opposed to node level considered here.
-}
+as well studying its behavior on the cluster level.
 
@@
 
-D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to {\color{red}Martin Kronbichler (TU Munich),} Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for \changeJP{fruitful discussions} on matrix-free operator evaluation approaches.
+D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to Martin Kronbichler (TU Munich), Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for fruitful discussions on matrix-free operator evaluation approaches.
 
@@
 
-% J-P.~Pelteret was supported by the European Research Council (ERC) through
-% the Advanced Grant 289049 MOCOPOLY
-
 P.~Steinmann acknowledges the support of the Cluster of Excellence Engineering of Advanced Materials (EAM) which made this collaboration possible, as well as funding by the EPSRC Strategic Support Package ``Engineering of Active Materials by Multiscale/Multiphysics Computational Mechanics''.
-
-% \todo{Add back of the envelope calculations by Veselin Dobrev? See {https://github.com/CEED/Forum/issues/1}.}
 

### 2019-04-30 doc/large_strain.tex
put tables inside the main tex so that arXiv is not tyring to build them separately

@@
 
-\input{parallelization_2d.tex}
-\input{parallelization_3d.tex}
+\begin{table}
+  \centering
+  \resizebox{\textwidth}{!}{
+  \begin{tabular}{|c|cc|ccc|ccc|ccc|}
+  \hline
+                & \multicolumn{2}{c|}{serial} & \multicolumn{3}{c|}{MPI} & \multicolumn{3}{c|}{SIMD} & \multicolumn{3}{c|}{MPI+SIMD}  \\
+  \hline
+  p             & time  & GFlop/s              & time & GFlop/s & speedup & time & GFlop/s & speedup & time & GFlop/s & speedup \\
+  \hline
+  2& \pgfmathprintnumber{0.71685} & \pgfmathprintnumber{0.8747712} & \pgfmathprintnumber{0.07427} & \pgfmathprintnumber{8.57318} & \pgfmathprintnumber{9.65194560388} & \pgfmathprintnumber{0.3939} & \pgfmathprintnumber{1.6528706} & \pgfmathprintnumber{1.81987814166} & \pgfmathprintnumber{0.04142} & \pgfmathprintnumber{16.0073657} & \pgfmathprintnumber{17.306856591} \\
+  4& \pgfmathprintnumber{0.47603} & \pgfmathprintnumber{1.3088965} & \pgfmathprintnumber{0.0511} & \pgfmathprintnumber{12.3860771} & \pgfmathprintnumber{9.3156555773} & \pgfmathprintnumber{0.25747} & \pgfmathprintnumber{2.3690572} & \pgfmathprintnumber{1.84887559716} & \pgfmathprintnumber{0.02786} & \pgfmathprintnumber{22.2420699} & \pgfmathprintnumber{17.0865039483} \\
+  6& \pgfmathprintnumber{0.24683} & \pgfmathprintnumber{1.5497601} & \pgfmathprintnumber{0.02636} & \pgfmathprintnumber{14.8125524} & \pgfmathprintnumber{9.36380880121} & \pgfmathprintnumber{0.13089} & \pgfmathprintnumber{2.8544567} & \pgfmathprintnumber{1.88578195431} & \pgfmathprintnumber{0.01487} & \pgfmathprintnumber{26.0509717} & \pgfmathprintnumber{16.5991930061} \\
+  \hline
+  \end{tabular}
+  }
+  \caption{Wall-clock time in seconds and performance in GFlops of Algorithm \ref{alg:mf_tensor4} in 2D for various combinations of polynomial degrees,
+  vectorization and parallelization.}
+  \label{tab:numbers_2d}
+\end{table}
+
+\begin{table}
+  \centering
+  \resizebox{\textwidth}{!}{
+  \begin{tabular}{|c|cc|ccc|ccc|ccc|}
+  \hline
+                & \multicolumn{2}{c|}{serial} & \multicolumn{3}{c|}{MPI} & \multicolumn{3}{c|}{SIMD} & \multicolumn{3}{c|}{MPI+SIMD}  \\
+  \hline
+  p             & time  & GFlop/s              & time & GFlop/s & speedup & time & GFlop/s & speedup & time & GFlop/s & speedup \\
+  \hline
+  2& \pgfmathprintnumber{1.82927} & \pgfmathprintnumber{1.4408182} & \pgfmathprintnumber{0.19184} & \pgfmathprintnumber{13.75528} & \pgfmathprintnumber{9.5353940784} & \pgfmathprintnumber{1.24738} & \pgfmathprintnumber{2.2390063} & \pgfmathprintnumber{1.46648976254} & \pgfmathprintnumber{0.13397} & \pgfmathprintnumber{21.0187594} & \pgfmathprintnumber{13.6543255953} \\
+  4& \pgfmathprintnumber{1.14988} & \pgfmathprintnumber{1.7483867} & \pgfmathprintnumber{0.1211} & \pgfmathprintnumber{16.6162812} & \pgfmathprintnumber{9.49529314616} & \pgfmathprintnumber{0.74974} & \pgfmathprintnumber{2.7527409} & \pgfmathprintnumber{1.53370501774} & \pgfmathprintnumber{0.08166} & \pgfmathprintnumber{25.4865287} & \pgfmathprintnumber{14.0813127602} \\
+  \hline
+  \end{tabular}
+  }
+  \caption{Wall-clock time in seconds and performance in GFlops of Algorithm \ref{alg:mf_tensor4} in 3D for various combinations of polynomial degrees,
+  vectorization and parallelization.}
+  \label{tab:numbers_3d}
+\end{table}
 
@@
 
-\input{weak_scaling_3d.tex}
+\begin{table}
+  \centering
+  \begin{tabular}{|r|r|c|r|c|}
+  \hline
+  cores  & $N_{DoF}$ & $N_{gref}$ & vmult [s] & CG [s / iteration] \\
+  \hline
+  20 & 4,442,880 & 3 & \pgfmathprintnumber{0.0675} & \pgfmathprintnumber{0.529553679131} \\
+  160 & 35,071,488 & 4 & \pgfmathprintnumber{0.0705} & \pgfmathprintnumber{0.562637362637} \\
+  1280 & 278,694,912 & 5 & \pgfmathprintnumber{0.083} & \pgfmathprintnumber{0.688792165397} \\
+  \hline
+  \end{tabular}
+  \caption{Weak scaling of Algorithm \ref{alg:mf_tensor4} in 3D for quadratic polynomial basis.}
+  \label{tab:weak_3d}
+\end{table}
 

### 2019-06-09 doc/large_strain.tex
remove unused figure
remove unused (?) package
cleanup commands
combine Framework and GMG

@@
-\RequirePackage[l2tabu, orthodox]{nag}
+%\RequirePackage[l2tabu, orthodox]{nag}
 
@@
 
-\def\gz  #1{           \mbox{$\boldsymbol{#1}$}}
-\def\Grad #1 {{\rm Grad} #1}
-\def\grad {\mbox{grad}}
-\def\Div {\mbox{Div\,}}
-\def\div {\mbox{div\,}}
-\def\d {{\rm d}}
-\def\D {{\rm D}}
-\def\mcl  #1{               {\mathcal #1}}
+\newcommand*{\gz}[1]{\boldsymbol{#1}}
+\newcommand*{\Grad}{\mathrm{Grad}}
+\newcommand*{\grad}{\mathrm{grad}}
+\renewcommand*{\d}{\mathrm{d}}
+\newcommand*{\D}{\mathrm{D}}
+\newcommand*{\mcl}[1]{\mathcal{#1}}
 \DeclareMathOperator{\trace}{tr}
-\newcommand{\diff}{\mathop{}\!\mathrm{d}}
 
@@
 
-The paper is organized as follows: In Section \ref{sec:theory} we briefly introduce the partial differential equations used in  finite-strain solid mechanics.
-Sections \ref{sec:fe} and \ref{sec: Description of numerical framework} cover the finite element discretization and provide details of the numerical framework with which this study is implemented.
-In \ref{sec:mf} we describe the different matrix-free operator implementations that are evaluated in this study, and in section \ref{sec:gmg} we propose a geometric multigrid preconditioner that is suitable for the matrix-free implementation of finite-strain hyperelasticity.
+The paper is organized as follows:
+In Section \ref{sec:theory} we briefly introduce the partial differential equations used in  finite-strain solid mechanics.
+Sections \ref{sec:fe} cover the finite element discretization.
+In \ref{sec:mf} we describe the different matrix-free operator implementations that are evaluated in this study.
+Section \ref{sec:framework} provides details of the numerical framework with which this study is implemented,
+where we also propose a method by which to construct level tangent operators for
+a geometric multigrid preconditioner that is suitable for the matrix-free implementation of finite-strain hyperelasticity.
+%
 The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}.
@@
 
-\section{Description of the utilized numerical framework
-\label{sec: Description of numerical framework}
-}
-
-The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
-This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aims to support exascale computing.
-In this work we employ \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
-%
-\begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.8\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{fine_level_2d.png}
-    % \caption{2D multigrid mesh with 2 global refinements}
-  \end{subfigure}
-  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
-  \label{fig:miehe_fine_level}
-\end{figure}
-%
-%
-Figure \ref{fig:miehe_fine_level} illustrates a distributed MPI decomposition on the fine mesh level for the 2D problem that will be introduced later in Section \ref{sec:example}.
-
-In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
-For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
-
-The discrete tangent operator described by \eqref{eq:algebraic_tangent} can be implemented in one of two ways, namely through a classical matrix-based approach or, alternatively, a matrix-free approach.
-For the matrix-based approach adopted in this study, the Trilinos \cite{Heroux2005} library was leveraged;
-the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
-This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is possible in  \texttt{deal.II}.
-
-The second implementation, which is discussed in detail in the following Sections, utilized a matrix-free approach in conjunction with a geometric-multigrid preconditioner.
-As will be detailed later, the pre-existing matrix-free framework that was used offers further opportunity for low-level parallelism that we exploit.
-
 \section{Matrix-free operator evaluation}
@@
 
-\section{Geometric multigrid preconditioning}
-\label{sec:gmg}
-
-Efficient and scalable matrix-vector products are not enough to obtain a method suitable for large computations.
-In particular, linear scaling with respect to the number $N$ of unknown DoFs is desirable.
-The efficiency of standard iterative solvers and preconditioners deteriorates for large systems of linear equations due to increasing iteration counts under mesh refinement.
-In contrast to other preconditioners, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes.
+\section{Description of the utilized numerical framework
+\label{sec:framework}
+}
+
+The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
+This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aims to support exascale computing.
+In this work we employ \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
+%
+\begin{figure}[!ht]
+  \centering
+  \begin{subfigure}[b]{0.8\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{fine_level_2d.png}
+  \end{subfigure}
+  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
+  \label{fig:miehe_fine_level}
+\end{figure}
+%
+%
+Figure \ref{fig:miehe_fine_level} illustrates a distributed MPI decomposition on the fine mesh level for the 2D problem that will be introduced later in Section \ref{sec:example}.
+
+In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
+For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
+
+The discrete tangent operator described by \eqref{eq:algebraic_tangent} can be implemented in one of two ways, namely through a classical matrix-based approach or, alternatively, a matrix-free approach.
+For the matrix-based approach adopted in this study, the Trilinos \cite{Heroux2005} library was leveraged;
+the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
+This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is possible in \texttt{deal.II}.
+
+The second implementation utilized a matrix-free approach in conjunction with a geometric-multigrid (GMG) preconditioner.
+GMG preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes.
 For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
-Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.
-
-For this work, we adopt the matrix-free level operators within the GMG.
+%
+%
+In this work, we adopt the matrix-free level operators within the GMG.
 In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial.
-One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
-  We note that this approach is limited to the case where for each quadrature point within an element the fourth-order tangent stiffness tensor is the same.
-}.
+One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}.
 In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
@@
 We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
-Essentially, level operators correspond to the linearization around the smoothed representation of displacement field\footnote{
+Essentially, level operators correspond to the linearization around the smoothed representation of displacement field.
 Note that this differs from the classical approach where the (tangent) operator $\gz A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $\gz A^{l}$ on level $l$ via $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, where $\gz I^l_{l+1}$ and $\gz I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
-}.
-
-To summarize, in this section we propose a method by which to construct level tangent operators for finite-strain elasticity to be employed within the GMG preconditioner.
-The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
-The efficiency of the preconditioner with the considered level operators will be assessed by numerical examples in the next section.
+
+The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library, implementational details are discussed in \cite{Clevenger2018}.
+%
+The efficiency of the GMG preconditioner with the considered level operators will be assessed by numerical examples in the next section.
 We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
+%
 Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
-
-For examples of other approaches that combine homogenization and multigrid methods, see \cite{Bayreuther2003,Fish1995} for applications using small strain elastic materials and \cite{Kaczmarczyk2010} with applications in fracture.
+%
+% For examples of other approaches that combine homogenization and multigrid methods, see \cite{Bayreuther2003,Fish1995} for applications using small strain elastic materials and \cite{Kaczmarczyk2010} with applications in fracture.
 
@@
     \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
-    % \caption{2D multigrid mesh with 2 global refinements}
   \end{subfigure}
@@
 \bibliography{bibliography}
+%\input{large_strain.bbl}
 

### 2019-06-09 doc/large_strain.tex
drop GMG figure

@@
 %
-\begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.8\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
-  \end{subfigure}
-  \caption{2D multigrid mesh after two global refinements distributed into four MPI processes. The color indicates the process owning the element.}%
-  \label{fig:miehe_gmg}
-\end{figure}
-%
-%
-The same distributed approach to mesh decomposition (as most evident on the finest level) is adopted on the multigrid levels.
-Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements for the mesh introduced in figure \ref{fig:miehe_fine_level}.
+% \begin{figure}[!ht]
+%   \centering
+%   \begin{subfigure}[b]{0.8\textwidth}
+%     \centering
+%     \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
+%   \end{subfigure}
+%   \caption{2D multigrid mesh after two global refinements distributed into four MPI processes. The color indicates the process owning the element.}%
+%   \label{fig:miehe_gmg}
+% \end{figure}
+% %
+% %
+% The same distributed approach to mesh decomposition (as most evident on the finest level) is adopted on the multigrid levels.
+% Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements for the mesh introduced in figure \ref{fig:miehe_fine_level}.
 %

### 2019-06-09 doc/large_strain.tex
make GMG part a little bit more concise

@@
 One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}.
-In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
-However, for a patch of elements consisting of the same material, the transfer operators reduce to the standard geometric transfer.
-Based on this observation we adopt the following approach: It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material.
+Such operators are, however, computationally very demanding.
+In this work we propose a different approach:
+% In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
+% However, for a patch of elements consisting of the same material, the transfer operators reduce to the standard geometric transfer.
+% Based on this observation we adopt the following approach:
+It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material.
 In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.

### 2019-06-09 doc/large_strain.tex
change how we mention Clevenger paper
cite Martin' work at the beginning of MF section

@@
 To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
-We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018}.
+We are only aware of a single paper that deals with small strain linear elasticity \cite{Clevenger2018}.
 One possible reason is that the tangent operators\footnote{Here and
@@
 On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
-For this reason, recent implementations often focus on so-called matrix-free approaches where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
+For this reason, recent implementations often focus on so-called matrix-free approaches \cite{kronbichler12, kronbichler2017fast} where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
 The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.

### 2019-06-09 doc/large_strain.tex
group bechmark tables (2D and 3D) to save space

@@
 %
-\begin{table}
+\begin{table}[!htb]
+  \begin{subtable}{.5\linewidth}
   \centering
@@
   \end{tabular}
-  \caption{Parameters for the 2D benchmark: $p$ is the polynomial degree,
-  $q$ is the number of quadrature points in 1D, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
-  }
+  \caption{2D}
   \label{tab:input_parameters_2d}
-\end{table}
-
-\begin{table}
+  \end{subtable}
+  \begin{subtable}{.5\linewidth}
   \centering
@@
   \end{tabular}
-  \caption{Parameters for the 3D benchmark: $p$ is the polynomial degree,
+  \caption{3D}
+  \label{tab:input_parameters_3d}
+  \end{subtable}
+  \caption{Parameters for the benchmark: $p$ is the polynomial degree,
   $q$ is the number of quadrature points in 1D, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
   }
-  \label{tab:input_parameters_3d}
 \end{table}

### 2019-06-09 doc/large_strain.tex
fourth degree -> fith degree (https://github.com/dealii/dealii/pull/7708)

@@
 The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
-The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
+The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fifth degree, provided
 by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.

### 2019-06-10 doc/large_strain.tex
remove double Bibliography
improve Intro

@@
 The paper is organized as follows:
-In Section \ref{sec:theory} we briefly introduce the partial differential equations used in  finite-strain solid mechanics.
-Sections \ref{sec:fe} cover the finite element discretization.
-In \ref{sec:mf} we describe the different matrix-free operator implementations that are evaluated in this study.
+In Section \ref{sec:theory} we briefly introduce the partial differential equations used in finite-strain solid mechanics,
+while section \ref{sec:fe} covers the finite element discretization.
+In Section \ref{sec:mf} we describe the different matrix-free operator implementations that are evaluated in this study.
 Section \ref{sec:framework} provides details of the numerical framework with which this study is implemented,
@@
 \else
-\section*{Bibliography}
+%\section*{Bibliography}
 \bibliographystyle{elsarticle-num}

### 2019-07-02 doc/large_strain.tex

@@
 
-The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
+The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling}, were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
 This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aims to support exascale computing.
-In this work we employ \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
+In this work we employ \texttt{deal.II}'s interface to the \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
 %
@@
 
-In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
+In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver, both of which are the focal points of investigation in this manuscript, has been achieved using MPI;
+various pre- and post-processing steps also leveraged parallelization.
 For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
@@
 the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
-This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is possible in \texttt{deal.II}.
+This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is available in \texttt{deal.II}.
 
@@
 
-The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library, implementational details are discussed in \cite{Clevenger2018}.
+The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library; implementational details are discussed in \cite{Clevenger2018}.
 %

### 2019-07-24 doc/large_strain.tex

@@
 \end{align}
-where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
+where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form at one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
 $w_q$ is the quadrature weight.
@@
 in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
-For the total matrix-vector product we get $\mathcal{O}(n^{2d})$,
+For the total matrix-vector product we get $\mathcal{O}(n^{2d})$, \todo{Should we include the number of elements in the total mat-vec?}
 and this is the same as for a regular matrix-vector product where the entries are already precomputed.
@@
 
-For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.
+For the performance of the matrix-free operator evaluation, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.
 %
@@
           evaluate $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ \tcp*{2nd order}
-          evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
+          evaluate $J = \textrm{det}(\gz F)$ \tcp*{scalar}
           evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
@@
           queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
-          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
+          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \textrm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
           queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
@@
           queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
-          evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
+          evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \textrm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
           queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
@@
 For the matrix-based approach adopted in this study, the Trilinos \cite{Heroux2005} library was leveraged;
-the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
+the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} sparse data structure from the Epetra package \cite{Heroux2005b} and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
 This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is available in \texttt{deal.II}.
@@
 we limit ourselves to comparisons made on a node level in this study.
-Examinations performed in this manner should already indicate the differences between the matrix-free and matrix-based approaches, and give some preliminary understanding of the scalability potential of the numerical implementation without confounding factors related to data exchange through interconnects via MPI.
+Examinations performed in this manner are most representative for the differences between the matrix-free and matrix-based approaches because problems of sufficient size are dominated by the local node-level performance with multigrid solvers.
+Multigrid solvers are well-known to be scalable given sufficiently small coarse-grid problems even to the largest supercomputers \cite{gholami2016,ibeid2018} as the communication is mostly between nearest neighbors.
 The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
@@
 
-We conduct two performance studies using the described discretization, the first being for only matrix-vector multiplication, and the second for the preconditioned iterative linear solver.
+We conduct two performance studies using the described discretization, the first being for matrix-vector multiplication only, and the second for the preconditioned iterative linear solver.
 The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library version 12.12.1.
@@
 The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
-is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
+is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.\todo{Does this work well (verified in code)? I would use $0.1\lambda_{\rm{max}}$, i.e., smoothing range 12.}
 On the coarsest level, we use the Chebyshev preconditioner as a solver \cite{Varga2009} to reduce the residual by three orders of magnitude.
@@
   \end{subfigure}
-  \caption{Emmy cluster, RRZE, matrix-vector multiplication.}%
+  \caption{Emmy cluster, 20 cores, RRZE, matrix-vector multiplication.}%
   \label{fig:benchmark_miehe_Emmy}
@@
 
+\todo{Is the IWR result also run with 20 MPI ranks? How are MPI ranks pinned to the 80 physical cores?}
+
 Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
-We are interested in the wall-clock time (average over 10 runs per each linear solver step) per DoF as a first metric.
+We are interested in the wall-clock time (average over 10 runs for each linear solver step) per DoF as a first metric.
 As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
@@
 The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
-Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces \cite{Arndt2017a} to capture the geometry as well as
+Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces \cite{dealII90} to capture the geometry as well as
 the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
@@
 
-When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}) caching strategy,
+When comparing the data for the ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}) caching strategy,
 we observe a marginal increase in computational intensity and almost unchanged performance.
@@
 (ii) `sum factorization' denotes application of sum factorization techniques (lines 4, 5 and 12);
-(iii) `quadrature loop' denotes operations performed on each quadrature (lines \mbox{6 -- 11}).
+(iii) `quadrature loop' denotes operations performed at each quadrature (lines \mbox{6 -- 11}).
 Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one

### 2019-08-03 doc/large_strain.tex
Reword introduction. Add table about algorithms. Add CSL figures

@@
   \begin{abstract}
-    The performance of finite element solvers on modern computer architectures is typically memory bound for sufficiently large problems.
-    The main cause for this is that loading matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
-    In order to improve the performance of iterative solvers within the high-performance computing context, so-called matrix-free methods are
-    widely adopted in the fluid mechanics community, where matrix-vector products are computed on-the-fly.
-
-    To date, there have been few (if any) assessments into the applicability of the matrix-free approach to problems in solid mechanics.
-    In this work, we perform an initial investigation on the application of the matrix-free approach to problems in quasi-static finite-strain hyperelasticity to determine whether it is viable for further extension.
-    Specifically, we study different numerical implementations of the finite element tangent operator, and determine whether generalized methods of incorporating complex constitutive behavior might be feasible.
+    This work investigates matrix-free algorithms for problems in quasi-static finite-strain hyperelasticity.
+Matrix-free algorithms have emerged as an attractive alternative to iterative solvers with sparse matrices in the fluid dynamics and wave propagation context because they significantly reduce the memory traffic, the limiting factor in classical finite element solvers.
+    Specifically, we study different matrix-free realizations of the finite element tangent operator, and determine whether generalized methods of incorporating complex constitutive behavior might be feasible.
     In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
@@
     The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner on a single node for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
-    We conclude that the application of matrix-free methods to finite-strain solid mechanics is promising, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.
+    We conclude that matrix-free methods for finite-strain solid mechanics are very promising, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.
   \end{abstract}
@@
 The performance of finite element solvers on modern computer architectures is typically memory bound for sufficiently large problems.
-The main cause for this is that loading pre-computed matrix elements from RAM into CPU cache is significantly slower than performing the arithmetic operations when solving the problem.
+The main cause for this is that loading pre-computed matrix elements from RAM memory into the CPU cores is significantly slower than actually performing the arithmetic operations.
 In order to improve the performance of iterative solvers, so-called matrix-free methods,
@@
 
-To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
-We are only aware of a single paper that deals with small strain linear elasticity \cite{Clevenger2018}.
-One possible reason is that the tangent operators\footnote{Here and
+Within the solid mechanics community, however, matrix-free methods are currently not widely adopted.
+One reason is that the tangent operators\footnote{Here and
 below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
 Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
-} (stemming from linearization of the nonlinear balance equations) are
-more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
+} derived from linearization of the nonlinear balance equations are
+more elaborate as compared to the common operators in linear and non-linear fluid mechanics.
+For matrix-free schemes using fast integration via sum factorization, the most common matrix-free variant for polynomial degrees $p\geq 2$, the main challenge is to find a representation of the linearization point, either in referential or spatial configurations, at the quadrature points that is cheap enough to be evaluated on the fly.
 Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
@@
 whereas higher order elements are rarely employed.
-This can potentially reduce the advantage of matrix-free methods using sum-factorization techniques, that are generally more competitive for higher order elements \cite{kronbichler12,kronbichler2017fast,muthing2017high}.
-However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}.
+This can potentially reduce the advantage of matrix-free methods using sum-factorization techniques, that are generally more competitive for higher order elements \cite{kronbichler12,kronbichler2017fast,muthing2017high,Fischer2019}.
+However, we note that, even with linear FEM, large scale computations with as many as $10^{12}$ unknowns are possible with a matrix-free approach but infeasible with sparse matrices as there is simply not enough memory to store the sparse tangent matrix even on large supercomputers \cite{Gmeiner2016}.
 Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.
 
-In this work, we perform a preliminary investigation of different implementations of the finite-strain hyperelastic tangent operator with respect to their computational cost
-and performance on a single node of a high performance computing cluster.
+In this work, we investigate different matrix-free representations of the
+finite-strain hyperelastic tangent operator with respect to the computational cost
+and the performance on a single node of a high performance computing cluster.
+Particular emphasis is on the tradeoff between computations at quadrature points versus
+pre-computation with a higher memory access.
 The \texttt{deal.II} \cite{dealII90} finite element library, which provides MPI parallelization together with a generalized matrix-free framework that incorporates SIMD vectorization and an interface to high-performance, MPI distributed matrix-based libraries, is used for this study.
@@
 where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$
-is involved due to the symmetry of $\gz \tau$.
+is involved due to the symmetry of $\gz \tau$. Here, `$\grad$' denotes the gradient with respect to the spatial configuration, as opposed to the gradient in the referential setting `$\Grad$'.
 Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
-that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.
+that is, $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.
 
@@
   \ForEach{ element $K \in \Omega^h$ }{
-        extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ \;
+        gather local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ \;
         evaluate the following tensors at each quadrature point using sum factorization :\\
         \enskip $\Grad \, \overline{\gz u^h}_K$ \tcp*{2nd order}
-        \enskip $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
         \enskip $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
@@
           evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
-          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
+          evaluate $\gz g_s = \text{symmetrize} (\gz g)$ \tcp*{2nd order symmetric}
           evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \textrm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
-          queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
+          queue $\boldsymbol{\mathcal{C}} \gz g_s + \boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : (\boldsymbol{\mathcal{C}}\gz g_s  + \boldsymbol{\mathcal{G}} \gz g)$ \;
         }
       evaluate queued contractions using sum factorization \;
-      distribute results to the destination vector
+      scatter results to the destination vector
   }
@@
   \ForEach{ element $K \in \Omega^h$ }{
-        extract local vector values on this element $\gz x_K$ \;
+        gather local vector values on this element $\gz x_K$ \;
         evaluate the following tensors at each quadrature point using sum factorization :\\
-        \enskip$\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
         \enskip$\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
@@
           evaluate $\boldsymbol{\mathcal{G}}\gz g := \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
-          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
+          evaluate $\gz g_s = \text{symmetrize} (\gz g)$ \tcp*{2nd order symmetric}
           evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \textrm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
-          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
+          queue $\boldsymbol{\mathcal{C}} \gz g_s + \boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : (\boldsymbol{\mathcal{C}}\gz g_s  + \boldsymbol{\mathcal{G}} \gz g)$ \;
         }
         evaluate queued contractions using sum factorization \;
-        distribute results to the destination vector
+        scatter results to the destination vector
   }
@@
           evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
-          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
+          evaluate $\gz g_s = \text{symmetrize} (\gz g)$ \tcp*{2nd order symmetric}
           evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
-          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
+          queue $\boldsymbol{\mathcal{C}} \gz g_s + \boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : (\boldsymbol{\mathcal{C}}\gz g_s  + \boldsymbol{\mathcal{G}} \gz g)$ \;
         }
         evaluate queued contractions using sum factorization \;
-        distribute results to the destination vector
+        scatter results to the destination vector
   }
@@
 The core of the sum factorization algorithms is provided by the
-\textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
+\textit{FEEvaluation} class, that supports all the necessary contractions such as $\overline{\grad} \gz N_i : (\boldsymbol{\mathcal{C}}\gz g_s  + \boldsymbol{\mathcal{G}} \gz g)$ to implement the three algorithms proposed above.
 Thanks to the object-oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
@@
 For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
+
+\begin{table}
+\centering
+\begin{tabular}{|l|ccc|}
+\hline
+ & Vector access & physics terms & metric terms grad/Grad \\
+ & double / node & double / q-point & double / q-point \\
+ \hline
+Algorithm 1 & $2d$ read, $d$ write & 1 & up to $d^2 + d^2 + 1$\\
+Algorithm 2 & $d$ read, $d$ write & 2 + $\frac{d(d+1)}{2}$ & up to $d^2+1$\\
+Algorithm 3 & $d$ read, $d$ write & $\frac{d(d+1)}{2}+\left(\frac{d(d+1)}{2}\right)^2$ & up to $d^2+1$\\
+\hline
+\end{tabular}
+\caption{Details of memory access for matrix-free evaluation with Algorithms~\ref{alg:mf_scalar}--\ref{alg:mf_tensor4}.}
+\label{tab:memory}
+\end{table}
+
+Table~\ref{tab:memory} summarizes the three algorithms in terms of their implied memory access for one operator evaluation.
+The vector access in Algorithm~\ref{alg:mf_scalar} is higher than in the other variants because both the source vector $\gz x$ and the linearization point from the current FE solution $\overline{\gz u^h}$ need to be accessed, each with the usual indirect addressing pertaining to continuous finite element fields \cite{kronbichler12}.
+The cached quantities listed in the algorithms, labeled ``physics terms'' in the table, are increasing in size for the algorithms, and are separate for each point.
+Finally, the evaluation of the gradients in the referential and spatial coordinates also involve a data access via the so-called metric terms, the Jacobian from the unit element to the referential or spatial configurations, respectively.
+Algorithm~\ref{alg:mf_scalar} involves both the Jacobian in the referential and the spatial frame, whereas the other two algorithms only involve the spatial (Eulerian) terms.
+Finally, the implementation framework also holds the determinant of the Jacobian for each point.
+Following \cite{kronbichler12}, some compression may be applied to the metric terms: On affine geometries, the Jacobian is the same at all quadrature points of a cell, and needs only be stored once.
+Therefore, the table lists those quantities as ``up to'' $d^2+1$ terms, which depends on the mesh and is a run-time decision in the code.
 
@@
 The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
-is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.\todo{Does this work well (verified in code)? I would use $0.1\lambda_{\rm{max}}$, i.e., smoothing range 12.}
+is set to target the range $[0.06\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
 On the coarsest level, we use the Chebyshev preconditioner as a solver \cite{Varga2009} to reduce the residual by three orders of magnitude.
@@
 
-Computations were performed on a single node of two clusters:
-A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM.
+Computations were performed on two systems: A dual-socket Intel Xeon Gold 6230 (``Cascade Lake'') with 20 cores per socket and 96 GB of DDR4-2933 memory (nominal bandwidth per socket: 141 GB/s).
+Turbo mode is enabled with a power limit of 135 watt per socket, which results in the processor running at 2.8 GHz for scalar code and 2.0 GHz for AVX-512-heavy code in all experiments.
+The GNU compiler version 9.1.0 with flags ``-O3 -march=skylake-avx512' and OpenMPI version 4.0.1 were used, with the code run in pure MPI mode.
+%A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM.
 Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
-The ``IWR'' results were obtained on a single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip.
+The ``IWR'' results were obtained on a single machine with eight Intel Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip.
 For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
-Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. one fully occupied node.
+Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. one socket for Cascade Lake and two sockets for ``IWR''.
 
@@
       \centering
-      \includegraphics[width=\textwidth]{Emmy_RRZE_timing2d.eps}
-      \caption{matrix-vector product (2D)}
-      \label{fig:benchmark_miehe_Emmy_vmult2}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.49\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{Emmy_RRZE_timing3d.eps}
-    \caption{matrix-vector product (3D)}
-    \label{fig:benchmark_miehe_Emmy_vmult3}
+      \includegraphics[width=\textwidth]{CSL_Munich_throughput2d.eps}
+      \caption{throughput matrix-vector product (2D)}
+      \label{fig:benchmark_miehe_CSL_vmult2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{CSL_Munich_throughput3d.eps}
+    \caption{throughput matrix-vector product (3D)}
+    \label{fig:benchmark_miehe_CSL_vmult3}
   \end{subfigure}
@@
       \centering
-      \includegraphics[width=\textwidth]{Emmy_RRZE_memory2d.eps}
+      \includegraphics[width=\textwidth]{CSL_Munich_timing2d.eps}
+      \caption{time matrix-vector product (2D)}
+      \label{fig:benchmark_miehe_CSL_vmult2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{CSL_Munich_timing3d.eps}
+    \caption{time matrix-vector product (3D)}
+    \label{fig:benchmark_miehe_CSL_vmult3}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[b]{0.49\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{CSL_Munich_memory2d.eps}
       \caption{memory consumption (2D)}
-      \label{fig:benchmark_miehe_Emmy_memory2}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.49\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{Emmy_RRZE_memory3d.eps}
+      \label{fig:benchmark_miehe_CSL_memory2}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{CSL_Munich_memory3d.eps}
     \caption{memory consumption (3D)}
-    \label{fig:benchmark_miehe_Emmy_memory3}
-  \end{subfigure}
-  \caption{Emmy cluster, 20 cores, RRZE, matrix-vector multiplication.}%
-  \label{fig:benchmark_miehe_Emmy}
+    \label{fig:benchmark_miehe_CSL_memory3}
+  \end{subfigure}
+  \caption{Cascade Lake, 1 socket with 20 cores, matrix-vector multiplication.}%
+  \label{fig:benchmark_miehe_CSL}
 \end{figure}
@@
       \centering
-      \includegraphics[width=\textwidth]{IWR_timing2d.eps}
+      \includegraphics[width=\textwidth]{IWR_newest_patched_timing2d.eps}
       \caption{matrix-vector product (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_timing3d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_timing3d.eps}
     \caption{matrix-vector product (3D)}
@@
       \centering
-      \includegraphics[width=\textwidth]{IWR_memory2d.eps}
+      \includegraphics[width=\textwidth]{IWR_newest_patched_memory2d.eps}
       \caption{memory consumption (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_memory3d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_memory3d.eps}
     \caption{memory consumption (3D)}
@@
 
-\todo{Is the IWR result also run with 20 MPI ranks? How are MPI ranks pinned to the 80 physical cores?}
-
-Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
+Figures \ref{fig:benchmark_miehe_CSL} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
 We are interested in the wall-clock time (average over 10 runs for each linear solver step) per DoF as a first metric.
@@
     \centering
-    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline_2d.pdf}
+    \includegraphics[width=\textwidth]{LIKWID_CSL_Munich_roofline_2d.pdf}
     \caption{2D}
@@
     \centering
-    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline_3d.pdf}
+    \includegraphics[width=\textwidth]{LIKWID_CSL_Munich_roofline_3d.pdf}
     \caption{3D}
@@
   \centering
-  \begin{subfigure}[b]{0.49\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_2d.pdf}
-      \caption{Roofline (2D)}
-      \label{fig:roofline_2d_tensor4}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.49\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_3d.pdf}
-    \caption{Roofline (3D)}
-    \label{fig:roofline_3d_tensor4}
-  \end{subfigure}
-  ~
-  \begin{subfigure}[b]{0.49\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_stackedbar_2d.pdf}
+  %\begin{subfigure}[b]{0.49\textwidth}
+  %    \centering
+  %    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_2d.pdf}
+  %    \caption{Roofline (2D)}
+  %    \label{fig:roofline_2d_tensor4}
+  %\end{subfigure}
+  %\begin{subfigure}[b]{0.49\textwidth}
+  %  \centering
+  %  \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_3d.pdf}
+  %  \caption{Roofline (3D)}
+  %  \label{fig:roofline_3d_tensor4}
+  %\end{subfigure}
+  %~
+  \begin{subfigure}[b]{0.49\textwidth}
+      \centering
+      \includegraphics[width=0.94\textwidth]{LIKWID_CSL_Munich_breakdown_stackedbar_2d.pdf}
       \caption{Computing times (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_stackedbar_3d.pdf}
+    \includegraphics[width=0.94\textwidth]{LIKWID_CSL_Munich_breakdown_stackedbar_3d.pdf}
     \caption{Computing times (3D)}

### 2019-08-04 doc/large_strain.tex
mention zeroing and MPI in algorithms

@@
   \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
+  zero destination vector $\gz y = 0$ \;
+  update ghost values for source vector $\gz x$ \;
   \ForEach{ element $K \in \Omega^h$ }{
@@
   }
+  compress operation on $\gz y$ \;
   \caption{Matrix-free tangent operator: cache scalar quantities only}
@@
   }
+  compress operation on $\gz y$ \;
   \caption{Matrix-free tangent operator: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
@@
   }
+  compress operation on $\gz y$ \;
   \caption{Matrix-free tangent operator: cache material part of the fourth-order spatial tangent stiffness tensor $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}

### 2019-08-04 doc/large_strain.tex
spellout symmetrize in algorithms

@@
           evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
-          evaluate $\gz g_s = \text{symmetrize} (\gz g)$ \tcp*{2nd order symmetric}
+          evaluate $\gz g_s = (\gz g + \gz g^T)/2$ \tcp*{2nd order symmetric}
           evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \textrm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
@@
           evaluate $\boldsymbol{\mathcal{G}}\gz g := \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
-          evaluate $\gz g_s = \text{symmetrize} (\gz g)$ \tcp*{2nd order symmetric}
+          evaluate $\gz g_s = (\gz g + \gz g^T)/2$ \tcp*{2nd order symmetric}
           evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \textrm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
@@
           evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
-          evaluate $\gz g_s = \text{symmetrize} (\gz g)$ \tcp*{2nd order symmetric}
+          evaluate $\gz g_s = (\gz g + \gz g^T)/2$ \tcp*{2nd order symmetric}
           evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}

### 2019-08-05 doc/large_strain.tex
Update experiment descriptions.
Add some references.
Switch style to new Wiley one for IJNME (I had to hardcode a few decisions).

--- 2019-08-03 doc/large_strain.tex
@@
 \newif\ifijnme %Declaration of ijnme
-% \ijnmetrue %Sets ijnme to true
-\ijnmefalse %Sets ijnme to false
+\ijnmetrue %Sets ijnme to true
+% \ijnmefalse %Sets ijnme to false
 
 \ifijnme
-\documentclass[times,doublespace]{nmeauth}
+\documentclass[AMA,STIX1COL]{WileyNJD-v2}
 \else
@@
 }
+
 \usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}
-
+\ifijnme
+\usepackage{subcaption}
+\else
 \usepackage[format=plain,indention=.5cm]{caption}
@@
 \renewcommand\thesubfigure{(\alph{subfigure})}
+\fi
 
@@
 
+\ifijnme
+\articletype{Research Article}%
+
+\received{1 January 2019}
+\revised{}
+\accepted{}
+\fi
+
 \begin{document}
@@
 \ifijnme
-\runningheads{D. Davydov et al.}{A matrix-free approach for finite-strain hyperelastic problems}
 \else
@@
 \ifijnme
-  \author{
-    Denis Davydov\affil{1} \corrauth ,
-    Jean-Paul Pelteret\affil{1},
-    Daniel Arndt\affil{2},
-    Paul Steinmann\affil{1,3}
-  }
-
-  \address{
-    \affilnum{1}Chair of Applied Mechanics,
-    Friedrich-Alexander-Universit\"{a}t Erlangen-N\"{u}rnberg,
-    Egerlandstr.\ 5, 91058 Erlangen, Germany \break
-    \affilnum{2}Interdisciplinary Center for Scientific Computing (IWR),
-    Heidelberg University,
-    Im Neuenheimer Feld 205, 69120 Heidelberg, Germany \break
-    \affilnum{3}Glasgow Computational Engineering Center (GCEC),
-        University of Glasgow, G12 8QQ Glasgow, United Kingdom \break
-    }
-
-  \corraddr{E-mail: denis.davydov@fau.de}
+\author[1]{Denis Davydov*}
+\author[1]{Jean-Paul Pelteret}
+\author[2]{Daniel Arndt}
+\author[3]{Martin Kronbichler}
+\author[1,4]{Paul Steinmann}
+
+\address[1]{\orgdiv{Chair of Applied Mechanics}, \orgname{Friedrich-Alexander-Universit\"{a}t Erlangen-N\"{u}rnberg}, \orgaddress{\country{Germany}}}
+
+\address[2]{\orgdiv{Interdisciplinary Center for Scientific Computing (IWR)}, \orgname{Heidelberg University}, \orgaddress{\country{Germany}}}
+
+\address[3]{\orgdiv{Institute for Computational Mechanics}, \orgname{Technical University of Munich}, \orgaddress{\country{Germany}}}
+
+\address[4]{\orgdiv{Glasgow Computational Engineering Center (GCEC)}, \orgname{University of Glasgow}, \orgaddress{\country{United Kingdom}}}
+
+\corres{*D. Davydov, Egerlandstr.\ 5, 91058 Erlangen, Germany. \email{denis.davydov@fau.de}}
+
+\presentaddress{Chair of Applied Mechanics, Friedrich-Alexander-Universit\"{a}t Erlangen-N\"{u}rnberg, Egerlandstr.\ 5, 91058 Erlangen, Germany.}
 
 \else
+
   \author[a]{Denis Davydov\corref{cor}}
@@
 \fi
-
-  \begin{abstract}
-    This work investigates matrix-free algorithms for problems in quasi-static finite-strain hyperelasticity.
+  \begin{abstract}[Summary]
+    {This work investigates matrix-free algorithms for problems in quasi-static finite-strain hyperelasticity.
 Matrix-free algorithms have emerged as an attractive alternative to iterative solvers with sparse matrices in the fluid dynamics and wave propagation context because they significantly reduce the memory traffic, the limiting factor in classical finite element solvers.
@@
     The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner on a single node for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
-    We conclude that matrix-free methods for finite-strain solid mechanics are very promising, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.
+    We conclude that matrix-free methods for finite-strain solid mechanics are very promising, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
   \end{abstract}
@@
 
+\jnlcitation{\cname{%
+\author{D. Davydov},
+\author{J.-P. Pelteret},
+\author{D. Arndt},
+\author{M. Kronbichler} and
+\author{P. Steinmann}} (\cyear{2019}),
+\ctitle{A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid}, \cjournal{International Journal for Numerical Methods in Engineering}, \cvol{2019;00:1--23}.}
+
   \maketitle
@@
 \else
+
     \begin{keyword}
@@
 In order to improve the performance of iterative solvers, so-called matrix-free methods,
-where matrix-vector products are formed on-the-fly \cite{kronbichler12,May2015, Krank2017, Brown2010, Gmeiner2016},
+where matrix-vector products are formed on-the-fly \cite{Brown2010,kronbichler12,May2015, Krank2017,  Gmeiner2016},
 are widely adopted in the fluid mechanics community.
 Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016, ljungkvist2017multigrid}.
+Among matrix-free methods, stencil-based variants are most popular for linear elements and tetrahedral shapes in particular~\cite{Gmeiner2016,Bauer2018}, whereas implementations based on integration with sum-factorization techniques are attractive for higher polynomial degrees $p\geq 2$ and tensor product shape functions~\cite{Brown2010,kronbichler12,May2015,Cantwell2011}.
 
@@
 more elaborate as compared to the common operators in linear and non-linear fluid mechanics.
-For matrix-free schemes using fast integration via sum factorization, the most common matrix-free variant for polynomial degrees $p\geq 2$, the main challenge is to find a representation of the linearization point, either in referential or spatial configurations, at the quadrature points that is cheap enough to be evaluated on the fly.
+For matrix-free schemes using fast integration via sum factorization, the main challenge is to find a representation of the linearization point, either in referential or spatial configurations, at the quadrature points that is cheap enough to be evaluated on the fly.
 Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
@@
 However, we note that, even with linear FEM, large scale computations with as many as $10^{12}$ unknowns are possible with a matrix-free approach but infeasible with sparse matrices as there is simply not enough memory to store the sparse tangent matrix even on large supercomputers \cite{Gmeiner2016}.
-Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.
+Therefore, we believe that for large-scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient solvers.
 
@@
 pre-computation with a higher memory access.
-The \texttt{deal.II} \cite{dealII90} finite element library, which provides MPI parallelization together with a generalized matrix-free framework that incorporates SIMD vectorization and an interface to high-performance, MPI distributed matrix-based libraries, is used for this study.
+The \texttt{deal.II} \cite{dealII90} finite element library, which provides MPI parallelization together with a generalized matrix-free framework that incorporates SIMD vectorization and an interface to high-performance MPI-distributed matrix-based libraries is used for this study.
 In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
@@
 \mcl E =
-\int_{\mcl B_0} \mcl \psi(\gz F) \, \d V \,
+\int_{\mcl B_0} \psi(\gz F) \, \d V \,
 - \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \, \d S \, ,
 \end{align}
-where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the referential configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
+where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the referential configuration, $\psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
 
@@
 \end{align}
-where $\gz P := \partial \mcl \psi / \partial \gz F$ is the Piola stress tensor.
+where $\gz P := \partial \psi / \partial \gz F$ is the Piola stress tensor.
 The double contraction in the first term can be re-written in terms of the symmetric Kirchhoff stress tensor $\gz \tau := \gz P \cdot \gz F^T$ as
@@
 is involved due to the symmetry of $\gz \tau$. Here, `$\grad$' denotes the gradient with respect to the spatial configuration, as opposed to the gradient in the referential setting `$\Grad$'.
-Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
+Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \psi / \partial \gz E \equiv 2 \partial \psi / \partial \gz C$,
 that is, $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.
@@
 in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
-For the total matrix-vector product we get $\mathcal{O}(n^{2d})$, \todo{Should we include the number of elements in the total mat-vec?}
-and this is the same as for a regular matrix-vector product where the entries are already precomputed.
+For the total matrix-vector product we get a cost of $\mathcal{O}(n^{2d})$ per element,
+the same as for a regular matrix-vector product where the entries are already precomputed.
 Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.
 
-\begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.45\textwidth}
+\begin{figure}
+  \centering
+  \begin{subfigure}[b]{0.3\textwidth}
       \centering
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.3\textwidth}
     \centering
@@
   \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
+  zero destination vector $\gz y = 0$ \;
+  update ghost values for source vector $\gz x$ with MPI \;
   \ForEach{ element $K \in \Omega^h$ }{
@@
   }
+  compress remote integral contributions into $\gz y$ via MPI \;
   \caption{Matrix-free tangent operator: cache scalar quantities only}
@@
   \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
+  zero destination vector $\gz y = 0$ \;
+  update ghost values for source vector $\gz x$ with MPI\;
   \ForEach{ element $K \in \Omega^h$ }{
@@
   }
+  compress remote integral contributions into $\gz y$ via MPI \;
   \caption{Matrix-free tangent operator: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
@@
   }
+  compress remote integral contributions into $\gz y$ via MPI \;
   \caption{Matrix-free tangent operator: cache material part of the fourth-order spatial tangent stiffness tensor $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
@@
 \centering
+\caption{Details of memory access for matrix-free evaluation with Algorithms~\ref{alg:mf_scalar}--\ref{alg:mf_tensor4}.}
+\label{tab:memory}
 \begin{tabular}{|l|ccc|}
@@
 \end{tabular}
-\caption{Details of memory access for matrix-free evaluation with Algorithms~\ref{alg:mf_scalar}--\ref{alg:mf_tensor4}.}
-\label{tab:memory}
 \end{table}
 
-Table~\ref{tab:memory} summarizes the three algorithms in terms of their implied memory access for one operator evaluation.
-The vector access in Algorithm~\ref{alg:mf_scalar} is higher than in the other variants because both the source vector $\gz x$ and the linearization point from the current FE solution $\overline{\gz u^h}$ need to be accessed, each with the usual indirect addressing pertaining to continuous finite element fields \cite{kronbichler12}.
-The cached quantities listed in the algorithms, labeled ``physics terms'' in the table, are increasing in size for the algorithms, and are separate for each point.
-Finally, the evaluation of the gradients in the referential and spatial coordinates also involve a data access via the so-called metric terms, the Jacobian from the unit element to the referential or spatial configurations, respectively.
-Algorithm~\ref{alg:mf_scalar} involves both the Jacobian in the referential and the spatial frame, whereas the other two algorithms only involve the spatial (Eulerian) terms.
-Finally, the implementation framework also holds the determinant of the Jacobian for each point.
-Following \cite{kronbichler12}, some compression may be applied to the metric terms: On affine geometries, the Jacobian is the same at all quadrature points of a cell, and needs only be stored once.
-Therefore, the table lists those quantities as ``up to'' $d^2+1$ terms, which depends on the mesh and is a run-time decision in the code.
-
-\section{Description of the utilized numerical framework
+Table~\ref{tab:memory} summarizes the three algorithms in terms of their implied
+memory access for one operator evaluation. The vector access in
+Algorithm~\ref{alg:mf_scalar} is higher than in the other variants because both
+the source vector $\gz x$ and the linearization point from the current FE
+solution $\overline{\gz u^h}$ need to be accessed, each with the usual indirect
+addressing pertaining to continuous finite element fields \cite{kronbichler12}.
+The cached quantities listed in the algorithms, labeled ``physics terms'' in the
+table, are increasing in size with more caching, and are separate data for each
+quadrature point. Finally, the evaluation of the gradients in the referential
+and spatial coordinates also involves a data access via the so-called metric
+terms, the inverse of the Jacobian from the unit element to the referential or
+spatial configurations, respectively. Algorithm~\ref{alg:mf_scalar} involves
+both the Jacobian in the referential and the spatial frame, whereas the other
+two algorithms only involve the spatial (Eulerian) terms. Finally, the
+implementation framework also holds the determinant of the Jacobian for each
+point. Following \cite{kronbichler12}, some compression may be applied to the
+metric terms: On affine geometries, the Jacobian is the same at all quadrature
+points of a cell, and needs only be stored once. Therefore, the table lists
+those quantities as ``up to'' $d^2+1$ terms, with a memory access that depends
+on the mesh via a run-time decision in the code.
+
+\section{Description of the utilized numerical framework}
 \label{sec:framework}
-}
-
-The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling}, were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
-This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aims to support exascale computing.
+
+The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling}, was implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
+This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aiming to support exascale computing.
 In this work we employ \texttt{deal.II}'s interface to the \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
@@
   \centering
-  \begin{subfigure}[b]{0.8\textwidth}
+  \begin{subfigure}[b]{0.4\textwidth}
     \centering
@@
 various pre- and post-processing steps also leveraged parallelization.
-For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
+For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently usedg.
 
@@
 Such operators are, however, computationally very demanding.
-In this work we propose a different approach:
+In this work we avoid these complications by assuming
 % In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
@@
 % Based on this observation we adopt the following approach:
-It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material.
+that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material.
 In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
@@
 
-The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library; implementational details are discussed in \cite{Clevenger2018}.
+The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators, and the coarse level iterative solver are provided by the \texttt{deal.II} library; implementational details are discussed in \cite{Clevenger2018}.
 %
@@
 %
-Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
+Finally, we note that a $p-$ version of multigrid \cite{Ronquist1987} would also be conceivable but is not considered here.
 %
@@
   \centering
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.35\textwidth}
       \centering
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.35\textwidth}
     \centering
@@
 
-In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}.
+In this section, we evaluate the proposed matrix-free operator algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner on a benchmark problem from \cite{Miehe2007}.
 The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}.
@@
 \begin{table}[!htb]
+  \caption{Parameters for the benchmark: $p$ is the polynomial degree,
+  $q$ is the number of quadrature points in 1D, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.}
   \begin{subtable}{.5\linewidth}
@@
   \end{subtable}
-  \caption{Parameters for the benchmark: $p$ is the polynomial degree,
-  $q$ is the number of quadrature points in 1D, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
-  }
 \end{table}
@@
 Examinations performed in this manner are most representative for the differences between the matrix-free and matrix-based approaches because problems of sufficient size are dominated by the local node-level performance with multigrid solvers.
-Multigrid solvers are well-known to be scalable given sufficiently small coarse-grid problems even to the largest supercomputers \cite{gholami2016,ibeid2018} as the communication is mostly between nearest neighbors.
+Multigrid solvers with small enough coarse-grid problems are well-known to be scalable even to the largest supercomputers \cite{gholami2016,ibeid2018} as the communication is mostly between nearest neighbors.
 The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
-We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach with a high polynomial order finite element.
+We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach with a high polynomial degree.
 Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound.
+Furthermore, also the data structures of the matrix-free approaches are too large to fit completely into caches.
 
@@
 On the coarsest level, we use the Chebyshev preconditioner as a solver \cite{Varga2009} to reduce the residual by three orders of magnitude.
-Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.
-
-Computations were performed on two systems: A dual-socket Intel Xeon Gold 6230 (``Cascade Lake'') with 20 cores per socket and 96 GB of DDR4-2933 memory (nominal bandwidth per socket: 141 GB/s).
+Since this smoother only uses diagonal elements of the operator, it is readily compatible with the matrix-free approach to leverage fast matrix-vector products, as opposed to most default smoothers which rely on an explicit matrix form.
+
+Computations were performed on two systems: A dual-socket Intel Xeon Gold 6230 (``Cascade Lake''), released in 2019, with 20 cores per socket and 96 GB of DDR4-2933 memory (nominal bandwidth per socket: 141 GB/s).
 Turbo mode is enabled with a power limit of 135 watt per socket, which results in the processor running at 2.8 GHz for scalar code and 2.0 GHz for AVX-512-heavy code in all experiments.
@@
 Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
-The ``IWR'' results were obtained on a single machine with eight Intel Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip.
+The ``IWR'' results were obtained on a single machine with eight Intel Xeon E7-8870 ``Westmere'' chips, released in 2011, (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip.
 For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
@@
 
-\begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.49\textwidth}
+Figures \ref{fig:benchmark_miehe_CSL} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem on the two hardware systems.
+We are interested in the wall-clock time (average over 10 runs for each linear solver step) per DoF as a first metric.
+As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
+as the matrix becomes more and more dense.
+In order to increase visibility, the figures also report the throughput of the matrix vector product, computed by the ratio of the number of degrees of freedom over the time for a matrix-vector product, in panels (a) and (d).
+The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
+Note that we consider only deformed meshes together with separate metric terms according to Table~\ref{tab:memory}.
+These deformed terms are the result of manifold descriptions of the boundaries and interfaces  to capture the geometry as well as
+the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration.
+
+\begin{figure}
+  \centering
+  \begin{subfigure}[b]{0.32\textwidth}
       \centering
@@
       \caption{throughput matrix-vector product (2D)}
-      \label{fig:benchmark_miehe_CSL_vmult2}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.49\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{CSL_Munich_throughput3d.eps}
-    \caption{throughput matrix-vector product (3D)}
-    \label{fig:benchmark_miehe_CSL_vmult3}
-  \end{subfigure}
-  ~
-  \begin{subfigure}[b]{0.49\textwidth}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\textwidth}
       \centering
@@
   \end{subfigure}
-  \begin{subfigure}[b]{0.49\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{CSL_Munich_memory2d.eps}
+      \caption{memory access (2D)}
+      \label{fig:benchmark_miehe_CSL_memory2}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[b]{0.32\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{CSL_Munich_throughput3d.eps}
+    \caption{throughput matrix-vector product (3D)}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\textwidth}
     \centering
@@
   \end{subfigure}
-  ~
-  \begin{subfigure}[b]{0.49\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{CSL_Munich_memory2d.eps}
-      \caption{memory consumption (2D)}
-      \label{fig:benchmark_miehe_CSL_memory2}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.49\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \centering
     \includegraphics[width=\textwidth]{CSL_Munich_memory3d.eps}
-    \caption{memory consumption (3D)}
+    \caption{memory access (3D)}
     \label{fig:benchmark_miehe_CSL_memory3}
   \end{subfigure}
-  \caption{Cascade Lake, 1 socket with 20 cores, matrix-vector multiplication.}%
+  \caption{Cascade Lake, one socket with 20 cores, matrix-vector multiplication.}%
   \label{fig:benchmark_miehe_CSL}
@@
 
-\begin{figure}[!ht]
-  \centering
-  \begin{subfigure}[b]{0.49\textwidth}
+\begin{figure}
+  \centering
+  \begin{subfigure}[b]{0.32\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{IWR_newest_patched_throughput2d.eps}
+      \caption{throughput matrix-vector product (2D)}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\textwidth}
       \centering
       \includegraphics[width=\textwidth]{IWR_newest_patched_timing2d.eps}
-      \caption{matrix-vector product (2D)}
+      \caption{time matrix-vector product (2D)}
       \label{fig:benchmark_miehe_IWR_vmult2}
   \end{subfigure}
-  \begin{subfigure}[b]{0.49\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{IWR_newest_patched_memory2d.eps}
+      \caption{memory access (2D)}
+      \label{fig:benchmark_miehe_IWR_memory2}
+  \end{subfigure}
+  ~
+  \begin{subfigure}[b]{0.32\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{IWR_newest_patched_throughput3d.eps}
+    \caption{throughput matrix-vector product (3D)}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.32\textwidth}
     \centering
     \includegraphics[width=\textwidth]{IWR_newest_patched_timing3d.eps}
-    \caption{matrix-vector product (3D)}
+    \caption{time matrix-vector product (3D)}
     \label{fig:benchmark_miehe_IWR_vmult3}
   \end{subfigure}
-  ~
-  \begin{subfigure}[b]{0.49\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{IWR_newest_patched_memory2d.eps}
-      \caption{memory consumption (2D)}
-      \label{fig:benchmark_miehe_IWR_memory2}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.49\textwidth}
+  \begin{subfigure}[b]{0.32\textwidth}
     \centering
     \includegraphics[width=\textwidth]{IWR_newest_patched_memory3d.eps}
-    \caption{memory consumption (3D)}
+    \caption{memory access (3D)}
     \label{fig:benchmark_miehe_IWR_memory3}
   \end{subfigure}
-  \caption{IWR cluster, matrix-vector multiplication}%
+  \caption{IWR cluster, 20 cores, matrix-vector multiplication}%
   \label{fig:benchmark_miehe_IWR}
@@
 
-Figures \ref{fig:benchmark_miehe_CSL} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
-We are interested in the wall-clock time (average over 10 runs for each linear solver step) per DoF as a first metric.
-As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
-as the matrix becomes more and more dense.
-The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
-Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces \cite{dealII90} to capture the geometry as well as
-the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
-Both factors increase the memory consumption and make element operations more expensive.
-Nonetheless,
-the matrix-free implementation is faster than the matrix-based counterpart already for bi-cubic Lagrangian basis in 2D and tri-quadratic in 3D.
-Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
+The matrix-free implementation is faster than the matrix-based counterpart already for a bi-quadractic Lagrangian basis in 2D and a tri-quadratic in 3D on both systems on both systems The ``MF tensor2'' and ``MF scalar'' strategies outperform the sparse matrix already for tri-linear elements of polynomial degree $p=1$.
+For example for the ``tensor2'' matrix-free implementation with $p=4$, we record a 5$\times$ higher throughput than the matrix-based evaluation of $p=1$.
+The advantage of the matrix-free implementation can be explained by the memory access per unknown shown in the panels (c) and (f) of the figures, as the time per unknown in panels (b) and (e) closely follows the memory access.
+Figures \ref{fig:benchmark_miehe_CSL_vmult2}, \ref{fig:benchmark_miehe_CSL_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
 described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.
 Two conclusions can be drawn from these results:
-The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
-There is little difference in terms of the wall-clock time per DoF between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'').
-This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
-Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
+Firstly, the algorithm where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'') is the fastest approach in both 2D and 3D and on both systems.
+When compared to the scalar caching implementation, the ``tensor2'' variant loads slightly less data, because only the spatial configuration is needed in terms of metric terms, as compared to both the spatial and the referential configuration in the scalar caching cache in order to evaluate the Kirchhoff stress $\gz \tau$.
+Besides the memory access, the scalar caching strategy also involves considerably more computations at quadrature points.
+On the newer Cascade Lake system with a Flop/Byte ratio of 9.1 for the theoretical peak values, the scalar caching strategy is closer to the ``tensor2'' variant.
+On the IWR machine, the Flop/Byte ratio is 2.9, which means that much fewer computations can be sustained for the same memory transfer.
+
+Secondly, the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') is slower than the ``tensor2'' approach, despite fewer arithmetic operations, on both systems.
+This shows that the memory access is the deciding metric for the ``tensor4'' strategy.
+Nonetheless, the ``tensor4'' strategy with $p\geq 2$ is considerably faster than matrix-based strategies: For $p=4$ in 3D, the matrix-free variant is three times as fast as a sparse matrix-vector product on \emph{linear elements} on Cascade Lake.
+This means we can apply this matrix-free variant to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
 It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
@@
 As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
-Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point (``tensor4''), the matrix-free approach takes much less memory than its matrix-based counterpart.
-In those graphs, we report the overall memory needed to cache additional quantities required for the tangent operator here considered,
-as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
-from unit to real cell \cite{kronbichler12}.
-Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deformed and undeformed
-configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.
-
-Finally, we analyze the node-level performance of the two approaches using the roofline model.
-Memory bandwidth (MBytes/s) and the number of floating point operations per second (MFLOP/s) are measured using the \texttt{MEM\_DP} group of the LIKWID \cite{likwid} tool, version 4.2.1.
-The measurements are done on a single socket of RRZE cluster by pinning all 10 MPI processes to this socket, whose
-memory bandwidth\footnote{As measured with the \textit{load\_avx} benchmark of \textit{likwid-bench} \cite{likwid}.} and peak performance are $B=47.2 \,\rm{GB/s}$ and $P=176\, \rm{GFlop/s}$, respectively.
-Turbo mode of the CPU was disabled.
+Figures \ref{fig:benchmark_miehe_CSL_memory2}, \ref{fig:benchmark_miehe_CSL_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point (``tensor4''), the matrix-free approach takes much less memory than its matrix-based counterpart.
+In those graphs, we report the overall memory needed to cache additional quantities required for the tangent operator according to Table~\ref{tab:memory}, as well as all metadata required by
+the \texttt{deal.II} \textit{MatrixFree} object.
+
+Finally, we analyze the node-level performance of the two approaches using the roofline performance model \cite{Williams2009}.
+The achieved memory throughput (MBytes/s) and the number of floating point operations executed per second (MFLOP/s) are measured using the \texttt{MEM\_DP} group of the LIKWID \cite{likwid} tool, version 4.2.1.
+The measurements are done on a single socket of the Cascade Lake system by pinning all 20 MPI processes to this socket, whose arithmetic peak performance at 2.0 GHz is $1280\, \text{GFlop/s}$ and measured memory bandwidth for pure read tasks is $120\, \text{GB/s}$ as well as $90\, \text{GB/s}$ for mixed data access of two loads per one store.
 
@@
   \end{subfigure}
-  \caption{Roofline model for a selection of polynomial degrees. `Peak DP' denotes the peak double precision performance ($P$),
+  \caption{Roofline model for a selection of polynomial degrees on Cascade Lake. `Peak DP' denotes the peak double precision performance ($P$),
   `w/o FMA' represents a ceiling without Fused Multiply-Add ($P/2$), finally `w/o SIMD' is a ceiling where both FMA
-    and SIMD vectorization are discarded ($P/8$).}%
+    and SIMD vectorization are discarded ($P/16$).}%
   \label{fig:roofline}
@@
 
-Figure \ref{fig:roofline} shows results of the roofline performance model.
-As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $\pgfmathprintnumber{8.2093245}\,\rm{GFlop/s}$ (4.6\% of the peak performance) in 2D.
+The results of the roofline analysis in Figure~\ref{fig:roofline} show that the matrix-based approach is fully memory-bandwidth bound with a very low arithmetic intensity of 0.16 and a performance of $\pgfmathprintnumber{17.36}\,\rm{GFlop/s}$ (1.4\% of the arithmetic peak) averaged over the 2D results.
 The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
-$\pgfmathprintnumber{18.8047083222}\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
-The achieved performance is about 10.6\% of the peak arithmetic performance. This is much lower than the
-70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}, but similar to what was obtained in \cite[Figure 14]{kronbichler2017fast} when not optimizing for Cartesian meshes.
-Unfortunately, performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.
-
-When comparing the data for the ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}) caching strategy,
-we observe a marginal increase in computational intensity and almost unchanged performance.
-This coincides with our expectation that the latter algorithm requires transferring fewer bytes to perform operations on a single quadrature point
-(compare line 9 in Algorithm \ref{alg:mf_tensor4} that loads a symmetric rank-4 tensor to the line 9 in Algorithm \ref{alg:mf_tensor2} that evaluates the
-same contraction by loading two scalars $c_1$ and $c_2$).
-On the other hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
-This is not surprising as at each quadrature point we need to re-evaluate the Kirchoff stress $\gz \tau$.
-Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
-second-order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
-As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}),
-however the ``scalar'' algorithm performs a large number of repetitive operations that are
-independent on the right hand side vector (steps 8--11 in Algorithm \ref{alg:mf_scalar}), which makes it the worst option among the three.
+$\pgfmathprintnumber{86.9}\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
+The achieved performance is about 6.8\% of the peak arithmetic performance. This result similar to the deformed mesh results of \cite[Figure 18]{kronbichler2017fast}.
+
+The roofline model also explains the difference in performance between the ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) and the ``tensor2'' caching strategy (Algorithm \ref{alg:mf_tensor2}): Both algorithms essentially run at the memory limit of around $100\, \text{GB/s}$, but the former involves a considerably higher memory transfer according to Table~\ref{tab:memory}.
+The ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates the highest computational intensity, but remains close to the memory bandwidth limit with more than $80\, \text{GB/s}$ except for the $p=2$ case.
+The higher arithmetic load is not surprising as at each quadrature point we need to re-evaluate the Kirchoff stress $\gz \tau$.
+Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations as listed in Table~\ref{tab:memory}, no advantage of the scalar caching can be deduced despite the higher arithmetic performance.
 
@@
 
-In order to have a better understanding of the behavior of the most promising ``tensor4'' implementation,
-we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
-(i) `read / write' denotes reading and writing from/to a global vector (lines 2 and 13);
+In order to better understand the behavior of the matrix-free implementations,
+we exemplarily collect a breakdown of the operator evaluation into various stages inside the element loop of Algorithm \ref{alg:mf_tensor4}:
+(i) `read / write' denotes reading and writing from/to a global vector (lines 2 and 13) with indirect addressing of gather/scatter type;
 (ii) `sum factorization' denotes application of sum factorization techniques (lines 4, 5 and 12);
 (iii) `quadrature loop' denotes operations performed at each quadrature (lines \mbox{6 -- 11}).
-Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one
-global refinement in 2D and no global mesh refinement in 3D.
-It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that the
-`read / write' steps take a negligible amount of time on each element.
-For higher order bases, the majority of time is spent within the quadrature loop.
-When analyzing the roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
-we observe that this part of the algorithm demonstrates rather low performance, which in our opinion is the source of the
-suboptimal performance of the whole matrix-free matrix-vector product (see Figure \ref{fig:roofline}).
-In this case the, quadrature loop consists of, essentially, a
-single contraction between two second order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
-double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
+Given that the measurements are within the cell loop and there is an extensive overlap between the memory transfer (due to hardware prefetching) and arithmetic operations, the data of Figure~\ref{fig:breakdown} is derived from measuring the cost of `sum factorization' by comparing the timings of part (i) and (ii) together against part (i) only.
+From Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4}, it becomes clear that sum factorization is cheap enough to be almost completely hidden behind the memory transfer.
+The majority of time is spent within the quadrature loop for all variants, with an increasing proportion for increasing degrees.
+Note also that higher degrees derive a higher throughput in terms of DoF/s in Figures~\ref{fig:benchmark_miehe_CSL} and~\ref{fig:benchmark_miehe_IWR}, as the memory access of quadrature point data scales as $(p+1)^d/p^d$ compared to the number of unknowns.
 
@@
 Furthermore, the improvement due to explicit intrinsics for the data used in the quadrature loop, is comparable to the ones in \cite[Figure 6]{kronbichler2017fastcomputer} where the speedup factor is $1.2-1.8$ for a similar architecture.
+\todo[inline]{Do we want to update Tables \ref{tab:numbers_2d} and \ref{tab:numbers_3d}?}
 
 To summarize this section, we can attribute the speed-up of the matrix-free finite-strain tangent operators for higher-order elements to two factors.
-The first one is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
-bound, see Figure \ref{fig:roofline}.
-The second one is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
+The first one is the higher algorithmic intensity and lower memory access as compared to the matrix-based implementation, see Figure \ref{fig:roofline}.
+Since both variants are in the memory-limited regime, the matrix-free implementation with its lower memory access is beneficial.
+The second benefit is the reduction in the number of floating point operations per DoF for higher order elements thanks to the sum factorization technique, as discussed in
 Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
-Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach with relatively high computational intensity.
+Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach, whereas Algorithm~\ref{alg:mf_tensor2} is faster due to a higher arithmetic intensity.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{Emmy_RRZE_cg2d.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_cg2d.eps}
     \caption{CG iterations (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{Emmy_RRZE_cg3d.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_cg3d.eps}
     \caption{CG iterations (3D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{Emmy_RRZE_solver2d.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_solver2d.eps}
     \caption{CG solution time (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{Emmy_RRZE_solver3d.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_solver3d.eps}
     \caption{CG solution time (3D)}
@@
   \end{subfigure}
-  \caption{Emmy cluster, RRZE, iterative solver.}%
+  \caption{Cascade Lake, one socket with 20 cores, iterative solver.}%
   \label{fig:benchmark_miehe_Emmy_cg}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_cg2d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_cg2d.eps}
     \caption{CG iterations (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_cg3d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_cg3d.eps}
     \caption{CG iterations (3D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_solver2d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_solver2d.eps}
     \caption{CG solution time (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_solver3d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_solver3d.eps}
     \caption{CG solution time (3D)}
@@
   \end{subfigure}
-  \caption{IWR cluster.}%
+  \caption{IWR cluster, 20 cores.}%
   \label{fig:benchmark_miehe_IWR_cg}
@@
 Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end,
-we consider the average number of CG iterations throughout the entire simulation (i.e. each Newton-Raphson iteration and each loading step).
+we consider the average number of CG iterations throughout the entire simulation, i.e., averaged over each Newton-Raphson iteration and each loading step.
 Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
- noticeable increase in the average number of CG iterations for different polynomial degrees
+ noticeable increase in the average number of CG iterations for increasing polynomial degrees
  for the GMG preconditioner.
@@
 the coarse level solver setup will greatly affect the performance of the preconditioner.
-In particular, we can improve the number of iterations by choosing a more accurate coarse level solver,
-however this leads to an overall more expensive linear solver.
-However, the settings used in this study are optimized for the considered problem in terms of the wall-time spent in the linear solver.
+In particular, we can improve the number of iterations by choosing a more accurate coarse level solver.
+%However, the settings used in this study are optimized for the considered problem in terms of the wall-time spent in the linear solver.
 The black-box AMG preconditioner
@@
 requires many more iterations for convergence.
-Consequently, we can conclude that for the here considered finite strain elasticity problem (i) the selected smoother is suitable, and that (ii) although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires much fewer CG iterations as compared to the algebraic multigrid.
-
-Additionally, we compare solution times of the preconditioned iterative solvers.
-Our goal is to determine whether the matrix-free approach with a GMG preconditioner can be faster than the matrix-based approach with a well-established implementation of an AMG preconditioner.
+Consequently, we can conclude that (i) the selected smoother is suitable for the present finite strain elasticity problem, and that (ii) although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires fewer CG iterations as compared to the algebraic multigrid baseline.
+
+Figures~\ref{fig:benchmark_miehe_Emmy_cg} and~\ref{fig:benchmark_miehe_IWR_cg} also report the solution times of the preconditioned iterative solvers.
+Our goal is to compare the matrix-free solver with a GMG preconditioner against a matrix-based approach with a well-established implementation of an AMG preconditioner.
 We note specifically that the comparison is restricted to the choice of level operators used with the pre-existing GMG framework, and not the framework itself.
@@
 Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
-Based on these results, we conclude that the matrix-free approach with the adopted here geometric multigrid preconditioner can deliver a very competitive solution strategy for engineering problems
+Based on these results, we conclude that the matrix-free approach represents a very competitive solution strategy for engineering problems
 with compressible hyperelastic finite strain material models.
 
-Extension of this study from the node level to the cluster level is beyond the scope of this work.
-However, as a step in this direction we have performed a preliminary study of the weak scalability of the implementation.
+An extension of this study from the node level to the cluster level is beyond the scope of this work.
+However, we have performed a preliminary study of the weak scalability of the implementation.
 Table \ref{tab:weak_3d} demonstrates a good weak scaling up to 1280 processes\footnote{%
@@
 also compare \cite[Figure 7]{Krank2017} for scaling results for up to 147,456 CPU cores and 34.4 billion degrees of freedom for an incompressible flow problem.
+\todo[inline]{Do we want to update?
+I can run on up to 8 nodes with CSL.
+This allows me to do 3 steps of weak scaling in 2D, 1 socket, 2 nodes, 8 nodes, and two steps in 3D, 1 node and 8 nodes.
+Not very large but enough to show our point?}
 
@@
 In this contribution, we proposed and numerically investigated three different matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
-To the best of our knowledge, this is the first work that applies matrix-free sum factorization techniques to partial differential equations arising in this case.
+To the best of our knowledge, this is the first work that evaluates matrix-free sum factorization techniques to the partial differential equations arising in this case.
 Among the three examined algorithms,
-the implementation that caches the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress was shown to be faster than the matrix-based approach for polynomial degrees higher than one in 3D and polynomial degrees higher than two in 2D.
-Given that we did not utilize the specific form of the material part of the stiffness tensor in this case, we conclude
-that the matrix-free implementation of the tangent operator can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
+the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the spatial configuration is faster than caching only a scalar at quadrature points or caching the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress.
+Given that we did not utilize the specific form of the material part of the stiffness tensor in the latter case, we conclude
+that the matrix-free implementation of the tangent operator is somewhat slower but more flexible because it can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
 
 The roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
-We identified that one of the bottlenecks is related to the quadrature loop on each element where double contractions of fourth-order and second-order symmetric tensor
-as well as single contraction of two second-order tensors is performed.
+Despite the higher arithmetic intensity, the matrix-free implementations are still memory bandwidth-limited and in particular by the data at quadrature points in terms of fourth-order and second-order symmetric tensors.
 
@@
 
-\ifijnme
-\acks
-\else
 \section*{Acknowledgements}
-\fi
 
 D.~Davydov acknowledges the financial support of the German Research Foundation (DFG), grant DA 1664/2-1.
-
-D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to Martin Kronbichler (TU Munich), Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for fruitful discussions on matrix-free operator evaluation approaches.
-
-D.~Arndt was supported by the German Research Foundation (DFG) under the project ``High-order discontinuous
+D.~Davydov, D.~Arndt and J-P.~Pelteret are grateful to Jed Brown (CU Boulder) and Veselin Dobrev (Lawrence Livermore National Laboratory) for fruitful discussions on matrix-free operator evaluation approaches.
+D.~Arndt and M.~Kronbichler were supported by the German Research Foundation (DFG) under the project ``High-order discontinuous
 Galerkin for the exa-scale'' (\mbox{ExaDG}) within the priority program ``Software
 for Exascale Computing'' (SPPEXA).
-
 P.~Steinmann acknowledges the support of the Cluster of Excellence Engineering of Advanced Materials (EAM) which made this collaboration possible, as well as funding by the EPSRC Strategic Support Package ``Engineering of Active Materials by Multiscale/Multiphysics Computational Mechanics''.
@@
 \ifijnme
-\bibliographystyle{wileyj}
+%\bibliographystyle{wileyj}
 \else

### 2019-08-05 Calculations/post_process.py

@@ -361,7 +361,8 @@ remove_creation_date(fig_prefix + 'memory3d.eps')
 
 # clear
 plt.clf()
-plt.ticklabel_format(style='sci', axis='y', scilimits=(0,0))
+plt.yscale('log', basey=10)
+plt.ylim(top=1e-2,bottom=1e-9)
 
 plt.plot(deg2d,solver2d_tr, 'rs--', label='Trilinos Solver')
 # plt.plot(deg2d,solver2d_sc, 'bo--', label='MF scalar')
@@ -374,7 +375,7 @@ plt.plot(deg2d,solver2d_t4_coarse, 'g^--', label='MF Coarse Solver')
 plt.xlabel('polynomial degree')
 plt.ylabel('wall time (s) / DoF')
 plt.plot(deg2d,assembly2d_tr, 'bp--', label='Trilinos Assembly')
-leg = plt.legend(loc='best', ncol=1)
+leg = plt.legend(loc='lower right', ncol=1, labelspacing=0.1)
 plt.tight_layout()
 plt.savefig(fig_prefix + 'solver2d.eps', format='eps')
 remove_creation_date(fig_prefix + 'solver2d.eps')
@@ -382,8 +383,9 @@ remove_creation_date(fig_prefix + 'solver2d.eps')
 # clear
 plt.clf()
 ax = plt.figure().gca()
+plt.yscale('log', basey=10)
+plt.ylim(top=1e-2,bottom=1e-9)
 ax.xaxis.set_major_locator(MaxNLocator(integer=True))
-plt.ticklabel_format(style='sci', axis='y', scilimits=(0,0))
 
 plt.plot(deg3d,solver3d_tr, 'rs--', label='Trilinos Solver')
 # plt.plot(deg3d,solver3d_sc, 'bo--', label='MF scalar')
@@ -396,7 +398,7 @@ plt.plot(deg3d,solver3d_t4_coarse, 'g^--', label='MF Coarse Solver')
 plt.plot(deg3d,assembly3d_tr, 'bp--', label='Trilinos Assembly')
 plt.xlabel('polynomial degree')
 plt.ylabel('wall time (s) / DoF')
-leg = plt.legend(loc='best', ncol=1)
+leg = plt.legend(loc='lower right', ncol=1, labelspacing=0.1)
 plt.tight_layout()
 plt.savefig(fig_prefix + 'solver3d.eps', format='eps')
 remove_creation_date(fig_prefix + 'solver3d.eps')

### 2019-08-05 Calculations/post_process_likwid_csl.py

@@ -60,7 +60,7 @@ sections = [
 clock_speed = 2.0  # GHz
 # memory bandwidth
 # likwid-bench -t load_avx -w S0:1GB:20:1:2
-B=96.2 # 50 GB/s single socket
+B=120 # 120 GB/s single socket
 
 print 'Peak performance calculations:'
 P = clock_speed
@@ -206,7 +206,7 @@ likwid_data.sort(key=lambda tup: (tup[3], tup[1]))
 #####################
 
 params = {'legend.fontsize': 10,
-          'font.size': 20}
+          'font.size': 14}
 plt.rcParams.update(params)
 
 plt.xscale('log', basex=2)
@@ -233,6 +233,7 @@ plt.plot(x,peak, roofline_style, label='_nolegend_')
 # various ceilings (w/o FMA, w/o FMA and vectorization):
 for p_ in [P/2, P/2/8]:
     plt.plot(x,Roofline(x,p_,B), roofline_style, label='_nolegend_')
+plt.plot([0.0725, 14.222],[6.53, 1280], roofline_style, label='_nolegend_')
 
 plt.fill_between(x, base, peak, where=peak>base, interpolate=True, zorder=1, color='aqua', alpha=0.1)
 
@@ -278,8 +279,9 @@ plt.axes().yaxis.set_major_formatter(mp.ticker.FuncFormatter(lambda x, pos: '{0}
 plt.axes().xaxis.set_major_formatter(mp.ticker.FuncFormatter(lambda x, pos: '1/{0}'.format(int(round(1./x))) if x < 1.0 else '{0}'.format(int(round(x))) ))
 
 ang = 45
-y_pos = 7.5 if args.breakdown else 16.5
+y_pos = 7.5 if args.breakdown else 37
 plt.text(xmin, y_pos, 'B={:.1f} GB/s'.format(B), rotation=ang, fontsize=14)
+plt.text(xmin, 12, 'B=90 GB/s'.format(B), rotation=ang, fontsize=14)
 
 x_pos = 9 if not args.breakdown else 16
 plt.text(x_pos,1400,'Peak DP', fontsize=14)

### 2019-08-06 doc/large_strain.tex
add fully referential tangent for our free energy

@@
   + \lambda \ln^{2}\left( J \right)
+\label{eq:neo_hookean_energy}
 \end{gather}
@@
 
+For completeness we mention that the fully referential linearization of the here considered constitutive equation \eqref{eq:neo_hookean_energy} results in the following equivalent bilinear form
+\begin{align}
+  A_{ij} \equiv a(\gz N_i, \gz N_j) =&
+  \int_{\mcl B_0}
+  \grad \gz N_i : D_{F}P : \grad \gz N_j
+  \d V
+  \label{eq:algebraic_tangent_dPdF}
+  \\
+  D_{F}P :=&
+  2 \lambda
+  \gz F^{-T} \otimes \gz F^{-T}
+  +
+  \left[\mu - 2 \lambda \ln\left( J \right) \right]
+  \gz F^{-T} \underline{\otimes} \gz F^{-1}
+  +
+  \mu \gz I \overline{\otimes} \gz I
+\end{align}
+where the upper and lower dyadic products of pairs of second-order tensors are respectively given by
+\begin{align*}
+    \gz A \overline{\otimes} \gz B &= A_{ik} B_{jl} \gz e_i \otimes \gz e_j \otimes \gz e_k \otimes \gz e_l \\
+    \gz A \underline{\otimes} \gz B &= A_{il} B_{jk} \gz e_i \otimes \gz e_j \otimes \gz e_k \otimes \gz e_l  \, .
+\end{align*}
+
 \section{Matrix-free operator evaluation}

### 2019-08-06 doc/large_strain.tex
remove commented out text, fix few typos and mark majority of text changes w.r.t. arXiv version using red color

@@
   \begin{abstract}[Summary]
-    {This work investigates matrix-free algorithms for problems in quasi-static finite-strain hyperelasticity.
+    {{\color{red}This work investigates matrix-free algorithms for problems in quasi-static finite-strain hyperelasticity.
 Matrix-free algorithms have emerged as an attractive alternative to iterative solvers with sparse matrices in the fluid dynamics and wave propagation context because they significantly reduce the memory traffic, the limiting factor in classical finite element solvers.
-    Specifically, we study different matrix-free realizations of the finite element tangent operator, and determine whether generalized methods of incorporating complex constitutive behavior might be feasible.
+    Specifically, we study different matrix-free realizations of the finite element tangent operator, and determine whether generalized methods of incorporating complex constitutive behavior might be feasible.}
     In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
@@
 Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016, ljungkvist2017multigrid}.
+{\color{red}
 Among matrix-free methods, stencil-based variants are most popular for linear elements and tetrahedral shapes in particular~\cite{Gmeiner2016,Bauer2018}, whereas implementations based on integration with sum-factorization techniques are attractive for higher polynomial degrees $p\geq 2$ and tensor product shape functions~\cite{Brown2010,kronbichler12,May2015,Cantwell2011}.
+}
 
@@
 more elaborate as compared to the common operators in linear and non-linear fluid mechanics.
+{\color{red}
 For matrix-free schemes using fast integration via sum factorization, the main challenge is to find a representation of the linearization point, either in referential or spatial configurations, at the quadrature points that is cheap enough to be evaluated on the fly.
+}
 Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
@@
 
+{\color{red}
 In this work, we investigate different matrix-free representations of the
@@
 and employ them to define a \mbox{geometric} multigrid preconditioner.
+}
 
@@
 
+{\color{red}
 For completeness we mention that the fully referential linearization of the here considered constitutive equation \eqref{eq:neo_hookean_energy} results in the following equivalent bilinear form
+}
 \begin{align}
@@
         gather local vector values on this element $\gz x_K$ \;
-        evaluate the following tensors at each quadrature point using sum factorization :\\
+        evaluate the following tensor at each quadrature point using sum factorization :\\
         \enskip$\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
@@
 
+{\color{red}
 Table~\ref{tab:memory} summarizes the three algorithms in terms of their implied
@@
 quadrature point. Finally, the evaluation of the gradients in the referential
-and spatial coordinates also involves a data access via the so-called metric
+and spatial configuration also involves a data access via the so-called metric
 terms, the inverse of the Jacobian from the unit element to the referential or
 spatial configurations, respectively. Algorithm~\ref{alg:mf_scalar} involves
-both the Jacobian in the referential and the spatial frame, whereas the other
+both the Jacobian in the referential and the spatial configuration, whereas the other
 two algorithms only involve the spatial (Eulerian) terms. Finally, the
@@
 on the mesh via a run-time decision in the code.
+}
 
@@
 The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling}, was implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
-This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aiming to support exascale computing.
+This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aims to support exascale computing.
 In this work we employ \texttt{deal.II}'s interface to the \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
@@
 various pre- and post-processing steps also leveraged parallelization.
-For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently usedg.
+For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used.
 
@@
 In this work we avoid these complications by assuming
-% In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
-% However, for a patch of elements consisting of the same material, the transfer operators reduce to the standard geometric transfer.
-% Based on this observation we adopt the following approach:
 that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material.
@@
 Finally, we note that a $p-$ version of multigrid \cite{Ronquist1987} would also be conceivable but is not considered here.
-%
-% For examples of other approaches that combine homogenization and multigrid methods, see \cite{Bayreuther2003,Fish1995} for applications using small strain elastic materials and \cite{Kaczmarczyk2010} with applications in fracture.
 
@@
 %
-% \begin{figure}[!ht]
-%   \centering
-%   \begin{subfigure}[b]{0.8\textwidth}
-%     \centering
-%     \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
-%   \end{subfigure}
-%   \caption{2D multigrid mesh after two global refinements distributed into four MPI processes. The color indicates the process owning the element.}%
-%   \label{fig:miehe_gmg}
-% \end{figure}
-% %
-% %
-% The same distributed approach to mesh decomposition (as most evident on the finest level) is adopted on the multigrid levels.
-% Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements for the mesh introduced in figure \ref{fig:miehe_fine_level}.
-%
 \begin{table}[!htb]
@@
 we limit ourselves to comparisons made on a node level in this study.
+{\color{red}
 Examinations performed in this manner are most representative for the differences between the matrix-free and matrix-based approaches because problems of sufficient size are dominated by the local node-level performance with multigrid solvers.
 Multigrid solvers with small enough coarse-grid problems are well-known to be scalable even to the largest supercomputers \cite{gholami2016,ibeid2018} as the communication is mostly between nearest neighbors.
+}
 The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
@@
 
+{\color{red}
 Computations were performed on two systems: A dual-socket Intel Xeon Gold 6230 (``Cascade Lake''), released in 2019, with 20 cores per socket and 96 GB of DDR4-2933 memory (nominal bandwidth per socket: 141 GB/s).
@@
 The GNU compiler version 9.1.0 with flags ``-O3 -march=skylake-avx512' and OpenMPI version 4.0.1 were used, with the code run in pure MPI mode.
-%A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM.
-Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
+}
 The ``IWR'' results were obtained on a single machine with eight Intel Xeon E7-8870 ``Westmere'' chips, released in 2011, (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip.
@@
 as the matrix becomes more and more dense.
+{\color{red}
 In order to increase visibility, the figures also report the throughput of the matrix vector product, computed by the ratio of the number of degrees of freedom over the time for a matrix-vector product, in panels (a) and (d).
+}
 The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
+{\color{red}
 Note that we consider only deformed meshes together with separate metric terms according to Table~\ref{tab:memory}.
@@
 the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration.
+}
 
@@
 
-The matrix-free implementation is faster than the matrix-based counterpart already for a bi-quadractic Lagrangian basis in 2D and a tri-quadratic in 3D on both systems on both systems The ``MF tensor2'' and ``MF scalar'' strategies outperform the sparse matrix already for tri-linear elements of polynomial degree $p=1$.
+The matrix-free implementation is faster than the matrix-based counterpart already for a bi-quadractic Lagrangian basis in 2D and a tri-quadratic in 3D on both systems.
+{\color{red}
+The ``MF tensor2'' and ``MF scalar'' strategies outperform the sparse matrix already for tri-linear elements of polynomial degree $p=1$.
 For example for the ``tensor2'' matrix-free implementation with $p=4$, we record a 5$\times$ higher throughput than the matrix-based evaluation of $p=1$.
 The advantage of the matrix-free implementation can be explained by the memory access per unknown shown in the panels (c) and (f) of the figures, as the time per unknown in panels (b) and (e) closely follows the memory access.
+}
 Figures \ref{fig:benchmark_miehe_CSL_vmult2}, \ref{fig:benchmark_miehe_CSL_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
@@
 Two conclusions can be drawn from these results:
+{\color{red}
 Firstly, the algorithm where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'') is the fastest approach in both 2D and 3D and on both systems.
-When compared to the scalar caching implementation, the ``tensor2'' variant loads slightly less data, because only the spatial configuration is needed in terms of metric terms, as compared to both the spatial and the referential configuration in the scalar caching cache in order to evaluate the Kirchhoff stress $\gz \tau$.
+When compared to the scalar caching implementation, the ``tensor2'' variant loads slightly less data, because only the spatial configuration is needed in terms of metric terms, as compared to both the spatial and the referential configuration in the scalar caching case in order to evaluate the Kirchhoff stress $\gz \tau$.
 Besides the memory access, the scalar caching strategy also involves considerably more computations at quadrature points.
@@
 On the IWR machine, the Flop/Byte ratio is 2.9, which means that much fewer computations can be sustained for the same memory transfer.
-
+}
+
+{\color{red}
 Secondly, the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') is slower than the ``tensor2'' approach, despite fewer arithmetic operations, on both systems.
@@
 This means we can apply this matrix-free variant to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
+}
 It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
@@
 Figures \ref{fig:benchmark_miehe_CSL_memory2}, \ref{fig:benchmark_miehe_CSL_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point (``tensor4''), the matrix-free approach takes much less memory than its matrix-based counterpart.
+{\color{red}
 In those graphs, we report the overall memory needed to cache additional quantities required for the tangent operator according to Table~\ref{tab:memory}, as well as all metadata required by
 the \texttt{deal.II} \textit{MatrixFree} object.
+}
 
@@
 The results of the roofline analysis in Figure~\ref{fig:roofline} show that the matrix-based approach is fully memory-bandwidth bound with a very low arithmetic intensity of 0.16 and a performance of $\pgfmathprintnumber{17.36}\,\rm{GFlop/s}$ (1.4\% of the arithmetic peak) averaged over the 2D results.
+{\color{red}
 The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
@@
 The achieved performance is about 6.8\% of the peak arithmetic performance. This result similar to the deformed mesh results of \cite[Figure 18]{kronbichler2017fast}.
-
+}
+
+{\color{red}
 The roofline model also explains the difference in performance between the ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) and the ``tensor2'' caching strategy (Algorithm \ref{alg:mf_tensor2}): Both algorithms essentially run at the memory limit of around $100\, \text{GB/s}$, but the former involves a considerably higher memory transfer according to Table~\ref{tab:memory}.
@@
 Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations as listed in Table~\ref{tab:memory}, no advantage of the scalar caching can be deduced despite the higher arithmetic performance.
+}
 
@@
   \centering
-  %\begin{subfigure}[b]{0.49\textwidth}
-  %    \centering
-  %    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_2d.pdf}
-  %    \caption{Roofline (2D)}
-  %    \label{fig:roofline_2d_tensor4}
-  %\end{subfigure}
-  %\begin{subfigure}[b]{0.49\textwidth}
-  %  \centering
-  %  \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_breakdown_roofline_3d.pdf}
-  %  \caption{Roofline (3D)}
-  %  \label{fig:roofline_3d_tensor4}
-  %\end{subfigure}
-  %~
   \begin{subfigure}[b]{0.49\textwidth}
@@
 we exemplarily collect a breakdown of the operator evaluation into various stages inside the element loop of Algorithm \ref{alg:mf_tensor4}:
-(i) `read / write' denotes reading and writing from/to a global vector (lines 2 and 13) with indirect addressing of gather/scatter type;
-(ii) `sum factorization' denotes application of sum factorization techniques (lines 4, 5 and 12);
-(iii) `quadrature loop' denotes operations performed at each quadrature (lines \mbox{6 -- 11}).
+(i) `read / write' denotes reading and writing from/to a global vector (lines 4 and 14) with indirect addressing of gather/scatter type;
+(ii) `sum factorization' denotes application of sum factorization techniques (lines 6 and 13);
+(iii) `quadrature loop' denotes operations performed at each quadrature (lines \mbox{7 -- 12});
+(iv) 'zero vector' denotes setting destination vector to zero (line 1);
+(v) 'MPI' denotes MPI communication (lines 2 and 16).
+{\color{red}
 Given that the measurements are within the cell loop and there is an extensive overlap between the memory transfer (due to hardware prefetching) and arithmetic operations, the data of Figure~\ref{fig:breakdown} is derived from measuring the cost of `sum factorization' by comparing the timings of part (i) and (ii) together against part (i) only.
@@
 Note also that higher degrees derive a higher throughput in terms of DoF/s in Figures~\ref{fig:benchmark_miehe_CSL} and~\ref{fig:benchmark_miehe_IWR}, as the memory access of quadrature point data scales as $(p+1)^d/p^d$ compared to the number of unknowns.
+}
 
@@
 To summarize this section, we can attribute the speed-up of the matrix-free finite-strain tangent operators for higher-order elements to two factors.
+{\color{red}
 The first one is the higher algorithmic intensity and lower memory access as compared to the matrix-based implementation, see Figure \ref{fig:roofline}.
 Since both variants are in the memory-limited regime, the matrix-free implementation with its lower memory access is beneficial.
+}
 The second benefit is the reduction in the number of floating point operations per DoF for higher order elements thanks to the sum factorization technique, as discussed in
@@
 In particular, we can improve the number of iterations by choosing a more accurate coarse level solver.
-%However, the settings used in this study are optimized for the considered problem in terms of the wall-time spent in the linear solver.
 The black-box AMG preconditioner
@@
 the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the spatial configuration is faster than caching only a scalar at quadrature points or caching the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress.
+{\color{red}
 Given that we did not utilize the specific form of the material part of the stiffness tensor in the latter case, we conclude
 that the matrix-free implementation of the tangent operator is somewhat slower but more flexible because it can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
+}
 
 The roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
+{\color{red}
 Despite the higher arithmetic intensity, the matrix-free implementations are still memory bandwidth-limited and in particular by the data at quadrature points in terms of fourth-order and second-order symmetric tensors.
+}
 

### 2019-08-06 doc/large_strain.tex
move dPdF into appendix

@@
 
-{\color{red}
-For completeness we mention that the fully referential linearization of the here considered constitutive equation \eqref{eq:neo_hookean_energy} results in the following equivalent bilinear form
-}
-\begin{align}
-  A_{ij} \equiv a(\gz N_i, \gz N_j) =&
-  \int_{\mcl B_0}
-  \grad \gz N_i : D_{F}P : \grad \gz N_j
-  \d V
-  \label{eq:algebraic_tangent_dPdF}
-  \\
-  D_{F}P :=&
-  2 \lambda
-  \gz F^{-T} \otimes \gz F^{-T}
-  +
-  \left[\mu - 2 \lambda \ln\left( J \right) \right]
-  \gz F^{-T} \underline{\otimes} \gz F^{-1}
-  +
-  \mu \gz I \overline{\otimes} \gz I
-\end{align}
-where the upper and lower dyadic products of pairs of second-order tensors are respectively given by
-\begin{align*}
-    \gz A \overline{\otimes} \gz B &= A_{ik} B_{jl} \gz e_i \otimes \gz e_j \otimes \gz e_k \otimes \gz e_l \\
-    \gz A \underline{\otimes} \gz B &= A_{il} B_{jk} \gz e_i \otimes \gz e_j \otimes \gz e_k \otimes \gz e_l  \, .
-\end{align*}
-
 \section{Matrix-free operator evaluation}
@@
 
+\appendix
+
+{\color{red}
+It is a common practice to write bilinear form in fully referential configuration.
+For the here considered constitutive equation \eqref{eq:neo_hookean_energy} this results in the following bilinear form
+}
+\begin{align}
+  A_{ij} \equiv a(\gz N_i, \gz N_j) =&
+  \int_{\mcl B_0}
+  \grad \gz N_i : {\rm D}_{\gz F}{\gz P} : \grad \gz N_j
+  \d V
+  \label{eq:algebraic_tangent_dPdF}
+  \\
+  {\rm D}_{\gz F}{\gz P} :=&
+  2 \lambda
+  \gz F^{-T} \otimes \gz F^{-T}
+  +
+  \left[\mu - 2 \lambda \ln\left( J \right) \right]
+  \gz F^{-T} \underline{\otimes} \gz F^{-1}
+  +
+  \mu \gz I \overline{\otimes} \gz I
+\end{align}
+where the upper and lower dyadic products of pairs of second-order tensors are respectively given by
+\begin{align*}
+    \gz A \overline{\otimes} \gz B &= A_{ik} B_{jl} \gz e_i \otimes \gz e_j \otimes \gz e_k \otimes \gz e_l \\
+    \gz A \underline{\otimes} \gz B &= A_{il} B_{jk} \gz e_i \otimes \gz e_j \otimes \gz e_k \otimes \gz e_l  \, .
+\end{align*}
+
+{\color{red}
+This formulation can also be straight forwardly adopted within the matrix-free framework of \texttt{deal.II} library.
+To that end a variation of the Algorithm \ref{alg:mf_tensor4} can be used where the gradients are evaluated with respect to the undeformed configuration and only a fourth-order referential tangent tensor ${\rm D}_{\gz F}{\gz P}$ is cached per quadrature point.
+}
+
+{\color{red}
+In the interest of brevity, we did not report any results for this approach,
+however we do support it in our code.
+Below give a brief summary of the measured performance.
+Clearly in this case we need to cache more data as compared to the Algorithm \ref{alg:mf_tensor4}, this results in both higher memory requirement per DoF as well as lower arithmetic intensity of the vmult operator.
+On both architectures we recorded a slightly larger wall time per DoF as compared to the Algorithm \ref{alg:mf_tensor4}, however still much faster than the matrix-based approach.
+From the LIKWID measurements we observed that this algorithm also results in lower performance on Cascade Lake.
+The advantage of this algorithm is that it does not require evaluation of gradients in deformed configuration, therefore the matrix-free setup time is slightly lower as compared to the other three algorithms that rely on \texttt{MappingQEulerian} class of the \texttt{deal.II} library.
+Overall we consider this approach to be a valid alternative to the Algorithm \ref{alg:mf_tensor4}
+that can be be applied to any constitutive model which operates with the fourth-order referential tangent stiffness tensor on the quadrature point level.
+}
+
 \ifijnme

### 2019-08-07 doc/large_strain.tex
Update tables

@@
 \begin{table}
+  \caption{Wall-clock time in seconds and performance in GFlops of Algorithms \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4} in 2D for various combinations of polynomial degrees,
+  vectorization and parallelization on Cascade Lake.}
+  \label{tab:numbers_2d}
   \centering
@@
   \hline
-                & \multicolumn{2}{c|}{serial} & \multicolumn{3}{c|}{MPI} & \multicolumn{3}{c|}{SIMD} & \multicolumn{3}{c|}{MPI+SIMD}  \\
+                & \multicolumn{2}{c|}{auto-vect serial} & \multicolumn{3}{c|}{auto-vect MPI 20 cores} & \multicolumn{3}{c|}{SIMD} & \multicolumn{3}{c|}{SIMD+MPI 20 cores}  \\
+                & \multicolumn{2}{c|}{3.8 GHz} & \multicolumn{3}{c|}{2.4 GHz} & \multicolumn{3}{c|}{3.7 GHz} & \multicolumn{3}{c|}{2.0 GHz} \\
   \hline
@@
   \hline
-  2& \pgfmathprintnumber{0.71685} & \pgfmathprintnumber{0.8747712} & \pgfmathprintnumber{0.07427} & \pgfmathprintnumber{8.57318} & \pgfmathprintnumber{9.65194560388} & \pgfmathprintnumber{0.3939} & \pgfmathprintnumber{1.6528706} & \pgfmathprintnumber{1.81987814166} & \pgfmathprintnumber{0.04142} & \pgfmathprintnumber{16.0073657} & \pgfmathprintnumber{17.306856591} \\
-  4& \pgfmathprintnumber{0.47603} & \pgfmathprintnumber{1.3088965} & \pgfmathprintnumber{0.0511} & \pgfmathprintnumber{12.3860771} & \pgfmathprintnumber{9.3156555773} & \pgfmathprintnumber{0.25747} & \pgfmathprintnumber{2.3690572} & \pgfmathprintnumber{1.84887559716} & \pgfmathprintnumber{0.02786} & \pgfmathprintnumber{22.2420699} & \pgfmathprintnumber{17.0865039483} \\
-  6& \pgfmathprintnumber{0.24683} & \pgfmathprintnumber{1.5497601} & \pgfmathprintnumber{0.02636} & \pgfmathprintnumber{14.8125524} & \pgfmathprintnumber{9.36380880121} & \pgfmathprintnumber{0.13089} & \pgfmathprintnumber{2.8544567} & \pgfmathprintnumber{1.88578195431} & \pgfmathprintnumber{0.01487} & \pgfmathprintnumber{26.0509717} & \pgfmathprintnumber{16.5991930061} \\
+& \multicolumn{11}{c|}{Algorithm~\ref{alg:mf_tensor2}}\\
+\hline
+2& \pgfmathprintnumber{0.67518} & \pgfmathprintnumber{2.1798691} & \pgfmathprintnumber{0.04811} & \pgfmathprintnumber{31.270185} & \pgfmathprintnumber{14.0340885471} & \pgfmathprintnumber{0.21837} & \pgfmathprintnumber{6.6211982} & \pgfmathprintnumber{3.09190822915} & \pgfmathprintnumber{0.02194} & \pgfmathprintnumber{67.8161088} & \pgfmathprintnumber{30.773928897} \\
+4& \pgfmathprintnumber{0.35754} & \pgfmathprintnumber{3.6114434} & \pgfmathprintnumber{0.02991} & \pgfmathprintnumber{43.3813195} & \pgfmathprintnumber{11.9538615848} & \pgfmathprintnumber{0.11568} & \pgfmathprintnumber{11.0068739} & \pgfmathprintnumber{3.09076763485} & \pgfmathprintnumber{0.01438} & \pgfmathprintnumber{89.247277} & \pgfmathprintnumber{24.8636995828} \\
+6& \pgfmathprintnumber{0.17724} & \pgfmathprintnumber{4.4013124} & \pgfmathprintnumber{0.01489} & \pgfmathprintnumber{52.73939} & \pgfmathprintnumber{11.9032907992} & \pgfmathprintnumber{0.05934} & \pgfmathprintnumber{12.9981312} & \pgfmathprintnumber{2.9868554095} & \pgfmathprintnumber{0.00732} & \pgfmathprintnumber{106.9726898} & \pgfmathprintnumber{24.2131147541} \\
+\hline
+& \multicolumn{11}{c|}{Algorithm~\ref{alg:mf_tensor4}}\\
+\hline
+2& \pgfmathprintnumber{0.62536} & \pgfmathprintnumber{2.623296} & \pgfmathprintnumber{0.04582} & \pgfmathprintnumber{36.1519843} & \pgfmathprintnumber{13.6481885639} & \pgfmathprintnumber{0.22937} & \pgfmathprintnumber{6.925918} & \pgfmathprintnumber{2.72642455421} & \pgfmathprintnumber{0.02532} & \pgfmathprintnumber{63.6564648} & \pgfmathprintnumber{24.6982622433} \\
+4& \pgfmathprintnumber{0.36549} & \pgfmathprintnumber{3.8534134} & \pgfmathprintnumber{0.03113} & \pgfmathprintnumber{45.7061894} & \pgfmathprintnumber{11.7407645358} & \pgfmathprintnumber{0.14346} & \pgfmathprintnumber{9.5663639} & \pgfmathprintnumber{2.54767879548} & \pgfmathprintnumber{0.01862} & \pgfmathprintnumber{74.2778413} & \pgfmathprintnumber{19.6288936627} \\
+6& \pgfmathprintnumber{0.18374} & \pgfmathprintnumber{4.5579733} & \pgfmathprintnumber{0.01572} & \pgfmathprintnumber{54.0989777} & \pgfmathprintnumber{11.6882951654} & \pgfmathprintnumber{0.07388} & \pgfmathprintnumber{11.097412} & \pgfmathprintnumber{2.4870059556} & \pgfmathprintnumber{0.00946} & \pgfmathprintnumber{88.2340296} & \pgfmathprintnumber{19.422832981} \\
   \hline
@@
   }
-  \caption{Wall-clock time in seconds and performance in GFlops of Algorithm \ref{alg:mf_tensor4} in 2D for various combinations of polynomial degrees,
-  vectorization and parallelization.}
-  \label{tab:numbers_2d}
 \end{table}
@@
 \begin{table}
+  \caption{Wall-clock time in seconds and performance in GFlops of Algorithms~\ref{alg:mf_tensor2} and \ref{alg:mf_tensor4} in 3D for various combinations of polynomial degrees,
+  vectorization and parallelization on Cascade Lake.}
   \centering
@@
   \hline
-  2& \pgfmathprintnumber{1.82927} & \pgfmathprintnumber{1.4408182} & \pgfmathprintnumber{0.19184} & \pgfmathprintnumber{13.75528} & \pgfmathprintnumber{9.5353940784} & \pgfmathprintnumber{1.24738} & \pgfmathprintnumber{2.2390063} & \pgfmathprintnumber{1.46648976254} & \pgfmathprintnumber{0.13397} & \pgfmathprintnumber{21.0187594} & \pgfmathprintnumber{13.6543255953} \\
-  4& \pgfmathprintnumber{1.14988} & \pgfmathprintnumber{1.7483867} & \pgfmathprintnumber{0.1211} & \pgfmathprintnumber{16.6162812} & \pgfmathprintnumber{9.49529314616} & \pgfmathprintnumber{0.74974} & \pgfmathprintnumber{2.7527409} & \pgfmathprintnumber{1.53370501774} & \pgfmathprintnumber{0.08166} & \pgfmathprintnumber{25.4865287} & \pgfmathprintnumber{14.0813127602} \\
+& \multicolumn{11}{c|}{Algorithm~\ref{alg:mf_tensor2}}\\
+\hline
+2& \pgfmathprintnumber{0.33065} & \pgfmathprintnumber{4.341083} & \pgfmathprintnumber{0.02814} & \pgfmathprintnumber{51.7484607} & \pgfmathprintnumber{11.750177683} & \pgfmathprintnumber{0.12258} & \pgfmathprintnumber{11.7249488} & \pgfmathprintnumber{2.6974220917} & \pgfmathprintnumber{0.01317} & \pgfmathprintnumber{112.4745656} & \pgfmathprintnumber{25.106302202} \\
+4& \pgfmathprintnumber{0.18685} & \pgfmathprintnumber{5.461242} & \pgfmathprintnumber{0.01589} & \pgfmathprintnumber{64.7662171} & \pgfmathprintnumber{11.7589679043} & \pgfmathprintnumber{0.0695} & \pgfmathprintnumber{14.7126397} & \pgfmathprintnumber{2.68848920863} & \pgfmathprintnumber{0.00803} & \pgfmathprintnumber{129.6590805} & \pgfmathprintnumber{23.2689912827} \\
+  \hline
+& \multicolumn{11}{c|}{Algorithm~\ref{alg:mf_tensor4}}\\
+  \hline
+2& \pgfmathprintnumber{0.38409} & \pgfmathprintnumber{4.4971589} & \pgfmathprintnumber{0.03472} & \pgfmathprintnumber{51.9008095} & \pgfmathprintnumber{11.0625} & \pgfmathprintnumber{0.22335} & \pgfmathprintnumber{7.7420018} & \pgfmathprintnumber{1.719677636} & \pgfmathprintnumber{0.0243} & \pgfmathprintnumber{72.555647} & \pgfmathprintnumber{15.8061728395} \\
+4& \pgfmathprintnumber{0.22053} & \pgfmathprintnumber{5.393482} & \pgfmathprintnumber{0.01898} & \pgfmathprintnumber{63.1076193} & \pgfmathprintnumber{11.6190727081} & \pgfmathprintnumber{0.12765} & \pgfmathprintnumber{9.3340763} & \pgfmathprintnumber{1.72761457109} & \pgfmathprintnumber{0.01464} & \pgfmathprintnumber{82.3893605} & \pgfmathprintnumber{15.0635245902} \\
 \hline
@@
   }
-  \caption{Wall-clock time in seconds and performance in GFlops of Algorithm \ref{alg:mf_tensor4} in 3D for various combinations of polynomial degrees,
-  vectorization and parallelization.}
   \label{tab:numbers_3d}
@@
 
-Table \ref{tab:numbers_2d} and Table \ref{tab:numbers_3d} show the speed-up obtained due to explicit SIMD vectorization intrinsics and due to MPI.
-For these rather small count of MPI processes, the results confirm that the performance gained is independent.
-Furthermore, the improvement due to explicit intrinsics for the data used in the quadrature loop, is comparable to the ones in \cite[Figure 6]{kronbichler2017fastcomputer} where the speedup factor is $1.2-1.8$ for a similar architecture.
-\todo[inline]{Do we want to update Tables \ref{tab:numbers_2d} and \ref{tab:numbers_3d}?}
+Table \ref{tab:numbers_2d} and Table \ref{tab:numbers_3d} show the speed-up obtained due to explicit SIMD vectorization intrinsics and due to the MPI parallelization.
+{\color{red}
+When run in serial, explicit SIMD vectorization via instrinsics provides a speedup of up to $3\times$ over the auto-vectorization case.
+This is less than the theoretical factor of 8, but in line with previous experiments presented in \cite[Figure 6]{kronbichler2017fastcomputer} with a speedup of around $1.5-2.5\times$ with 8-wide vectorization.
+The reason for sub-optimal speedup is mostly related to the memory access in the gather/scatter operations as well as the limited memory bandwidth.
+When going to the parallel case with 20 cores, the less-than-optimal speedup can be attributed to two different effect.
+For the auto-vectorized case, the speedup is limited due to different turbo clock frequencies, which permit a speedup of a factor 12.6.\footnote{Note that we intentionally run with Turbo mode on to permit the processor to come close to the power limit for all settings and present more realistic limits.} For the SIMD+MPI case, the main reason is the memory bandwidth: As can be seen from Figures~\ref{fig:roofline_2d} and~\ref{fig:roofline_3d}, both algorithms are eventually memory-bandwidth limited.
+Since Algorithm~\ref{alg:mf_tensor4} has a lower arithmetic intensity, the speedup over the serial, auto-vectorized case is less than that of Algorithm~\ref{alg:mf_tensor2}.
+}
 
@@
 However, we have performed a preliminary study of the weak scalability of the implementation.
-Table \ref{tab:weak_3d} demonstrates a good weak scaling up to 1280 processes\footnote{%
-This corresponds to 64 nodes, which is the maximum we could request on the RRZE Emmy HPC cluster.
-} of the wall time per matrix-vector product and per iteration in the CG solver for the 3D problem with a tri-quadratic basis.
+Tables~\ref{tab:weak_2d} \ref{tab:weak_3d} demonstrate a good weak scaling up to 320 cores, i.e., 8 nodes of dual-socket Cascade Lake, of the wall time per matrix-vector product and per iteration in the CG solver for the 2D problem with $p=4$ and the 3D problem with a tri-quadratic basis.
 This confirms that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for cluster-sized problems as well,
 also compare \cite[Figure 7]{Krank2017} for scaling results for up to 147,456 CPU cores and 34.4 billion degrees of freedom for an incompressible flow problem.
-\todo[inline]{Do we want to update?
-I can run on up to 8 nodes with CSL.
-This allows me to do 3 steps of weak scaling in 2D, 1 socket, 2 nodes, 8 nodes, and two steps in 3D, 1 node and 8 nodes.
-Not very large but enough to show our point?}
 
@@
   \hline
-  20 & 4,442,880 & 3 & \pgfmathprintnumber{0.0675} & \pgfmathprintnumber{0.529553679131} \\
-  160 & 35,071,488 & 4 & \pgfmathprintnumber{0.0705} & \pgfmathprintnumber{0.562637362637} \\
-  1280 & 278,694,912 & 5 & \pgfmathprintnumber{0.083} & \pgfmathprintnumber{0.688792165397} \\
+  20 & 2,887,680 & 5 & \pgfmathprintnumber{0.0033900} & \pgfmathprintnumber{0.045574} \\
+  80 & 11,542,528 & 6 & \pgfmathprintnumber{0.0035350} & \pgfmathprintnumber{0.047788} \\
+  320 & 46,153,728 & 7 & \pgfmathprintnumber{0.0037879} & \pgfmathprintnumber{0.052012} \\
   \hline
   \end{tabular}
-  \caption{Weak scaling of Algorithm \ref{alg:mf_tensor4} in 3D for quadratic polynomial basis.}
+  \caption{Weak scaling of Algorithm \ref{alg:mf_tensor2} in 2D for a quartic polynomial basis.}
+  \label{tab:weak_2d}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|r|r|c|r|c|}
+  \hline
+  cores  & $N_{DoF}$ & $N_{gref}$ & vmult [s] & CG [s / iteration] \\
+  \hline
+  40 & 4,442,880 & 3 & \pgfmathprintnumber{0.0070667} & \pgfmathprintnumber{0.064328} \\
+  320 & 35,071,488 & 4 & \pgfmathprintnumber{0.0075333} & \pgfmathprintnumber{0.071448} \\
+  \hline
+  \end{tabular}
+  \caption{Weak scaling of Algorithm \ref{alg:mf_tensor2} in 3D for a quadratic polynomial basis.}
   \label{tab:weak_3d}

### 2019-08-07 Calculations/post_process_likwid_simd.py

@@ -16,7 +16,7 @@ parser.add_argument('--dim', metavar='dim', default=2, type=int,
 args = parser.parse_args()
 
 # expected blockspeed:
-clock_speed = 2.2  # GHz
+clock_speed = 2.0  # GHz
 
 # serial | MPI | SIMD | MPI+SIMD
 suffixes = [
@@ -59,23 +59,27 @@ for idx, s in enumerate(suffixes):
         # we are only interested in tensor2 (Algorithm 3) for the currently chosen dimension
         if not (dim == args.dim and '_tensor4' in fname):
             continue
+        if ('_tensor4_ns' in fname):
+            continue
 
         print 'dim={0} p={1} q={2} file={3}'.format(dim,p,q,fname)
 
         # these files contain a single region
         result = parse_likwid_file(f,'LIKWID_MARKER_CLOSE')['vmult_MF']
 
+        print result
+
         # depending on the run, the LIKWID output may not have Sum table
         if 'Metric_Sum' not in result:
             data = result['Metric']
-            flops = data['MFLOP/s'][0]
+            flops = data['DP MFLOP/s'][0]
             runtime = data['Runtime (RDTSC) [s]'][0]
-            assert abs(float(data['Clock [MHz]'][0]) - clock_speed * 1000) < clock_speed
+            #assert abs(float(data['Clock [MHz]'][0]) - clock_speed * 1000) < clock_speed
         else:
             data = result['Metric_Sum']
-            flops = data['MFLOP/s STAT'][0]                # take Sum
+            flops = data['DP MFLOP/s STAT'][0]                # take Sum
             runtime = data['Runtime (RDTSC) [s] STAT'][2]  # take Max
-            assert abs(float(data['Clock [MHz] STAT'][2]) - clock_speed * 1000) < clock_speed
+            #assert abs(float(data['Clock [MHz] STAT'][2]) - clock_speed * 1000) < clock_speed
 
         runtime = float(runtime) / 10  # FIXME: hard-code that we run 10 times
         flops = float(flops) / 1000  # MFLOP/s -> GFLOP/s
@@ -125,7 +129,7 @@ p             & time  & GFlop/s              & time & GFlop/s & speedup & time &
 \hline
 \end{tabular}
 }
-\caption{Wall-clock time in seconds and performance in GFlops of Algorithm \\ref{alg:mf_tensor4} in 2D for various combinations of polynomial degrees,
[…]

### 2019-08-07 doc/large_strain.tex
Update intro

@@
 Particular emphasis is on the tradeoff between computations at quadrature points versus
-pre-computation with a higher memory access.
+pre-computation with a higher memory access in terms of the representation of the linearization point.
 The \texttt{deal.II} \cite{dealII90} finite element library, which provides MPI parallelization together with a generalized matrix-free framework that incorporates SIMD vectorization and an interface to high-performance MPI-distributed matrix-based libraries is used for this study.

### 2019-08-07 include/mf_nh_operator.h
add Martin's explanation as incode comments

@@ -506,6 +506,16 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::cache()
               else if (mf_caching == MFCaching::scalar_referential)
                 {
                   cached_scalar(cell, q) = scalar;
+                  // MK:
+                  // This is to avoid the phi_reference.read_dof_values() call and the full
+                  // phi_reference.evaluate(false, true) calls. With this quadrature point information,
+                  // I only need to call a cheap "collocation gradient" function, which is likely the
+                  // best compromise in terms of caching some data versus computing: read_dof_values is
+                  // expensive because it is not fully vectorized (indirect addressing gather access)
+                  // and once you start to store things element-by-element you can eliminate the
+                  // interpolation from nodes to quadrature points (that happens in the usual matrix-free
+                  // interpolations as well). So my code makes use of some internal workings of the matrix-free
+                  // framework that one would need to clean up in case one really wanted to make user-friendly programs
                   for (unsigned int d=0; d<dim; ++d)
                     cached_second_scalar(cell, q + d * phi_reference.n_q_points)
                       = phi_reference.begin_values()[q + d * phi_reference.n_q_points];
@@ -1210,6 +1220,15 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::do_operation_on_cell(
         }
       else if (cell_mat->formulation == 1 && mf_caching == MFCaching::scalar_referential)
         // the least amount of cache and the most calculations
+        // MK:
+        // What I implemented here is essentially the same as the usual scalar variant.
+        // The only thing I had to change was to replace grad x (and submit_gradient() in the spatial frame)
+        // by Grad x (get_gradient() in referential frame) and then multiplying by F^{-T}.
+        // And this is the F that gets computed in the scalar case as well,
+        // I just hardcoded it as this was the way I thought about it in the implementation,
+        // but one could of course use the phi_reference.get_gradient() function.
+        // And similarly by F^{-1} for submit_gradient(). In other words,
+        // I simply pulled out the Eulerian motion of the grid into F; nothing else changed.
         {
           const NumberType *cached_position =
             &cached_second_scalar(cell, 0);

### 2019-08-07 include/mf_nh_operator.h
fix compilation with slightly older version of dealii

@@ -772,7 +772,6 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::local_diagonal_cell(
   const unsigned int &,
   const std::pair<unsigned int, unsigned int> &cell_range) const
 {
-  // FIXME: I don't use data input, can this be bad?
   const VectorizedArray<number> one  = make_vectorized_array<number>(1.);
   const VectorizedArray<number> zero = make_vectorized_array<number>(0.);
 
@@ -1230,6 +1229,7 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::do_operation_on_cell(
         // And similarly by F^{-1} for submit_gradient(). In other words,
         // I simply pulled out the Eulerian motion of the grid into F; nothing else changed.
         {
+          const NumberType one = make_vectorized_array<number>(1.);
           const NumberType *cached_position =
             &cached_second_scalar(cell, 0);
           constexpr unsigned int n_q_points = Utilities::pow(n_q_points_1d, dim);
@@ -1266,7 +1266,7 @@ NeoHookOperator<dim, fe_degree, n_q_points_1d, number>::do_operation_on_cell(
                         sum2 += inv_jac[e][f] * x_grads[(d * dim + f)*n_q_points + q];
                       grad_Nx_v[d][e] = sum2;
                     }
-                  F[d][d] += NumberType(1.0);
+                  F[d][d] += one;
                 }
               const SymmetricTensor<2, dim, NumberType> b =
                 Physics::Elasticity::Kinematics::b(F);

### 2019-08-08 doc/large_strain.tex
Go through paper and improve text

@@
   \begin{abstract}[Summary]
-    {{\color{red}This work investigates matrix-free algorithms for problems in quasi-static finite-strain hyperelasticity.
-Matrix-free algorithms have emerged as an attractive alternative to iterative solvers with sparse matrices in the fluid dynamics and wave propagation context because they significantly reduce the memory traffic, the limiting factor in classical finite element solvers.
+    {{\color{red}This work investigates matrix-free algorithms for problems in quasi-static finite-strain hyperelasticity. Iterative solvers with matrix-free operator evaluation have emerged as 
+an attractive alternative to sparse matrices in the fluid dynamics and wave 
+propagation communities because they significantly reduce the memory traffic, 
+the limiting factor in classical finite element solvers.
     Specifically, we study different matrix-free realizations of the finite element tangent operator, and determine whether generalized methods of incorporating complex constitutive behavior might be feasible.}
@@
     The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner on a single node for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
-    We conclude that matrix-free methods for finite-strain solid mechanics are very promising, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
+    We {\color{red}find} that matrix-free methods for finite-strain solid mechanics are very promising, {\color{red}outperforming linear matrix-based schemes by two to five times}, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
   \end{abstract}
@@
 
-The performance of finite element solvers on modern computer architectures is typically memory bound for sufficiently large problems.
-The main cause for this is that loading pre-computed matrix elements from RAM memory into the CPU cores is significantly slower than actually performing the arithmetic operations.
+The performance of finite element solvers on modern computer architectures is typically {\color{red} limited by the bandwidth from RAM memory} for sufficiently large problems.
+{\color{red}This is because} loading pre-computed matrix elements from memory into the CPU cores is significantly slower than actually performing the arithmetic operations{\color{red}, dictated by hardware properties}.
 In order to improve the performance of iterative solvers, so-called matrix-free methods,
-where matrix-vector products are formed on-the-fly \cite{Brown2010,kronbichler12,May2015, Krank2017,  Gmeiner2016},
+where matrix-vector products are formed on-the-fly rather than relying on precomputed matrix entries \cite{Brown2010,kronbichler12,May2015, Krank2017,  Gmeiner2016},
 are widely adopted in the fluid mechanics community.
@@
 {\color{red}
-Among matrix-free methods, stencil-based variants are most popular for linear elements and tetrahedral shapes in particular~\cite{Gmeiner2016,Bauer2018}, whereas implementations based on integration with sum-factorization techniques are attractive for higher polynomial degrees $p\geq 2$ and tensor product shape functions~\cite{Brown2010,kronbichler12,May2015,Cantwell2011}.
+Among matrix-free methods, stencil-based variants are most popular for linear elements and tetrahedral shapes~\cite{Gmeiner2016,Bauer2018}, whereas implementations based on integration with sum-factorization techniques are attractive for higher polynomial degrees $p\geq 2$ and tensor product shape functions~\cite{Brown2010,kronbichler12,May2015,Cantwell2011}.
 }
@@
 Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
-The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
+The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted,
 whereas higher order elements are rarely employed.
-This can potentially reduce the advantage of matrix-free methods using sum-factorization techniques, that are generally more competitive for higher order elements \cite{kronbichler12,kronbichler2017fast,muthing2017high,Fischer2019}.
+This can potentially reduce the advantage of matrix-free methods using sum-factorization techniques, which are generally more competitive for higher order elements \cite{kronbichler12,kronbichler2017fast,muthing2017high,Fischer2019}.
 However, we note that, even with linear FEM, large scale computations with as many as $10^{12}$ unknowns are possible with a matrix-free approach but infeasible with sparse matrices as there is simply not enough memory to store the sparse tangent matrix even on large supercomputers \cite{Gmeiner2016}.
@@
 {\color{red}
-In this work, we investigate different matrix-free representations of the
+This work aims to fill this gap by investigating different matrix-free representations of the
 finite-strain hyperelastic tangent operator with respect to the computational cost
@@
 pre-computation with a higher memory access in terms of the representation of the linearization point.
-The \texttt{deal.II} \cite{dealII90} finite element library, which provides MPI parallelization together with a generalized matrix-free framework that incorporates SIMD vectorization and an interface to high-performance MPI-distributed matrix-based libraries is used for this study.
+The \texttt{deal.II} finite element library \cite{dealII90}, which provides MPI parallelization together with a generalized matrix-free framework that incorporates SIMD vectorization and an interface to high-performance MPI-distributed matrix-based libraries, is used for this study.
 In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
@@
 while section \ref{sec:fe} covers the finite element discretization.
-In Section \ref{sec:mf} we describe the different matrix-free operator implementations that are evaluated in this study.
-Section \ref{sec:framework} provides details of the numerical framework with which this study is implemented,
+In Section \ref{sec:mf} we describe the different matrix-free implementations that are evaluated in this study.
+Section \ref{sec:framework} provides details of the numerical framework {\color{red}this study relies on},
 where we also propose a method by which to construct level tangent operators for
@@
 The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}.
-Lastly, the results are summarized in Section \ref{sec:summary}.
+Finally, the results are summarized in Section \ref{sec:summary}.
 
@@
 where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called the deformation gradient and $\gz I$ is the second-order unit tensor.
-The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian $J = \det \gz F > 0$ has to be positive.
+The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian {\color{red}determinant} $J = \det \gz F > 0$ has to be positive.
 We shall denote the push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by a linear transformation $\chi$ defined as
@@
   = \frac{\mu}{2} \left[ \trace{\mathbf{C}} - \trace{\mathbf{I}} - 2 \ln\left( J \right) \right]
-  + \lambda \ln^{2}\left( J \right)
+  + \lambda \ln^{2}\left( J \right) ,
 \label{eq:neo_hookean_energy}
@@
 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}}
-  &= \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1} \\
+  &= \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1} ,\\
 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
   &= \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
-  + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
+  + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1} .
 \end{align}
@@
   = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
-  + 2 \lambda \mathbf{I} \otimes \mathbf{I}
+  + 2 \lambda \mathbf{I} \otimes \mathbf{I} ,
 \end{gather}
@@
 We now introduce a FE triangulation $\mathcal{B}^h_0$ of $\mcl B_0$ and
-the associated FE space of continuous piecewise elements with polynomial space of fixed degree. % : $V^h \subset H^1 (\mcl B_0)$.
+the associated FE space of continuous piecewise elements with polynomial space of fixed degree $p$. % : $V^h \subset H^1 (\mcl B_0)$.
 The displacement fields are given in
@@
 \begin{align}
-  \sum_{j \in \mcl I} A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
+  \sum_{j \in \mcl I} A_{ij} \Delta u_j &= - F_i , \label{eq:linear_system} \\
   A_{ij} \equiv a(\gz N_i, \gz N_j) &=
@@
   }_{\displaystyle \tilde{a}(\gz N_i, \gz N_j)}
-  \d V
+  \d V ,
   \label{eq:algebraic_tangent}
@@
 In this case, the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
-On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
+On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the associated arithmetic operations.
 For this reason, recent implementations often focus on so-called matrix-free approaches \cite{kronbichler12, kronbichler2017fast} where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
 The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
-For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
+For iterative solvers, this is possible since it is sufficient to compute matrix-vector products.
 %
@@
 \end{align}
-where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form at one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
+where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form at one quadrature point $\gz \xi_q$, $J^K_q$ is the {\color{red}determinant of the} Jacobian of the mapping from the isoparametric element to the element $K$ and
 $w_q$ is the quadrature weight.
@@
 \end{align*}
-Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
+Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one{\color{red}, $n=p+1$}, the evaluation of all of the shape functions ($n^d$)
 in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
@@
 Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, this results in an algorithm with an arithmetic complexity of $\mathcal{O}(n^{d+1})$.
-Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for the 3D case, and where the discretization and evaluation employs high polynomial degree basis functions with the corresponding quadrature rule.
+Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for the 3D case, and where the discretization and evaluation employs higher degree basis functions with the corresponding quadrature rule.
 More information on matrix-free techniques can be found in \cite{kronbichler12,vos10}.
@@
 %
-In order to obtain gradients with respect to the spatial configuration,
-that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem derived with a Lagrangian formulation,
-we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
+In order to obtain the gradients with respect to the spatial configuration,
+which appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem derived with a Lagrangian formulation,
+we use the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
 In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
@@
 
-In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point.
+In order to avoid {\color{red}the repeated} calculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point.
 In this case, we also avoid the need to evaluate gradients of the displacement field with respect to the referential configuration.
-This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
+This algorithm utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
 Therefore, the action of the material part of the fourth-order spatial tangent stiffness tensor $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
-To that end, we cache two scalars that depend on the Jacobian $J$.
+To that end, we cache two scalars that depend on the Jacobian {\color{red}determinant} $J$.
 
@@
 Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material part of the fourth-order spatial tangent stiffness tensor.
-In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\overline{\grad^s} \gz x$ on-the-fly.
+In this case, we cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\overline{\grad^s} \gz x$ on-the-fly.
 
@@
 is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
-The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements\footnote{%
+The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements{\color{red}, as opposed to the operations within an element}\footnote{%
 In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
@@
 This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
-In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type
+In particular, \texttt{deal.II} provides the generic templated class \textit{VectorizedArray} that defines a unified interface to a vectorized data type
 and overloads basic arithmetic operations based on compiler intrinsics.
@@
 Thanks to the object-oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
-Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
-For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
+Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the locally owned elements.
+For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well as SIMD vectorization) and performance tests, see \cite{kronbichler12}.
 
@@
 The cached quantities listed in the algorithms, labeled ``physics terms'' in the
-table, are increasing in size with more caching, and are separate data for each
+table, are increasing in size with more caching, and are stored for each
 quadrature point. Finally, the evaluation of the gradients in the referential
@@
 
-The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling}, was implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
+The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling}, has been implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
 This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aims to support exascale computing.
@@
 The discrete tangent operator described by \eqref{eq:algebraic_tangent} can be implemented in one of two ways, namely through a classical matrix-based approach or, alternatively, a matrix-free approach.
-For the matrix-based approach adopted in this study, the Trilinos \cite{Heroux2005} library was leveraged;
+For the matrix-based approach adopted in this study, the Trilinos \cite{Heroux2005} library has been leveraged;
 the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} sparse data structure from the Epetra package \cite{Heroux2005b} and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
@@
 
-The second implementation utilized a matrix-free approach in conjunction with a geometric-multigrid (GMG) preconditioner.
+The second implementation utilizes a matrix-free approach in conjunction with a geometric-multigrid (GMG) preconditioner.
 GMG preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes.
@@
 %
-Finally, we note that a $p-$ version of multigrid \cite{Ronquist1987} would also be conceivable but is not considered here.
+Finally, we note that a $p$ version of multigrid \cite{Ronquist1987} would also be conceivable but is not considered here.
 
@@
 
-In this section, we evaluate the proposed matrix-free operator algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner on a benchmark problem from \cite{Miehe2007}.
+In this section, we evaluate the proposed matrix-free algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner on a benchmark problem from \cite{Miehe2007}.
 The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}.
@@
 Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
-Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
+Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ directions for the 2D and the 3D problems, respectively.
 The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \sqrt{2} \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
@@
   \caption{Parameters for the benchmark: $p$ is the polynomial degree,
-  $q$ is the number of quadrature points in 1D, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.}
+  $q$ is the number of quadrature points in 1D, $N_{\text{gref}}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{\text{DoF}}$ is the number of DoFs.}
   \begin{subtable}{.5\linewidth}
+  \caption{2D}
+  \label{tab:input_parameters_2d}
   \centering
@@
   \hline
-    $p$ & $q$ & $N_{gref}$ & $N_{el}$ & $N_{DoF}$ \\
+    $p$ & $q$ & $N_{\text{gref}}$ & $N_{el}$ & $N_{\text{DoF}}$ \\
   \hline
@@
   \end{tabular}
-  \caption{2D}
-  \label{tab:input_parameters_2d}
   \end{subtable}
   \begin{subtable}{.5\linewidth}
+  \caption{3D}
+  \label{tab:input_parameters_3d}
   \centering
@@
   \end{tabular}
-  \caption{3D}
-  \label{tab:input_parameters_3d}
   \end{subtable}
@@
 Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound.
-Furthermore, also the data structures of the matrix-free approaches are too large to fit completely into caches.
+{\color{red}Furthermore, the data structures of the matrix-free approaches are also too large to fit completely into caches.}
 
@@
 The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
-The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fifth degree, provided
+The geometric multigrid algorithm uses a V-cycle with pre- and post-smoothing using Chebyshev polynomials \cite{Varga2009} of fifth degree, provided
 by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
@@
 Computations were performed on two systems: A dual-socket Intel Xeon Gold 6230 (``Cascade Lake''), released in 2019, with 20 cores per socket and 96 GB of DDR4-2933 memory (nominal bandwidth per socket: 141 GB/s).
-Turbo mode is enabled with a power limit of 135 watt per socket, which results in the processor running at 2.8 GHz for scalar code and 2.0 GHz for AVX-512-heavy code in all experiments.
+Turbo mode is enabled with a power limit of 135 watt per socket, which results in the processor running at 2.8 GHz for scalar code and 2.0 GHz for AVX-512-heavy code in the experiments.
 The GNU compiler version 9.1.0 with flags ``-O3 -march=skylake-avx512' and OpenMPI version 4.0.1 were used, with the code run in pure MPI mode.
@@
 We are interested in the wall-clock time (average over 10 runs for each linear solver step) per DoF as a first metric.
-As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
+As expected, the matrix-vector multiplication becomes very expensive for higher {\color{red}polynomial degrees} for sparse matrix-based approaches,
 as the matrix becomes more and more dense.
@@
 
-The matrix-free implementation is faster than the matrix-based counterpart already for a bi-quadractic Lagrangian basis in 2D and a tri-quadratic in 3D on both systems.
+The matrix-free implementation is faster than the matrix-based counterpart already for a bi-quadractic Lagrangian basis in 2D and a tri-quadratic basis in 3D on both systems.
 {\color{red}
@@
 The majority of time is spent within the quadrature loop for all variants, with an increasing proportion for increasing degrees.
-Note also that higher degrees derive a higher throughput in terms of DoF/s in Figures~\ref{fig:benchmark_miehe_CSL} and~\ref{fig:benchmark_miehe_IWR}, as the memory access of quadrature point data scales as $(p+1)^d/p^d$ compared to the number of unknowns.
+Note also that higher degrees achieve a higher throughput in terms of DoF/s in Figures~\ref{fig:benchmark_miehe_CSL} and~\ref{fig:benchmark_miehe_IWR}, as the memory access of quadrature point data scales as $(p+1)^d/p^d$ compared to the number of unknowns.
 }
@@
 When run in serial, explicit SIMD vectorization via instrinsics provides a speedup of up to $3\times$ over the auto-vectorization case.
-This is less than the theoretical factor of 8, but in line with previous experiments presented in \cite[Figure 6]{kronbichler2017fastcomputer} with a speedup of around $1.5-2.5\times$ with 8-wide vectorization.
+This is less than the theoretical factor of 8, but in line with previous experiments presented in Figure 6 of \cite{kronbichler2017fastcomputer} with a speedup of around $1.5-2.5\times$ with 8-wide vectorization.
 The reason for sub-optimal speedup is mostly related to the memory access in the gather/scatter operations as well as the limited memory bandwidth.
-When going to the parallel case with 20 cores, the less-than-optimal speedup can be attributed to two different effect.
-For the auto-vectorized case, the speedup is limited due to different turbo clock frequencies, which permit a speedup of a factor 12.6.\footnote{Note that we intentionally run with Turbo mode on to permit the processor to come close to the power limit for all settings and present more realistic limits.} For the SIMD+MPI case, the main reason is the memory bandwidth: As can be seen from Figures~\ref{fig:roofline_2d} and~\ref{fig:roofline_3d}, both algorithms are eventually memory-bandwidth limited.
+When going to the parallel case with 20 cores, the less-than-optimal speedup can be attributed to two different effects.
+For the auto-vectorized case, the different (turbo) clock frequencies permit a speedup of a factor 12.6 at most, close to our results.\footnote{Note that we intentionally run with Turbo mode on to permit the processor to come close to the power limit for all settings and present more realistic performance limits.} For the SIMD+MPI case, the main reason is the memory bandwidth: As can be seen from Figures~\ref{fig:roofline_2d} and~\ref{fig:roofline_3d}, both algorithms are eventually memory-bandwidth limited.
 Since Algorithm~\ref{alg:mf_tensor4} has a lower arithmetic intensity, the speedup over the serial, auto-vectorized case is less than that of Algorithm~\ref{alg:mf_tensor2}.
@@
 The second benefit is the reduction in the number of floating point operations per DoF for higher order elements thanks to the sum factorization technique, as discussed in
-Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
+Section \ref{sec:mf} and Section 2.4 of \cite{kronbichler12}.
 Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach, whereas Algorithm~\ref{alg:mf_tensor2} is faster due to a higher arithmetic intensity.
@@
 However, we have performed a preliminary study of the weak scalability of the implementation.
-Tables~\ref{tab:weak_2d} \ref{tab:weak_3d} demonstrate a good weak scaling up to 320 cores, i.e., 8 nodes of dual-socket Cascade Lake, of the wall time per matrix-vector product and per iteration in the CG solver for the 2D problem with $p=4$ and the 3D problem with a tri-quadratic basis.
+{\color{red}Tables~\ref{tab:weak_2d} and \ref{tab:weak_3d} demonstrate a good weak scaling up to 320 cores, i.e., 8 nodes of dual-socket Cascade Lake, of the wall time per matrix-vector product and per iteration in the CG solver for the 2D problem with $p=4$ and the 3D problem with a tri-quadratic basis.}
 This confirms that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for cluster-sized problems as well,
@@
   \hline
-  cores  & $N_{DoF}$ & $N_{gref}$ & vmult [s] & CG [s / iteration] \\
+  cores  & $N_{\text{DoF}}$ & $N_{\text{gref}}$ & vmult [s] & CG [s / iteration] \\
   \hline
@@
 Among the three examined algorithms,
-the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the spatial configuration is faster than caching only a scalar at quadrature points or caching the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress.
+{\color{red}the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the spatial configuration is faster than caching only a scalar at quadrature points or caching the material part} of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress.
 {\color{red}
@@
 {\color{red}
-Despite the higher arithmetic intensity, the matrix-free implementations are still memory bandwidth-limited and in particular by the data at quadrature points in terms of fourth-order and second-order symmetric tensors.
+Despite the higher arithmetic intensity, the matrix-free implementations are still memory bandwidth-limited and in particular by the data required at quadrature points in terms of fourth-order and second-order symmetric tensors.
 }
@@
 {\color{red}
-It is a common practice to write bilinear form in fully referential configuration.
-For the here considered constitutive equation \eqref{eq:neo_hookean_energy} this results in the following bilinear form
+It is a common practice to write the bilinear form completely in the referential configuration.
+For the constitutive equation \eqref{eq:neo_hookean_energy} considered in this study, this results in the following bilinear form
 }
@@
 {\color{red}
-In the interest of brevity, we did not report any results for this approach,
+In the interest of brevity, no results are reported for this approach,
 however we do support it in our code.
-Below give a brief summary of the measured performance.
-Clearly in this case we need to cache more data as compared to the Algorithm \ref{alg:mf_tensor4}, this results in both higher memory requirement per DoF as well as lower arithmetic intensity of the vmult operator.
+As compared to Algorithm~\ref{alg:mf_tensor4}, we need to cache more data, which results in both higher memory requirement per DoF as well as a lower arithmetic intensity of the operator evaluation.
 On both architectures we recorded a slightly larger wall time per DoF as compared to the Algorithm \ref{alg:mf_tensor4}, however still much faster than the matrix-based approach.
 From the LIKWID measurements we observed that this algorithm also results in lower performance on Cascade Lake.
-The advantage of this algorithm is that it does not require evaluation of gradients in deformed configuration, therefore the matrix-free setup time is slightly lower as compared to the other three algorithms that rely on \texttt{MappingQEulerian} class of the \texttt{deal.II} library.
+The advantage of this algorithm is that it does not require evaluation of gradients in the deformed configuration, which leads to a slightly lower setup time as compared to the other three algorithms that rely on \texttt{MappingQEulerian} class of the \texttt{deal.II} library.
 Overall we consider this approach to be a valid alternative to the Algorithm \ref{alg:mf_tensor4}
 that can be be applied to any constitutive model which operates with the fourth-order referential tangent stiffness tensor on the quadrature point level.
+
+A scalar variant of the algorithm in referential coordinates, related to Algorithm~\ref{alg:mf_scalar}, can also be devised.
+In this setting, all quantities are evaluated in referential coordinates, necessitating only a single set of metric terms according to Table~\ref{tab:memory}.
+For the derivatives of the current solution an additional multiplication by $\gz F^{-T}$ for both the solution as well as the test function is necessary, with $\gz F^{-T}$ evaluated on the fly after line 10 in Algorithm~\ref{alg:mf_scalar}.
+Despite the additional calculations, this leads to a faster calculation than Algorithm~\ref{alg:mf_scalar} on Cascade Lake, running with 162 GFlop/s and 68 GB/s for $p=6$ in 2D or 225 GFlop/s and 59 GB/s for $p=4$ in 3D.
+However, execution is still slower than Algorithm~\ref{alg:mf_tensor2}, despite the lower memory access.
+This is because the bottleneck is now mostly within the core, although below the arithmetic throughput limit.
+This is mainly caused by the cost of indirect addressing in the vector access as well as division for inverting $\gz F$ on the fly.
 }

### 2019-08-08 Calculations/post_process.py
Do not plot tensor4_ns in the figures.

@@ -268,7 +268,7 @@ plt.plot(deg2d,time2d_tr, 'rs--', label='Trilinos')
 plt.plot(deg2d,time2d_sc, 'bo--', label='MF scalar')
 plt.plot(deg2d,time2d_t2, 'g^--', label='MF tensor2')
 plt.plot(deg2d,time2d_t4, 'cv--', label='MF tensor4')
-plt.plot(deg2d,time2d_t4_ns, 'mD--', label='MF tensor4 P')
+#plt.plot(deg2d,time2d_t4_ns, 'mD--', label='MF tensor4 P')
 plt.xlabel('polynomial degree')
 plt.ylabel('vmult wall time (s) / DoF')
 leg = plt.legend(loc='best', ncol=1)
@@ -285,7 +285,7 @@ plt.plot(deg3d,time3d_tr, 'rs--', label='Trilinos')
 plt.plot(deg3d,time3d_sc, 'bo--', label='MF scalar')
 plt.plot(deg3d,time3d_t2, 'g^--', label='MF tensor2')
 plt.plot(deg3d,time3d_t4, 'cv--', label='MF tensor4')
-plt.plot(deg3d,time3d_t4_ns, 'mD--', label='MF tensor4 P')
+#plt.plot(deg3d,time3d_t4_ns, 'mD--', label='MF tensor4 P')
 plt.xlabel('polynomial degree')
 plt.ylabel('vmult wall time (s) / DoF')
 leg = plt.legend(loc='best', ncol=1)
@@ -300,10 +300,10 @@ plt.plot(deg2d,through2d_tr, 'rs--', label='Trilinos')
 plt.plot(deg2d,through2d_sc, 'bo--', label='MF scalar')
 plt.plot(deg2d,through2d_t2, 'g^--', label='MF tensor2')
 plt.plot(deg2d,through2d_t4, 'cv--', label='MF tensor4')
-plt.plot(deg2d,through2d_t4_ns, 'mD--', label='MF tensor4 P')
+#plt.plot(deg2d,through2d_t4_ns, 'mD--', label='MF tensor4 P')
 plt.xlabel('polynomial degree')
 plt.ylabel('vmult DoF / s')
-leg = plt.legend(loc='best', ncol=1)
+leg = plt.legend(loc='lower right', ncol=1)
 plt.tight_layout()
 plt.savefig(fig_prefix + 'throughput2d.eps', format='eps')
[…]

### 2019-08-08 Calculations/post_process_likwid_csl.py

@@ -110,6 +110,11 @@ for f in files:
         regions = [
             'vmult_MF' if 'MF_CG' in f else 'vmult_Trilinos'
         ]
+
+    # skip tensor4_ns for plotting
+    if '_tensor4_ns' in fname:
+        continue
+
     label = ''
     color = ''  # use colors consistent with post_process.py
     if 'MF_CG' in fname:
@@ -205,7 +210,7 @@ likwid_data.sort(key=lambda tup: (tup[3], tup[1]))
 #       PLOT        #
 #####################
 
-params = {'legend.fontsize': 10,
+params = {'legend.fontsize': 12,
           'font.size': 14}
 plt.rcParams.update(params)
 

### 2019-08-08 doc/large_strain.tex
fix few typos and whitespaces

@@
 
-The matrix-free implementation is faster than the matrix-based counterpart already for a bi-quadractic Lagrangian basis in 2D and a tri-quadratic basis in 3D on both systems.
+The matrix-free implementation is faster than the matrix-based counterpart already for a bi-quadratic Lagrangian basis in 2D and a tri-quadratic basis in 3D on both systems.
 {\color{red}
@@
 The ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates the highest computational intensity, but remains close to the memory bandwidth limit with more than $80\, \text{GB/s}$ except for the $p=2$ case.
-The higher arithmetic load is not surprising as at each quadrature point we need to re-evaluate the Kirchoff stress $\gz \tau$.
+The higher arithmetic load is not surprising as at each quadrature point we need to re-evaluate the Kirchhoff stress $\gz \tau$.
 Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations as listed in Table~\ref{tab:memory}, no advantage of the scalar caching can be deduced despite the higher arithmetic performance.
@@
 {\color{red}
-When run in serial, explicit SIMD vectorization via instrinsics provides a speedup of up to $3\times$ over the auto-vectorization case.
+When run in serial, explicit SIMD vectorization via intrinsics provides a speedup of up to $3\times$ over the auto-vectorization case.
 This is less than the theoretical factor of 8, but in line with previous experiments presented in Figure 6 of \cite{kronbichler2017fastcomputer} with a speedup of around $1.5-2.5\times$ with 8-wide vectorization.

### 2019-08-12 Calculations/post_process.py
Update plot scripts

@@ -260,7 +260,7 @@ mem3d_t4_ns = [tup[3]/tup[1] for tup in mf3d_data_tensor4_ns]
 # file location
 fig_prefix = os.path.join(os.getcwd(), '../doc/' + os.path.basename(os.path.normpath(prefix)) + '_')
 
-params = {'legend.fontsize': 20,
+params = {'legend.fontsize': 18,
           'font.size': 20}
 plt.rcParams.update(params)
 
@@ -363,6 +363,9 @@ remove_creation_date(fig_prefix + 'memory3d.eps')
 plt.clf()
 plt.yscale('log', basey=10)
 plt.ylim(top=1e-2,bottom=1e-9)
+#ax.xaxis.set_major_locator(MaxNLocator(integer=True))
+params = {'legend.fontsize': 14, 'font.size': 20}
+plt.rcParams.update(params)
 
 plt.plot(deg2d,solver2d_tr, 'rs--', label='Trilinos Solver')
 # plt.plot(deg2d,solver2d_sc, 'bo--', label='MF scalar')
@@ -386,6 +389,9 @@ ax = plt.figure().gca()
 plt.yscale('log', basey=10)
 plt.ylim(top=1e-2,bottom=1e-9)
 ax.xaxis.set_major_locator(MaxNLocator(integer=True))
+params = {'legend.fontsize': 14,
+          'font.size': 16}
+plt.rcParams.update(params)
 
 plt.plot(deg3d,solver3d_tr, 'rs--', label='Trilinos Solver')
 # plt.plot(deg3d,solver3d_sc, 'bo--', label='MF scalar')
@@ -405,6 +411,9 @@ remove_creation_date(fig_prefix + 'solver3d.eps')
 
 # clear
 plt.clf()
+params = {'legend.fontsize': 14,
+          'font.size': 16}
+plt.rcParams.update(params)
 
 plt.plot(deg2d,cg2d_tr, 'rs--', label='Trilinos')
 plt.plot(deg2d,cg2d_t4, 'cv--', label='MF')  # tensor4')
@@ -417,6 +426,9 @@ remove_creation_date(fig_prefix + 'cg2d.eps')
 
 # clear
 plt.clf()
+params = {'legend.fontsize': 14,
+          'font.size': 16}
+plt.rcParams.update(params)
 
 ax = plt.figure().gca()
 ax.xaxis.set_major_locator(MaxNLocator(integer=True))

### 2019-08-12 Calculations/post_process_likwid_csl_breakdown.py

@@ -196,9 +196,10 @@ for i in range(n_stack):
 
   bars[i] = b[0]
 
-plt.legend(bars, stack_labels, loc='upper left')
+plt.legend(bars, stack_labels, loc='upper center', ncol=3, bbox_to_anchor=(0.5, 1.))
 
 plt.ylabel('fraction of wall time')
+plt.gca().set_ylim([0, 1.2])
 plt.xticks(ind_plt, ind_str)
 plt.xlabel('polynomial degree')
 

### 2019-08-12 doc/large_strain.tex
Update affiliations

@@
 \author[1]{Jean-Paul Pelteret}
-\author[2]{Daniel Arndt}
-\author[3]{Martin Kronbichler}
-\author[1,4]{Paul Steinmann}
+\author[2,3]{Daniel Arndt}
+\author[4]{Martin Kronbichler}
+\author[1,5]{Paul Steinmann}
 
@@
 
-\address[3]{\orgdiv{Institute for Computational Mechanics}, \orgname{Technical University of Munich}, \orgaddress{\country{Germany}}}
-
-\address[4]{\orgdiv{Glasgow Computational Engineering Center (GCEC)}, \orgname{University of Glasgow}, \orgaddress{\country{United Kingdom}}}
+\address[3]{\orgdiv{Computational Engineering and Energy Sciences Group},
+         \orgname{Oak Ridge National Laboratory}, \orgaddress{Oak Ridge, TN, USA}}
+
+\address[4]{\orgdiv{Institute for Computational Mechanics}, \orgname{Technical University of Munich}, \orgaddress{\country{Germany}}}
+
+\address[5]{\orgdiv{Glasgow Computational Engineering Center (GCEC)}, \orgname{University of Glasgow}, \orgaddress{\country{United Kingdom}}}
 
@@
 \else
-
  \author[a]{Denis Davydov\corref{cor}}
@@
 
-  \author[b]{Daniel Arndt}
-  \ead{daniel.arndt@iwr.uni-heidelberg.de}
-
-  \author[a,c]{Paul Steinmann}
+  \author[b,c]{Daniel Arndt
+   \footnotetext{%
+   This manuscript has been authored by UT-Battelle, LLC under Contract No.
+   DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government
+   retains and the publisher, by accepting the article for publication, acknowledges
+   that the United States Government retains a non-exclusive, paid-up, irrevocable,
+   worldwide license to publish or reproduce the published form of this manuscript,
+   or allow others to do so, for United States Government purposes. The Department
+   of Energy will provide public access to these results of federally sponsored
+   research in accordance with the DOE Public Access Plan
+   (http://energy.gov/downloads/doe-public-access-plan).}}
+  \ead{arndtd@ornl.gov}
+
+  \author[d]{Martin Kronbichler}
+  \ead{kronbichler@lnm.mw.tum.de}
+
+  \author[a,e]{Paul Steinmann}
   \ead{paul.steinmann@fau.de}
@@
 
-  \address[c]{Glasgow Computational Engineering Center (GCEC),
+  \address[c]{Computational Engineering and Energy Sciences Group,
+         Oak Ridge National Laboratory;
+         Oak Ridge, TN, USA}
+
+   \address[d]{Institute for Computational Mechanics, Technical University of Munich; Garching, Germany}
+
+  \address[e]{Glasgow Computational Engineering Center (GCEC),
       University of Glasgow, G12 8QQ Glasgow, United Kingdom
@@
     The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner on a single node for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
-    We {\color{red}find} that matrix-free methods for finite-strain solid mechanics are very promising, {\color{red}outperforming linear matrix-based schemes by two to five times}, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
+    We {\color{red}find} that matrix-free methods for finite-strain solid mechanics are very promising, {\color{red}outperforming linear matrix-based schemes by two to five times}, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}\footnotetext{%
+   This manuscript has been authored by UT-Battelle, LLC under Contract No.
+   DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government
+   retains and the publisher, by accepting the article for publication, acknowledges
+   that the United States Government retains a non-exclusive, paid-up, irrevocable,
+   worldwide license to publish or reproduce the published form of this manuscript,
+   or allow others to do so, for United States Government purposes. The Department
+   of Energy will provide public access to these results of federally sponsored
+   research in accordance with the DOE Public Access Plan
+   (http://energy.gov/downloads/doe-public-access-plan).}
   \end{abstract}

### 2019-08-12 doc/large_strain.tex
Use updated images and fix minor typos

@@
 The core of the sum factorization algorithms is provided by the
-\textit{FEEvaluation} class, that supports all the necessary contractions such as $\overline{\grad} \gz N_i : (\boldsymbol{\mathcal{C}}\gz g_s  + \boldsymbol{\mathcal{G}} \gz g)$ to implement the three algorithms proposed above.
+\textit{FEEvaluation} class, that supports all the necessary contractions such as $\overline{\grad} \, \gz N_i : (\boldsymbol{\mathcal{C}}\gz g_s  + \boldsymbol{\mathcal{G}} \gz g)$ to implement the three algorithms proposed above.
 Thanks to the object-oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
@@
 We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
-Essentially, level operators correspond to the linearization around the smoothed representation of displacement field.
+Essentially, level operators correspond to the linearization around the smoothed representation of the displacement field.
 Note that this differs from the classical approach where the (tangent) operator $\gz A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $\gz A^{l}$ on level $l$ via $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, where $\gz I^l_{l+1}$ and $\gz I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
@@
       \centering
-      \includegraphics[width=\textwidth]{CSL_Munich_throughput2d.eps}
+      \includegraphics[width=\textwidth]{CSL_Munich_throughput2d-nolables.eps}
       \caption{throughput matrix-vector product (2D)}
@@
       \centering
-      \includegraphics[width=\textwidth]{CSL_Munich_memory2d.eps}
+      \includegraphics[width=\textwidth]{CSL_Munich_memory2d-nolables.eps}
       \caption{memory access (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{CSL_Munich_throughput3d.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_throughput3d-nolables.eps}
     \caption{throughput matrix-vector product (3D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{CSL_Munich_memory3d.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_memory3d-nolables.eps}
     \caption{memory access (3D)}
@@
       \centering
-      \includegraphics[width=\textwidth]{IWR_newest_patched_throughput2d.eps}
+      \includegraphics[width=\textwidth]{IWR_newest_patched_throughput2d-nolables.eps}
       \caption{throughput matrix-vector product (2D)}
@@
       \centering
-      \includegraphics[width=\textwidth]{IWR_newest_patched_memory2d.eps}
+      \includegraphics[width=\textwidth]{IWR_newest_patched_memory2d-nolables.eps}
       \caption{memory access (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_newest_patched_throughput3d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_throughput3d-nolables.eps}
     \caption{throughput matrix-vector product (3D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_newest_patched_memory3d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_memory3d-nolables.eps}
     \caption{memory access (3D)}
@@
 {\color{red}
-This formulation can also be straight forwardly adopted within the matrix-free framework of \texttt{deal.II} library.
+This formulation can also be straight forwardly adopted within the matrix-free framework of the \texttt{deal.II} library.
 To that end a variation of the Algorithm \ref{alg:mf_tensor4} can be used where the gradients are evaluated with respect to the undeformed configuration and only a fourth-order referential tangent tensor ${\rm D}_{\gz F}{\gz P}$ is cached per quadrature point.

### 2019-08-12 doc/large_strain.tex
remove the comment on coarse solver
minor TeX improvements

@@
 
-\usepackage{todonotes}
+\usepackage{pgfplots}
 \usepackage{bm}
@@
   $q$ is the number of quadrature points in 1D, $N_{\text{gref}}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{\text{DoF}}$ is the number of DoFs.}
-  \begin{subtable}{.5\linewidth}
+  \begin{subtable}{.49\linewidth}
   \caption{2D}
@@
   \end{subtable}
-  \begin{subtable}{.5\linewidth}
+  \begin{subtable}{.49\linewidth}
   \caption{3D}
@@
     \caption{CG iterations (2D)}
-    \label{fig:benchmark_miehe_Emmy_cg2}
-  \end{subfigure}
+    \label{fig:benchmark_miehe_CSL_cg2}
+  \end{subfigure}
+  ~
   \begin{subfigure}[b]{0.49\textwidth}
@@
     \caption{CG iterations (3D)}
-    \label{fig:benchmark_miehe_Emmy_cg3}
+    \label{fig:benchmark_miehe_CSL_cg3}
   \end{subfigure}
@@
   \end{subfigure}
+  ~
   \begin{subfigure}[b]{0.49\textwidth}
@@
 we consider the average number of CG iterations throughout the entire simulation, i.e., averaged over each Newton-Raphson iteration and each loading step.
-Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
+Figures \ref{fig:benchmark_miehe_CSL_cg2}, \ref{fig:benchmark_miehe_CSL_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
  noticeable increase in the average number of CG iterations for increasing polynomial degrees
  for the GMG preconditioner.
-Although not reported in this section, we identified that
-the coarse level solver setup will greatly affect the performance of the preconditioner.
-In particular, we can improve the number of iterations by choosing a more accurate coarse level solver.
 The black-box AMG preconditioner

### 2019-08-13 doc/large_strain.tex
a few minor fixes

@@
   \begin{abstract}[Summary]
-    {{\color{red}This work investigates matrix-free algorithms for problems in quasi-static finite-strain hyperelasticity. Iterative solvers with matrix-free operator evaluation have emerged as
+    {{\color{red}This work investigates matrix-free algorithms for problems in quasi-static finite-strain hyperelasticity.
+    Iterative solvers with matrix-free operator evaluation have emerged as
 an attractive alternative to sparse matrices in the fluid dynamics and wave
@@
 where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$
-is involved due to the symmetry of $\gz \tau$. Here, `$\grad$' denotes the gradient with respect to the spatial configuration, as opposed to the gradient in the referential setting `$\Grad$'.
+is involved due to the symmetry of $\gz \tau$.
+{\color{red}
+Here, `$\grad$' denotes the gradient with respect to the spatial configuration, as opposed to the gradient in the referential setting `$\Grad$'.
+}
 Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \psi / \partial \gz E \equiv 2 \partial \psi / \partial \gz C$,
@@
 table, are increasing in size with more caching, and are stored for each
-quadrature point. Finally, the evaluation of the gradients in the referential
+quadrature point. The evaluation of the gradients in the referential
 and spatial configuration also involves a data access via the so-called metric
@@
 {\color{red}
-The ``MF tensor2'' and ``MF scalar'' strategies outperform the sparse matrix already for tri-linear elements of polynomial degree $p=1$.
+The ``MF tensor2'' and ``MF scalar'' strategies outperform the sparse matrix already for tri-linear elements of polynomial degree $p=1$ in 3D on Cascade Lake architecture.
 For example for the ``tensor2'' matrix-free implementation with $p=4$, we record a 5$\times$ higher throughput than the matrix-based evaluation of $p=1$.
@@
 Firstly, the algorithm where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'') is the fastest approach in both 2D and 3D and on both systems.
-When compared to the scalar caching implementation, the ``tensor2'' variant loads slightly less data, because only the spatial configuration is needed in terms of metric terms, as compared to both the spatial and the referential configuration in the scalar caching case in order to evaluate the Kirchhoff stress $\gz \tau$.
+When compared to the scalar caching implementation, the ``tensor2'' variant loads slightly less data, because only the spatial configuration is needed for the metric terms, as compared to both the spatial and the referential configuration in the scalar caching case in order to evaluate the Kirchhoff stress $\gz \tau$.
 Besides the memory access, the scalar caching strategy also involves considerably more computations at quadrature points.
@@
 {\color{red}
-The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
+The matrix-free implementations of the finite-strain elastic tangent operator (on average) lead to a much better performance of
 $\pgfmathprintnumber{86.9}\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
-The achieved performance is about 6.8\% of the peak arithmetic performance. This result similar to the deformed mesh results of \cite[Figure 18]{kronbichler2017fast}.
+The achieved performance is about 6.8\% of the peak arithmetic performance. This result is similar to the deformed mesh results of \cite[Figure 18]{kronbichler2017fast}.
 }
@@
 When going to the parallel case with 20 cores, the less-than-optimal speedup can be attributed to two different effects.
-For the auto-vectorized case, the different (turbo) clock frequencies permit a speedup of a factor 12.6 at most, close to our results.\footnote{Note that we intentionally run with Turbo mode on to permit the processor to come close to the power limit for all settings and present more realistic performance limits.} For the SIMD+MPI case, the main reason is the memory bandwidth: As can be seen from Figures~\ref{fig:roofline_2d} and~\ref{fig:roofline_3d}, both algorithms are eventually memory-bandwidth limited.
+For the auto-vectorized case, the different (turbo) clock frequencies permit a speedup of a factor 12.6 at most, close to our results\footnote{Note that we intentionally run with Turbo mode on to permit the processor to come close to the power limit for all settings and present more realistic performance limits.}.
+For the SIMD+MPI case, the main reason is the memory bandwidth: As can be seen from Figures~\ref{fig:roofline_2d} and~\ref{fig:roofline_3d}, both algorithms are eventually memory-bandwidth limited.
 Since Algorithm~\ref{alg:mf_tensor4} has a lower arithmetic intensity, the speedup over the serial, auto-vectorized case is less than that of Algorithm~\ref{alg:mf_tensor2}.
@@
 In this contribution, we proposed and numerically investigated three different matrix-free implementations of tangent operators for finite-strain elasticity with heterogeneous materials.
-To the best of our knowledge, this is the first work that evaluates matrix-free sum factorization techniques to the partial differential equations arising in this case.
+To the best of our knowledge, this is the first work that evaluates matrix-free sum factorization techniques on the partial differential equations arising in this case.
 Among the three examined algorithms,
@@
 For the derivatives of the current solution an additional multiplication by $\gz F^{-T}$ for both the solution as well as the test function is necessary, with $\gz F^{-T}$ evaluated on the fly after line 10 in Algorithm~\ref{alg:mf_scalar}.
-Despite the additional calculations, this leads to a faster calculation than Algorithm~\ref{alg:mf_scalar} on Cascade Lake, running with 162 GFlop/s and 68 GB/s for $p=6$ in 2D or 225 GFlop/s and 59 GB/s for $p=4$ in 3D.
+Despite the additional evaluations, this leads to a faster calculation than Algorithm~\ref{alg:mf_scalar} on Cascade Lake, running with 162 GFlop/s and 68 GB/s for $p=6$ in 2D or 225 GFlop/s and 59 GB/s for $p=4$ in 3D.
 However, execution is still slower than Algorithm~\ref{alg:mf_tensor2}, despite the lower memory access.

### 2019-08-14 Calculations/post_process.py
update README and plotting scripts; actual contents of Table3/4 are now in a separate file as generated by script (numbers did not change, as expected)

@@ -18,7 +18,7 @@ def remove_creation_date(file_name):
 parser = argparse.ArgumentParser(
     description='Post-Process timing/memory info and plot figures.',
     formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-parser.add_argument('prefix', metavar='prefix', default='Emmy_RRZE', nargs='?',
+parser.add_argument('prefix', metavar='prefix', default='CSL_Munich', nargs='?',
                     help='A folder to look for benchmark results')
 args = parser.parse_args()
 
@@ -272,7 +272,6 @@ plt.plot(deg2d,time2d_t4, 'cv--', label='MF tensor4')
 plt.xlabel('polynomial degree')
 plt.ylabel('vmult wall time (s) / DoF')
 leg = plt.legend(loc='best', ncol=1)
-#plt.tight_layout()
 plt.savefig(fig_prefix + 'timing2d.eps', format='eps', bbox_inches = 'tight')
 remove_creation_date(fig_prefix + 'timing2d.eps')
 
@@ -289,7 +288,6 @@ plt.plot(deg3d,time3d_t4, 'cv--', label='MF tensor4')
 plt.xlabel('polynomial degree')
 plt.ylabel('vmult wall time (s) / DoF')
 leg = plt.legend(loc='best', ncol=1)
-#plt.tight_layout()
 plt.savefig(fig_prefix + 'timing3d.eps', format='eps', bbox_inches = 'tight')
 remove_creation_date(fig_prefix + 'timing3d.eps')
 
@@ -303,8 +301,7 @@ plt.plot(deg2d,through2d_t4, 'cv--', label='MF tensor4')
 #plt.plot(deg2d,through2d_t4_ns, 'mD--', label='MF tensor4 P')
 plt.xlabel('polynomial degree')
 plt.ylabel('vmult DoF / s')
-leg = plt.legend(loc='lower right', ncol=1)
-#plt.tight_layout()
+# leg = plt.legend(loc='lower right', ncol=1)
 plt.savefig(fig_prefix + 'throughput2d.eps', format='eps', bbox_inches = 'tight')
 remove_creation_date(fig_prefix + 'throughput2d.eps')
 
@@ -320,8 +317,7 @@ plt.plot(deg3d,through3d_t4, 'cv--', label='MF tensor4')
[…]

### 2019-08-14 Calculations/post_process_likwid_csl.py

@@ -18,7 +18,7 @@ def remove_creation_date(file_name):
 parser = argparse.ArgumentParser(
     description='Post-Process timing/memory info and plot figures.',
     formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-parser.add_argument('--prefix', metavar='prefix', default='LIKWID_Emmy_RRZE',
+parser.add_argument('--prefix', metavar='prefix', default='LIKWID_CSL_Munich',
                     help='A folder to look for benchmark results')
 parser.add_argument('--dim', metavar='dim', default=2, type=int,
                     help='Dimension (2 or 3)')

### 2019-08-14 Calculations/post_process_likwid_simd.py

@@ -8,9 +8,11 @@ from utilities import *
 parser = argparse.ArgumentParser(
     description='Post-Process timing/memory info and create a table.',
     formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-parser.add_argument('--prefix', metavar='prefix', default='LIKWID_Emmy_RRZE',
+parser.add_argument('--prefix', metavar='prefix', default='LIKWID_CSL_Munich',
                     help='A folder prefix to look for benchmark results')
 parser.add_argument('--dim', metavar='dim', default=2, type=int,
+                    help='Dimension (2 or 3)')
+parser.add_argument('--alg', metavar='alg', default='tensor4',
                     help='Dimension (2 or 3)')
 
 args = parser.parse_args()
@@ -51,15 +53,14 @@ for idx, s in enumerate(suffixes):
     for f in files:
         # guess parameters from the file name:
         fname = os.path.basename(f)
+        dname = os.path.dirname(f)
         strings = fname.split('_')
         dim = int(re.findall(get_regex_pattern(),strings[2])[0])
         p   = int(re.findall(get_regex_pattern(),strings[3])[0])
         q   = int(re.findall(get_regex_pattern(),strings[3])[1])
 
         # we are only interested in tensor2 (Algorithm 3) for the currently chosen dimension
-        if not (dim == args.dim and '_tensor4' in fname):
-            continue
-        if ('_tensor4_ns' in fname):
+        if not (dim == args.dim and dname.endswith(args.alg)):
             continue
 
         print 'dim={0} p={1} q={2} file={3}'.format(dim,p,q,fname)
@@ -99,22 +100,22 @@ for idx, s in enumerate(suffixes):
             table_data[row][col+2] = table_data[row][1] / runtime
 
 # Finally write out a latex multicolumn table
-file_name = os.path.join(os.getcwd(), '../doc/parallelization_{0}d.tex'.format(args.dim))
+file_name = os.path.join(os.getcwd(), '../doc/parallelization_{0}d_{1}.tex'.format(args.dim,args.alg))
 print 'Saving table in ' + file_name
 
 with open(file_name, 'w') as f:
-    # start with the header
-    f.write("""\
-\\begin{table}
-\centering
-\\resizebox{\\textwidth}{!}{
[…]

### 2019-08-14 doc/large_strain.tex
break lines so that we don't confuse git as much
update README and plotting scripts; actual contents of Table3/4 are now in a separate file as generated by script (numbers did not change, as expected)
cleanup python script, remove unused figures
rework main post_process.py to use .pdf and remove all unused figures

@@
     The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner on a single node for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
-    We {\color{red}find} that matrix-free methods for finite-strain solid mechanics are very promising, {\color{red}outperforming linear matrix-based schemes by two to five times}, and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}\footnotetext{%
+    We {\color{red}find} that matrix-free methods for finite-strain solid mechanics are very promising, {\color{red}outperforming linear matrix-based schemes by two to five times},
+    and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}\footnotetext{%
    This manuscript has been authored by UT-Battelle, LLC under Contract No.
@@
 {\color{red}
-Among matrix-free methods, stencil-based variants are most popular for linear elements and tetrahedral shapes~\cite{Gmeiner2016,Bauer2018}, whereas implementations based on integration with sum-factorization techniques are attractive for higher polynomial degrees $p\geq 2$ and tensor product shape functions~\cite{Brown2010,kronbichler12,May2015,Cantwell2011}.
+Among matrix-free methods, stencil-based variants are most popular for linear elements and tetrahedral shapes~\cite{Gmeiner2016,Bauer2018},
+whereas implementations based on integration with sum-factorization techniques are attractive for higher polynomial degrees $p\geq 2$
+and tensor product shape functions~\cite{Brown2010,kronbichler12,May2015,Cantwell2011}.
 }
@@
 One reason is that the tangent operators\footnote{Here and
-below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
+below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$,
+where $N$ is the number of unknown degrees of freedom in the FE discretization.
 Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
@@
 {\color{red}
-For matrix-free schemes using fast integration via sum factorization, the main challenge is to find a representation of the linearization point, either in referential or spatial configurations, at the quadrature points that is cheap enough to be evaluated on the fly.
+For matrix-free schemes using fast integration via sum factorization, the main challenge is to find a representation of the linearization point,
+either in referential or spatial configurations, at the quadrature points that is cheap enough to be evaluated on the fly.
 }
 Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
-The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted,
+The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking),
+or mixed or enhanced formulations (that avoid locking) are often adopted,
 whereas higher order elements are rarely employed.
-This can potentially reduce the advantage of matrix-free methods using sum-factorization techniques, which are generally more competitive for higher order elements \cite{kronbichler12,kronbichler2017fast,muthing2017high,Fischer2019}.
-However, we note that, even with linear FEM, large scale computations with as many as $10^{12}$ unknowns are possible with a matrix-free approach but infeasible with sparse matrices as there is simply not enough memory to store the sparse tangent matrix even on large supercomputers \cite{Gmeiner2016}.
+This can potentially reduce the advantage of matrix-free methods using sum-factorization techniques,
+which are generally more competitive for higher order elements \cite{kronbichler12,kronbichler2017fast,muthing2017high,Fischer2019}.
+However, we note that, even with linear FEM, large scale computations with as many as $10^{12}$ unknowns are possible with a matrix-free approach
+but infeasible with sparse matrices as there is simply not enough memory to store the sparse tangent matrix even on large supercomputers \cite{Gmeiner2016}.
 Therefore, we believe that for large-scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient solvers.
@@
 pre-computation with a higher memory access in terms of the representation of the linearization point.
-The \texttt{deal.II} finite element library \cite{dealII90}, which provides MPI parallelization together with a generalized matrix-free framework that incorporates SIMD vectorization and an interface to high-performance MPI-distributed matrix-based libraries, is used for this study.
+The \texttt{deal.II} finite element library \cite{dealII90}, which provides MPI parallelization together with a generalized matrix-free framework
+that incorporates SIMD vectorization and an interface to high-performance MPI-distributed matrix-based libraries, is used for this study.
 In order to improve the convergence of iterative solvers, we also propose a method by which to construct level matrix-free tangent operators
@@
 %
-The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}.
+The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with
+an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}.
 Finally, the results are summarized in Section \ref{sec:summary}.
@@
 \end{align}
-where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the referential configuration, $\psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
+where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the referential configuration,
+$\psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary
+conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
 
@@
 \end{equation}
-Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order material part of the spatial tangent stiffness tensor is the push forward of the material part of the referential tangent stiffness tensor $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
+Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$,
+and the fourth-order material part of the spatial tangent stiffness tensor is the push forward of the material part of the referential tangent stiffness tensor
+$J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
 The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and is therefore called material part of the directional derivative.
@@
 where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively.
-Derived using statistical thermodynamics applied to cross-linked polymers, the Neo-Hookean model \cite{Treloar1975a,Treloar1976a} is commonly used to model rubber-like materials in the finite-strain regime.
+Derived using statistical thermodynamics applied to cross-linked polymers,
+the Neo-Hookean model \cite{Treloar1975a,Treloar1976a} is commonly used to model rubber-like materials in the finite-strain regime.
 It can be shown that the first and second derivatives of the strain energy function are given by
@@
 The Newton-Raphson solution approach is adopted for the nonlinear problem considered here.
-Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is obtained as the solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent}):
+Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is obtained as the solution to the following
+system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent}):
 \begin{align}
@@
 and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
-Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the spatial configuration, whereas the integration is done in the referential configuration.
+Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the spatial configuration,
+whereas the integration is done in the referential configuration.
 
@@
 \label{sec:mf}
-Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
-In this case, the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
+Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled
+(that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
+In this case, the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector,
+and is usually implemented with the aid of specialized linear algebra packages.
 On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the associated arithmetic operations.
-For this reason, recent implementations often focus on so-called matrix-free approaches \cite{kronbichler12, kronbichler2017fast} where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
+For this reason, recent implementations often focus on so-called matrix-free approaches \cite{kronbichler12, kronbichler2017fast} where the matrix is not
+assembled but rather the result of its operation on a vector is directly calculated.
 The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
@@
 \end{align}
-where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form at one quadrature point $\gz \xi_q$, $J^K_q$ is the {\color{red}determinant of the} Jacobian of the mapping from the isoparametric element to the element $K$ and
+where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form at one quadrature point $\gz \xi_q$, $J^K_q$ is the
+{\color{red}determinant of the} Jacobian of the mapping from the isoparametric element to the element $K$ and
 $w_q$ is the quadrature weight.
@@
 \end{align*}
-Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one{\color{red}, $n=p+1$}, the evaluation of all of the shape functions ($n^d$)
+Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one{\color{red}, $n=p+1$},
+the evaluation of all of the shape functions ($n^d$)
 in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
@@
 %
-Note that $x_j \equiv X_{j_1\,j_2}$; the former accesses the element source vector using a linear index $j$, whereas the latter uses the associated multi-index accesses $\{j_1\,j_2\}$.
+Note that $x_j \equiv X_{j_1\,j_2}$; the former accesses the element source vector using a linear index $j$,
+whereas the latter uses the associated multi-index accesses $\{j_1\,j_2\}$.
 %
 These four loops all have the same structure.
-We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
-Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, this results in an algorithm with an arithmetic complexity of $\mathcal{O}(n^{d+1})$.
-Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for the 3D case, and where the discretization and evaluation employs higher degree basis functions with the corresponding quadrature rule.
+We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or
+keep the quadrature point fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
+Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or
+values in degrees of freedoms simultaneously, this results in an algorithm with an arithmetic complexity of $\mathcal{O}(n^{d+1})$.
+Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach,
+especially for the 3D case, and where the discretization and evaluation employs higher degree basis functions with the corresponding quadrature rule.
 More information on matrix-free techniques can be found in \cite{kronbichler12,vos10}.
 
-For the performance of the matrix-free operator evaluation, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.
+For the performance of the matrix-free operator evaluation, it is crucial to find a good intermediate point
+between precomputing everything (caching) and evaluating all the quantities on-the-fly.
 %
@@
 we use the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
-In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
+In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping
+to the spatial configuration given a displacement field.
 %
@@
 
-Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach, each of which have a different level of abstraction, algorithmic complexity, and memory requirements.
+Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent})
+using the matrix-free approach, each of which have a different level of abstraction, algorithmic complexity, and memory requirements.
 They are introduced in order of lowest memory footprint to largest:
@@
 
-Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to additions and multiplications.
-As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the tangent operator.
-This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the referential configuration.
+Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to
+additions and multiplications. As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point
+in order to apply the tangent operator. This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the referential configuration.
 
@@
 This algorithm utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
-Therefore, the action of the material part of the fourth-order spatial tangent stiffness tensor $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
+Therefore, the action of the material part of the fourth-order spatial tangent stiffness tensor $J \boldsymbol{\mathcal{C}}$ on the
+second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
 To that end, we cache two scalars that depend on the Jacobian {\color{red}determinant} $J$.
@@
 Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material part of the fourth-order spatial tangent stiffness tensor.
-In this case, we cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\overline{\grad^s} \gz x$ on-the-fly.
+In this case, we cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and
+quadrature point, and perform the double contraction with the second-order symmetric tensor $\overline{\grad^s} \gz x$ on-the-fly.
 
@@
 is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
-The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements{\color{red}, as opposed to the operations within an element}\footnote{%
-In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
-This motivates the caching strategy employed in Algorithm 3, as the pre-caching of the chosen data ensures a context in which SIMD parallelization of subsequent operations is guaranteed.
+The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements{\color{red},
+as opposed to the operations within an element}\footnote{%
+In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that
+follow the same code path during evaluation or the kinetic quantities and tangents.
+This motivates the caching strategy employed in Algorithm 3, as the pre-caching of the chosen data ensures a context in which
+SIMD parallelization of subsequent operations is guaranteed.
 Algorithms 1 and 2 sacrifice some of this generality for decreased memory requirements.
@@
 Thanks to the object-oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
-Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the locally owned elements.
-For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well as SIMD vectorization) and performance tests, see \cite{kronbichler12}.
+Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique,
+where each MPI process is responsible for applying the operator only on the locally owned elements.
+For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II}
+(including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well as SIMD vectorization) and performance tests, see \cite{kronbichler12}.
 
@@
 
-The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling}, has been implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
-This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aims to support exascale computing.
-In this work we employ \texttt{deal.II}'s interface to the \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
+The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in
+Section \ref{sec: Constitutive modelling}, has been implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
+This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the
+xSDK (Extreme-scale Scientific Software Development Kit) ecosystem \citep{Bartlett2017b} aims to support exascale computing.
+In this work we employ \texttt{deal.II}'s interface to the \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition,
+whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
 %
@@
 
-In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver, both of which are the focal points of investigation in this manuscript, has been achieved using MPI;
+In practice, this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver,
+both of which are the focal points of investigation in this manuscript, has been achieved using MPI;
 various pre- and post-processing steps also leveraged parallelization.
@@
 
-The discrete tangent operator described by \eqref{eq:algebraic_tangent} can be implemented in one of two ways, namely through a classical matrix-based approach or, alternatively, a matrix-free approach.
+The discrete tangent operator described by \eqref{eq:algebraic_tangent} can be implemented in one of two ways,
+namely through a classical matrix-based approach or, alternatively, a matrix-free approach.
 For the matrix-based approach adopted in this study, the Trilinos \cite{Heroux2005} library has been leveraged;
-the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} sparse data structure from the Epetra package \cite{Heroux2005b} and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
+the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} sparse data structure from the Epetra package \cite{Heroux2005b} and
+the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
 This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is available in \texttt{deal.II}.
@@
 The second implementation utilizes a matrix-free approach in conjunction with a geometric-multigrid (GMG) preconditioner.
-GMG preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes.
+GMG preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the
+number of mesh refinements by smoothing the residual on a hierarchy of meshes.
 For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
@@
 In this work, we adopt the matrix-free level operators within the GMG.
-In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial.
+In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{Finite dimensional linear operators from a vector space associated
+with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial.
 One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}.
@@
 that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material.
-In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
-We define the level operators as follows: Recall that the fine scale tangent operator \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$.
-We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
+In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and
+thus admits usage of standard geometric transfer operators.
+We define the level operators as follows: Recall that the fine scale tangent operator \eqref{eq:algebraic_tangent} is obtained by
+linearization around the current displacement field $\gz u^h$. We then restrict this displacement field to all multigrid levels $\{l\}$ and
+evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
 Essentially, level operators correspond to the linearization around the smoothed representation of the displacement field.
-Note that this differs from the classical approach where the (tangent) operator $\gz A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $\gz A^{l}$ on level $l$ via $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, where $\gz I^l_{l+1}$ and $\gz I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
-
-The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators, and the coarse level iterative solver are provided by the \texttt{deal.II} library; implementational details are discussed in \cite{Clevenger2018}.
+Note that this differs from the classical approach where the (tangent) operator $\gz A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator
+$\gz A^{l}$ on level $l$ via $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, where $\gz I^l_{l+1}$ and $\gz I^{l+1}_l$ are global
+prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
+
+The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators, and the
+coarse level iterative solver are provided by the \texttt{deal.II} library; implementational details are discussed in \cite{Clevenger2018}.
 %
 The efficiency of the GMG preconditioner with the considered level operators will be assessed by numerical examples in the next section.
-We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
+We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is
+the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
 %
@@
 
-In this section, we evaluate the proposed matrix-free algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner on a benchmark problem from \cite{Miehe2007}.
-The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}.
-The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red).
-The size of the square domain is $10^{-3} \rm{mm}$.
+In this section, we evaluate the proposed matrix-free algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner
+on a benchmark problem from \cite{Miehe2007}. The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}.
+The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red). The size of the square domain is $10^{-3} \rm{mm}$.
 The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
@@
 The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \sqrt{2} \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
-With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$.
-The relative convergence criteria for the linear solver is $10^{-6}$.
+With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$
+whereas the relative and absolute tolerances for the residual forces is $10^{-8}$. The relative convergence criteria for the linear solver is $10^{-6}$.
 %
@@
   \caption{Parameters for the benchmark: $p$ is the polynomial degree,
-  $q$ is the number of quadrature points in 1D, $N_{\text{gref}}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{\text{DoF}}$ is the number of DoFs.}
+  $q$ is the number of quadrature points in 1D, $N_{\text{gref}}$ is the number of global mesh refinements,
+  $N_{el}$ is the number of elements and $N_{\text{DoF}}$ is the number of DoFs.}
   \begin{subtable}{.49\linewidth}
@@
 {\color{red}
-Examinations performed in this manner are most representative for the differences between the matrix-free and matrix-based approaches because problems of sufficient size are dominated by the local node-level performance with multigrid solvers.
+Examinations performed in this manner are most representative for the differences between the matrix-free and matrix-based approaches because problems
+of sufficient size are dominated by the local node-level performance with multigrid solvers.
 Multigrid solvers with small enough coarse-grid problems are well-known to be scalable even to the largest supercomputers \cite{gholami2016,ibeid2018} as the communication is mostly between nearest neighbors.
 }
-The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
+The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements,
+as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
 We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach with a high polynomial degree.
-Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound.
+Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization)
+will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound.
 {\color{red}Furthermore, the data structures of the matrix-free approaches are also too large to fit completely into caches.}
@@
 We conduct two performance studies using the described discretization, the first being for matrix-vector multiplication only, and the second for the preconditioned iterative linear solver.
-The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library version 12.12.1.
+The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the
+algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library version 12.12.1.
 The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
@@
 On the coarsest level, we use the Chebyshev preconditioner as a solver \cite{Varga2009} to reduce the residual by three orders of magnitude.
-Since this smoother only uses diagonal elements of the operator, it is readily compatible with the matrix-free approach to leverage fast matrix-vector products, as opposed to most default smoothers which rely on an explicit matrix form.
-
-{\color{red}
-Computations were performed on two systems: A dual-socket Intel Xeon Gold 6230 (``Cascade Lake''), released in 2019, with 20 cores per socket and 96 GB of DDR4-2933 memory (nominal bandwidth per socket: 141 GB/s).
+Since this smoother only uses diagonal elements of the operator, it is readily compatible with the matrix-free approach to leverage fast matrix-vector products,
+as opposed to most default smoothers which rely on an explicit matrix form.
+
+{\color{red}
+Computations were performed on two systems: A dual-socket Intel Xeon Gold 6230 (``Cascade Lake''), released in 2019,
+with 20 cores per socket and 96 GB of DDR4-2933 memory (nominal bandwidth per socket: 141 GB/s).
 Turbo mode is enabled with a power limit of 135 watt per socket, which results in the processor running at 2.8 GHz for scalar code and 2.0 GHz for AVX-512-heavy code in the experiments.
@@
 }
-The ``IWR'' results were obtained on a single machine with eight Intel Xeon E7-8870 ``Westmere'' chips, released in 2011, (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip.
-For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
+The ``IWR'' results were obtained on a single machine with eight Intel Xeon E7-8870 ``Westmere'' chips, released in 2011,
+(10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip. For the simulation, GCC version 8.1.0 with flags
+``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
 Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. one socket for Cascade Lake and two sockets for ``IWR''.
@@
 {\color{red}
-In order to increase visibility, the figures also report the throughput of the matrix vector product, computed by the ratio of the number of degrees of freedom over the time for a matrix-vector product, in panels (a) and (d).
+In order to increase visibility, the figures also report the throughput of the matrix vector product,
+computed by the ratio of the number of degrees of freedom over the time for a matrix-vector product, in panels (a) and (d).
 }
@@
       \centering
-      \includegraphics[width=\textwidth]{CSL_Munich_throughput2d-nolables.eps}
+      \includegraphics[width=\textwidth]{CSL_Munich_throughput2d.pdf}
       \caption{throughput matrix-vector product (2D)}
@@
       \centering
-      \includegraphics[width=\textwidth]{CSL_Munich_timing2d.eps}
+      \includegraphics[width=\textwidth]{CSL_Munich_timing2d.pdf}
       \caption{time matrix-vector product (2D)}
@@
       \centering
-      \includegraphics[width=\textwidth]{CSL_Munich_memory2d-nolables.eps}
+      \includegraphics[width=\textwidth]{CSL_Munich_memory2d.pdf}
       \caption{memory access (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{CSL_Munich_throughput3d-nolables.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_throughput3d.pdf}
     \caption{throughput matrix-vector product (3D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{CSL_Munich_timing3d.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_timing3d.pdf}
     \caption{time matrix-vector product (3D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{CSL_Munich_memory3d-nolables.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_memory3d.pdf}
     \caption{memory access (3D)}
@@
       \centering
-      \includegraphics[width=\textwidth]{IWR_newest_patched_throughput2d-nolables.eps}
+      \includegraphics[width=\textwidth]{IWR_newest_patched_throughput2d.pdf}
       \caption{throughput matrix-vector product (2D)}
@@
       \centering
-      \includegraphics[width=\textwidth]{IWR_newest_patched_timing2d.eps}
+      \includegraphics[width=\textwidth]{IWR_newest_patched_timing2d.pdf}
       \caption{time matrix-vector product (2D)}
@@
       \centering
-      \includegraphics[width=\textwidth]{IWR_newest_patched_memory2d-nolables.eps}
+      \includegraphics[width=\textwidth]{IWR_newest_patched_memory2d.pdf}
       \caption{memory access (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_newest_patched_throughput3d-nolables.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_throughput3d.pdf}
     \caption{throughput matrix-vector product (3D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_newest_patched_timing3d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_timing3d.pdf}
     \caption{time matrix-vector product (3D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_newest_patched_memory3d-nolables.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_memory3d.pdf}
     \caption{memory access (3D)}
@@
 For example for the ``tensor2'' matrix-free implementation with $p=4$, we record a 5$\times$ higher throughput than the matrix-based evaluation of $p=1$.
-The advantage of the matrix-free implementation can be explained by the memory access per unknown shown in the panels (c) and (f) of the figures, as the time per unknown in panels (b) and (e) closely follows the memory access.
-}
-Figures \ref{fig:benchmark_miehe_CSL_vmult2}, \ref{fig:benchmark_miehe_CSL_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
+The advantage of the matrix-free implementation can be explained by the memory access per unknown shown in the panels (c) and (f) of the figures,
+as the time per unknown in panels (b) and (e) closely follows the memory access.
+}
+Figures \ref{fig:benchmark_miehe_CSL_vmult2}, \ref{fig:benchmark_miehe_CSL_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3}
+clearly show the influence of the different caching strategies,
 described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.
@@
 {\color{red}
-Firstly, the algorithm where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'') is the fastest approach in both 2D and 3D and on both systems.
-When compared to the scalar caching implementation, the ``tensor2'' variant loads slightly less data, because only the spatial configuration is needed for the metric terms, as compared to both the spatial and the referential configuration in the scalar caching case in order to evaluate the Kirchhoff stress $\gz \tau$.
+Firstly, the algorithm where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement
+the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'')
+is the fastest approach in both 2D and 3D and on both systems.
+When compared to the scalar caching implementation, the ``tensor2'' variant loads slightly less data, because only the spatial configuration is needed for the metric terms,
+as compared to both the spatial and the referential configuration in the scalar caching case in order to evaluate the Kirchhoff stress $\gz \tau$.
 Besides the memory access, the scalar caching strategy also involves considerably more computations at quadrature points.
@@
 {\color{red}
-Secondly, the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') is slower than the ``tensor2'' approach, despite fewer arithmetic operations, on both systems.
+Secondly, the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') is slower than the ``tensor2'' approach,
+despite fewer arithmetic operations, on both systems.
 This shows that the memory access is the deciding metric for the ``tensor4'' strategy.
-Nonetheless, the ``tensor4'' strategy with $p\geq 2$ is considerably faster than matrix-based strategies: For $p=4$ in 3D, the matrix-free variant is three times as fast as a sparse matrix-vector product on \emph{linear elements} on Cascade Lake.
-This means we can apply this matrix-free variant to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
+Nonetheless, the ``tensor4'' strategy with $p\geq 2$ is considerably faster than matrix-based strategies:
+For $p=4$ in 3D, the matrix-free variant is three times as fast as a sparse matrix-vector product on \emph{linear elements} on Cascade Lake.
+This means we can apply this matrix-free variant to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor
+in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
 }
@@
 As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
-Figures \ref{fig:benchmark_miehe_CSL_memory2}, \ref{fig:benchmark_miehe_CSL_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point (``tensor4''), the matrix-free approach takes much less memory than its matrix-based counterpart.
-{\color{red}
-In those graphs, we report the overall memory needed to cache additional quantities required for the tangent operator according to Table~\ref{tab:memory}, as well as all metadata required by
+Figures \ref{fig:benchmark_miehe_CSL_memory2}, \ref{fig:benchmark_miehe_CSL_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}
+demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point (``tensor4''), the matrix-free approach takes much less memory than its matrix-based counterpart.
+{\color{red}
+In those graphs, we report the overall memory needed to cache additional quantities required for the tangent operator according to Table~\ref{tab:memory},
+as well as all metadata required by
 the \texttt{deal.II} \textit{MatrixFree} object.
@@
 Finally, we analyze the node-level performance of the two approaches using the roofline performance model \cite{Williams2009}.
-The achieved memory throughput (MBytes/s) and the number of floating point operations executed per second (MFLOP/s) are measured using the \texttt{MEM\_DP} group of the LIKWID \cite{likwid} tool, version 4.2.1.
-The measurements are done on a single socket of the Cascade Lake system by pinning all 20 MPI processes to this socket, whose arithmetic peak performance at 2.0 GHz is $1280\, \text{GFlop/s}$ and measured memory bandwidth for pure read tasks is $120\, \text{GB/s}$ as well as $90\, \text{GB/s}$ for mixed data access of two loads per one store.
+The achieved memory throughput (MBytes/s) and the number of floating point operations executed per second (MFLOP/s)
+are measured using the \texttt{MEM\_DP} group of the LIKWID \cite{likwid} tool, version 4.2.1.
+The measurements are done on a single socket of the Cascade Lake system by pinning all 20 MPI processes to this socket,
+whose arithmetic peak performance at 2.0 GHz is $1280\, \text{GFlop/s}$ and measured memory bandwidth for pure read tasks is
+$120\, \text{GB/s}$ as well as $90\, \text{GB/s}$ for mixed data access of two loads per one store.
 
@@
 
-The results of the roofline analysis in Figure~\ref{fig:roofline} show that the matrix-based approach is fully memory-bandwidth bound with a very low arithmetic intensity of 0.16 and a performance of $\pgfmathprintnumber{17.36}\,\rm{GFlop/s}$ (1.4\% of the arithmetic peak) averaged over the 2D results.
+The results of the roofline analysis in Figure~\ref{fig:roofline} show that the matrix-based approach is fully memory-bandwidth bound with a
+very low arithmetic intensity of 0.16 and a performance of $\pgfmathprintnumber{17.36}\,\rm{GFlop/s}$ (1.4\% of the arithmetic peak) averaged over the 2D results.
 {\color{red}
@@
 {\color{red}
-The roofline model also explains the difference in performance between the ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) and the ``tensor2'' caching strategy (Algorithm \ref{alg:mf_tensor2}): Both algorithms essentially run at the memory limit of around $100\, \text{GB/s}$, but the former involves a considerably higher memory transfer according to Table~\ref{tab:memory}.
-The ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates the highest computational intensity, but remains close to the memory bandwidth limit with more than $80\, \text{GB/s}$ except for the $p=2$ case.
+The roofline model also explains the difference in performance between the ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) and the ``tensor2'' caching strategy (Algorithm \ref{alg:mf_tensor2}):
+Both algorithms essentially run at the memory limit of around $100\, \text{GB/s}$, but the former involves a considerably higher memory transfer according to Table~\ref{tab:memory}.
+The ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates the highest computational intensity, but remains close to the memory bandwidth limit
+with more than $80\, \text{GB/s}$ except for the $p=2$ case.
 The higher arithmetic load is not surprising as at each quadrature point we need to re-evaluate the Kirchhoff stress $\gz \tau$.
-Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations as listed in Table~\ref{tab:memory}, no advantage of the scalar caching can be deduced despite the higher arithmetic performance.
+Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations as listed in Table~\ref{tab:memory},
+no advantage of the scalar caching can be deduced despite the higher arithmetic performance.
 }
@@
 {\color{red}
-Given that the measurements are within the cell loop and there is an extensive overlap between the memory transfer (due to hardware prefetching) and arithmetic operations, the data of Figure~\ref{fig:breakdown} is derived from measuring the cost of `sum factorization' by comparing the timings of part (i) and (ii) together against part (i) only.
-From Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4}, it becomes clear that sum factorization is cheap enough to be almost completely hidden behind the memory transfer.
+Given that the measurements are within the cell loop and there is an extensive overlap between the memory transfer (due to hardware prefetching) and arithmetic operations,
+the data of Figure~\ref{fig:breakdown} is derived from measuring the cost of `sum factorization' by comparing the timings of
+part (i) and (ii) together against part (i) only.
+From Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4}, it becomes clear that sum factorization is cheap enough
+to be almost completely hidden behind the memory transfer.
 The majority of time is spent within the quadrature loop for all variants, with an increasing proportion for increasing degrees.
-Note also that higher degrees achieve a higher throughput in terms of DoF/s in Figures~\ref{fig:benchmark_miehe_CSL} and~\ref{fig:benchmark_miehe_IWR}, as the memory access of quadrature point data scales as $(p+1)^d/p^d$ compared to the number of unknowns.
+Note also that higher degrees achieve a higher throughput in terms of DoF/s in Figures~\ref{fig:benchmark_miehe_CSL} and~\ref{fig:benchmark_miehe_IWR},
+as the memory access of quadrature point data scales as $(p+1)^d/p^d$ compared to the number of unknowns.
 }
@@
 \hline
-2& \pgfmathprintnumber{0.67518} & \pgfmathprintnumber{2.1798691} & \pgfmathprintnumber{0.04811} & \pgfmathprintnumber{31.270185} & \pgfmathprintnumber{14.0340885471} & \pgfmathprintnumber{0.21837} & \pgfmathprintnumber{6.6211982} & \pgfmathprintnumber{3.09190822915} & \pgfmathprintnumber{0.02194} & \pgfmathprintnumber{67.8161088} & \pgfmathprintnumber{30.773928897} \\
-4& \pgfmathprintnumber{0.35754} & \pgfmathprintnumber{3.6114434} & \pgfmathprintnumber{0.02991} & \pgfmathprintnumber{43.3813195} & \pgfmathprintnumber{11.9538615848} & \pgfmathprintnumber{0.11568} & \pgfmathprintnumber{11.0068739} & \pgfmathprintnumber{3.09076763485} & \pgfmathprintnumber{0.01438} & \pgfmathprintnumber{89.247277} & \pgfmathprintnumber{24.8636995828} \\
-6& \pgfmathprintnumber{0.17724} & \pgfmathprintnumber{4.4013124} & \pgfmathprintnumber{0.01489} & \pgfmathprintnumber{52.73939} & \pgfmathprintnumber{11.9032907992} & \pgfmathprintnumber{0.05934} & \pgfmathprintnumber{12.9981312} & \pgfmathprintnumber{2.9868554095} & \pgfmathprintnumber{0.00732} & \pgfmathprintnumber{106.9726898} & \pgfmathprintnumber{24.2131147541} \\
+\input{parallelization_2d_tensor2.tex}
 \hline
@@
 \hline
-2& \pgfmathprintnumber{0.62536} & \pgfmathprintnumber{2.623296} & \pgfmathprintnumber{0.04582} & \pgfmathprintnumber{36.1519843} & \pgfmathprintnumber{13.6481885639} & \pgfmathprintnumber{0.22937} & \pgfmathprintnumber{6.925918} & \pgfmathprintnumber{2.72642455421} & \pgfmathprintnumber{0.02532} & \pgfmathprintnumber{63.6564648} & \pgfmathprintnumber{24.6982622433} \\
-4& \pgfmathprintnumber{0.36549} & \pgfmathprintnumber{3.8534134} & \pgfmathprintnumber{0.03113} & \pgfmathprintnumber{45.7061894} & \pgfmathprintnumber{11.7407645358} & \pgfmathprintnumber{0.14346} & \pgfmathprintnumber{9.5663639} & \pgfmathprintnumber{2.54767879548} & \pgfmathprintnumber{0.01862} & \pgfmathprintnumber{74.2778413} & \pgfmathprintnumber{19.6288936627} \\
-6& \pgfmathprintnumber{0.18374} & \pgfmathprintnumber{4.5579733} & \pgfmathprintnumber{0.01572} & \pgfmathprintnumber{54.0989777} & \pgfmathprintnumber{11.6882951654} & \pgfmathprintnumber{0.07388} & \pgfmathprintnumber{11.097412} & \pgfmathprintnumber{2.4870059556} & \pgfmathprintnumber{0.00946} & \pgfmathprintnumber{88.2340296} & \pgfmathprintnumber{19.422832981} \\
+\input{parallelization_2d_tensor4.tex}
   \hline
@@
 \hline
-2& \pgfmathprintnumber{0.33065} & \pgfmathprintnumber{4.341083} & \pgfmathprintnumber{0.02814} & \pgfmathprintnumber{51.7484607} & \pgfmathprintnumber{11.750177683} & \pgfmathprintnumber{0.12258} & \pgfmathprintnumber{11.7249488} & \pgfmathprintnumber{2.6974220917} & \pgfmathprintnumber{0.01317} & \pgfmathprintnumber{112.4745656} & \pgfmathprintnumber{25.106302202} \\
-4& \pgfmathprintnumber{0.18685} & \pgfmathprintnumber{5.461242} & \pgfmathprintnumber{0.01589} & \pgfmathprintnumber{64.7662171} & \pgfmathprintnumber{11.7589679043} & \pgfmathprintnumber{0.0695} & \pgfmathprintnumber{14.7126397} & \pgfmathprintnumber{2.68848920863} & \pgfmathprintnumber{0.00803} & \pgfmathprintnumber{129.6590805} & \pgfmathprintnumber{23.2689912827} \\
+\input{parallelization_3d_tensor2.tex}
   \hline
@@
   \hline
-2& \pgfmathprintnumber{0.38409} & \pgfmathprintnumber{4.4971589} & \pgfmathprintnumber{0.03472} & \pgfmathprintnumber{51.9008095} & \pgfmathprintnumber{11.0625} & \pgfmathprintnumber{0.22335} & \pgfmathprintnumber{7.7420018} & \pgfmathprintnumber{1.719677636} & \pgfmathprintnumber{0.0243} & \pgfmathprintnumber{72.555647} & \pgfmathprintnumber{15.8061728395} \\
-4& \pgfmathprintnumber{0.22053} & \pgfmathprintnumber{5.393482} & \pgfmathprintnumber{0.01898} & \pgfmathprintnumber{63.1076193} & \pgfmathprintnumber{11.6190727081} & \pgfmathprintnumber{0.12765} & \pgfmathprintnumber{9.3340763} & \pgfmathprintnumber{1.72761457109} & \pgfmathprintnumber{0.01464} & \pgfmathprintnumber{82.3893605} & \pgfmathprintnumber{15.0635245902} \\
+\input{parallelization_3d_tensor4.tex}
 \hline
@@
 When going to the parallel case with 20 cores, the less-than-optimal speedup can be attributed to two different effects.
-For the auto-vectorized case, the different (turbo) clock frequencies permit a speedup of a factor 12.6 at most, close to our results\footnote{Note that we intentionally run with Turbo mode on to permit the processor to come close to the power limit for all settings and present more realistic performance limits.}.
-For the SIMD+MPI case, the main reason is the memory bandwidth: As can be seen from Figures~\ref{fig:roofline_2d} and~\ref{fig:roofline_3d}, both algorithms are eventually memory-bandwidth limited.
-Since Algorithm~\ref{alg:mf_tensor4} has a lower arithmetic intensity, the speedup over the serial, auto-vectorized case is less than that of Algorithm~\ref{alg:mf_tensor2}.
+For the auto-vectorized case, the different (turbo) clock frequencies permit a speedup of a factor 12.6 at most, close to our
+results\footnote{Note that we intentionally run with Turbo mode on to permit the processor to come close to the power limit for all settings and present more realistic performance limits.}.
+For the SIMD+MPI case, the main reason is the memory bandwidth: As can be seen from Figures~\ref{fig:roofline_2d} and~\ref{fig:roofline_3d},
+both algorithms are eventually memory-bandwidth limited. Since Algorithm~\ref{alg:mf_tensor4} has a lower arithmetic intensity,
+the speedup over the serial, auto-vectorized case is less than that of Algorithm~\ref{alg:mf_tensor2}.
 }
@@
 {\color{red}
-The first one is the higher algorithmic intensity and lower memory access as compared to the matrix-based implementation, see Figure \ref{fig:roofline}.
-Since both variants are in the memory-limited regime, the matrix-free implementation with its lower memory access is beneficial.
+The first one is the higher algorithmic intensity and lower memory access as compared to the matrix-based implementation,
+see Figure \ref{fig:roofline}. Since both variants are in the memory-limited regime, the matrix-free implementation with its lower memory access is beneficial.
 }
@@
 Section \ref{sec:mf} and Section 2.4 of \cite{kronbichler12}.
-Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach, whereas Algorithm~\ref{alg:mf_tensor2} is faster due to a higher arithmetic intensity.
+Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach,
+whereas Algorithm~\ref{alg:mf_tensor2} is faster due to a higher arithmetic intensity.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{CSL_Munich_cg2d.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_cg2d.pdf}
     \caption{CG iterations (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{CSL_Munich_cg3d.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_cg3d.pdf}
     \caption{CG iterations (3D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{CSL_Munich_solver2d.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_solver2d.pdf}
     \caption{CG solution time (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{CSL_Munich_solver3d.eps}
+    \includegraphics[width=\textwidth]{CSL_Munich_solver3d.pdf}
     \caption{CG solution time (3D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_newest_patched_cg2d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_cg2d.pdf}
     \caption{CG iterations (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_newest_patched_cg3d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_cg3d.pdf}
     \caption{CG iterations (3D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_newest_patched_solver2d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_solver2d.pdf}
     \caption{CG solution time (2D)}
@@
     \centering
-    \includegraphics[width=\textwidth]{IWR_newest_patched_solver3d.eps}
+    \includegraphics[width=\textwidth]{IWR_newest_patched_solver3d.pdf}
     \caption{CG solution time (3D)}
@@
 requires many more iterations for convergence.
-Consequently, we can conclude that (i) the selected smoother is suitable for the present finite strain elasticity problem, and that (ii) although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires fewer CG iterations as compared to the algebraic multigrid baseline.
+Consequently, we can conclude that (i) the selected smoother is suitable for the present finite strain elasticity problem, and that
+(ii) although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$,
+they still result in a multigrid preconditioner which requires fewer CG iterations as compared to the algebraic multigrid baseline.
 
@@
 We note specifically that the comparison is restricted to the choice of level operators used with the pre-existing GMG framework, and not the framework itself.
-Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3} confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for all polynomial degrees.
+Figures \ref{fig:benchmark_miehe_Emmy_sol2}, \ref{fig:benchmark_miehe_Emmy_sol3}, \ref{fig:benchmark_miehe_IWR_sol2} and \ref{fig:benchmark_miehe_IWR_sol3}
+confirm that the efficiency of the proposed GMG preconditioner translates into faster solution times for all polynomial degrees.
 Most importantly, the wall time per DoF for the matrix-free solver with GMG preconditioner stays almost constant both for 2D and 3D problems.
 In comparison, the wall time of the matrix-based iterative solver grows rapidly with increasing polynomial degree.
-Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products, studied in the previous section.
+Clearly this is related both to the performance of the preconditioner from the linear algebra perspective as well as the matrix-vector products,
+studied in the previous section.
 Based on these results, we conclude that the matrix-free approach represents a very competitive solution strategy for engineering problems
@@
 However, we have performed a preliminary study of the weak scalability of the implementation.
-{\color{red}Tables~\ref{tab:weak_2d} and \ref{tab:weak_3d} demonstrate a good weak scaling up to 320 cores, i.e., 8 nodes of dual-socket Cascade Lake, of the wall time per matrix-vector product and per iteration in the CG solver for the 2D problem with $p=4$ and the 3D problem with a tri-quadratic basis.}
+{\color{red}Tables~\ref{tab:weak_2d} and \ref{tab:weak_3d} demonstrate a good weak scaling up to 320 cores, i.e., 8 nodes of dual-socket Cascade Lake,
+of the wall time per matrix-vector product and per iteration in the CG solver for the 2D problem with $p=4$ and the 3D problem with a tri-quadratic basis.}
 This confirms that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for cluster-sized problems as well,
@@
 Among the three examined algorithms,
-{\color{red}the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the spatial configuration is faster than caching only a scalar at quadrature points or caching the material part} of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress.
+{\color{red}the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the spatial configuration is faster than
+caching only a scalar at quadrature points or caching the material part} of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress.
 {\color{red}
 Given that we did not utilize the specific form of the material part of the stiffness tensor in the latter case, we conclude
-that the matrix-free implementation of the tangent operator is somewhat slower but more flexible because it can be applied to any constitutive model which operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
+that the matrix-free implementation of the tangent operator is somewhat slower but more flexible because it can be applied to any constitutive model which
+operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
 }
@@
 {\color{red}
-Despite the higher arithmetic intensity, the matrix-free implementations are still memory bandwidth-limited and in particular by the data required at quadrature points in terms of fourth-order and second-order symmetric tensors.
+Despite the higher arithmetic intensity, the matrix-free implementations are still memory bandwidth-limited and in particular by the data
+required at quadrature points in terms of fourth-order and second-order symmetric tensors.
 }
@@
 This formulation can also be straight forwardly adopted within the matrix-free framework of the \texttt{deal.II} library.
-To that end a variation of the Algorithm \ref{alg:mf_tensor4} can be used where the gradients are evaluated with respect to the undeformed configuration and only a fourth-order referential tangent tensor ${\rm D}_{\gz F}{\gz P}$ is cached per quadrature point.
+To that end a variation of the Algorithm \ref{alg:mf_tensor4} can be used where the gradients are evaluated with respect to the undeformed configuration
+and only a fourth-order referential tangent tensor ${\rm D}_{\gz F}{\gz P}$ is cached per quadrature point.
 }
@@
 however we do support it in our code.
-As compared to Algorithm~\ref{alg:mf_tensor4}, we need to cache more data, which results in both higher memory requirement per DoF as well as a lower arithmetic intensity of the operator evaluation.
-On both architectures we recorded a slightly larger wall time per DoF as compared to the Algorithm \ref{alg:mf_tensor4}, however still much faster than the matrix-based approach.
+As compared to Algorithm~\ref{alg:mf_tensor4}, we need to cache more data, which results in both higher memory requirement per DoF
+as well as a lower arithmetic intensity of the operator evaluation.
+On both architectures we recorded a slightly larger wall time per DoF as compared to the Algorithm \ref{alg:mf_tensor4},
+however still much faster than the matrix-based approach.
 From the LIKWID measurements we observed that this algorithm also results in lower performance on Cascade Lake.
-The advantage of this algorithm is that it does not require evaluation of gradients in the deformed configuration, which leads to a slightly lower setup time as compared to the other three algorithms that rely on \texttt{MappingQEulerian} class of the \texttt{deal.II} library.
+The advantage of this algorithm is that it does not require evaluation of gradients in the deformed configuration, which leads to a slightly lower setup time
+as compared to the other three algorithms that rely on \texttt{MappingQEulerian} class of the \texttt{deal.II} library.
 Overall we consider this approach to be a valid alternative to the Algorithm \ref{alg:mf_tensor4}
@@
 In this setting, all quantities are evaluated in referential coordinates, necessitating only a single set of metric terms according to Table~\ref{tab:memory}.
-For the derivatives of the current solution an additional multiplication by $\gz F^{-T}$ for both the solution as well as the test function is necessary, with $\gz F^{-T}$ evaluated on the fly after line 10 in Algorithm~\ref{alg:mf_scalar}.
-Despite the additional evaluations, this leads to a faster calculation than Algorithm~\ref{alg:mf_scalar} on Cascade Lake, running with 162 GFlop/s and 68 GB/s for $p=6$ in 2D or 225 GFlop/s and 59 GB/s for $p=4$ in 3D.
+For the derivatives of the current solution an additional multiplication by $\gz F^{-T}$ for both the solution as well as the test function is necessary,
+with $\gz F^{-T}$ evaluated on the fly after line 10 in Algorithm~\ref{alg:mf_scalar}.
+Despite the additional evaluations, this leads to a faster calculation than Algorithm~\ref{alg:mf_scalar} on Cascade Lake,
+running with 162 GFlop/s and 68 GB/s for $p=6$ in 2D or 225 GFlop/s and 59 GB/s for $p=4$ in 3D.
 However, execution is still slower than Algorithm~\ref{alg:mf_tensor2}, despite the lower memory access.

### 2019-08-14 Calculations/post_process.py

@@ -1,17 +1,12 @@
 import re
 import os
 import argparse
-from matplotlib.pyplot import figure, show
+from matplotlib.pyplot import figure
 import matplotlib.pyplot as plt
 import matplotlib as mp
 import numpy as np
 import fileinput
 from matplotlib.ticker import MaxNLocator
-
-def remove_creation_date(file_name):
-    for line in fileinput.input(file_name, inplace=True):
-        if not 'CreationDate' in line:
-            print line,
 
 
 # define command line arguments
@@ -260,10 +255,27 @@ mem3d_t4_ns = [tup[3]/tup[1] for tup in mf3d_data_tensor4_ns]
 # file location
 fig_prefix = os.path.join(os.getcwd(), '../doc/' + os.path.basename(os.path.normpath(prefix)) + '_')
 
-params = {'legend.fontsize': 20,
-          'font.size': 20}
+#
+#
+#  MATPLOTLIB
+#
+#
+
+# 6.1:
+params   = {'legend.fontsize': 16,
+            'font.size': 16}
+# 6.2:
+params2 = {'legend.fontsize': 14,
+           'font.size': 16}
+
 plt.rcParams.update(params)
 
+#
+# Plot Section 6.1
+#
+
+ax = plt.figure().gca()
+ax.xaxis.set_major_locator(MaxNLocator(integer=True))
 plt.plot(deg2d,time2d_tr, 'rs--', label='Trilinos')
 plt.plot(deg2d,time2d_sc, 'bo--', label='MF scalar')
 plt.plot(deg2d,time2d_t2, 'g^--', label='MF tensor2')
@@ -271,9 +283,8 @@ plt.plot(deg2d,time2d_t4, 'cv--', label='MF tensor4')
 #plt.plot(deg2d,time2d_t4_ns, 'mD--', label='MF tensor4 P')
 plt.xlabel('polynomial degree')
 plt.ylabel('vmult wall time (s) / DoF')
-#leg = plt.legend(loc='best', ncol=1)
-plt.savefig(fig_prefix + 'timing2d-nolables.eps', format='eps', bbox_inches = 'tight')
-remove_creation_date(fig_prefix + 'timing2d-nolables.eps')
+leg = plt.legend(loc='best', ncol=1)
[…]

### 2019-08-15 Calculations/post_process.py
fix y ticks

@@ -6,8 +6,22 @@ import matplotlib.pyplot as plt
 import matplotlib as mp
 import numpy as np
 import fileinput
+import matplotlib.ticker
 from matplotlib.ticker import MaxNLocator
+from matplotlib.ticker import FormatStrFormatter
 
+# https://stackoverflow.com/questions/42656139/set-scientific-notation-with-fixed-exponent-and-significant-digits-for-multiple
+class OOMFormatter(matplotlib.ticker.ScalarFormatter):
+    def __init__(self, order=0, fformat="%1.1f", offset=True, mathText=True):
+        self.oom = order
+        self.fformat = fformat
+        matplotlib.ticker.ScalarFormatter.__init__(self,useOffset=offset,useMathText=mathText)
+    def _set_orderOfMagnitude(self, nothing):
+        self.orderOfMagnitude = self.oom
+    def _set_format(self, vmin, vmax):
+        self.format = self.fformat
+        if self._useMathText:
+            self.format = '$%s$' % matplotlib.ticker._mathdefault(self.format)
 
 # define command line arguments
 parser = argparse.ArgumentParser(
@@ -273,8 +287,12 @@ plt.rcParams.update(params)
 #
 # Plot Section 6.1
 #
+ratio = 0.7
+timing_exp = -8 if "CSL" in args.prefix else -7
 
 ax = plt.figure().gca()
+ax.yaxis.set_major_formatter(OOMFormatter(timing_exp, "%1.1f"))
+ax.ticklabel_format(axis='y', style='sci', scilimits=(0,0))
 ax.xaxis.set_major_locator(MaxNLocator(integer=True))
 plt.plot(deg2d,time2d_tr, 'rs--', label='Trilinos')
 plt.plot(deg2d,time2d_sc, 'bo--', label='MF scalar')
@@ -284,12 +302,15 @@ plt.plot(deg2d,time2d_t4, 'cv--', label='MF tensor4')
 plt.xlabel('polynomial degree')
 plt.ylabel('vmult wall time (s) / DoF')
 leg = plt.legend(loc='best', ncol=1)
+ax.set_aspect(1.0/ax.get_data_ratio()*ratio)
[…]

### 2019-08-16 doc/large_strain.tex
fix post_process_weak.py and use it for Weak scaling tables, no changes, just peace of mind
consistently use: referential and current configurations (based on Prof. Steinmann corrections in other manuscripts)

@@
 For matrix-free schemes using fast integration via sum factorization, the main challenge is to find a representation of the linearization point,
-either in referential or spatial configurations, at the quadrature points that is cheap enough to be evaluated on the fly.
+either in referential or current configurations, at the quadrature points that is cheap enough to be evaluated on the fly.
 }
@@
 \subsection{Weak form}
-The deformation of a body $\mcl B$ from the referential configuration $\mcl B_0$ to the spatial configuration $\mcl B_t$ at time $t$
+The deformation of a body $\mcl B$ from the referential configuration $\mcl B_0$ to the {\color{red}current} configuration $\mcl B_t$ at time $t$
 is defined via the deformation map $\gz \varphi_t: \mcl B_0 \rightarrow \mcl B_t$, which places material points of $\mcl B$ into the Euclidean space $\mathbb E^3$.
@@
 {\color{red}
-Here, `$\grad$' denotes the gradient with respect to the spatial configuration, as opposed to the gradient in the referential setting `$\Grad$'.
+Here, `$\grad$' denotes the gradient with respect to the current configuration, as opposed to the gradient in the referential setting `$\Grad$'.
 }
@@
 and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
-Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the spatial configuration,
+Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the {\color{red}current} configuration,
 whereas the integration is done in the referential configuration.
@@
 %
-In order to obtain the gradients with respect to the spatial configuration,
+In order to obtain the gradients with respect to the {\color{red}current} configuration,
 which appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem derived with a Lagrangian formulation,
@@
 In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping
-to the spatial configuration given a displacement field.
-%
-Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the spatial configuration.
+to the {\color{red}current} configuration given a displacement field.
+%
+Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the {\color{red}current} configuration.
 Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.
@@
 quadrature point. The evaluation of the gradients in the referential
-and spatial configuration also involves a data access via the so-called metric
+and current configuration also involves a data access via the so-called metric
 terms, the inverse of the Jacobian from the unit element to the referential or
-spatial configurations, respectively. Algorithm~\ref{alg:mf_scalar} involves
-both the Jacobian in the referential and the spatial configuration, whereas the other
+current configurations, respectively. Algorithm~\ref{alg:mf_scalar} involves
+both the Jacobian in the referential and the current configuration, whereas the other
 two algorithms only involve the spatial (Eulerian) terms. Finally, the
@@
 These deformed terms are the result of manifold descriptions of the boundaries and interfaces  to capture the geometry as well as
-the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration.
+the linearization-related mapping required for evaluation of gradients with respect to the current configuration.
 }
@@
 is the fastest approach in both 2D and 3D and on both systems.
-When compared to the scalar caching implementation, the ``tensor2'' variant loads slightly less data, because only the spatial configuration is needed for the metric terms,
+When compared to the scalar caching implementation, the ``tensor2'' variant loads slightly less data, because only the current configuration is needed for the metric terms,
 as compared to both the spatial and the referential configuration in the scalar caching case in order to evaluate the Kirchhoff stress $\gz \tau$.
@@
 The higher arithmetic load is not surprising as at each quadrature point we need to re-evaluate the Kirchhoff stress $\gz \tau$.
-Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations as listed in Table~\ref{tab:memory},
+Given that in this case we need to evaluate gradients with respect to both referential and current configurations as listed in Table~\ref{tab:memory},
 no advantage of the scalar caching can be deduced despite the higher arithmetic performance.
@@
   \hline
-  20 & 2,887,680 & 5 & \pgfmathprintnumber{0.0033900} & \pgfmathprintnumber{0.045574} \\
-  80 & 11,542,528 & 6 & \pgfmathprintnumber{0.0035350} & \pgfmathprintnumber{0.047788} \\
-  320 & 46,153,728 & 7 & \pgfmathprintnumber{0.0037879} & \pgfmathprintnumber{0.052012} \\
+  \input{weak_scaling_2d.tex}
   \hline
@@
   \hline
-  40 & 4,442,880 & 3 & \pgfmathprintnumber{0.0070667} & \pgfmathprintnumber{0.064328} \\
-  320 & 35,071,488 & 4 & \pgfmathprintnumber{0.0075333} & \pgfmathprintnumber{0.071448} \\
+  \input{weak_scaling_3d.tex}
   \hline
@@
 Among the three examined algorithms,
-{\color{red}the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the spatial configuration is faster than
+{\color{red}the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the current configuration is faster than
 caching only a scalar at quadrature points or caching the material part} of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress.
@@
 This formulation can also be straight forwardly adopted within the matrix-free framework of the \texttt{deal.II} library.
-To that end a variation of the Algorithm \ref{alg:mf_tensor4} can be used where the gradients are evaluated with respect to the undeformed configuration
-and only a fourth-order referential tangent tensor ${\rm D}_{\gz F}{\gz P}$ is cached per quadrature point.
+To that end a variation of the Algorithm \ref{alg:mf_tensor4} can be used where the gradients are evaluated with respect to the referential configuration
+and only a fourth-order tangent tensor ${\rm D}_{\gz F}{\gz P}$ is cached per quadrature point.
 }
@@
 From the LIKWID measurements we observed that this algorithm also results in lower performance on Cascade Lake.
-The advantage of this algorithm is that it does not require evaluation of gradients in the deformed configuration, which leads to a slightly lower setup time
+The advantage of this algorithm is that it does not require evaluation of gradients in the current configuration, which leads to a slightly lower setup time
 as compared to the other three algorithms that rely on \texttt{MappingQEulerian} class of the \texttt{deal.II} library.

### 2019-08-16 doc/large_strain.tex
cite 9.1 and add a link to performance improvements

@@
 pre-computation with a higher memory access in terms of the representation of the linearization point.
-The \texttt{deal.II} finite element library \cite{dealII90}, which provides MPI parallelization together with a generalized matrix-free framework
+The \texttt{deal.II} finite element library \cite{dealII91}, which provides MPI parallelization together with a generalized matrix-free framework
 that incorporates SIMD vectorization and an interface to high-performance MPI-distributed matrix-based libraries, is used for this study.
@@
 which appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem derived with a Lagrangian formulation,
-we use the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
+we use the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII91} library.
 In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping
@@
 Note that the SIMD vectorization
-in \texttt{deal.II} \cite{dealII90}
+in \texttt{deal.II} \cite{dealII91}
 is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
@@
 The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in
-Section \ref{sec: Constitutive modelling}, has been implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
+Section \ref{sec: Constitutive modelling}, has been implemented using\footnote{{\color{red}Including additional performance improvements for \texttt{Tensor} classes \url{https://github.com/dealii/dealii/pull/8428}.}} \texttt{deal.II} version 9.1 \cite{dealII91}, an open-source finite element library.
 This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the

### 2019-08-16 doc/large_strain.tex
cite Lukacz

@@
 with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial.
-One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}.
+One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007, Kaczmarczyk2010}.
 Such operators are, however, computationally very demanding.

### 2019-08-16 doc/large_strain.tex
caption above tabular
combine two weak tables

@@
 However, we have performed a preliminary study of the weak scalability of the implementation.
-{\color{red}Tables~\ref{tab:weak_2d} and \ref{tab:weak_3d} demonstrate a good weak scaling up to 320 cores, i.e., 8 nodes of dual-socket Cascade Lake,
+{\color{red}Table~\ref{tab:weak} demonstrate a good weak scaling up to 320 cores, i.e., 8 nodes of dual-socket Cascade Lake,
 of the wall time per matrix-vector product and per iteration in the CG solver for the 2D problem with $p=4$ and the 3D problem with a tri-quadratic basis.}
@@
 
-\begin{table}
-  \centering
+\begin{table}[!htb]
+  \centering
+  \caption{Weak scaling of Algorithm \ref{alg:mf_tensor2}.}
+  \label{tab:weak}
+  \begin{subtable}{.49\linewidth}
+  \centering
+  \caption{2D and quartic polynomial basis}
+  \resizebox{\textwidth}{!}{
   \begin{tabular}{|r|r|c|r|c|}
@@
   \end{tabular}
-  \caption{Weak scaling of Algorithm \ref{alg:mf_tensor2} in 2D for a quartic polynomial basis.}
-  \label{tab:weak_2d}
-\end{table}
-
-\begin{table}
-  \centering
+  }
+  \end{subtable}
+  \begin{subtable}{.49\linewidth}
+  \centering
+  \caption{3D and quadratic polynomial basis}
+  \resizebox{\textwidth}{!}{
   \begin{tabular}{|r|r|c|r|c|}
@@
   \end{tabular}
-  \caption{Weak scaling of Algorithm \ref{alg:mf_tensor2} in 3D for a quadratic polynomial basis.}
-  \label{tab:weak_3d}
+  }
+  \end{subtable}
 \end{table}

### 2019-09-15 doc/large_strain.tex
comment out DOE footnote

@@
   \author[b,c]{Daniel Arndt
-   \footnotetext{%
-   This manuscript has been authored by UT-Battelle, LLC under Contract No.
-   DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government
-   retains and the publisher, by accepting the article for publication, acknowledges
-   that the United States Government retains a non-exclusive, paid-up, irrevocable,
-   worldwide license to publish or reproduce the published form of this manuscript,
-   or allow others to do so, for United States Government purposes. The Department
-   of Energy will provide public access to these results of federally sponsored
-   research in accordance with the DOE Public Access Plan
-   (http://energy.gov/downloads/doe-public-access-plan).}}
+  %  \footnotetext{%
+  %  This manuscript has been authored by UT-Battelle, LLC under Contract No.
+  %  DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government
+  %  retains and the publisher, by accepting the article for publication, acknowledges
+  %  that the United States Government retains a non-exclusive, paid-up, irrevocable,
+  %  worldwide license to publish or reproduce the published form of this manuscript,
+  %  or allow others to do so, for United States Government purposes. The Department
+  %  of Energy will provide public access to these results of federally sponsored
+  %  research in accordance with the DOE Public Access Plan
+  %  (http://energy.gov/downloads/doe-public-access-plan).}
+  }
   \ead{arndtd@ornl.gov}
@@
     We {\color{red}find} that matrix-free methods for finite-strain solid mechanics are very promising, {\color{red}outperforming linear matrix-based schemes by two to five times},
-    and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}\footnotetext{%
-   This manuscript has been authored by UT-Battelle, LLC under Contract No.
-   DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government
-   retains and the publisher, by accepting the article for publication, acknowledges
-   that the United States Government retains a non-exclusive, paid-up, irrevocable,
-   worldwide license to publish or reproduce the published form of this manuscript,
-   or allow others to do so, for United States Government purposes. The Department
-   of Energy will provide public access to these results of federally sponsored
-   research in accordance with the DOE Public Access Plan
-   (http://energy.gov/downloads/doe-public-access-plan).}
+    and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
+  %   \footnotetext{%
+  %  This manuscript has been authored by UT-Battelle, LLC under Contract No.
+  %  DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government
+  %  retains and the publisher, by accepting the article for publication, acknowledges
+  %  that the United States Government retains a non-exclusive, paid-up, irrevocable,
+  %  worldwide license to publish or reproduce the published form of this manuscript,
+  %  or allow others to do so, for United States Government purposes. The Department
+  %  of Energy will provide public access to these results of federally sponsored
+  %  research in accordance with the DOE Public Access Plan
+  %  (http://energy.gov/downloads/doe-public-access-plan).}
   \end{abstract}

### 2019-09-15 doc/large_strain.tex
incorporate Prof. Steinmann's edits

@@
 {\color{red}
-For matrix-free schemes using fast integration via sum factorization, the main challenge is to find a representation of the linearization point,
-either in referential or current configurations, at the quadrature points that is cheap enough to be evaluated on the fly.
+For matrix-free schemes using fast integration via sum factorization, the main challenge is to find a representation of the tangent operator
+at the quadrature points,
+either in the referential or current configuration,
+that is cheap enough to be evaluated on the fly.
 }
@@
 Particular emphasis is on the tradeoff between computations at quadrature points versus
-pre-computation with a higher memory access in terms of the representation of the linearization point.
+pre-computation with a higher memory access in terms of the representation of the tangent operator.
 The \texttt{deal.II} finite element library \cite{dealII91}, which provides MPI parallelization together with a generalized matrix-free framework
@@
 %
-The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with
+The performance of the matrix-free tangent operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with
 an algebraic multigrid preconditioner for a representative numerical example of a heterogeneous hyperelastic material simulated in two and three dimensions in Section \ref{sec:example}.
@@
 The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian {\color{red}determinant} $J = \det \gz F > 0$ has to be positive.
-We shall denote the push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by a linear transformation $\chi$ defined as
+We shall denote the (contravariant) push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by a linear transformation $\chi$ defined as
 \begin{align}
@@
 where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the referential configuration,
-$\psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary
+$\psi$ denotes the strain-energy density per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary
 conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.
@@
 }
-Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \psi / \partial \gz E \equiv 2 \partial \psi / \partial \gz C$,
+Note that $\gz \tau$ is the (contravariant) push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \psi / \partial \gz E \equiv 2 \partial \psi / \partial \gz C$,
 that is, $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.
@@
 
-In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
+In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution point (a field) $\overline{\gz u}$ is required such that
 \begin{align}
@@
 Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$,
-and the fourth-order material part of the spatial tangent stiffness tensor is the push forward of the material part of the referential tangent stiffness tensor
+and the fourth-order material part of the spatial tangent stiffness tensor is the (contravariant) push-forward of the material part of the referential tangent stiffness tensor
 $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
@@
 J \boldsymbol{\mathcal{C}} : \left( \bullet \right)
-  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
+  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)^s
   + 2 \lambda \trace\left( \bullet \right) \mathbf{I}.
@@
 \end{align}
-Here, $\gz A$ is the discrete tangent operator, $\gz F$ is the discrete gradient of the potential energy
+Here, $\gz A$ with elements $A_{ij}$ is the discrete tangent operator, $\gz F$ with elements $F_i$ is the discrete gradient of the potential energy
 and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
 Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the {\color{red}current} configuration,
-whereas the integration is done in the referential configuration.
+whereas the integration is performed in the referential configuration.
 
@@
 %
-Note that $x_j \equiv X_{j_1\,j_2}$; the former accesses the element source vector using a linear index $j$,
+Note that here $x_j \equiv X_{j_1\,j_2}$; the former accesses the element source vector using a linear index $j$,
 whereas the latter uses the associated multi-index accesses $\{j_1\,j_2\}$.
@@
 is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
-The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements{\color{red},
+The main reason is that, apart from the point-wise computation of the stresses and corresponding tangent operators, the operations are typically the same on all elements{\color{red},
 as opposed to the operations within an element}\footnote{%
@@
 Algorithm~\ref{alg:mf_scalar} is higher than in the other variants because both
-the source vector $\gz x$ and the linearization point from the current FE
+the source vector $\gz x$ and the current FE
 solution $\overline{\gz u^h}$ need to be accessed, each with the usual indirect
@@
 terms, the inverse of the Jacobian from the unit element to the referential or
-current configurations, respectively. Algorithm~\ref{alg:mf_scalar} involves
+current configuration, respectively. Algorithm~\ref{alg:mf_scalar} involves
 both the Jacobian in the referential and the current configuration, whereas the other
@@
 We define the level operators as follows: Recall that the fine scale tangent operator \eqref{eq:algebraic_tangent} is obtained by
-linearization around the current displacement field $\gz u^h$. We then restrict this displacement field to all multigrid levels $\{l\}$ and
+linearization around the current displacement field $\overline{\gz u^h}$. We then restrict this displacement field to all multigrid levels $\{l\}$ and
 evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
@@
 as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
-We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach with a high polynomial degree.
+We chose a problem size on the order of $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach with a high polynomial degree.
 Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization)
@@
 It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
-as long as the resulting tangent operator is expressed via \eqref{eq:algebraic_tangent}.
+as long as the resulting tangent operator is expressed via the format in \eqref{eq:algebraic_tangent}.
 
@@
 {\color{red}
-In those graphs, we report the overall memory needed to cache additional quantities required for the tangent operator according to Table~\ref{tab:memory},
+In those graphs, we report the overall memory needed to cache additional quantities required for the tangent operator according to Table~\ref{tab:memory}
 as well as all metadata required by
@@
 The higher arithmetic load is not surprising as at each quadrature point we need to re-evaluate the Kirchhoff stress $\gz \tau$.
-Given that in this case we need to evaluate gradients with respect to both referential and current configurations as listed in Table~\ref{tab:memory},
+Given that in this case we need to evaluate gradients with respect to both the referential and the current configuration as listed in Table~\ref{tab:memory},
 no advantage of the scalar caching can be deduced despite the higher arithmetic performance.
@@
 with standard geometric transfer operations between each level.
-The GMG was applied to heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
+The GMG was applied to a heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
 The numerical studies indicate that the proposed preconditioner

### 2019-09-15 doc/large_strain.tex
remove red

@@
   \begin{abstract}[Summary]
-    {{\color{red}This work investigates matrix-free algorithms for problems in quasi-static finite-strain hyperelasticity.
+    {This work investigates matrix-free algorithms for problems in quasi-static finite-strain hyperelasticity.
     Iterative solvers with matrix-free operator evaluation have emerged as
@@
 the limiting factor in classical finite element solvers.
-    Specifically, we study different matrix-free realizations of the finite element tangent operator, and determine whether generalized methods of incorporating complex constitutive behavior might be feasible.}
+    Specifically, we study different matrix-free realizations of the finite element tangent operator, and determine whether generalized methods of incorporating complex constitutive behavior might be feasible.
     In order to improve the convergence behavior of iterative solvers, we also propose a method by which to construct level tangent operators
@@
     The performance of the matrix-free operator and the geometric multigrid preconditioner is compared to the matrix-based implementation with an algebraic multigrid preconditioner on a single node for a representative numerical example of a heterogeneous hyperelastic material in two and three dimensions.
-    We {\color{red}find} that matrix-free methods for finite-strain solid mechanics are very promising, {\color{red}outperforming linear matrix-based schemes by two to five times},
+    We find that matrix-free methods for finite-strain solid mechanics are very promising, outperforming linear matrix-based schemes by two to five times,
     and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
@@
 
-The performance of finite element solvers on modern computer architectures is typically {\color{red} limited by the bandwidth from RAM memory} for sufficiently large problems.
-{\color{red}This is because} loading pre-computed matrix elements from memory into the CPU cores is significantly slower than actually performing the arithmetic operations{\color{red}, dictated by hardware properties}.
+The performance of finite element solvers on modern computer architectures is typically limited by the bandwidth from RAM memory for sufficiently large problems.
+This is because loading pre-computed matrix elements from memory into the CPU cores is significantly slower than actually performing the arithmetic operations, dictated by hardware properties.
 In order to improve the performance of iterative solvers, so-called matrix-free methods,
@@
 Such methods can also be efficiently implemented on GPUs \cite{Abdelfattah2016, ljungkvist2017multigrid}.
-{\color{red}
 Among matrix-free methods, stencil-based variants are most popular for linear elements and tetrahedral shapes~\cite{Gmeiner2016,Bauer2018},
@@
 and tensor product shape functions~\cite{Brown2010,kronbichler12,May2015,Cantwell2011}.
-}
 
@@
 more elaborate as compared to the common operators in linear and non-linear fluid mechanics.
-{\color{red}
 For matrix-free schemes using fast integration via sum factorization, the main challenge is to find a representation of the tangent operator
@@
 that is cheap enough to be evaluated on the fly.
-}
 Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
@@
 
-{\color{red}
 This work aims to fill this gap by investigating different matrix-free representations of the
@@
 and employ them to define a \mbox{geometric} multigrid preconditioner.
-}
 
@@
 In Section \ref{sec:mf} we describe the different matrix-free implementations that are evaluated in this study.
-Section \ref{sec:framework} provides details of the numerical framework {\color{red}this study relies on},
+Section \ref{sec:framework} provides details of the numerical framework this study relies on,
 where we also propose a method by which to construct level tangent operators for
@@
 \subsection{Weak form}
-The deformation of a body $\mcl B$ from the referential configuration $\mcl B_0$ to the {\color{red}current} configuration $\mcl B_t$ at time $t$
+The deformation of a body $\mcl B$ from the referential configuration $\mcl B_0$ to the current configuration $\mcl B_t$ at time $t$
 is defined via the deformation map $\gz \varphi_t: \mcl B_0 \rightarrow \mcl B_t$, which places material points of $\mcl B$ into the Euclidean space $\mathbb E^3$.
@@
 where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called the deformation gradient and $\gz I$ is the second-order unit tensor.
-The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian {\color{red}determinant} $J = \det \gz F > 0$ has to be positive.
+The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian determinant $J = \det \gz F > 0$ has to be positive.
 We shall denote the (contravariant) push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by a linear transformation $\chi$ defined as
@@
 is involved due to the symmetry of $\gz \tau$.
-{\color{red}
 Here, `$\grad$' denotes the gradient with respect to the current configuration, as opposed to the gradient in the referential setting `$\Grad$'.
-}
 Note that $\gz \tau$ is the (contravariant) push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \psi / \partial \gz E \equiv 2 \partial \psi / \partial \gz C$,
@@
 and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
-Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the {\color{red}current} configuration,
+Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the current configuration,
 whereas the integration is performed in the referential configuration.
@@
 where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form at one quadrature point $\gz \xi_q$, $J^K_q$ is the
-{\color{red}determinant of the} Jacobian of the mapping from the isoparametric element to the element $K$ and
+determinant of the Jacobian of the mapping from the isoparametric element to the element $K$ and
 $w_q$ is the quadrature weight.
@@
 \end{align*}
-Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one{\color{red}, $n=p+1$},
+Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, $n=p+1$,
 the evaluation of all of the shape functions ($n^d$)
@@
 %
-In order to obtain the gradients with respect to the {\color{red}current} configuration,
+In order to obtain the gradients with respect to the current configuration,
 which appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem derived with a Lagrangian formulation,
@@
 In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping
-to the {\color{red}current} configuration given a displacement field.
-%
-Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the {\color{red}current} configuration.
+to the current configuration given a displacement field.
+%
+Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the current configuration.
 Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.
@@
 
-In order to avoid {\color{red}the repeated} calculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point.
+In order to avoid the repeated calculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point.
 In this case, we also avoid the need to evaluate gradients of the displacement field with respect to the referential configuration.
@@
 second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
-To that end, we cache two scalars that depend on the Jacobian {\color{red}determinant} $J$.
+To that end, we cache two scalars that depend on the Jacobian determinant $J$.
 
@@
 is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
-The main reason is that, apart from the point-wise computation of the stresses and corresponding tangent operators, the operations are typically the same on all elements{\color{red},
-as opposed to the operations within an element}\footnote{%
+The main reason is that, apart from the point-wise computation of the stresses and corresponding tangent operators, the operations are typically the same on all elements,
+as opposed to the operations within an element\footnote{%
 In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that
@@
 
-{\color{red}
 Table~\ref{tab:memory} summarizes the three algorithms in terms of their implied
@@
 on the mesh via a run-time decision in the code.
-}
 
@@
 The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in
-Section \ref{sec: Constitutive modelling}, has been implemented using\footnote{{\color{red}Including additional performance improvements for \texttt{Tensor} classes \url{https://github.com/dealii/dealii/pull/8428}.}} \texttt{deal.II} version 9.1 \cite{dealII91}, an open-source finite element library.
+Section \ref{sec: Constitutive modelling}, has been implemented using\footnote{Including additional performance improvements for \texttt{Tensor} classes \url{https://github.com/dealii/dealii/pull/8428}.} \texttt{deal.II} version 9.1 \cite{dealII91}, an open-source finite element library.
 This library offers a number of paradigms by which to parallelize a finite-element code, and as a member of the
@@
 we limit ourselves to comparisons made on a node level in this study.
-{\color{red}
 Examinations performed in this manner are most representative for the differences between the matrix-free and matrix-based approaches because problems
@@
 Multigrid solvers with small enough coarse-grid problems are well-known to be scalable even to the largest supercomputers \cite{gholami2016,ibeid2018} as the communication is mostly between nearest neighbors.
-}
 The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements,
@@
 will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound.
-{\color{red}Furthermore, the data structures of the matrix-free approaches are also too large to fit completely into caches.}
+Furthermore, the data structures of the matrix-free approaches are also too large to fit completely into caches.
 
@@
 
-{\color{red}
 Computations were performed on two systems: A dual-socket Intel Xeon Gold 6230 (``Cascade Lake''), released in 2019,
@@
 The GNU compiler version 9.1.0 with flags ``-O3 -march=skylake-avx512' and OpenMPI version 4.0.1 were used, with the code run in pure MPI mode.
-}
 The ``IWR'' results were obtained on a single machine with eight Intel Xeon E7-8870 ``Westmere'' chips, released in 2011,
@@
 We are interested in the wall-clock time (average over 10 runs for each linear solver step) per DoF as a first metric.
-As expected, the matrix-vector multiplication becomes very expensive for higher {\color{red}polynomial degrees} for sparse matrix-based approaches,
+As expected, the matrix-vector multiplication becomes very expensive for higher polynomial degrees for sparse matrix-based approaches,
 as the matrix becomes more and more dense.
-{\color{red}
 In order to increase visibility, the figures also report the throughput of the matrix vector product,
 computed by the ratio of the number of degrees of freedom over the time for a matrix-vector product, in panels (a) and (d).
-}
 The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
-{\color{red}
 Note that we consider only deformed meshes together with separate metric terms according to Table~\ref{tab:memory}.
@@
 the linearization-related mapping required for evaluation of gradients with respect to the current configuration.
-}
 
@@
 The matrix-free implementation is faster than the matrix-based counterpart already for a bi-quadratic Lagrangian basis in 2D and a tri-quadratic basis in 3D on both systems.
-{\color{red}
 The ``MF tensor2'' and ``MF scalar'' strategies outperform the sparse matrix already for tri-linear elements of polynomial degree $p=1$ in 3D on Cascade Lake architecture.
@@
 as the time per unknown in panels (b) and (e) closely follows the memory access.
-}
 Figures \ref{fig:benchmark_miehe_CSL_vmult2}, \ref{fig:benchmark_miehe_CSL_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3}
@@
 Two conclusions can be drawn from these results:
-{\color{red}
 Firstly, the algorithm where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement
@@
 On the IWR machine, the Flop/Byte ratio is 2.9, which means that much fewer computations can be sustained for the same memory transfer.
-}
-
-{\color{red}
+
 Secondly, the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') is slower than the ``tensor2'' approach,
@@
 in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
-}
 It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
@@
 demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point (``tensor4''), the matrix-free approach takes much less memory than its matrix-based counterpart.
-{\color{red}
 In those graphs, we report the overall memory needed to cache additional quantities required for the tangent operator according to Table~\ref{tab:memory}
@@
 the \texttt{deal.II} \textit{MatrixFree} object.
-}
 
@@
 very low arithmetic intensity of 0.16 and a performance of $\pgfmathprintnumber{17.36}\,\rm{GFlop/s}$ (1.4\% of the arithmetic peak) averaged over the 2D results.
-{\color{red}
 The matrix-free implementations of the finite-strain elastic tangent operator (on average) lead to a much better performance of
@@
 The achieved performance is about 6.8\% of the peak arithmetic performance. This result is similar to the deformed mesh results of \cite[Figure 18]{kronbichler2017fast}.
-}
-
-{\color{red}
+
 The roofline model also explains the difference in performance between the ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) and the ``tensor2'' caching strategy (Algorithm \ref{alg:mf_tensor2}):
@@
 no advantage of the scalar caching can be deduced despite the higher arithmetic performance.
-}
 
@@
 (v) 'MPI' denotes MPI communication (lines 2 and 16).
-{\color{red}
 Given that the measurements are within the cell loop and there is an extensive overlap between the memory transfer (due to hardware prefetching) and arithmetic operations,
@@
 as the memory access of quadrature point data scales as $(p+1)^d/p^d$ compared to the number of unknowns.
-}
 
@@
 Table \ref{tab:numbers_2d} and Table \ref{tab:numbers_3d} show the speed-up obtained due to explicit SIMD vectorization intrinsics and due to the MPI parallelization.
-{\color{red}
 When run in serial, explicit SIMD vectorization via intrinsics provides a speedup of up to $3\times$ over the auto-vectorization case.
@@
 the speedup over the serial, auto-vectorized case is less than that of Algorithm~\ref{alg:mf_tensor2}.
-}
 
 To summarize this section, we can attribute the speed-up of the matrix-free finite-strain tangent operators for higher-order elements to two factors.
-{\color{red}
 The first one is the higher algorithmic intensity and lower memory access as compared to the matrix-based implementation,
 see Figure \ref{fig:roofline}. Since both variants are in the memory-limited regime, the matrix-free implementation with its lower memory access is beneficial.
-}
 The second benefit is the reduction in the number of floating point operations per DoF for higher order elements thanks to the sum factorization technique, as discussed in
@@
 However, we have performed a preliminary study of the weak scalability of the implementation.
-{\color{red}Table~\ref{tab:weak} demonstrate a good weak scaling up to 320 cores, i.e., 8 nodes of dual-socket Cascade Lake,
-of the wall time per matrix-vector product and per iteration in the CG solver for the 2D problem with $p=4$ and the 3D problem with a tri-quadratic basis.}
+Table~\ref{tab:weak} demonstrate a good weak scaling up to 320 cores, i.e., 8 nodes of dual-socket Cascade Lake,
+of the wall time per matrix-vector product and per iteration in the CG solver for the 2D problem with $p=4$ and the 3D problem with a tri-quadratic basis.
 This confirms that the matrix-free and multigrid frameworks implemented in the \texttt{deal.II} library can provide a competitive solution for cluster-sized problems as well,
@@
 Among the three examined algorithms,
-{\color{red}the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the current configuration is faster than
-caching only a scalar at quadrature points or caching the material part} of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress.
-{\color{red}
+the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the current configuration is faster than
+caching only a scalar at quadrature points or caching the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress.
 Given that we did not utilize the specific form of the material part of the stiffness tensor in the latter case, we conclude
@@
 operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
-}
 
 The roofline model indicates that the matrix-free finite-strain elasticity tangent operators have an order magnitude higher algorithmic intensity.
-{\color{red}
 Despite the higher arithmetic intensity, the matrix-free implementations are still memory bandwidth-limited and in particular by the data
 required at quadrature points in terms of fourth-order and second-order symmetric tensors.
-}
 
@@
 
-{\color{red}
 It is a common practice to write the bilinear form completely in the referential configuration.
 For the constitutive equation \eqref{eq:neo_hookean_energy} considered in this study, this results in the following bilinear form
-}
 \begin{align}
@@
 
-{\color{red}
 This formulation can also be straight forwardly adopted within the matrix-free framework of the \texttt{deal.II} library.
@@
 and only a fourth-order tangent tensor ${\rm D}_{\gz F}{\gz P}$ is cached per quadrature point.
-}
-
-{\color{red}
+
 In the interest of brevity, no results are reported for this approach,
@@
 This is mainly caused by the cost of indirect addressing in the vector access as well as division for inverting $\gz F$ on the fly.
-}
 

### 2019-09-15 doc/large_strain.tex
Update doc/large_strain.tex
Co-Authored-By: Martin Kronbichler <kronbichler@lnm.mw.tum.de>

@@
     We find that matrix-free methods for finite-strain solid mechanics are very promising, outperforming linear matrix-based schemes by two to five times,
-    and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
+    and that it is possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
   %   \footnotetext{%

### 2019-09-16 doc/large_strain.tex

--- 2019-09-15 doc/large_strain.tex
@@
   \author[b,c]{Daniel Arndt
-  %  \footnotetext{%
-  %  This manuscript has been authored by UT-Battelle, LLC under Contract No.
-  %  DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government
-  %  retains and the publisher, by accepting the article for publication, acknowledges
-  %  that the United States Government retains a non-exclusive, paid-up, irrevocable,
-  %  worldwide license to publish or reproduce the published form of this manuscript,
-  %  or allow others to do so, for United States Government purposes. The Department
-  %  of Energy will provide public access to these results of federally sponsored
-  %  research in accordance with the DOE Public Access Plan
-  %  (http://energy.gov/downloads/doe-public-access-plan).}
-  }
+   \footnotetext{%
+   This manuscript has been authored by UT-Battelle, LLC under Contract No.
+   DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government
+   retains and the publisher, by accepting the article for publication, acknowledges
+   that the United States Government retains a non-exclusive, paid-up, irrevocable,
+   worldwide license to publish or reproduce the published form of this manuscript,
+   or allow others to do so, for United States Government purposes. The Department
+   of Energy will provide public access to these results of federally sponsored
+   research in accordance with the DOE Public Access Plan
+   (http://energy.gov/downloads/doe-public-access-plan).}}
   \ead{arndtd@ornl.gov}
@@
     and that is it possible to develop numerically efficient implementations that are independent of the hyperelastic constitutive law.}
-  %   \footnotetext{%
-  %  This manuscript has been authored by UT-Battelle, LLC under Contract No.
-  %  DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government
-  %  retains and the publisher, by accepting the article for publication, acknowledges
-  %  that the United States Government retains a non-exclusive, paid-up, irrevocable,
-  %  worldwide license to publish or reproduce the published form of this manuscript,
-  %  or allow others to do so, for United States Government purposes. The Department
-  %  of Energy will provide public access to these results of federally sponsored
-  %  research in accordance with the DOE Public Access Plan
-  %  (http://energy.gov/downloads/doe-public-access-plan).}
+    \footnotetext{%
+   This manuscript has been authored by UT-Battelle, LLC under Contract No.
+   DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government
+   retains and the publisher, by accepting the article for publication, acknowledges
+   that the United States Government retains a non-exclusive, paid-up, irrevocable,
+   worldwide license to publish or reproduce the published form of this manuscript,
+   or allow others to do so, for United States Government purposes. The Department
+   of Energy will provide public access to these results of federally sponsored
+   research in accordance with the DOE Public Access Plan
+   (http://energy.gov/downloads/doe-public-access-plan).}
   \end{abstract}

### 2019-09-16 doc/large_strain.tex
add grant number for EPSRC

@@
 for Exascale Computing'' (SPPEXA).
-P.~Steinmann acknowledges the support of the Cluster of Excellence Engineering of Advanced Materials (EAM) which made this collaboration possible, as well as funding by the EPSRC Strategic Support Package ``Engineering of Active Materials by Multiscale/Multiphysics Computational Mechanics''.
+P.~Steinmann acknowledges the support of the Cluster of Excellence Engineering of Advanced Materials (EAM) which made this collaboration possible, as well as funding by the EPSRC Strategic Support Package ``Engineering of Active Materials by Multiscale/Multiphysics Computational Mechanics'' (EP/R008531/1).
 

### 2019-09-17 doc/large_strain.tex

@@
 D.~Arndt and M.~Kronbichler were supported by the German Research Foundation (DFG) under the project ``High-order discontinuous
-Galerkin for the exa-scale'' (\mbox{ExaDG}) within the priority program ``Software
+Galerkin for the exa-scale'' (\mbox{ExaDG}), grant 279336170 within the priority program ``Software
 for Exascale Computing'' (SPPEXA).

### 2020-01-24 doc/IJNME/reply_to_reviewers.md
star preparing reply to reviewers

# A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

We thank the anonymous reviewers for the valuable comments and suggestions to the manuscript. We did our best to incorporate them into the revision. Below we state our response point-by-point.
To facilitate the review process, all significant changes are marked in red in the updated version of the manuscript.

----

## Referee: 1

> Minor editing is required in figures (axis titles).

> A uniform convention should be adopted for author names in the list of references.

## Referee: 2

> In this paper, the authors develop a matrix-free method for nonlinear elasticity. The paper is very well written, both the methods (Alg. 1-3) and the numerical tests are convincing.

> Here are some minor remarks:

> 1. Eq. (4) on page 3:
>  a) The period "." is wrong. Please make a comma
>  b) And \forall \delta u reads strange wittout function space. I would write \forall \text{admissible directions } \delta u

> 2. To non-familiar Heidelberg-persons, IWR might be unknown. Please explain IWR on page 12.

> 3. Section 6: Out of curiosity: how many nonlinear Newton steps are required to solve the problem?

> 4. In the conclusions, I would specifically refer again to Alg. 1,2,3 and give some suggestions which Alg. 1 finally is prefererred and why.
> On page 16, just before Section 6.2, there are some suggestions. Are these the final conclusions w.r.t. the proposed Alg.?

### 2020-01-25 doc/IJNME/reply_to_reviewers.md
Comment on author names
Work on some of the review comments

@@
 
+The multiple first names were abbreviated the same way. This is the only non-uniformity we could find for the author names.
+
 ## Referee: 2
@@
 
+We have fixed these problems.
+Note that we removed \forall \delta u because it is explained in the text that follows, "vanishes for all admissible directions \delta u which satisfy homogeneous Dirichlet boundary conditions".
+
 > 2. To non-familiar Heidelberg-persons, IWR might be unknown. Please explain IWR on page 12.
+
+We have replaced the system name "IWR" by "Westmere", which is the name of the CPU architecture and consistent with the use of "Cascade Lake" for the other system.
 
@@
 
+FIXME
+
 > 4. In the conclusions, I would specifically refer again to Alg. 1,2,3 and give some suggestions which Alg. 1 finally is prefererred and why.
 > On page 16, just before Section 6.2, there are some suggestions. Are these the final conclusions w.r.t. the proposed Alg.?
+
+FIXME

### 2020-01-25 doc/large_strain.tex
Work on some of the review comments

@@
 \D_{\displaystyle \delta \gz u} \mcl E :=
-\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, \qquad \forall \delta \gz u \, .
+\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0
 \label{eq:stationary}
 \end{align}
-vanishes for all directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
+vanishes for all {\color{red}admissible} directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
 %
@@
 Turbo mode is enabled with a power limit of 135 watt per socket, which results in the processor running at 2.8 GHz for scalar code and 2.0 GHz for AVX-512-heavy code in the experiments.
-The GNU compiler version 9.1.0 with flags ``-O3 -march=skylake-avx512' and OpenMPI version 4.0.1 were used, with the code run in pure MPI mode.
-The ``IWR'' results were obtained on a single machine with eight Intel Xeon E7-8870 ``Westmere'' chips, released in 2011,
+The GNU compiler version 9.1.0 with flags ``-O3 -march=skylake-avx512'' and OpenMPI version 4.0.1 were used, with the code run in pure MPI mode.
+The {\color{red}``Westmere''} results were obtained on a single machine with eight Intel Xeon E7-8870 ``Westmere'' chips, released in 2011,
 (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip. For the simulation, GCC version 8.1.0 with flags
 ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
-Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. one socket for Cascade Lake and two sockets for ``IWR''.
+Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e., one socket for Cascade Lake and two sockets for {\color{red}Westmere}.
 
@@
   \end{subfigure}
-  \caption{IWR cluster, 20 cores, matrix-vector multiplication}%
+  \caption{{\color{red}Westmere} cluster, 20 cores, matrix-vector multiplication}%
   \label{fig:benchmark_miehe_IWR}
@@
 On the newer Cascade Lake system with a Flop/Byte ratio of 9.1 for the theoretical peak values, the scalar caching strategy is closer to the ``tensor2'' variant.
-On the IWR machine, the Flop/Byte ratio is 2.9, which means that much fewer computations can be sustained for the same memory transfer.
+On the {\color{red}Westmere} machine, the Flop/Byte ratio is 2.9, which means that much fewer computations can be sustained for the same memory transfer.
 

### 2020-01-26 Calculations/post_process_newton_iterations.py
reply on non-linear newton steps

@@ -1,0 +1,79 @@
+import re
+import os
+import argparse
+from matplotlib.pyplot import figure
+import matplotlib.pyplot as plt
+import matplotlib as mp
+import numpy as np
+import fileinput
+import matplotlib.ticker
+from matplotlib.ticker import MaxNLocator
+from matplotlib.ticker import FormatStrFormatter
+
+# define command line arguments
+parser = argparse.ArgumentParser(
+    description='Post-Process average number of newton iterations.',
+    formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+parser.add_argument('prefix', metavar='prefix', default='IWR_newest_patched', nargs='?',
+                    help='A folder to look for benchmark results')
+args = parser.parse_args()
+
+prefix = args.prefix if args.prefix.startswith('/') else os.path.join(os.getcwd(), args.prefix)
+
+files = [os.path.join(prefix, k,'output') for k in os.listdir(prefix) if os.path.isfile(os.path.join(prefix, k,'output'))]
+
+print 'Gather data from {0}'.format(prefix)
+print 'found {0} files'.format(len(files))
+
+pattern = r'[+\-]?(?:[0-9]\d*)(?:\.\d*)?(?:[eE][+\-]?\d+)?'
+
+newton2d = []
+newton3d = []
+
+for f in files:
+    # Newton iterations should be same everywhere, take one flavour
+    if 'gmg_tensor4' not in f:
+        continue
+    if 'gmg_tensor4_ns' in f:
+        continue
+
+    # guess parameters from directory name
+    dirname = f.split('/')[-2]
+    strings = dirname.split('_')
+    dim = int(re.findall(pattern,strings[1])[0])
+    p   = int(re.findall(pattern,strings[2])[0])
+    q   = int(re.findall(pattern,strings[2])[1])
+
+    print 'dim={0} p={1} q={2} dir={3}'.format(dim,p,q,dirname)
+
+    # count newton iterations
+    fin = open(f, 'r')
+
+    newtonIterations = 0
+    counter = 0
+    for line in fin:
+        # Converged in 4 Newton iterations
+        if 'Converged in' in line:
+            thisIt = int(re.findall(pattern,line)[0])
+            counter = counter + 1
+            newtonIterations = newtonIterations + thisIt
+
+    averageNewtonIterations = float(newtonIterations)/counter
+    # print(averageNewtonIterations)
+
[…]

### 2020-01-26 doc/IJNME/reply_to_reviewers.md
one more IWR->Westmere
reply on non-linear newton steps
algorithms in conclusion

@@
 > Minor editing is required in figures (axis titles).
+
+FIXME
 
@@
 
-FIXME
+With our settings it takes 4 Newton-Raphson iterations to converge within a load step for the 2D problem and 3 iterations for the 3D setup.
 
@@
 
-FIXME
+If linearization of the chosen material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor, then Algorithm 2 is recommended.
+For applications within general frameworks with different geometrically non-linear constitutive models, Algorithm 3 provides the necessary flexibility while still being considerably faster than matrix-based strategies for polynomial degrees higher than one.
+
+Note that the higher arithmetic load for Algorithm 1 is not surprising as at each quadrature point we need to re-evaluate the Kirchhoff stress, and evaluate gradients with respect to both the referential and the current configuration.
+Thus no advantage of the scalar caching can be deduced despite the higher arithmetic performance.
+
+We have augmented Conclusion section to better convey this message.

### 2020-01-26 doc/large_strain.tex
one more IWR->Westmere
reply on non-linear newton steps
algorithms in conclusion

@@
 With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$
-whereas the relative and absolute tolerances for the residual forces is $10^{-8}$. The relative convergence criteria for the linear solver is $10^{-6}$.
+whereas the relative and absolute tolerances for the residual forces is $10^{-8}$.
+{\color{red}
+With the chosen criteria it takes 4 Newton-Raphson iterations to converge within a load step for the 2D problem and 3 iterations for the 3D setup.
+}
+The relative convergence criteria for the linear solver is $10^{-6}$.
 %
@@
   \end{subfigure}
-  \caption{IWR cluster, 20 cores.}%
+  \caption{{\color{red}Westmere} cluster, 20 cores.}%
   \label{fig:benchmark_miehe_IWR_cg}
@@
 Among the three examined algorithms,
-the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the current configuration is faster than
-caching only a scalar at quadrature points or caching the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress.
-Given that we did not utilize the specific form of the material part of the stiffness tensor in the latter case, we conclude
-that the matrix-free implementation of the tangent operator is somewhat slower but more flexible because it can be applied to any constitutive model which
+the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the current configuration ({\color{red}Algorithm \ref{alg:mf_tensor2}}) is faster than
+caching only a scalar at quadrature points ({\color{red}Algorithm \ref{alg:mf_scalar}}) or caching the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress ({\color{red}Algorithm \ref{alg:mf_tensor4}}).
+{\color{red}
+Algorithm \ref{alg:mf_tensor2} is recommended when linearization of the adopted material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor.
+Algorithm \ref{alg:mf_tensor4} does
+}
+not utilize the specific form of the material part of the stiffness tensor,
+{\color{red}
+in this case
+}
+the matrix-free implementation of the tangent operator is somewhat slower but more flexible because it can be applied to any constitutive model which
 operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
+{\color{red}
+This implementation is still considerably faster than matrix-based strategies for polynomial degrees higher than one.
+}
 

### 2020-01-26 doc/large_strain.tex
Minor text updates when reading through the document

@@
 
-The Newton-Raphson solution approach is adopted for the nonlinear problem considered here.
+The Newton--Raphson solution approach is adopted for the nonlinear problem considered here.
 Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is obtained as the solution to the following
@@
   compress remote integral contributions into $\gz y$ via MPI \;
-  \caption{Matrix-free tangent operator: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
+  \caption{Matrix-free tangent operator: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at {\color{red}quadrature points} when executed.}
   \label{alg:mf_tensor2}
@@
 the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} sparse data structure from the Epetra package \cite{Heroux2005b} and
-the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
+the linear solver is preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
 This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is available in \texttt{deal.II}.
@@
 {\color{red}
-With the chosen criteria it takes 4 Newton-Raphson iterations to converge within a load step for the 2D problem and 3 iterations for the 3D setup.
+With the chosen criteria it takes 4 Newton--Raphson iterations to converge within a load step for the 2D problem and 3 iterations for the 3D setup.
 }
@@
 as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
-We chose a problem size on the order of $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach with a high polynomial degree.
+We choose a problem size on the order of $10^6$ DoFs for each test case as this is an upper limit due to the memory requirements of the matrix-based approach with a high polynomial degree.
 Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization)
@@
 Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end,
-we consider the average number of CG iterations throughout the entire simulation, i.e., averaged over each Newton-Raphson iteration and each loading step.
+we consider the average number of CG iterations throughout the entire simulation, i.e., averaged over each Newton--Raphson iteration and each loading step.
 Figures \ref{fig:benchmark_miehe_CSL_cg2}, \ref{fig:benchmark_miehe_CSL_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
@@
 {\color{red}
-Algorithm \ref{alg:mf_tensor2} is recommended when linearization of the adopted material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor.
+Algorithm \ref{alg:mf_tensor2} is recommended when linearization of the adopted material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor {\color{red}by a second-order symmetric tensor}.
 Algorithm \ref{alg:mf_tensor4} does
 }
-not utilize the specific form of the material part of the stiffness tensor,
+not utilize the specific form of the material part of the stiffness tensor.
 {\color{red}
-in this case
+In this case
 }
 the matrix-free implementation of the tangent operator is somewhat slower but more flexible because it can be applied to any constitutive model which
-operates with the material part of the fourth-order spatial tangent stiffness tensor on the quadrature point level.
+operates with the material part of the fourth-order spatial tangent stiffness tensor at the quadrature points.
 {\color{red}
@@
 with standard geometric transfer operations between each level.
-The GMG was applied to a heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
+The GMG has been applied to a heterogeneous material assuming that the coarsest level can provide an adequate discretization of the heterogeneity.
 The numerical studies indicate that the proposed preconditioner
@@
 leads to a solution approach that is faster than the matrix-based AMG for all polynomial degrees studied here.
-Most importantly, the wall time per degree of freedom for solving the problem is close to being constant.
+Most importantly, the wall time per degree of freedom for solving the problem is close to being constant {\color{red}for polynomial degrees between one and four}.
 On the other hand, the matrix-based implementation with AMG becomes prohibitively expensive for higher-order bases.
 We conclude that the matrix-free implementations of tangent operators for finite-strain elasticity together with
-the geometric multigrid preconditioner is a very competitive solution strategy, as compared to more traditional matrix-based approach.
+the geometric multigrid preconditioner is a very competitive solution strategy, as compared to traditional matrix-based approach.
 Our future work will be focused on adopting the proposed matrix-free multigrid solution approach to $\operatorname{FE}^2$ homogenization

### 2020-01-27 doc/large_strain.tex
by -> on

@@
 {\color{red}
-Algorithm \ref{alg:mf_tensor2} is recommended when linearization of the adopted material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor {\color{red}by a second-order symmetric tensor}.
+Algorithm \ref{alg:mf_tensor2} is recommended when linearization of the adopted material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on a second-order symmetric tensor.
 Algorithm \ref{alg:mf_tensor4} does

### 2020-01-29 doc/IJNME/reply_to_reviewers.md
reply on axis titles

@@
 
-FIXME
+We have re-examined axis titles in all figures. We did not find any typos, but increased the font size of some of them to make more readable.
 
@@
 
-The multiple first names were abbreviated the same way. This is the only non-uniformity we could find for the author names.
+The multiple first names were abbreviated the same way. This is the only non-uniformity we could find for the authors' names.
+
+----
 
@@
 
-With our settings it takes 4 Newton-Raphson iterations to converge within a load step for the 2D problem and 3 iterations for the 3D setup.
+With our settings, it takes 4 Newton-Raphson iterations to converge within a load step for the 2D problem and 3 iterations for the 3D setup.
 
-> 4. In the conclusions, I would specifically refer again to Alg. 1,2,3 and give some suggestions which Alg. 1 finally is prefererred and why.
+> 4. In the conclusions, I would specifically refer again to Alg. 1,2,3 and give some suggestions which Alg. 1 finally is preferred and why.
 > On page 16, just before Section 6.2, there are some suggestions. Are these the final conclusions w.r.t. the proposed Alg.?
 
-If linearization of the chosen material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor, then Algorithm 2 is recommended.
+If the linearization of the chosen material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor, then Algorithm 2 is recommended.
 For applications within general frameworks with different geometrically non-linear constitutive models, Algorithm 3 provides the necessary flexibility while still being considerably faster than matrix-based strategies for polynomial degrees higher than one.
@@
 
-We have augmented Conclusion section to better convey this message.
+We have augmented the Conclusion section to better convey this message.

### 2020-01-29 doc/large_strain.tex

@@
 {\color{red}
-Algorithm \ref{alg:mf_tensor2} is recommended when linearization of the adopted material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on a second-order symmetric tensor.
+Algorithm \ref{alg:mf_tensor2} is recommended when linearization of the adopted material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on a second-order symmetric tensor by two scalar quantities.
 Algorithm \ref{alg:mf_tensor4} does

### 2020-02-10 doc/large_strain.tex

--- 2020-01-27 doc/large_strain.tex
@@
 \end{align}
-vanishes for all {\color{red}admissible} directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
+vanishes for all admissible directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
 %
@@
   compress remote integral contributions into $\gz y$ via MPI \;
-  \caption{Matrix-free tangent operator: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at {\color{red}quadrature points} when executed.}
+  \caption{Matrix-free tangent operator: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at quadrature points when executed.}
   \label{alg:mf_tensor2}
@@
 whereas the relative and absolute tolerances for the residual forces is $10^{-8}$.
-{\color{red}
 With the chosen criteria it takes 4 Newton--Raphson iterations to converge within a load step for the 2D problem and 3 iterations for the 3D setup.
-}
 The relative convergence criteria for the linear solver is $10^{-6}$.
@@
 The GNU compiler version 9.1.0 with flags ``-O3 -march=skylake-avx512'' and OpenMPI version 4.0.1 were used, with the code run in pure MPI mode.
-The {\color{red}``Westmere''} results were obtained on a single machine with eight Intel Xeon E7-8870 ``Westmere'' chips, released in 2011,
+The ``Westmere'' results were obtained on a single machine with eight Intel Xeon E7-8870 ``Westmere'' chips, released in 2011,
 (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip. For the simulation, GCC version 8.1.0 with flags
 ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
-Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e., one socket for Cascade Lake and two sockets for {\color{red}Westmere}.
+Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e., one socket for Cascade Lake and two sockets for Westmere.
 
@@
   \end{subfigure}
-  \caption{{\color{red}Westmere} cluster, 20 cores, matrix-vector multiplication}%
+  \caption{Westmere cluster, 20 cores, matrix-vector multiplication}%
   \label{fig:benchmark_miehe_IWR}
@@
 On the newer Cascade Lake system with a Flop/Byte ratio of 9.1 for the theoretical peak values, the scalar caching strategy is closer to the ``tensor2'' variant.
-On the {\color{red}Westmere} machine, the Flop/Byte ratio is 2.9, which means that much fewer computations can be sustained for the same memory transfer.
+On the Westmere machine, the Flop/Byte ratio is 2.9, which means that much fewer computations can be sustained for the same memory transfer.
 
@@
   \end{subfigure}
-  \caption{{\color{red}Westmere} cluster, 20 cores.}%
+  \caption{Westmere cluster, 20 cores.}%
   \label{fig:benchmark_miehe_IWR_cg}
@@
 Among the three examined algorithms,
-the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the current configuration ({\color{red}Algorithm \ref{alg:mf_tensor2}}) is faster than
-caching only a scalar at quadrature points ({\color{red}Algorithm \ref{alg:mf_scalar}}) or caching the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress ({\color{red}Algorithm \ref{alg:mf_tensor4}}).
-{\color{red}
-Algorithm \ref{alg:mf_tensor2} is recommended when linearization of the adopted material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on a second-order symmetric tensor.
+the implementation that caches the second-order Kirchhoff stress tensor and evaluates terms in the current configuration (Algorithm \ref{alg:mf_tensor2}) is faster than
+caching only a scalar at quadrature points (Algorithm \ref{alg:mf_scalar}) or caching the material part of the fourth-order spatial tangent stiffness tensor together with the second-order Kirchhoff stress (Algorithm \ref{alg:mf_tensor4}).
+Algorithm \ref{alg:mf_tensor2} is recommended when linearization of the adopted material model allows to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on a second-order symmetric tensor by two scalar quantities.
 Algorithm \ref{alg:mf_tensor4} does
-}
 not utilize the specific form of the material part of the stiffness tensor.
-{\color{red}
 In this case
-}
 the matrix-free implementation of the tangent operator is somewhat slower but more flexible because it can be applied to any constitutive model which
 operates with the material part of the fourth-order spatial tangent stiffness tensor at the quadrature points.
-{\color{red}
 This implementation is still considerably faster than matrix-based strategies for polynomial degrees higher than one.
-}
 
@@
 leads to a solution approach that is faster than the matrix-based AMG for all polynomial degrees studied here.
-Most importantly, the wall time per degree of freedom for solving the problem is close to being constant {\color{red}for polynomial degrees between one and four}.
+Most importantly, the wall time per degree of freedom for solving the problem is close to being constant for polynomial degrees between one and four.
 On the other hand, the matrix-based implementation with AMG becomes prohibitively expensive for higher-order bases.
